### 2016-01-15 doc/manuscript.Rmd
Added possible outline, methods, tabs+figs to manuscritp

---
bibliography: 
    - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
csl:
    - /home/luke/Documents/bibliography/styles/american-medical-association-no-et-al.csl
output:
    word_document:
      reference_docx: /home/luke/Templates/journal-templates/ajcn-template.docx
      fig_height: 6 
      fig_width: 5 
      fig_caption: true
---

# Title page #

Title:

Author:

Affiliations:

Disclaimers:

Correspondence:

Funding support:

# Abstract #

<!-- outline

Overall story: while nefa are known risk factors for diabetes, not sure about
the composition. we find that the composition doesn't matter so much as the
overall pool, specific to BCD. however, 16:0, etc, are the only two fa that have
an assoc, suggesting that irrespective of other fa increases or decreases, these
two fa are the fa that contribute to the harm seen in the overall nefa. using
cluster analysis, we see that those with low bcd have a pattern of fa that skews
to higher amounts of 16:0 etc and marginally less in other fa (maybe not enough
to observe on an individual scale). while other studies have seen a role of nefa
in ir, we do not. while this warrants further longitudinal research, this may
suggest that given prev studies on ir have been acute in nature, over the long
term the role of nefa on ir may not be substantial. this duality in nefa is also
seen in bcd, where acutely it improves bc, but chronically no.

Possible reason why IR is not shown... other studies have been cross-sectional..
Thus a potential confounder is the role that obesity plays in IR and IR on FFA
release... So, longitudinally, FFA would not predict IR since it is IR that is
suppressing NEFA.

total nefa evidence, well-studied. but there is diversity. no studies on it.

Some evidence to suggest that FFA delivery may be higher in obese, but that conc
may or may not be. Also that research on increasing FFA and IR comes from
experimental (probably fairly short term) research.

methods -> gee, then lda (or similar) to confirm the role of 16:0 etc rather
than any individual FA

- as GEE results suggest that % does nothing, but in conc 16:0 etc has assoc...
is it that 16:0 is harmful or is it the pool? is it that as 16:0 inc, there is
harm, irrespective of whether other FA increase or decrease as well?
- need to do LDA or some such to id the pattern of why... (don't show this in
paper, but examine PCA of it too)

figs/tabs: 
- tab:basic char
- fig:distr
- fig:nefaCorr (sup?)
- fig:varCorr (sup?)
- fig:geeconc%
- fig:lda

supfigs/tabs:
- tab:geeconc%
- tab:distr

THOUGHTS:

- sleep restriction can impact NEFA, but only higher during night and a bit
during early-morning. Should keep this in mind.
- Adjust for TAG in IR models?

-->

<!--
## Load data ##
-->

```{r setup, message = FALSE, echo = FALSE}
invisible(sapply(list.files('R', pattern = '*.R', full.names = TRUE), source))
run_setup()
load_data()
dim(ds)
```

```{r caption_setup}
## Include captions below using `captioner` package
figNums <- captioner(prefix = 'Figure')
supFigNums <- captioner(prefix = 'Supplementary Figure')
tabNums <- captioner(prefix = 'Table')
supTabNums <- captioner(prefix = 'Supplementary Table')
```

# Background #

Three paragraphs:

1. Total NEFA on diabetes, pathology, evidence
2. NEFA diverse, individual NEFA on pathology, bit more evidence from total, go
into experimental of species
3. Limitations, gaps, inconsistencies, objectives and hypotheses

# Subjects and Methods #

The Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort is a longitudinal observational study of
participants with one or more risk factors for type 2 diabetes
mellitus, including obesity, hypertension, family history of diabetes,
and/or a history of gestational diabetes or birth of a macrosomic
infant.  Participants aged 30 years and older were recruited for the baseline visit from Toronto and London, Ontario, Canada into the PROMISE cohort (n=736)
between 2004-2006.  Follow-up examinations occur every
three years, with currently three examination visits completed to date
(2004-2006, 2007-2009, and 2010-2013).  The current study used data on
participants who did not have diabetes at baseline, who returned for
one or more follow-up examinations, and who had samples available for
fatty acid measurements (n=477).  A diagram of the sample size at each
visit is shown in `r supFigRef('flow')`.  At each examination,
participants undergo extensive metabolic characterization,
anthropometric measurements, and structured questionnaires on
lifestyle and sociodemographics.  Research ethics approval was
obtained from Mount Sinai Hospital and the University of Western
Ontario.  Data collection methods were standardized across the 2
centres and research nurses were centrally trained.  <!--Over the
6-yrs, {{n return each visit}} returned for the 6-yr clinic visit
({{pct loss to followup}}% loss to follow-up).  By the 6-yr visit, {{n
converted to IFG}} converted into IFG, {{n converted to IGT}}
converted into IGT, and {{n converted to DM}} into diabetes.-->

## Blood measure assessment

At each clinic examination, participants had an 7-12 hour fasting blood sample 
taken and then had a 75g oral glucose tolerance test (OGTT) with a 30 minute and
2 hour blood draw. All blood samples were processed and frozen at -70°C. 
Alanine
aminotransferase (ALT) was measured using standard laboratory
procedures.  Cholesterol, HDL, and triacylglycerols (TAG) were
measured using Roche Modular's enzymatic colometric tests
(Mississauga, ON).  Both specific insulin and glucose were derived
from the OGTT at fasting, 30 minute, and 2 hour time points.  Specific
insulin was measured using the Elecsys 1010 (Roche Diagnostics, Basel,
Switzerland) immunoassay analyzer and electrochemiluminescence
immunoassay.  This assay shows 0.05% cross-reactivity to intact human
pro-insulin and the Des 31,32 circulating split form (Linco Res. Inc),
and has a CV of 9.3%.  Glucose was determined using an enzymatic
hexokinase (Roche Modular, Roche Diagnostics) with a detection range
of 0.11 (2 mg/dL) to 41.6 mmol/L.  The inter-assay %CV is <1.1% and
intra-assay %CV is < 1.9%.  All assays were performed at the Banting
and Best Diabetes Centre Core Lab at the University of Toronto.
Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and
diabetes were categorized using the 2006 WHO criteria [@WHO2006].

NEFA composition was quantified using stored fasting serum samples from the
baseline visit, which had been frozen at -70°C for 4-6 years and had not been
exposed to any freeze-thaw cycles.  Serum fatty acids have been documented to be
stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount
of {{internal standard name}} (17:0) was added to the serum as an internal 
standard prior to extracting total lipids according to the method of Folch
[@Folch1957a].  Each serum lipid fraction was isolated using thin layer 
chromatography, which were  visualized under UV light after lightly spraying 
with 8-anilino-1-naphthalene sulfonic acid (0.1% wt/vol) and then converted to
fatty acid methyl esters with 14% boron trifluoride in methanol at 100°C for 1
h.  Fatty acid methyl esters were separated and quantified using a Varian-430
gas chromatograph (Varian, Lake Forest, CA, USA) equipped with a Varian Factor
Four capillary column (VF-23ms; 30 m × 0.25 mm i.d. × 0.25 μm film thickness)
and a flame ionization detector, which were injected in splitless mode. Fatty
acid concentrations (nmol/ml) were calculated by proportional comparison of gas
chromatography peak areas to that of the internal standards [@Nishi2014a;
@Chen2011a].  There were 22 fatty acids measured in the NEFA fraction.

## Outcome variables

Insulin sensitivity and beta-cell function indices were computed using the OGTT
glucose and insulin.  Insulin sensitivity was assessed using the inverted
homeostasis model assessment of insulin resistance (1/HOMA-IR) [@Matthews1985]
and the Insulin Sensitivity Index (ISI) [@Matsuda1999]. HOMA-IR largely reflects
hepatic insulin resistance, while ISI reflects whole-body insulin sensitivity
[@AbdulGhani2007].  Beta-cell function was assessed using the Insulinogenic
Index [@Wareham1995] over HOMA-IR (IGI/IR) and the Insulin Secretion-Sensitivity
Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR is a measure of the first phase
insulin secretion while ISSI-2 is analogous to the disposition index. Each index
has been validated against gold standard measures [@Matthews1985; @Matsuda1999;
@Retnakaran2009].

## Anthropometrics and sociodemographics

Height, weight, and waist circumference (WC) were measured at all clinic
examinations using standard procedures.  WC was measured at the natural waist,
defined as the narrowest part of the torso between the umbilicus and the xiphoid
process.  Height was measured using a stadiometer and weight was measured on a 
medical balance beam scale. BMI was calculated by dividing weight (kg) by height
(m) squared. Sociodemographic information, such as age, sex, and ethnicity, were
determined using standardized questionnaires administered at each examination
visit.

## Statistical analysis

The primary outcome variables for the PROMISE cohort and this analyses were 
1/HOMA-IR, ISI, IGI/IR, and ISSI-2; outcome variables were log-transformed for
the statistical modeling. The primary predictor variables for this study were
the 22 NEFA as a mole percent (mol%) of the total fraction and as a 
concentration (nmol/mL).

Pearson correlation coefficients were computed for the PL and CE fatty acids
with other continuous variables. {{keep??}}

For the primary analysis, generalized estimating equation (GEE) models 
[@Zeger1986a] were used to determine the longitudinal associations between the 
outcome variables and the predictor variables.  Given the longitudinal design
and auto-regressive (AR1) correlation matrix was chosen for the GEE models,
though other matrices (eg. exchangeable) had similar fit (data not shown)
{{confirm}}. GEE is well suited to cohort studies given it's capacity to handle
missed visits. The predictor variables and continuous covariates were scaled 
(mean centered and standardized). The NEFA variables were classified as 
*time-independent* (held constant) as they were measured only at the baseline 
visit, while the outcome variables and covariates were set as *time-dependent*. 
No imputation was conducted on missing values.

Covariates selection was based on previous literature, suggestions from directed
acyclic graph modeling [@Greenland1999a], and the quasi-likelihood information
criterion (QIC). Various models compared using QIC are shown in **`r supTabRef('tabQIC')`**.  
The final GEE model (M#) we selected had time, sex,
ethnicity, baseline age, waist circumference, {{...}}, and family history of
diabetes as covariates.  GEE parameter estimates were exponentiated.
<!--(exponentiated, minus 1, and multiplied by 100)--> After scaling,
log-transforming, and exponentiating, the GEE parameter estimates are
interpreted as an expected percent difference in the outcome variable for every
SD increase in the predictor variable given the covariates are held constant. 
An time by predictor term was included in each of the GEE models to test for
time interactions.

After running the GEE models for the longitudinal analysis, linear discriminant
analysis (LDA) {{or QDA}} was used to identify the patterns of NEFA composition for
tertiles of the metabolic disorders. Jacknifed cross-validation method used in
the LDA analysis. Only the baseline visit data was used for the LDA analysis and
only the outcome and predictor variables could be included in the analysis.

All statistical tests were computed using R `r getRversion()`
[@Rbase], along with the R packages geepack `r packageVersion('geepack')` for
the GEE modeling [@Hoejsgaard2006a] and MASS `r packageVersion('MASS')` for LDA.
The R code for this manuscript is available at {{link}}. Results were
statistically significance at p<0.05, after adjusting for the False Discovery
Rate [@Benjamini1995a] for multiple testing.

# Results #

# Discussion #

## Acknowledgements ##

# References #

# Tables #

```{r tabBasic}
# table(ds$Ethnicity, ds$f.VN)
# table(ds$f.VN)
summary(ds$Sex)
ds %>%
    outline_table(c('ne_Total', 'HOMA', 'ISI', 'IGIIR', 'ISSI2',
                    'TAG', 'Chol', 'BMI', 'Waist', 'HDL', 'FamHistDiab',
                    'Age', 'Ethnicity', 'Sex', 'MET', 'IFG', 'IGT',
                    'DM'), 'f.VN') %>% 
#     add_rows(c('HOMA', 'ISI', 'IGIIR', 'ISSI2'), stat_medianIQR, digits = 1) %>% 
#     add_rows(c('TAG', 'Chol', 'HDL', 'ne_Total', 'BMI', 'Waist', 'Age'), 
#              stat_meanSD, digits = 1) %>% 
    add_rows(c('Ethnicity', 'Sex')) %>% 
    rename_rows(renaming_table_rows) %>%
    rename_columns('Measure', 'Baseline', '3-yr', '6-yr') %>%
    construct_table()

```

# Figures #

```{r figDist}
ds %>%
    filter(VN == 0) %>%
    select(SID, matches('^ne\\d')) %>%
    filter(complete.cases(.)) %>% 
    gather(Measure, Value, -SID) %>% 
    mutate(Measure = renaming_fa(Measure) %>% 
               gsub('ne_', '', .) %>%
               factor(., levels = unique(.))) %>% 
    spread(Measure, Value) %>% 
    select(-SID) %>% 
    trance('boxes_dots') %>% 
    visualize(dots = FALSE,
              xlab = 'Concentration (nmol/mL)',
              ylab = 'Non-esterified fatty acid') %>% 
    vision_simple(base_size = 10)
```

```{r geeData}

outcomes <- c('linvHOMA', 'lISI', 'lIGIIR', 'lISSI2')
ne_pct <- grep('^pct_ne', names(ds), value = TRUE)
ne_conc <- c('ne_Total', grep('^ne\\d\\d', names(ds), value = TRUE))
covariates <- c('Waist', 'VN', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'FamHistDiab')

gee_df <- full_join(
    ds %>%
        select(-matches('pct_ne|^ne'),-ne_Total),
    ## Scale all the fatty acids
    ds %>%
        filter(VN == 0) %>%
        select(SID, ne_Total, matches('pct_ne|^ne')) %>%
        mutate_each(funs(as.numeric(scale(.))),-SID),
    by = 'SID') %>%
    group_by(VN) %>%
    mutate(
        Waist = as.numeric(scale(Waist)),
        ALT = as.numeric(scale(ALT)),
        BaseAge = as.numeric(scale(BaseAge))
        ) %>% 
    ungroup() %>%
    arrange(SID, VN)

gee_results <- bind_rows(
    analyze_gee(gee_df, outcomes, ne_pct, covariates, 'mol%'),
    analyze_gee(gee_df, outcomes, ne_conc, covariates, 'nmol/mL')
    ) %>% 
    mutate()

gee_results_unadj <- bind_rows(
    analyze_gee(gee_df, outcomes, ne_pct, covariates, 'mol%'),
    analyze_gee(gee_df, outcomes, ne_conc, covariates, 'nmol/mL')
    )

```

```{r figGee}
gee_results %>%
    trance('main_effect') %>%
    visualize(groups = 'unit~Yterms',
              xlab = 'Percent difference with 95% CI in the outcomes\nfor each SD increase in fatty acid',
              ylab = 'Non-esterified fatty acids') %>% 
    vision_simple(legend_position = 'bottom') +
    theme(
        axis.text.y = element_text(face = ifelse(
        gee_results$Xterms == 'Total', "bold", "plain"
        )),
        legend.key.width = grid::unit(0.75, "line"),
        legend.key.height = grid::unit(0.75, "line"),
        panel.margin = grid::unit(0.75, "lines")
        ) +
    scale_alpha_discrete(name = 'P-value', range = c(0.4, 1.0))
```

# Supplemental Material #

```{r supTabQIC}

subds <- ds %>%
  select(SID, VN, lISSI2, pl_Total, Sex, Ethnicity, MET, ALT,
         BaseAge, Waist, BMI, AlcoholPerWk, TobaccoUse, SelfEdu,
         Occupation, totalNE, FamHistDiab) %>%
  mutate(SelfEdu = as.factor(SelfEdu),
         Occupation = as.factor(Occupation),
         TobaccoUse = as.factor(TobaccoUse),
         AlcoholPerWk = as.factor(AlcoholPerWk)) %>%
  filter(complete.cases(.))

levels(subds$SelfEdu)[1:2] <- c('2', '2')
subds$SelfEdu <- droplevels(subds$SelfEdu)

M0 <- geeglm(lISSI2 ~ pl_Total + VN, data = subds,
                id = SID, family = gaussian,
                corstr = 'exchangeable')
M1 <- update(M0, . ~ . + Sex + Ethnicity + BaseAge)
M2 <- update(M0, . ~ . + Sex + Ethnicity + BaseAge + Waist)
M3 <- update(M0, . ~ . + Sex + Ethnicity + BaseAge + BMI)
M4 <- update(M0, . ~ . + Sex + Ethnicity + BaseAge + Waist + MET +
               AlcoholPerWk + TobaccoUse + SelfEdu + Occupation +
               FamHistDiab)
M5 <- update(M0, . ~ . + Sex + Ethnicity + BaseAge + MET + Waist +
               AlcoholPerWk + TobaccoUse + SelfEdu + Occupation +
               FamHistDiab + totalNE + ALT)
M6 <- update(M0, . ~ . + Sex + Ethnicity + Waist + totalNE)
M7 <- update(M0, . ~ . + Sex + Ethnicity + Waist + totalNE + ALT)
M8 <- update(M0, . ~ . + Sex + Ethnicity + Waist + totalNE + ALT + BaseAge)
M9 <- update(M0, . ~ . + Sex + Ethnicity + Waist + totalNE + ALT + BaseAge + FamHistDiab)

set.caption(supTabRef('tabQIC', 'Comparing generalized estimating equation
models adjusting for different covariates using
Quasi-Likelihood Information Criterion.'))
MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M9,
          rank = QIC) %>%
  tableModelSelection()

tableModelSelection <- function(selection.data) {
    selection.data %>%
      add_rownames() %>%
      select(Model = rowname, QIC, Delta = delta) %>%
      pander(digits = 2)
}

```

### 2016-01-27 R/fetch_data.R
Added printing of fetching process and included other vars in data

@@ -15,7 +15,7 @@
         filter(is.na(Canoe)) %>%
         tbl_df()
 
-    print(paste('Dimensions are:', dim(ds.prep)))
+    print(paste0('Original dataset rows are ', dim(ds.prep)[1], ' and columns are ', dim(ds.prep)[2]))
 
     ##' Munge and wrangle the data into the final version.
     ds <-
@@ -26,7 +26,7 @@
         select(
             SID, VN, BMI, Waist, HOMA, ISI, IGIIR, ISSI2, TAG, LDL, HDL, Chol,
             ALT, CRP, FamHistDiab, matches('meds'), Age, Sex, Ethnicity,
-            IFG, IGT, DM, MET, BaseAge
+            IFG, IGT, DM, MET, BaseAge, AlcoholPerWk, TobaccoUse, SelfEdu, Occupation
         ) %>%
         mutate(
             FamHistDiab =
@@ -37,7 +37,8 @@
             linvHOMA = log(invHOMA),
             lISI = log(ISI),
             lIGIIR = log(IGIIR),
-            lISSI2 = log(ISSI2)
+            lISSI2 = log(ISSI2),
+            MedsLipidsChol = ifelse(is.na(MedsLipidsChol), 0, MedsLipidsChol)
         ) %>%
         full_join(
             .,
@@ -68,8 +69,8 @@
         mutate(
             VN = plyr::mapvalues(VN, c(1, 3, 6), c(0, 1, 2)),
             f.VN = factor(VN, c(0, 1, 2), c('yr0', 'yr3', 'yr6')),
-            Dysgly = IFG + IGT + DM,
-            Sex = plyr::mapvalues(Sex, c('F', ))
+            Dysgly = plyr::mapvalues(as.character(IFG + IGT + DM), c('0', '1'), c('No', 'Yes')),
+            Sex = plyr::mapvalues(Sex, c('F', 'M'), c('Female', 'Male')),
             Ethnicity =
                 plyr::mapvalues(
                     Ethnicity,
@@ -92,7 +93,7 @@
         ) %>%
         arrange(SID, VN)
 
-    print(paste0('Dimensions of working dataset: ',dim(ds)))
+    print(paste0('Working dataset rows are ', dim(ds)[1], ' and columns are ', dim(ds)[2]))
 
     # There are no duplicate rows
     if(any(duplicated(ds[c('SID', 'VN')])))

### 2016-01-27 R/functions.R
Added custom functions for gee

@@ -30,7 +30,7 @@ renaming_outcomes <- function(x) {
 
 renaming_fa <- function(x) {
     x %>%
-        gsub('\\D\\D(\\d\\d)(\\d)', '\\1:\\2', .) %>%
+        gsub('.*(\\d\\d)(\\d)', '\\1:\\2', .) %>%
         gsub('n(\\d)$', 'n-\\1', .) %>%
         gsub('D(\\d\\d)$', 'D-\\1', .) %>%
         gsub('^pct_', '', .) %>%
@@ -48,17 +48,84 @@ renaming_list <- function(x) {
         renaming_outcomes()
 }
 
-analyze_gee <- function(data, y, x, covar, unit) {
-    data %>%
+analyze_gee <- function(data, y, x, covar, unit, adj.p = FALSE, adj.p.method = 'BH', int = FALSE) {
+
+    if (int) {
+        extract_term <- ':'
+    } else {
+        extract_term <- 'Xterm$'
+    }
+    out <- data %>%
         mason::design('gee', family = gaussian, corstr = 'ar1') %>%
-        mason::lay_base('SID', y, x, covar) %>%
+        {
+            if (int) {
+                mason::lay_base(., 'SID', y, x, covar, intvar = 'VN')
+
+            } else {
+                mason::lay_base(., 'SID', y, x, covar)
+            }
+        } %>%
         mason::build() %>%
         mason::polish(
-            'Xterm$', adjust.p = FALSE,
+            extract_term, adjust.p = FALSE,
             transform.beta.funs = function(x)
                 (exp(x) - 1) * 100,
             rename.vars.funs = renaming_list
         ) %>%
-        dplyr::mutate(unit = as.factor(unit))
+        dplyr::mutate(Xterms = Xterms %>% factor(., unique(.)),
+                      unit = as.factor(unit))
+
+    if (adj.p) {
+        out <- out %>%
+            dplyr::mutate(p.value = p.adjust(p.value, method = adj.p.method))
+    }
+
+    return(out)
+
 }
 
+get_gee_data <- function(ds) {
+    full_join(
+    ds %>%
+        select(-matches('pct_ne|^ne'),-ne_Total),
+    ## Scale all the fatty acids
+    ds %>%
+        filter(VN == 0) %>%
+        select(SID, ne_Total, matches('pct_ne|^ne')) %>%
+        mutate_each(funs(as.numeric(scale(.))),-SID),
+    by = 'SID') %>%
+    group_by(VN) %>%
+    mutate(
[…]

### 2016-01-27 R/setup.R
Added more packages to load for setup

@@ -8,6 +8,7 @@ run_setup <- function() {
     # Run the following packages for setup
     .load_packages()
     .set_options()
+    .load_packages()
 }
 
 .load_packages <- function() {
@@ -15,15 +16,22 @@ run_setup <- function() {
     # installed.
     #
     # Add your packages you want to use here, using the following format:
+    # - For pre-bundled packages
+    library(MASS)
+    # - For CRAN packages:
     # if (!require(package)) install.packages('package')
     if (!require(fdrtool)) install.packages('fdrtool')
+    if (!require(captioner)) install.packages('captioner')
     if (!require(dplyr)) install.packages('dplyr')
     if (!require(tidyr)) install.packages('tidyr')
     if (!require(ggplot2)) install.packages('ggplot2')
     if (!require(pander)) install.packages('pander')
     if (!require(knitr)) install.packages('knitr')
     if (!require(DiagrammeR)) install.packages('DiagrammeR')
-    # Or, for github packages:
+    if (!require(MuMIn)) install.packages('MuMIn')
+    if (!require(geepack)) install.packages('geepack')
+    if (!require(magrittr)) install.packages('magrittr')
+    # - For github packages:
     # if (!require(package)) packrat::install_github('username/package')
     if (!require(mason)) {
         if (!require(curl)) install.packages('curl')

### 2016-01-27 doc/manuscript.Rmd
Started manuscript by adding methods, intro outline, and some results

@@
 bibliography: 
+    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
     - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
-    - /home/luke/Documents/bibliography/bib/bib-db/nefaDiabetes.bib
 csl:
@@
 
+Popular hypothesis: obese -> FFA -> IR.. may not be the case
+
 total nefa evidence, well-studied. but there is diversity. no studies on it.
@@
 experimental (probably fairly short term) research.
+
+Association between higher FFA and muscle/hepatic IR, but challenging as adipose
+tissue can also be IR. So is it the higher FFA -> IR or is it IR -> FFA?
 
@@
 
-<!--
-## Load data ##
--->
+```{r setup, collapse=TRUE}
+knitr::opts_knit$set(root.dir = '../')
+knitr::opts_chunk$set(echo = FALSE)
+```
 
-```{r setup, message = FALSE, echo = FALSE}
-invisible(sapply(list.files('R', pattern = '*.R', full.names = TRUE), source))
-run_setup()
+```{r loadData, message=FALSE}
+source('.Rprofile')
+suppressMessages(run_setup())
 load_data()
@@
 ## Include captions below using `captioner` package
-figNums <- captioner(prefix = 'Figure')
-supFigNums <- captioner(prefix = 'Supplementary Figure')
-tabNums <- captioner(prefix = 'Table')
-supTabNums <- captioner(prefix = 'Supplementary Table')
+fig <- captioner(prefix = 'Figure')
+cite_f <- pryr::partial(fig, display = 'cite')
+sfig <- captioner(prefix = 'Supplementary Figure')
+cite_sf <- pryr::partial(sfig, display = 'cite')
+tab <- captioner(prefix = 'Table')
+cite_t <- pryr::partial(tab, display = 'cite')
+stab <- captioner(prefix = 'Supplementary Table')
+cite_st <- pryr::partial(stab, display = 'cite')
+# usage: cite_st('basicChar')
+
+tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 time points.')
+fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006)')
+fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function using generalized estimating equations over the 6 years in the PROMISE cohort.
+Adjusted for {{covariates}}.
+Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
+P-values presented were adjusted for the false discovery rate.')
 ```
@@
 
-The Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort is a longitudinal observational study of
-participants with one or more risk factors for type 2 diabetes
-mellitus, including obesity, hypertension, family history of diabetes,
-and/or a history of gestational diabetes or birth of a macrosomic
-infant.  Participants aged 30 years and older were recruited for the baseline visit from Toronto and London, Ontario, Canada into the PROMISE cohort (n=736)
-between 2004-2006.  Follow-up examinations occur every
-three years, with currently three examination visits completed to date
-(2004-2006, 2007-2009, and 2010-2013).  The current study used data on
-participants who did not have diabetes at baseline, who returned for
-one or more follow-up examinations, and who had samples available for
-fatty acid measurements (n=477).  A diagram of the sample size at each
-visit is shown in `r supFigRef('flow')`.  At each examination,
-participants undergo extensive metabolic characterization,
-anthropometric measurements, and structured questionnaires on
-lifestyle and sociodemographics.  Research ethics approval was
-obtained from Mount Sinai Hospital and the University of Western
-Ontario.  Data collection methods were standardized across the 2
-centres and research nurses were centrally trained.  <!--Over the
-6-yrs, {{n return each visit}} returned for the 6-yr clinic visit
-({{pct loss to followup}}% loss to follow-up).  By the 6-yr visit, {{n
-converted to IFG}} converted into IFG, {{n converted to IGT}}
-converted into IGT, and {{n converted to DM}} into diabetes.-->
+The Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort is a
+longitudinal observational study of participants with one or more risk factors
+for type 2 diabetes mellitus, including obesity, hypertension, family history of
+diabetes, and/or a history of gestational diabetes or birth of a macrosomic 
+infant.  Participants aged 30 years and older were recruited for the baseline
+visit from Toronto and London, Ontario, Canada into the PROMISE cohort (n=736) 
+between 2004-2006.  Follow-up examinations occur every three years, with
+currently three examination visits completed to date (2004-2006, 2007-2009, and
+2010-2013).  The current study used data on participants who did not have
+diabetes at baseline, who returned for one or more follow-up examinations, and
+who had samples available for fatty acid measurements (n=477).  A diagram of the
+sample size at each visit is shown in `r supFigRef('flow')`.  At each
+examination, participants undergo extensive metabolic characterization, 
+anthropometric measurements, and structured questionnaires on lifestyle and
+sociodemographics.  Research ethics approval was obtained from Mount Sinai
+Hospital and the University of Western Ontario.  Data collection methods were
+standardized across the 2 centres and research nurses were centrally trained. 
+<!--Over the 6-yrs, {{n return each visit}} returned for the 6-yr clinic visit 
+({{pct loss to followup}}% loss to follow-up).  By the 6-yr visit, {{n converted
+to IFG}} converted into IFG, {{n converted to IGT}} converted into IGT, and {{n
+converted to DM}} into diabetes.-->
 
@@
 taken and then had a 75g oral glucose tolerance test (OGTT) with a 30 minute and
-2 hour blood draw. All blood samples were processed and frozen at -70°C. 
-Alanine
-aminotransferase (ALT) was measured using standard laboratory
-procedures.  Cholesterol, HDL, and triacylglycerols (TAG) were
-measured using Roche Modular's enzymatic colometric tests
-(Mississauga, ON).  Both specific insulin and glucose were derived
-from the OGTT at fasting, 30 minute, and 2 hour time points.  Specific
-insulin was measured using the Elecsys 1010 (Roche Diagnostics, Basel,
-Switzerland) immunoassay analyzer and electrochemiluminescence
-immunoassay.  This assay shows 0.05% cross-reactivity to intact human
-pro-insulin and the Des 31,32 circulating split form (Linco Res. Inc),
-and has a CV of 9.3%.  Glucose was determined using an enzymatic
-hexokinase (Roche Modular, Roche Diagnostics) with a detection range
-of 0.11 (2 mg/dL) to 41.6 mmol/L.  The inter-assay %CV is <1.1% and
-intra-assay %CV is < 1.9%.  All assays were performed at the Banting
-and Best Diabetes Centre Core Lab at the University of Toronto.
-Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and
-diabetes were categorized using the 2006 WHO criteria [@WHO2006].
+2 hour blood draw. All blood samples were processed and frozen at -70°C. Alanine
+aminotransferase (ALT) was measured using standard laboratory procedures. 
+Cholesterol, HDL, and triacylglycerols (TAG) were measured using Roche Modular's
+enzymatic colometric tests (Mississauga, ON).  Both specific insulin and glucose
+were derived from the OGTT at fasting, 30 minute, and 2 hour time points. 
+Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics, Basel, 
+Switzerland) immunoassay analyzer and electrochemiluminescence immunoassay. 
+This assay shows 0.05% cross-reactivity to intact human pro-insulin and the Des
+31,32 circulating split form (Linco Res. Inc), and has a CV of 9.3%.  Glucose
+was determined using an enzymatic hexokinase (Roche Modular, Roche Diagnostics)
+with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The inter-assay %CV is
+<1.1% and intra-assay %CV is < 1.9%.  All assays were performed at the Banting 
+and Best Diabetes Centre Core Lab at the University of Toronto. Impaired fasting
+glucose (IFG), impaired glucose tolerance (IGT), and diabetes were categorized
+using the 2006 WHO criteria [@WHO2006].
 
@@
 
-After running the GEE models for the longitudinal analysis, linear discriminant
-analysis (LDA) {{or QDA}} was used to identify the patterns of NEFA composition for
-tertiles of the metabolic disorders. Jacknifed cross-validation method used in
-the LDA analysis. Only the baseline visit data was used for the LDA analysis and
-only the outcome and predictor variables could be included in the analysis.
+Sensitivity analysis adjusting for clinically measured TG, as TG at the 6 year
+visit was only done on a subset of participants.
+
+While the GEE modeling takes into consideration the longitudinal design of the
+data, a limitation is that it can not analyze the inherent multivariate nature
+of the composition of the NEFA fraction. As such, after running the GEE models
+for the longitudinal analysis, linear discriminant analysis (LDA) {{or QDA}} was
+used to identify the patterns of NEFA composition for tertiles of the metabolic
+disorders. LDA is not suited for longitudinal data and so only the baseline
+visit data was used and only the outcome and predictor variables could be
+included in the analysis. Data from the later visit dates were used to confirm
+the results of the LDA analysis.
 
@@
 
-# Results #
+# Results
 
-# Discussion #
+```{r geeData, cache=TRUE}
+
+outcomes_is <- c('linvHOMA', 'lISI')
+outcomes_bcf <- c('lIGIIR', 'lISSI2')
+outcomes <- c(outcomes_is, outcomes_bcf)
+ne_pct <- grep('^pct_ne', names(ds), value = TRUE)
+ne_conc <- c('ne_Total', grep('^ne\\d\\d', names(ds), value = TRUE))
+covariates_is <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT')
+covariates_bcf <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 
+                    'MET', 'AlcoholPerWk')
+
+gee_df <- get_gee_data(ds)
+gee_results <- bind_rows(
+    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', adj.p = TRUE)
+    )
+
+gee_results_unadj <- bind_rows(
+    analyze_gee(gee_df, outcomes, ne_pct, 'VN', 'mol%', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes, ne_conc, 'VN', 'nmol/mL', adj.p = TRUE)
+    ) %>% 
+    mutate(Yterms = factor(Yterms, unique(Yterms)))
+```
+
+```{r ldaData}
+# include also the 3-yr and 6-yr data for training/testing.
+lda_df <- ds %>%
+    filter(VN == 0) %>%
+    mutate(BCF = rstatsToolkit::aide.tertile(lISSI2) %>%
+               factor(., labels = c('1st tertile: log(ISSI-2)', 
+                                    '2nd tertile: log(ISSI-2)', 
+                                    '3rd tertile: log(ISSI-2)'), 
+                      ordered = TRUE)) %>%
+    select(BCF, matches('^ne\\d\\d')) %>%
+    mutate_each(funs(as.numeric(scale(.))), -BCF) %>%
+    na.omit() %>% 
+    arrange(BCF)
+lda_results <- lda(BCF ~ ., data = lda_df)
+```
+
+```{r resultsInline, dependson='geeData'}
+# extract gee beta coefficient for indicating magnitude
+pctChg.o <- ds %>%
+  select(f.VN, HOMA, ISI, IGIIR, ISSI2) %>%
+  gather(Measure, Value, -f.VN) %>%
+  na.omit() %>%
+  group_by(Measure, f.VN) %>%
+  summarise(med = median(Value)) %>%
+  spread(f.VN, med) %>%
+  mutate(pctChg = ((`6-yrs` - `0-yr`) / `0-yr`) * 100) %>%
+  select(pctChg) %>%
+  abs() %>%
+  round(1) %>%
+  {paste0(min(.), '% to ', max(.), '%')}
+
+cumPctPL <- calcCumPctFA(ds, 'pct_pl', 90)
+calcCumPctFA <- function(data, variable.pattern, filter.value) {
+    data %>%
+      filter(VN == 0) %>%
+      select(matches(variable.pattern)) %>%
+      gather(fat, value) %>%
+      group_by(fat) %>%
+      summarise(pct = mean(value, na.rm = TRUE)) %>%
+      arrange(desc(pct)) %>%
+      mutate(cumpct = cumsum(pct) %>% round(1) %>% format(nsmall = 1),
+             pct = round(pct, 1) %>% format(nsmall = 1)) %>%
+      filter(cumpct <= filter.value)
+}
+extractBeta <- function(data) {
+    data %>%
+      mutate(estCI = paste0('(beta: ', trim(format(round(estimate, 1), nsmall = 1)),
+                            ' CI: ',
+                            trim(format(round(conf.low, 1), nsmall = 1)),
+                            ', ',
+                            trim(format(round(conf.high, 1), nsmall = 1)),
+                            ')'),
+             #indep.estCI = paste(indep, estCI),
+             #dep = paste(dep %>% gsub('log\\((.*)\\)', '\\1', .), estCI),
+             dep = paste(dep %>% gsub('log\\((.*)\\)', '\\1', .))
+             ) %>%
+      filter(p.value <= 0.05) %>%
+      arrange(dep, desc(estimate)) %>%
+      select(dep, indep, estCI)
+}
+```
+
+## Basic characteristics of the PROMISE cohort
+
+## Distribution of NEFA
+
+## Correlations of NEFA with basic characteristics
+
+## GEE
+
+As shown in {{supFigGeeUnadj}}, several associations were seen in the unadjusted
+GEE models, particularly for the concentration data. Full model adjustment
+attenuated most of these associations, with total NEFA, 16:0, and 18:2n-6 (all
+as nmol/mL) had negative associations between IGI/IR and ISSI-2 ({{figGee}}).
+The magnitude of association for each of these variables was fairly consistent
+for each beta-cell function measure. For every one SD decrease in any of these
+three NEFA variables, there was a predicted {{magnitude}}% lower IGI/IR and a
+{{magnitude}}% lower ISSI-2 at each clinic visit. None of the NEFA variables as
+a mol% had any association with the outcomes.
+
+Adjusting for TAG attenuates all associations with IGI/IR and ISSI-2 (data not
+shown). This may be due to either or both of two factors, mediation of TAG on
+the pathway between NEFA and beta-cell function or the lower sample size at the
+6 year visit from the incomplete measurement of TAG due to cuts in funding.
+
+There were no significant interaction effects between time and the individual
+NEFA on any of the outcome measures (data not shown).
+
+## Discriminant analysis
+
+# Discussion
+
+Brief overall conclusion...
+
+Given we found associations with the nmol/mL data, but none in the mol% data,
+this suggests to us that the total size of the NEFA fraction is more important
+in the pathogenesis of diabetes than is the specific FA composition. A
+limitation of the mol% data is the inherent dependency between the FA as an
+increase in one FA by definition means other FA must decrease. In the context of
+this dependency, the associations we found for 16:0 and 18:2n-6 as nmol/mL
+suggests that these NEFA may be contributing to the harmful association seen for
+the total NEFA fraction, irrespective of changes in the other NEFA.
+
+We found that higher palmitic acid (16:0) and linoleic acid (18:2n-6) associated
+with lower beta-cell function by {{number?}}. Expand on experimental studies,
+pathology/biology, mechanisms, etc.
+
+There is substantial experimental research on the role of increasing NEFA and
+increasing IR {{cite}}. Two major limitation of these studies are that they tend
+to be fairly short term ({{length of time}}) and they tend to increase NEFA to
+supraphysiological concentrations {{confirm}}. Epidemiological research has also
+shown a link between higher NEFA and higher IR {{cite}}. However, most of these
+studies are cross-sectional in nature and can not address causal direction.
+Adipose tissue can become insulin resistant and thus hormone-sensitive lipase
+may not respond to the inhibitory action of insulin {{cite}}. As such, IR may
+temporally occur sooner than elevations in blood NEFA. In this longitudinal
+analysis, we saw no association between NEFA composition and neither hepatic IR
+(HOMA-IR) nor whole-body insulin sensitivity (ISI). Possible explanations for
+this could be that over several years NEFA does not influence hepatic or muscle
+IR or that it is IR that elevates NEFA and as such NEFA would not predict IR
+over time.
+
+In the sensitivity analysis adjusting for TAG, all associations with beta-cell
+function were attenuated. While there is strong experimental evidence on the
+role of chronically elevated NEFA, particularly palmitic acid, on impaired
+beta-cell function, within a free-living population NEFA may be mediating it's
+association through TAG. In normal metabolism, NEFA enters the liver and assists
+in the production of TAG that is to be processed into very-low density
+lipoproteins (VLDL) {{cite}}. Elevated NEFA may be mediating it's association
+through hypertriglyceridemia, a known risk factor for diabetes {{cite}}. A major
+caveat with the TAG-adjusted GEE models is the reduced sample size in the 6 year
+visit, as only a subset (n={{sample}}) of participants had TAG measured.
+
+Comment on LDA analysis... ...LDA backs up some of the GEE model results,
+showing weak discriminatory ability of the NEFA pattern on identifying beta-cell
+function... ...While the overall discriminatory ability of the NEFA pattern was
+poor, there were certain NEFA that loaded differently on each level of beta-cell
+function....
+
+Limitations, etc...
+
+Final, brief conclusion...
 
@@
               ylab = 'Non-esterified fatty acid') %>% 
-    vision_simple(base_size = 10)
+    vision_simple() +
+    theme(axis.ticks.y = element_blank())
 ```
 
-```{r geeData}
-
-outcomes <- c('linvHOMA', 'lISI', 'lIGIIR', 'lISSI2')
-ne_pct <- grep('^pct_ne', names(ds), value = TRUE)
-ne_conc <- c('ne_Total', grep('^ne\\d\\d', names(ds), value = TRUE))
-covariates <- c('Waist', 'VN', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'FamHistDiab')
-
-gee_df <- full_join(
-    ds %>%
-        select(-matches('pct_ne|^ne'),-ne_Total),
-    ## Scale all the fatty acids
-    ds %>%
-        filter(VN == 0) %>%
-        select(SID, ne_Total, matches('pct_ne|^ne')) %>%
-        mutate_each(funs(as.numeric(scale(.))),-SID),
-    by = 'SID') %>%
-    group_by(VN) %>%
-    mutate(
-        Waist = as.numeric(scale(Waist)),
-        ALT = as.numeric(scale(ALT)),
-        BaseAge = as.numeric(scale(BaseAge))
-        ) %>% 
-    ungroup() %>%
-    arrange(SID, VN)
-
-gee_results <- bind_rows(
-    analyze_gee(gee_df, outcomes, ne_pct, covariates, 'mol%'),
-    analyze_gee(gee_df, outcomes, ne_conc, covariates, 'nmol/mL')
-    ) %>% 
-    mutate()
-
-gee_results_unadj <- bind_rows(
-    analyze_gee(gee_df, outcomes, ne_pct, covariates, 'mol%'),
-    analyze_gee(gee_df, outcomes, ne_conc, covariates, 'nmol/mL')
-    )
-
+```{r figGEE, dependson='geeData'}
+forest_plot(gee_results)
 ```
 
-```{r figGee}
-gee_results %>%
-    trance('main_effect') %>%
-    visualize(groups = 'unit~Yterms',
-              xlab = 'Percent difference with 95% CI in the outcomes\nfor each SD increase in fatty acid',
-              ylab = 'Non-esterified fatty acids') %>% 
-    vision_simple(legend_position = 'bottom') +
-    theme(
-        axis.text.y = element_text(face = ifelse(
-        gee_results$Xterms == 'Total', "bold", "plain"
-        )),
-        legend.key.width = grid::unit(0.75, "line"),
-        legend.key.height = grid::unit(0.75, "line"),
-        panel.margin = grid::unit(0.75, "lines")
-        ) +
-    scale_alpha_discrete(name = 'P-value', range = c(0.4, 1.0))
+```{r figLDA}
+visualize(lda_results, 'means', label.rename.fun = renaming_fa) %>% 
+    vision_simple() +
+    labs(y = 'Non-esterified fatty acids',
+         x = 'Standardized mean of the fatty acid (SD of the nmol/mL)')
 ```
@@
 
-```{r supTabQIC}
+```{r supFigGeeUnadj, dependson='geeData'}
+forest_plot(gee_results_unadj)
+```
 
-subds <- ds %>%
-  select(SID, VN, lISSI2, pl_Total, Sex, Ethnicity, MET, ALT,
-         BaseAge, Waist, BMI, AlcoholPerWk, TobaccoUse, SelfEdu,
-         Occupation, totalNE, FamHistDiab) %>%
-  mutate(SelfEdu = as.factor(SelfEdu),
-         Occupation = as.factor(Occupation),
-         TobaccoUse = as.factor(TobaccoUse),
-         AlcoholPerWk = as.factor(AlcoholPerWk)) %>%
-  filter(complete.cases(.))
-
-levels(subds$SelfEdu)[1:2] <- c('2', '2')
-subds$SelfEdu <- droplevels(subds$SelfEdu)
-
-M0 <- geeglm(lISSI2 ~ pl_Total + VN, data = subds,
-                id = SID, family = gaussian,
-                corstr = 'exchangeable')
-M1 <- update(M0, . ~ . + Sex + Ethnicity + BaseAge)
-M2 <- update(M0, . ~ . + Sex + Ethnicity + BaseAge + Waist)
-M3 <- update(M0, . ~ . + Sex + Ethnicity + BaseAge + BMI)
-M4 <- update(M0, . ~ . + Sex + Ethnicity + BaseAge + Waist + MET +
-               AlcoholPerWk + TobaccoUse + SelfEdu + Occupation +
-               FamHistDiab)
-M5 <- update(M0, . ~ . + Sex + Ethnicity + BaseAge + MET + Waist +
-               AlcoholPerWk + TobaccoUse + SelfEdu + Occupation +
-               FamHistDiab + totalNE + ALT)
-M6 <- update(M0, . ~ . + Sex + Ethnicity + Waist + totalNE)
-M7 <- update(M0, . ~ . + Sex + Ethnicity + Waist + totalNE + ALT)
-M8 <- update(M0, . ~ . + Sex + Ethnicity + Waist + totalNE + ALT + BaseAge)
-M9 <- update(M0, . ~ . + Sex + Ethnicity + Waist + totalNE + ALT + BaseAge + FamHistDiab)
-
-set.caption(supTabRef('tabQIC', 'Comparing generalized estimating equation
-models adjusting for different covariates using
-Quasi-Likelihood Information Criterion.'))
-MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M9,
-          rank = QIC) %>%
-  tableModelSelection()
-
-tableModelSelection <- function(selection.data) {
-    selection.data %>%
-      add_rownames() %>%
-      select(Model = rowname, QIC, Delta = delta) %>%
-      pander(digits = 2)
-}
+```{r supTabQIC, eval=FALSE}
 
 ```
+
+```{r supTabGEE}
+
+```

### 2016-01-27 doc/gee.Rmd
Added exploring of gee, lda, and qic

---
title: "GEE models"
author: "Luke W. Johnston"
output: 
    rmarkdown::html_vignette:
        keep_md: true
        toc: true
---

# Set up and load dataset

```{r setup, message = TRUE, collapse=TRUE}
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(collapse = TRUE)
```

```{r}
source('.Rprofile')
suppressMessages(run_setup())
load_data()
dim(ds)
ds %>% 
    select(VN, TAG) %>% 
    na.omit() %>% 
    group_by(VN) %>% 
    summarize(n = n())
```

# Prep and wrangle data into GEE form

```{r geeData}
outcomes <- c('linvHOMA', 'lISI', 'lIGIIR', 'lISSI2')
outcomes_is <- c('linvHOMA', 'lISI')
outcomes_bcf <- c('lIGIIR', 'lISSI2')
ne_pct <- grep('^pct_ne', names(ds), value = TRUE)
ne_conc <- c('ne_Total', grep('^ne\\d\\d', names(ds), value = TRUE))
covariates0 <- c('VN')
covariates_is <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT')
covariates_bcf <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 
                    'MET', 'AlcoholPerWk')
covariates_is_tag <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'TAG')
covariates_bcf_tag <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 
                        'ALT', 'MET', 'AlcoholPerWk', 'TAG')

gee_df <- get_gee_data(ds)
```

# Unadjusted GEE model

```{r geeResults_Unadj}
gee_results0 <- bind_rows(
    analyze_gee(gee_df, outcomes, ne_pct, covariates0, 'mol%'),
    analyze_gee(gee_df, outcomes, ne_conc, covariates0, 'nmol/mL')
    )
gee_results0 %>% 
    mutate(p.value = p.adjust(p.value, 'BH')) %>% 
    filter(p.value < 0.1) %>% 
    select(Yterms, Xterms, unit, estimate, p.value) %>% 
    mutate_each(funs(round(., 2)), estimate, p.value) %>% 
    pander('Unadjusted')
```

# Adjusted with or without TAG for IS

```{r geeResults_IS}
gee_results_is <- bind_rows(
    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%'),
    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL'),
    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is_tag, 'mol% (w/ tag)'),
    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is_tag, 'nmol/mL (w/ tag)')
    )
gee_results_is %>%
    group_by(unit) %>% 
    mutate(p.value = p.adjust(p.value, 'BH')) %>% 
    ungroup() %>% 
    filter(p.value < 0.1) %>% 
    select(Yterms, Xterms, unit, estimate, p.value) %>% 
    mutate_each(funs(round(., 2)), estimate, p.value) %>% 
    pander('Without or with TAG for IS')
```

# Adjusted with or without TAG for BCF

```{r geeResults_BCF}
gee_results_bcf <- bind_rows(
    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%'),
    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL'),
    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf_tag, 'mol% (w/ tag)'),
    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf_tag, 'nmol/mL (w/ tag)')
    )
gee_results_bcf %>% 
    group_by(unit) %>% 
    mutate(p.value = p.adjust(p.value, 'BH')) %>% 
    ungroup() %>% 
    filter(p.value < 0.05) %>% 
    select(Yterms, Xterms, unit, estimate, p.value) %>% 
    mutate_each(funs(round(., 2)), estimate, p.value) %>% 
    pander('Without or with TAG for BCF')
```

# Interaction for IS

```{r geeResults_IS_int}
gee_results_is_int <- bind_rows(
    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', int = TRUE),
    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', int = TRUE),
    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is_tag, 'mol% (w/ tag)', int = TRUE),
    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is_tag, 'nmol/mL (w/ tag)', int = TRUE)
    )
gee_results_is_int %>%
    group_by(unit) %>% 
    mutate(p.value = p.adjust(p.value, 'BH')) %>% 
    ungroup() %>% 
    select(p.value) %>% 
    plot()
```

# Interaction for BCF

```{r geeResults_BCF_int}
gee_results_bcf_int <- bind_rows(
    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', int = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', int = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf_tag, 'mol% (w/ tag)', int = TRUE),
    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf_tag, 'nmol/mL (w/ tag)', int = TRUE)
    )
gee_results_bcf_int %>% 
    group_by(unit) %>% 
    mutate(p.value = p.adjust(p.value, 'BH')) %>% 
    ungroup() %>% 
    select(p.value) %>% 
    plot()
```

### 2016-01-27 doc/lda.Rmd
Added exploring of gee, lda, and qic

---
title: "LDA exploration"
author: "Luke W. Johnston"
date: "January 19, 2016"
output: 
    rmarkdown::html_vignette:
        toc: true
---

# Load up packages and data

```{r setup, collapse=TRUE}
knitr::opts_knit$set(root.dir = '../')
knitr::opts_chunk$set(collapse = TRUE, fig.height = 8, fig.width = 10)
```

```{r loadData, message=FALSE}
source('.Rprofile')
suppressMessages(run_setup())
load_data()
dim(ds)
```

# Prepare the data for the LDA

```{r ldaData}
lda_ds <- ds %>%
    filter(VN == 0) %>%
    mutate(BCF = rstatsToolkit::aide.tertile(lISSI2) %>%
               factor(., labels = c('1st tertile', '2nd tertile', '3rd tertile'), ordered = TRUE)) %>%
    select(BCF, matches('^ne\\d\\d')) %>%
    mutate_each(funs(as.numeric(scale(.))), -BCF) %>%
    na.omit() %>% 
    arrange(BCF)

# include also the 3-yr and 6-yr data for training/testing.

#fit <- qda(BCF ~ ., data = lda_ds)
fit <- lda(BCF ~ ., data = lda_ds)
fit.cv <- lda(BCF ~ ., data = lda_ds, CV = TRUE)
prop_lda <- round(fit$svd^2/sum(fit$svd^2) * 100, 2)

# proportion explained by each LD
prop_lda
```

```{r testingLDA}
fit.class <- predict(fit, newdata = lda_ds[, -1])$class %>% 
    factor(., ordered = TRUE)
table(fit.class, lda_ds[[1]])
table(fit.cv$class, lda_ds[[1]])
```

```{r FAperBCF}
p_data <- fit$means %>%
    as.data.frame() %>%
    add_rownames('Group') %>%
    gather(FA, Mean,-Group) %>%
    mutate(Group = Group %>% factor(., c('1st tertile', '2nd tertile', '3rd tertile'), ordered = TRUE)) %>%
    arrange(Group, Mean) %>%
    mutate(FA = paste0(Group, FA) %>%
               factor(., unique(.)))
FA.orig <- p_data$FA %>% 
    as.character()
FA.label <- p_data$FA %>%
    gsub('^.*(\\d\\d)(\\d)', '\\1:\\2', .) %>%
    gsub('n(\\d)', 'n-\\1', .) %>%
    as.character()
p_data %>%
    ggplot(aes(x = Mean, y = FA)) +
    geom_point() +
    geom_vline(xintercept = 0) +
    facet_wrap(~ Group, scale = 'free_y') +
    scale_y_discrete(breaks = FA.orig, label = FA.label) +
    ggthemes::theme_tufte(12, 'sans') +
    theme(panel.margin = unit(2, 'lines'),
          panel.grid = element_line(colour = 'black'),
          panel.grid.minor.x = element_blank(),
          panel.grid.major.x = element_blank())

```

```{r FAperLD}
p_data2 <- fit$scaling %>%
    as.data.frame() %>%
    add_rownames('Var') %>%
    gather(LD, Value, -Var) %>%
    arrange(LD, Value) %>% 
    mutate(Var = factor(paste0(LD, Var), unique(paste0(LD, Var))))
FA.orig <- p_data2$Var %>% 
    as.character()
FA.label <- p_data2$Var %>%
    gsub('^.*(\\d\\d)(\\d)', '\\1:\\2', .) %>%
    gsub('n(\\d)', 'n-\\1', .) %>%
    as.character()
p_data2 %>% 
    ggplot(aes(x = Value, y = Var)) +
    geom_point() +
    geom_vline(xintercept = 0) +
    facet_wrap(~LD, scale = 'free_y') +
    scale_y_discrete(breaks = FA.orig, label = FA.label) +
    ggthemes::theme_tufte(12, 'sans') +
    theme(panel.margin = unit(2, 'lines'),
          panel.grid = element_line(colour = 'black'),
          panel.grid.minor.x = element_blank(),
          panel.grid.major.x = element_blank())
```

```{r LDbyBCF}
p_fit <- predict(object = fit, newdata = lda_ds)
# predict on follow up visits too.
p_ds <- data.frame(BCF = lda_ds[,"BCF"], lda = p_fit$x)
ggplot(p_ds, aes(lda.LD1, lda.LD2, group = BCF)) +
    geom_vline(xintercept = 0, linetype = 'dotted') +
    geom_hline(yintercept = 0, linetype = 'dotted') +
    stat_density2d(aes(alpha = ..level.., colour = BCF), size = 2) +
    labs(x = paste0("LD1 (", prop_lda[1], "%)"),
         y = paste0('LD2 (', prop_lda[2], '%)')) +
    ggthemes::theme_tufte(12, 'sans') +
    scale_alpha(guide = 'none') +
    theme(legend.title = element_blank(),
          legend.position = c(1,0),
          legend.justification = c(1,0))
```

### 2016-01-28 doc/lda.Rmd
Added seer to LDA + compared to later visits

@@
 ```{r ldaData}
-lda_ds <- ds %>%
-    filter(VN == 0) %>%
+ds.prep <- ds %>%
+    select(VN, lISSI2, matches('^ne\\d\\d')) %>% 
+    group_by(VN) %>% 
     mutate(BCF = rstatsToolkit::aide.tertile(lISSI2) %>%
                factor(., labels = c('1st tertile', '2nd tertile', '3rd tertile'), ordered = TRUE)) %>%
-    select(BCF, matches('^ne\\d\\d')) %>%
-    mutate_each(funs(as.numeric(scale(.))), -BCF) %>%
+    select(BCF, everything()) %>% 
+    select(-lISSI2) %>% 
+    mutate_each(funs(as.numeric(scale(.))), -BCF, -VN) %>%
+    ungroup()
+
+df_Dysgly <- ds %>% 
+    filter(VN == 0) %>% 
+    mutate(Dysgly = factor(IGT + IFG + DM, labels = c('no', 'yes'))) %>% 
+    select(Dysgly, matches('^ne\\d\\d')) %>%
+    mutate_each(funs(as.numeric(scale(.))), -Dysgly) %>% 
+    na.omit()
+
+df_y1 <- ds.prep %>% 
+    filter(VN == 0) %>%
+    select(-VN) %>% 
+    na.omit() %>%
+    arrange(BCF)
+df_y3 <- ds.prep %>% 
+    filter(VN == 1) %>% 
+    select(-VN) %>% 
+    na.omit() %>% 
+    arrange(BCF)
+df_y6 <- ds.prep %>% 
+    filter(VN == 2) %>% 
+    select(-VN) %>% 
     na.omit() %>% 
@@
 # include also the 3-yr and 6-yr data for training/testing.
-
-#fit <- qda(BCF ~ ., data = lda_ds)
-fit <- lda(BCF ~ ., data = lda_ds)
-fit.cv <- lda(BCF ~ ., data = lda_ds, CV = TRUE)
-prop_lda <- round(fit$svd^2/sum(fit$svd^2) * 100, 2)
-
-# proportion explained by each LD
-prop_lda
+fit_y1 <- lda(BCF ~ ., data = df_y1)
+fit_y3 <- lda(BCF ~ ., data = df_y3)
+fit_y6 <- lda(BCF ~ ., data = df_y6)
+fit_dg <- lda(Dysgly ~ ., data = df_Dysgly)
+# compare with CV lda.
+fitcv_y1 <- lda(BCF ~ ., data = df_y1, CV = TRUE)
+fitcv_y3 <- lda(BCF ~ ., data = df_y3, CV = TRUE)
+fitcv_y6 <- lda(BCF ~ ., data = df_y6, CV = TRUE)
 ```
@@
 ```{r testingLDA}
-fit.class <- predict(fit, newdata = lda_ds[, -1])$class %>% 
+fit.class_y1 <- predict(fit_y1, newdata = df_y1[, -1])$class %>% 
     factor(., ordered = TRUE)
-table(fit.class, lda_ds[[1]])
-table(fit.cv$class, lda_ds[[1]])
+fit.class_y3 <- predict(fit_y3, newdata = df_y3[, -1])$class %>% 
+    factor(., ordered = TRUE)
+fit.class_y6 <- predict(fit_y6, newdata = df_y6[, -1])$class %>% 
+    factor(., ordered = TRUE)
+
+table(fit.class_y1, df_y1[[1]])
+table(fitcv_y1$class, df_y1[[1]])
+
+table(fit.class_y3, df_y3[[1]])
+table(fitcv_y3$class, df_y3[[1]])
+
+table(fit.class_y6, df_y6[[1]])
+table(fitcv_y6$class, df_y6[[1]])
+
+# Amount of participants who were misclassified
+table(fit.class_y1, df_y1[[1]]) %>% 
+    as.data.frame() %>% 
+    filter(fit.class_y1 != Var2) %>% 
+    summarize(count = sum(Freq))
+
+# Amount of participants who were correctly classified
+table(fit.class_y1, df_y1[[1]]) %>% 
+    as.data.frame() %>% 
+    filter(fit.class_y1 == Var2) %>% 
+    fit.class_y1 <- predict(fit_y1, newdata = df_y1[, -1])$class %>% 
+    factor(., ordered = TRUE)summarize(count = sum(Freq))
+
+fit.class_dg <- predict(fit_dg, newdata = df_Dysgly[, -1])$class %>% 
+    factor(., ordered = TRUE)
+table(fit.class_dg, df_Dysgly[[1]])
+# Not enough sample size for dysglycemia
 ```
@@
 ```{r FAperBCF}
-p_data <- fit$means %>%
-    as.data.frame() %>%
-    add_rownames('Group') %>%
-    gather(FA, Mean,-Group) %>%
-    mutate(Group = Group %>% factor(., c('1st tertile', '2nd tertile', '3rd tertile'), ordered = TRUE)) %>%
-    arrange(Group, Mean) %>%
-    mutate(FA = paste0(Group, FA) %>%
-               factor(., unique(.)))
-FA.orig <- p_data$FA %>% 
-    as.character()
-FA.label <- p_data$FA %>%
+seer::visualize(
+    fit_y1, 'means', label.rename.fun = function(x)
+        x %>%
         gsub('^.*(\\d\\d)(\\d)', '\\1:\\2', .) %>%
-    gsub('n(\\d)', 'n-\\1', .) %>%
-    as.character()
-p_data %>%
-    ggplot(aes(x = Mean, y = FA)) +
-    geom_point() +
-    geom_vline(xintercept = 0) +
-    facet_wrap(~ Group, scale = 'free_y') +
-    scale_y_discrete(breaks = FA.orig, label = FA.label) +
-    ggthemes::theme_tufte(12, 'sans') +
-    theme(panel.margin = unit(2, 'lines'),
-          panel.grid = element_line(colour = 'black'),
-          panel.grid.minor.x = element_blank(),
-          panel.grid.major.x = element_blank())
-
+        gsub('n(\\d)', 'n-\\1', .)
+    ) %>% 
+    seer::vision_simple()
+seer::visualize(
+    fit_y3, 'means', label.rename.fun = function(x)
+        x %>%
+        gsub('^.*(\\d\\d)(\\d)', '\\1:\\2', .) %>%
+        gsub('n(\\d)', 'n-\\1', .)
+    ) %>% 
+    seer::vision_simple()
+seer::visualize(
+    fit_y6, 'means', label.rename.fun = function(x)
+        x %>%
+        gsub('^.*(\\d\\d)(\\d)', '\\1:\\2', .) %>%
+        gsub('n(\\d)', 'n-\\1', .)
+    ) %>% 
+    seer::vision_simple()
 ```
@@
 ```{r FAperLD}
-p_data2 <- fit$scaling %>%
-    as.data.frame() %>%
-    add_rownames('Var') %>%
-    gather(LD, Value, -Var) %>%
-    arrange(LD, Value) %>% 
-    mutate(Var = factor(paste0(LD, Var), unique(paste0(LD, Var))))
-FA.orig <- p_data2$Var %>% 
-    as.character()
-FA.label <- p_data2$Var %>%
+seer::visualize(
+    fit_y1, 'scaling', label.rename.fun = function(x)
+        x %>%
         gsub('^.*(\\d\\d)(\\d)', '\\1:\\2', .) %>%
-    gsub('n(\\d)', 'n-\\1', .) %>%
-    as.character()
-p_data2 %>% 
-    ggplot(aes(x = Value, y = Var)) +
-    geom_point() +
-    geom_vline(xintercept = 0) +
-    facet_wrap(~LD, scale = 'free_y') +
-    scale_y_discrete(breaks = FA.orig, label = FA.label) +
-    ggthemes::theme_tufte(12, 'sans') +
-    theme(panel.margin = unit(2, 'lines'),
-          panel.grid = element_line(colour = 'black'),
-          panel.grid.minor.x = element_blank(),
-          panel.grid.major.x = element_blank())
+        gsub('n(\\d)', 'n-\\1', .)
+    ) %>% 
+    seer::vision_simple()
+seer::visualize(
+    fit_y3, 'scaling', label.rename.fun = function(x)
+        x %>%
+        gsub('^.*(\\d\\d)(\\d)', '\\1:\\2', .) %>%
+        gsub('n(\\d)', 'n-\\1', .)
+    ) %>% 
+    seer::vision_simple()
+seer::visualize(
+    fit_y6, 'scaling', label.rename.fun = function(x)
+        x %>%
+        gsub('^.*(\\d\\d)(\\d)', '\\1:\\2', .) %>%
+        gsub('n(\\d)', 'n-\\1', .)
+    ) %>% 
+    seer::vision_simple()
 ```
@@
 ```{r LDbyBCF}
-p_fit <- predict(object = fit, newdata = lda_ds)
-# predict on follow up visits too.
-p_ds <- data.frame(BCF = lda_ds[,"BCF"], lda = p_fit$x)
-ggplot(p_ds, aes(lda.LD1, lda.LD2, group = BCF)) +
-    geom_vline(xintercept = 0, linetype = 'dotted') +
-    geom_hline(yintercept = 0, linetype = 'dotted') +
-    stat_density2d(aes(alpha = ..level.., colour = BCF), size = 2) +
-    labs(x = paste0("LD1 (", prop_lda[1], "%)"),
-         y = paste0('LD2 (', prop_lda[2], '%)')) +
-    ggthemes::theme_tufte(12, 'sans') +
-    scale_alpha(guide = 'none') +
-    theme(legend.title = element_blank(),
-          legend.position = c(1,0),
-          legend.justification = c(1,0))
+seer::visualize(fit_y1, orig.data = df_y1, group.name = 'BCF') %>% 
+    seer::vision_simple()
+seer::visualize(fit_y3, orig.data = df_y3, group.name = 'BCF') %>% 
+    seer::vision_simple()
+seer::visualize(fit_y6, orig.data = df_y6, group.name = 'BCF') %>% 
+    seer::vision_simple()
 ```

### 2016-01-28 R/functions.R
Added function for plot of dist, calc pct, and getting lda data

@@ -111,7 +111,25 @@ get_gee_data <- function(ds) {
     arrange(SID, VN)
 }
 
-forest_plot <- function(data) {
+get_lda_data <- function(data) {
+    data %>%
+        filter(VN == 0) %>%
+        mutate(BCF = rstatsToolkit::aide.tertile(lISSI2) %>%
+                   factor(
+                       ., labels = c(
+                           '1st tertile: log(ISSI-2)',
+                           '2nd tertile: log(ISSI-2)',
+                           '3rd tertile: log(ISSI-2)'
+                       ),
+                       ordered = TRUE
+                   )) %>%
+        select(BCF, matches('^ne\\d\\d')) %>%
+        mutate_each(funs(as.numeric(scale(.))),-BCF) %>%
+        na.omit() %>%
+        arrange(BCF)
+}
+
+plot_gee_results <- function(data) {
     data %>%
         seer::trance('main_effect') %>%
         seer::visualize(groups = 'unit~Yterms',
@@ -128,4 +146,37 @@ forest_plot <- function(data) {
             strip.background = element_blank()
         ) +
         ggplot2::scale_alpha_discrete(name = 'P-value', range = c(0.4, 1.0))
+}
+
+plot_nefa_distribution <- function(data) {
+    data %>%
+        dplyr::filter(VN == 0) %>%
+        dplyr::select(SID, matches('^ne\\d')) %>%
+        dplyr::filter(complete.cases(.)) %>%
+        tidyr::gather(Measure, Value,-SID) %>%
+        dplyr::mutate(Measure = renaming_fa(Measure) %>%
+                   gsub('ne_', '', .) %>%
+                   factor(., levels = unique(.))) %>%
+        tidyr::spread(Measure, Value) %>%
+        dplyr::select(-SID) %>%
+        seer::trance('boxes_dots') %>%
+        serr::visualize(dots = FALSE,
+                  xlab = 'Concentration (nmol/mL)',
+                  ylab = 'Non-esterified fatty acid') %>%
+        seer::vision_simple() +
+        ggplot2::theme(axis.ticks.y = element_blank())
+
+}
+
+calculate_percent_contribution <- function(data) {
+    data %>%
+        filter(VN == 0) %>%
+        select(matches('pct_ne')) %>%
+        gather(fat, value) %>%
+        group_by(fat) %>%
+        summarise(pct = round(mean(value, na.rm = TRUE), 1)) %>%
[…]

### 2016-01-28 doc/manuscript.Rmd
Added results for lda, dev discuss, added caps and figs

@@
 harm, irrespective of whether other FA increase or decrease as well?
-- need to do LDA or some such to id the pattern of why... (don't show this in
-paper, but examine PCA of it too)
 
@@
 - Adjust for TAG in IR models?
+- Key is that this is fasting, not postprandial!
 
@@
 P-values presented were adjusted for the false discovery rate.')
+fig_lda <- fig('lda', 'Linear discriminant analysis to identify patterns of NEFA composition between tertiles of log(ISSI-2) in 463 PROMISE participants at the baseline visit (2004-2006).')
+
+sfig_consort <- sfig('consort', 'CONSORT diagram....')
+sfig_lda <- sfig('lda', 'Linear discriminant factors with explained variance against tertiles of log(ISSI-2) in 463 PROMISE participants at the baseline visit (2004-2006).')
 ```
@@
 who had samples available for fatty acid measurements (n=477).  A diagram of the
-sample size at each visit is shown in `r supFigRef('flow')`.  At each
+sample size at each visit is shown in `r cite_sf('consort')`.  At each 
 examination, participants undergo extensive metabolic characterization, 
@@
 
-At each clinic examination, participants had an 7-12 hour fasting blood sample 
-taken and then had a 75g oral glucose tolerance test (OGTT) with a 30 minute and
-2 hour blood draw. All blood samples were processed and frozen at -70°C. Alanine
-aminotransferase (ALT) was measured using standard laboratory procedures. 
-Cholesterol, HDL, and triacylglycerols (TAG) were measured using Roche Modular's
+At each examination, participants had an 8-12 hour fasting blood sample 
+taken followed by a 75g oral glucose tolerance test (OGTT) with a 30 minute and
+2 hour blood draw. All blood samples were processed and frozen at -70°C. {{Alanine
+aminotransferase (ALT) was measured using standard laboratory procedures.}} 
+Cholesterol, HDL, and triglycerides (TG) were measured using Roche Modular's
 enzymatic colometric tests (Mississauga, ON).  Both specific insulin and glucose
@@
 Insulin sensitivity and beta-cell function indices were computed using the OGTT 
-glucose and insulin.  Insulin sensitivity was assessed using the inverted
-homeostasis model assessment of insulin resistance (1/HOMA-IR) [@Matthews1985]
-and the Insulin Sensitivity Index (ISI) [@Matsuda1999]. HOMA-IR largely reflects
-hepatic insulin resistance, while ISI reflects whole-body insulin sensitivity
-[@AbdulGhani2007].  Beta-cell function was assessed using the Insulinogenic
-Index [@Wareham1995] over HOMA-IR (IGI/IR) and the Insulin Secretion-Sensitivity
-Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR is a measure of the first phase
-insulin secretion while ISSI-2 is analogous to the disposition index. Each index
-has been validated against gold standard measures [@Matthews1985; @Matsuda1999;
-@Retnakaran2009].
+glucose and insulin.  Insulin sensitivity was assessed using 1/HOMA-IR (1
+divided by HOMA-IR) [@Matthews1985] and the Insulin Sensitivity Index (ISI)
+[@Matsuda1999]. HOMA-IR largely reflects hepatic insulin resistance, while ISI
+reflects whole-body insulin sensitivity [@AbdulGhani2007].  Beta-cell function
+was assessed using the Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR)
+and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR
+is a measure of the first phase insulin secretion while ISSI-2 is analogous to
+the disposition index. Each index has been validated against gold standard
+measures [@Matthews1985; @Matsuda1999; @Retnakaran2009].
 
@@
 defined as the narrowest part of the torso between the umbilicus and the xiphoid
-process.  Height was measured using a stadiometer and weight was measured on a 
-medical balance beam scale. BMI was calculated by dividing weight (kg) by height
-(m) squared. Sociodemographic information, such as age, sex, and ethnicity, were
-determined using standardized questionnaires administered at each examination
-visit.
+process. BMI was calculated by dividing weight (kg) by height (m) squared.
+Sociodemographic information, such as age, sex, and ethnicity, were determined
+using questionnaires administered at each examination.
 
@@
 
-Covariates selection was based on previous literature, suggestions from directed
-acyclic graph modeling [@Greenland1999a], and the quasi-likelihood information
-criterion (QIC). Various models compared using QIC are shown in **`r supTabRef('tabQIC')`**.  
-The final GEE model (M#) we selected had time, sex,
-ethnicity, baseline age, waist circumference, {{...}}, and family history of
-diabetes as covariates.  GEE parameter estimates were exponentiated.
-<!--(exponentiated, minus 1, and multiplied by 100)--> After scaling,
-log-transforming, and exponentiating, the GEE parameter estimates are
-interpreted as an expected percent difference in the outcome variable for every
-SD increase in the predictor variable given the covariates are held constant. 
-An time by predictor term was included in each of the GEE models to test for
-time interactions.
+Covariates selection was based on previous literature, directed acyclic graph
+[@Greenland1999a] recommendations, and quasi-likelihood information criterion
+(QIC). `r cite_sf('qic')` shows the covariates compared using QIC and 
+`r cite_sf('dag')` shows the specified directed acyclic graph. The final GEE model
+(M#) we selected had time, sex, ethnicity, baseline age, waist circumference,
+{{...}}, and family history of diabetes as covariates. <!--(exponentiated, minus
+1, and multiplied by 100)--> After scaling, log-transforming, and exponentiating
+the GEE estimates, the GEE results are interpreted as an expected percent
+difference in the outcome variable for every SD increase in the predictor
+variable given the covariates are held constant. An time by predictor term was
+included in each of the GEE models to test for time interactions. Given that TG
+is a risk factor for diabetes and that NEFA contribute to TG production, TG may
+act as a mediator between NEFA and the outcomes. However, only a subset of the
+participants at the 6-yr visit had clinically measured TG. To determine the role
+of TG in the association between NEFA and the outcomes, TG was included in the
+GEE model as a sensitivity analysis.
 
-Sensitivity analysis adjusting for clinically measured TG, as TG at the 6 year
-visit was only done on a subset of participants.
+While GEE controls for the longitudinal design of the data, a limitation is that
+it can not analyze the inherent multivariate nature of the composition of the
+NEFA fraction. As such, after running the GEE models for the longitudinal
+analysis, linear discriminant analysis (LDA) was used to identify the patterns
+of NEFA composition for tertiles of the metabolic disorders. Since LDA can't
+handle longitudinal data, only the baseline data was used and only the outcome
+and predictor variables could be included in the analysis. Data from the later
+visit dates were used to confirm the results of the LDA analysis.
 
-While the GEE modeling takes into consideration the longitudinal design of the
-data, a limitation is that it can not analyze the inherent multivariate nature
-of the composition of the NEFA fraction. As such, after running the GEE models
-for the longitudinal analysis, linear discriminant analysis (LDA) {{or QDA}} was
-used to identify the patterns of NEFA composition for tertiles of the metabolic
-disorders. LDA is not suited for longitudinal data and so only the baseline
-visit data was used and only the outcome and predictor variables could be
-included in the analysis. Data from the later visit dates were used to confirm
-the results of the LDA analysis.
-
-All statistical tests were computed using R `r getRversion()`
-[@Rbase], along with the R packages geepack `r packageVersion('geepack')` for
-the GEE modeling [@Hoejsgaard2006a] and MASS `r packageVersion('MASS')` for LDA.
-The R code for this manuscript is available at {{link}}. Results were
-statistically significance at p<0.05, after adjusting for the False Discovery
-Rate [@Benjamini1995a] for multiple testing.
+All analyses were performed using R `r getRversion()` [@Rbase], along with the R
+packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a] and 
+MASS `r packageVersion('MASS')` for LDA. The R code for this manuscript is
+available at {{link}}. Results were statistically significance at p<0.05, after
+adjusting for multiple testing using the False Discovery Rate [@Benjamini1995a].
 
@@
 ```{r geeData, cache=TRUE}
-
 outcomes_is <- c('linvHOMA', 'lISI')
@@
 # include also the 3-yr and 6-yr data for training/testing.
-lda_df <- ds %>%
-    filter(VN == 0) %>%
-    mutate(BCF = rstatsToolkit::aide.tertile(lISSI2) %>%
-               factor(., labels = c('1st tertile: log(ISSI-2)', 
-                                    '2nd tertile: log(ISSI-2)', 
-                                    '3rd tertile: log(ISSI-2)'), 
-                      ordered = TRUE)) %>%
-    select(BCF, matches('^ne\\d\\d')) %>%
-    mutate_each(funs(as.numeric(scale(.))), -BCF) %>%
-    na.omit() %>% 
-    arrange(BCF)
+lda_df <- get_lda_data(ds)
 lda_results <- lda(BCF ~ ., data = lda_df)
@@
 
-```{r resultsInline, dependson='geeData'}
+```{r inlineGEE, dependson='geeData'}
 # extract gee beta coefficient for indicating magnitude
@@
   {paste0(min(.), '% to ', max(.), '%')}
+```
 
-cumPctPL <- calcCumPctFA(ds, 'pct_pl', 90)
-calcCumPctFA <- function(data, variable.pattern, filter.value) {
-    data %>%
-      filter(VN == 0) %>%
-      select(matches(variable.pattern)) %>%
-      gather(fat, value) %>%
-      group_by(fat) %>%
-      summarise(pct = mean(value, na.rm = TRUE)) %>%
-      arrange(desc(pct)) %>%
-      mutate(cumpct = cumsum(pct) %>% round(1) %>% format(nsmall = 1),
-             pct = round(pct, 1) %>% format(nsmall = 1)) %>%
-      filter(cumpct <= filter.value)
-}
+```{r inlineLDA}
+fit.class <- predict(lda_results, newdata = lda_df[, -1])$class %>% 
+    factor(., ordered = TRUE)
+
+# Amount of participants who were misclassified
+misclass <- table(fit.class, lda_df[[1]]) %>% 
+    as.data.frame() %>% 
+    filter(fit.class != Var2) %>% 
+    summarize(count = sum(Freq)) %>% 
+    as.numeric()
+
+# Amount of participants who were correctly classified
+good_class <- table(fit.class, lda_df[[1]]) %>% 
+    as.data.frame() %>% 
+    filter(fit.class == Var2) %>% 
+    summarize(count = sum(Freq)) %>% 
+    as.numeric()
+
+misclass_npct <- paste0(misclass, ' (', round(misclass / (good_class + misclass) * 100, 0), '%)')
+```
+
+```{r inlineMisc}
+pctne <- calculate_percent_contribution(ds)
+
 extractBeta <- function(data) {
@@
 ## Distribution of NEFA
+
+`r cite_f('nefaDist')` shows the distribution of the NEFA composition. Four
+individual NEFA made up the majority (`r sum(pct_nefa$pct)`%) of the total NEFA
+fraction. The largest contributor was `r pctne$c[1]`, followed by `r
+pctne$c[2]`, `r pctne$c[3]`, and `r pctne$c[4]`.
 
@@
 
+There was poor discriminatory ability in identifying patterns between tertiles
+of ISSI-2, with only `r misclass_npct` of participants in the baseline
+visit who were classified correctly. In general, participants in the 3rd tertile
+(a high ISSI-2) had less of all NEFA, though the mean differences between
+tertiles were minor (`r cite_f('lda')`). The separation of tertiles of
+log(ISSI-2) and linear discriminants is shown in `r cite_sf('lda')`.
+
+These results confirm the findings from the longitudinal GEE modeling of
+individual NEFA variables.
+
+Results between visits were not substantially different.
+
 # Discussion
 
-Brief overall conclusion...
+In a Canadian population of adults who are at-risk for diabetes, we found that
+higher total NEFA independently predicted a lower beta-cell function over 6
+years. In particular, the specific composition of the NEFA fraction had no
+association with any of the metabolic outcomes within this cohort. These
+findings suggest that the size of the total fraction, rather than the
+composition, influences the pathogenesis of diabetes.
 
@@
 function... ...While the overall discriminatory ability of the NEFA pattern was
-poor, there were certain NEFA that loaded differently on each level of beta-cell
-function....
+poor, there was a general pattern that most NEFA were lower in individuals with
+higher beta-cell function. However, the differences between groups were minor
+and confirms the findings from the GEE modeling.
 
@@
 
-```{r figDist}
-ds %>%
-    filter(VN == 0) %>%
-    select(SID, matches('^ne\\d')) %>%
-    filter(complete.cases(.)) %>% 
-    gather(Measure, Value, -SID) %>% 
-    mutate(Measure = renaming_fa(Measure) %>% 
-               gsub('ne_', '', .) %>%
-               factor(., levels = unique(.))) %>% 
-    spread(Measure, Value) %>% 
-    select(-SID) %>% 
-    trance('boxes_dots') %>% 
-    visualize(dots = FALSE,
-              xlab = 'Concentration (nmol/mL)',
-              ylab = 'Non-esterified fatty acid') %>% 
-    vision_simple() +
-    theme(axis.ticks.y = element_blank())
+```{r figDist, fig.cap=fig_dist}
+plot_nefa_distribution(ds)
 ```
 
-```{r figGEE, dependson='geeData'}
-forest_plot(gee_results)
+```{r figGEE, dependson='geeData', fig.cap=fig_gee}
+plot_gee_results(gee_results)
 ```
 
-```{r figLDA}
+```{r figLDA, fig.cap=fig_lda}
 visualize(lda_results, 'means', label.rename.fun = renaming_fa) %>% 
@@
 ```{r supFigGeeUnadj, dependson='geeData'}
-forest_plot(gee_results_unadj)
+plot_gee_results(gee_results_unadj)
+```
+
+```{r supFigLDA}
+visualize(lda_results, orig.data = lda_df, group.name = 'BCF') %>% 
+    vision_simple()
 ```

### 2016-02-04 R/functions.R
Added functions for plotting, calculating misclass, outcomes, etc

@@ -71,9 +71,7 @@ analyze_gee <- function(data, y, x, covar, unit, adj.p = FALSE, adj.p.method = '
             transform.beta.funs = function(x)
                 (exp(x) - 1) * 100,
             rename.vars.funs = renaming_list
-        ) %>%
-        dplyr::mutate(Xterms = Xterms %>% factor(., unique(.)),
-                      unit = as.factor(unit))
+        )
 
     if (adj.p) {
         out <- out %>%
@@ -130,22 +128,26 @@ get_lda_data <- function(data) {
 }
 
 plot_gee_results <- function(data) {
+    data <- data %>%
+        mutate(Xterms = Xterms %>% factor(., unique(.)))
     data %>%
         seer::trance('main_effect') %>%
         seer::visualize(groups = 'unit~Yterms',
                         xlab = 'Percent difference with 95% CI in the outcomes\nfor each SD increase in fatty acid',
                         ylab = 'Non-esterified fatty acids') %>%
-        seer::vision_simple(legend_position = 'bottom') +
+        seer::vision_simple(legend.position = 'bottom') +
         ggplot2::theme(
-            axis.text.y = element_text(face = ifelse(
-                gee_results$Xterms == 'Total', "bold", "plain"
+            axis.text = element_text(face = ifelse(
+                levels(data$Xterms) == 'Total', "bold", "plain"
             )),
             legend.key.width = grid::unit(0.75, "line"),
             legend.key.height = grid::unit(0.75, "line"),
             panel.margin = grid::unit(0.75, "lines"),
             strip.background = element_blank()
         ) +
-        ggplot2::scale_alpha_discrete(name = 'P-value', range = c(0.4, 1.0))
+        ggplot2::scale_alpha_discrete(name = 'FDR-adjusted p-value', range = c(0.4, 1.0)) +
+        ggplot2::scale_size_discrete(name = 'FDR-adjusted p-value') +
+        ggplot2::facet_grid(unit~Yterms, scale = 'free_y', switch = 'y')
 }
 
 plot_nefa_distribution <- function(data) {
@@ -160,7 +162,7 @@ plot_nefa_distribution <- function(data) {
         tidyr::spread(Measure, Value) %>%
         dplyr::select(-SID) %>%
         seer::trance('boxes_dots') %>%
-        serr::visualize(dots = FALSE,
+        seer::visualize(dots = FALSE,
[…]

### 2016-02-04 doc/manuscript.Rmd
Added to manuscript intro, result, and discussion seections

@@
 
-Title:
+Title: Total non-esterified fatty acids, but not composition, predict
+progression in beta-cell dysfunction
 
-Author:
+Author: Luke W. Johnston, Richard P. Bazinet, Anthony J. Hanley
 
@@
 
-Possible reason why IR is not shown... other studies have been cross-sectional..
-Thus a potential confounder is the role that obesity plays in IR and IR on FFA
-release... So, longitudinally, FFA would not predict IR since it is IR that is
-suppressing NEFA.
-
-Popular hypothesis: obese -> FFA -> IR.. may not be the case
-
-total nefa evidence, well-studied. but there is diversity. no studies on it.
-
-Some evidence to suggest that FFA delivery may be higher in obese, but that conc
-may or may not be. Also that research on increasing FFA and IR comes from
-experimental (probably fairly short term) research.
-
-Association between higher FFA and muscle/hepatic IR, but challenging as adipose
-tissue can also be IR. So is it the higher FFA -> IR or is it IR -> FFA?
-
-methods -> gee, then lda (or similar) to confirm the role of 16:0 etc rather
-than any individual FA
-
 - as GEE results suggest that % does nothing, but in conc 16:0 etc has assoc...
@@
 
-THOUGHTS:
-
-- sleep restriction can impact NEFA, but only higher during night and a bit
-during early-morning. Should keep this in mind.
-- Adjust for TAG in IR models?
-- Key is that this is fasting, not postprandial!
+TODO:
+- Adjust for baseline prediabetes? SenAnalysis?
+- Do PLS-LDA instead of LDA, since it is apparently better for this type of
+analysis/data
 
@@
 
-```{r setup, collapse=TRUE}
+```{r knit_setup, collapse=TRUE}
 knitr::opts_knit$set(root.dir = '../')
@@
 
-```{r loadData, message=FALSE}
+```{r setup, message=FALSE}
 source('.Rprofile')
@@
 load_data()
+ds <- ds %>% 
+    mutate(lALT = log(ALT),
+           lTAG = log(TAG))
 dim(ds)
@@
 
-sfig_consort <- sfig('consort', 'CONSORT diagram....')
+sfig_consort <- sfig('consort', 'CONSORT diagram of sample size at each examination visit.')
+stab_qic <- stab('qic', 'Comparing generalized estimating equation models adjusting for different covariates using Quasi-Likelihood Information Criterion.')
+sfig_corr <- sfig('corr', 'Pearson correlation heatmap of non-esterified fatty acids (nmol/mL) and basic PROMISE participant characteristics for the baseline visit (2004-2006)')
+sfig_gee <- sfig('gee', 'Unadjusted generalized estimating equation modeling of the longitudinal association of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort.
+Only adjusted for time.
+Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
+P-values presented were adjusted for the false discovery rate.')
 sfig_lda <- sfig('lda', 'Linear discriminant factors with explained variance against tertiles of log(ISSI-2) in 463 PROMISE participants at the baseline visit (2004-2006).')
@@
 
-# Background #
+# Background
 
-Three paragraphs:
+Total non-esterified fatty acids (NEFA) have been well documented to influence
+the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that
+exposure of cell or animal to high concentrations of NEFA can induce insulin
+resistance in insulin-sensitive tissue such as the muscle or liver, impair
+beta-cell production of insulin, and cause hyperglycemia {{confirm...cite}}.
+Epidemiological and clinical studies have observed similar findings, showing
+that elevated total plasma NEFA associates with risk for diabetes {{cite}}.
 
-1. Total NEFA on diabetes, pathology, evidence
-2. NEFA diverse, individual NEFA on pathology, bit more evidence from total, go
-into experimental of species
-3. Limitations, gaps, inconsistencies, objectives and hypotheses
+Most previous experimental work uses palmitic acid (16:0), oleic acid (18:1n-9),
+or a standard oil such as soybean {{confirm, cite}} and generalize as total
+NEFA. However, fatty acids comprise multiple classes of molecules with diverse 
+physiological functions, with few studies comparing different fatty acids. For
+instance, one study on beta-cell {{plates? lines?}} suggests that
+eicosapentaenoic acid (20:5n-3) can protect against the lipotoxic effect of
+palmitic acid {{xiao? cite}}.
 
-# Subjects and Methods #
+Prospective with diabetes [@Ilyasova2010a]
 
-The Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort is a 
-longitudinal observational study of participants with one or more risk factors 
-for type 2 diabetes mellitus, including obesity, hypertension, family history of
-diabetes, and/or a history of gestational diabetes or birth of a macrosomic 
-infant.  Participants aged 30 years and older were recruited for the baseline 
-visit from Toronto and London, Ontario, Canada into the PROMISE cohort (n=736) 
-between 2004-2006.  Follow-up examinations occur every three years, with 
-currently three examination visits completed to date (2004-2006, 2007-2009, and 
-2010-2013).  The current study used data on participants who did not have 
-diabetes at baseline, who returned for one or more follow-up examinations, and 
-who had samples available for fatty acid measurements (n=477).  A diagram of the
-sample size at each visit is shown in `r cite_sf('consort')`.  At each 
-examination, participants undergo extensive metabolic characterization, 
-anthropometric measurements, and structured questionnaires on lifestyle and 
-sociodemographics.  Research ethics approval was obtained from Mount Sinai 
-Hospital and the University of Western Ontario.  Data collection methods were 
-standardized across the 2 centres and research nurses were centrally trained. 
-<!--Over the 6-yrs, {{n return each visit}} returned for the 6-yr clinic visit 
-({{pct loss to followup}}% loss to follow-up).  By the 6-yr visit, {{n converted
-to IFG}} converted into IFG, {{n converted to IGT}} converted into IGT, and {{n 
-converted to DM}} into diabetes.-->
+Lifestyle interventions aimed at managing or preventing diabetes show a
+concomitant reduction in NEFA and insulin resistance {{cite}}.
 
-## Blood measure assessment
+Lack of longitudinal data is particularly important given that NEFA release from
+adipocytes is inhibited by insulin, so if there is IR, more NEFA are released,
+making it difficult to disentangle which caused which.
 
-At each examination, participants had an 8-12 hour fasting blood sample 
-taken followed by a 75g oral glucose tolerance test (OGTT) with a 30 minute and
-2 hour blood draw. All blood samples were processed and frozen at -70°C. {{Alanine
-aminotransferase (ALT) was measured using standard laboratory procedures.}} 
-Cholesterol, HDL, and triglycerides (TG) were measured using Roche Modular's
-enzymatic colometric tests (Mississauga, ON).  Both specific insulin and glucose
-were derived from the OGTT at fasting, 30 minute, and 2 hour time points. 
-Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics, Basel, 
-Switzerland) immunoassay analyzer and electrochemiluminescence immunoassay. 
-This assay shows 0.05% cross-reactivity to intact human pro-insulin and the Des
-31,32 circulating split form (Linco Res. Inc), and has a CV of 9.3%.  Glucose
-was determined using an enzymatic hexokinase (Roche Modular, Roche Diagnostics)
-with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The inter-assay %CV is
-<1.1% and intra-assay %CV is < 1.9%.  All assays were performed at the Banting 
-and Best Diabetes Centre Core Lab at the University of Toronto. Impaired fasting
-glucose (IFG), impaired glucose tolerance (IGT), and diabetes were categorized
-using the 2006 WHO criteria [@WHO2006].
+Despite the fairly extensive literature on NEFA and diabetes, there are
+substantial research gaps. Importantly, most of the experimental work using
+individual NEFA species has been in animal and cell lines or if in humans has
+been acute, infusion studies. For epidemiological studies, few have examined the
+role of total, but not individual, NEFA on the pathogenesis of diabetes and
+fewer still have been longitudinal, which is key to understanding causal
+associations and given the long timescale in diabetes pathophysiology.
+Presently, no longitudinal study has been conducted that has examined the role
+of the composition of the NEFA fraction on the pathogenesis of diabetes. Our
+research objective was to examine the longitudinal role of the NEFA composition
+on insulin sensitivity and beta-cell function. Some specific hypotheses included
+that higher palmitic acid while lower polyunsaturated fatty acids such as
+eicosapentaenoic acid would associate with lower insulin sensitivity and
+beta-cell function.
+
+# Subjects and Methods
+
+Participants from London and Toronto, Canada, were recruited into the 
+Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
+for recruitment into PROMISE included having one or more risk factors for type 2
+diabetes mellitus, such as obesity, hypertension, family history of diabetes, 
+and/or a history of gestational diabetes or birth of a macrosomic infant. 
+Participants aged 30 years and older (n=726) conducted the baseline visit
+between 2004-2006. Follow-up examinations occur every three years, with three
+examination visits completed to date (2004-2006, 2007-2009, and 2010-2013).  The
+current study used data on participants who did not have diabetes at baseline,
+who returned for one or more follow-up examinations, and who had samples
+available for fatty acid measurements (n=477).  A diagram of the sample size at
+each visit is shown in `r cite_sf('consort')`.  At each examination,
+participants undergo metabolic characterization, anthropometric measurements,
+and questionnaires on lifestyle and sociodemographics.  Research ethics approval
+was obtained from Mount Sinai Hospital and the University of Western Ontario. 
+Data collection methods were standardized across the 2 centres and research
+nurses were centrally trained.
+
+## Blood measure assessments
+
+At each examination, an 8-12 hour fasting blood sample was draw from each
+participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
+minute and 2 hour blood draw. All blood samples were processed and frozen at
+-70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
+procedures. Cholesterol, HDL, and triglycerides (TG) were measured using Roche
+Modular's enzymatic colometric tests (Mississauga, ON).  Both specific insulin
+and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
+points. Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics,
+Basel, Switzerland) immunoassay analyzer and electrochemiluminescence
+immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
+and the Des 31,32 circulating split form (Linco Res. Inc), and has a CV of 9.3%.
+Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
+Diagnostics) with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The
+inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
+performed at the Banting and Best Diabetes Centre Core Lab at the University of
+Toronto. Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and
+diabetes were categorized using the 2006 WHO criteria [@WHO2006].
 
@@
 stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
-of {{internal standard name}} (17:0) was added to the serum as an internal 
+of heptadecanoic acid (17:0) {{confirm}} was added to the serum as an internal 
 standard prior to extracting total lipids according to the method of Folch 
 [@Folch1957a].  Each serum lipid fraction was isolated using thin layer 
-chromatography, which were  visualized under UV light after lightly spraying 
-with 8-anilino-1-naphthalene sulfonic acid (0.1% wt/vol) and then converted to
-fatty acid methyl esters with 14% boron trifluoride in methanol at 100°C for 1
-h.  Fatty acid methyl esters were separated and quantified using a Varian-430
-gas chromatograph (Varian, Lake Forest, CA, USA) equipped with a Varian Factor
-Four capillary column (VF-23ms; 30 m × 0.25 mm i.d. × 0.25 μm film thickness)
-and a flame ionization detector, which were injected in splitless mode. Fatty
-acid concentrations (nmol/ml) were calculated by proportional comparison of gas
-chromatography peak areas to that of the internal standards [@Nishi2014a;
-@Chen2011a].  There were 22 fatty acids measured in the NEFA fraction.
+chromatography, which were visualized under UV light after lightly spraying with
+8-anilino-1-naphthalene sulfonic acid (0.1% wt/vol) and then converted to fatty
+acid methyl esters with 14% boron trifluoride in methanol at 100°C for 1 h. 
+Fatty acid methyl esters were separated and quantified using a Varian-430 gas
+chromatograph (Varian, Lake Forest, CA, USA) equipped with a Varian Factor Four
+capillary column and a flame ionization detector, which were injected in
+splitless mode. Fatty acid concentrations (nmol/ml) were calculated by
+proportional comparison of gas chromatography peak areas to that of the internal
+standards [@Nishi2014a; @Chen2011a].  There were 22 fatty acids measured in the
+NEFA fraction.
 
@@
 Sociodemographic information, such as age, sex, and ethnicity, were determined
-using questionnaires administered at each examination.
+using questionnaires administered at each examination. In the lifestyle
+questionnaire, physical activity was determined using a version of the
+Modifiable Activity Questionnaire (MAQ) [@Kriska1990].  The MAQ collects 
+information on leisure and occupational activity, including intensity, 
+frequency, and duration, over the past year.  Each reported activity from the 
+MAQ was weighted by its metabolic intensity allowing for the estimation of 
+MET-hours per week.
 
@@
 the 22 NEFA as a mole percent (mol%) of the total fraction and as a 
-concentration (nmol/mL).
-
-Pearson correlation coefficients were computed for the PL and CE fatty acids
-with other continuous variables. {{keep??}}
+concentration (nmol/mL). Pearson correlation coefficients were computed for the
+individual NEFA and with other continuous variables.
 
@@
 and auto-regressive (AR1) correlation matrix was chosen for the GEE models, 
-though other matrices (eg. exchangeable) had similar fit (data not shown)
-{{confirm}}. GEE is well suited to cohort studies given it's capacity to handle
-missed visits. The predictor variables and continuous covariates were scaled 
-(mean centered and standardized). The NEFA variables were classified as 
-*time-independent* (held constant) as they were measured only at the baseline 
-visit, while the outcome variables and covariates were set as *time-dependent*. 
-No imputation was conducted on missing values.
+though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
+is well suited to cohort studies given it's capacity to handle missed visits.
+The predictor variables and continuous covariates were scaled (mean centered and
+standardized). The NEFA variables were classified as *time-independent* (held
+constant) as they were measured only at the baseline visit, while the outcome
+variables and covariates were set as *time-dependent*. No imputation was
+conducted on missing values.
 
@@
 [@Greenland1999a] recommendations, and quasi-likelihood information criterion 
-(QIC). `r cite_sf('qic')` shows the covariates compared using QIC and 
-`r cite_sf('dag')` shows the specified directed acyclic graph. The final GEE model
-(M#) we selected had time, sex, ethnicity, baseline age, waist circumference,
-{{...}}, and family history of diabetes as covariates. <!--(exponentiated, minus
-1, and multiplied by 100)--> After scaling, log-transforming, and exponentiating
-the GEE estimates, the GEE results are interpreted as an expected percent
-difference in the outcome variable for every SD increase in the predictor
-variable given the covariates are held constant. An time by predictor term was
-included in each of the GEE models to test for time interactions. Given that TG
-is a risk factor for diabetes and that NEFA contribute to TG production, TG may
-act as a mediator between NEFA and the outcomes. However, only a subset of the
-participants at the 6-yr visit had clinically measured TG. To determine the role
-of TG in the association between NEFA and the outcomes, TG was included in the
-GEE model as a sensitivity analysis.
+(QIC). `r cite_st('qic')` shows the covariates compared using QIC and `r 
+cite_sf('dag')` shows the specified directed acyclic graph. The final GEE model 
+we selected differed between insulin sensitivity and beta-cell function models. 
+For insulin sensitivity, we adjusted for time, sex, ethnicity, baseline age, WC,
+ALT, MET, alcohol intake, and family history of diabetes (M6). For beta-cell
+function, the last few models were comparable in performance, so we chose the
+simplest model (M8), adjusting for time, sex, ethnicity, baseline age, WC, ALT,
+and family history of diabetes. After scaling, log-transforming, and
+exponentiating the GEE estimates, the GEE results are interpreted as an expected
+percent difference in the outcome variable for every SD increase in the
+predictor variable given the covariates are held constant. While models with a
+time interaction did not perform better than others for the QIC (M6.1 and M8.1
+in `r cite_st('qic')`, we ran models with a time by predictor term for each of
+the NEFA to test if there was a time interaction. Given that TG is a risk factor
+for diabetes and that NEFA contribute to TG production, TG may act as a mediator
+between NEFA and the outcomes. However, only a subset of the participants at the
+6-yr visit had clinically measured TG. To determine the role of TG in the
+association between NEFA and the outcomes, TG was included in the GEE model as a
+sensitivity analysis.
 
@@
 
-```{r geeData, cache=TRUE}
+```{r varSet}
 outcomes_is <- c('linvHOMA', 'lISI')
@@
 ne_conc <- c('ne_Total', grep('^ne\\d\\d', names(ds), value = TRUE))
-covariates_is <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT')
-covariates_bcf <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 
-                    'MET', 'AlcoholPerWk')
+covariates_is <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'MET', 'AlcoholPerWk')
+covariates_bcf <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'FamHistDiab')
 
 gee_df <- get_gee_data(ds)
+```
+
+```{r geeData, cache=TRUE}
 gee_results <- bind_rows(
@@
     ) 
+gee_results %>% str
+# Use substr after this above to make the Xterms go in order.
 
@@
 
-```{r inlineGEE, dependson='geeData'}
-# extract gee beta coefficient for indicating magnitude
-pctChg.o <- ds %>%
-  select(f.VN, HOMA, ISI, IGIIR, ISSI2) %>%
-  gather(Measure, Value, -f.VN) %>%
-  na.omit() %>%
-  group_by(Measure, f.VN) %>%
-  summarise(med = median(Value)) %>%
-  spread(f.VN, med) %>%
-  mutate(pctChg = ((`6-yrs` - `0-yr`) / `0-yr`) * 100) %>%
-  select(pctChg) %>%
-  abs() %>%
-  round(1) %>%
-  {paste0(min(.), '% to ', max(.), '%')}
-```
+```{r inlineResults}
+pctne <- calculate_percent_nefa_contribution(ds) 
+misclass_npct <- calculate_misclass_npct(lda_results, lda_df)
+pctChg.o <- calculate_outcomes_pct_change(ds)
 
-```{r inlineLDA}
-fit.class <- predict(lda_results, newdata = lda_df[, -1])$class %>% 
-    factor(., ordered = TRUE)
-
-# Amount of participants who were misclassified
-misclass <- table(fit.class, lda_df[[1]]) %>% 
-    as.data.frame() %>% 
-    filter(fit.class != Var2) %>% 
-    summarize(count = sum(Freq)) %>% 
-    as.numeric()
-
-# Amount of participants who were correctly classified
-good_class <- table(fit.class, lda_df[[1]]) %>% 
-    as.data.frame() %>% 
-    filter(fit.class == Var2) %>% 
-    summarize(count = sum(Freq)) %>% 
-    as.numeric()
-
-misclass_npct <- paste0(misclass, ' (', round(misclass / (good_class + misclass) * 100, 0), '%)')
-```
-
-```{r inlineMisc}
-pctne <- calculate_percent_contribution(ds)
-
-extractBeta <- function(data) {
-    data %>%
-      mutate(estCI = paste0('(beta: ', trim(format(round(estimate, 1), nsmall = 1)),
-                            ' CI: ',
-                            trim(format(round(conf.low, 1), nsmall = 1)),
-                            ', ',
-                            trim(format(round(conf.high, 1), nsmall = 1)),
-                            ')'),
-             #indep.estCI = paste(indep, estCI),
-             #dep = paste(dep %>% gsub('log\\((.*)\\)', '\\1', .), estCI),
-             dep = paste(dep %>% gsub('log\\((.*)\\)', '\\1', .))
-             ) %>%
-      filter(p.value <= 0.05) %>%
-      arrange(dep, desc(estimate)) %>%
-      select(dep, indep, estCI)
-}
 ```
@@
 
-## Distribution of NEFA
+The primary outcome variables declined between `r pctChg.o` over the 6 years.
 
 `r cite_f('nefaDist')` shows the distribution of the NEFA composition. Four 
-individual NEFA made up the majority (`r sum(pct_nefa$pct)`%) of the total NEFA
-fraction. The largest contributor was `r pctne$c[1]`, followed by `r
-pctne$c[2]`, `r pctne$c[3]`, and `r pctne$c[4]`.
+individual NEFA made up the vast majority (`r sum(pct_nefa$pct)`%) of the total
+NEFA fraction. The largest contributor was `r pctne$c[1]`, followed by
+`r pctne$c[2]`, `r pctne$c[3]`, and `r pctne$c[4]`.
 
-## Correlations of NEFA with basic characteristics
+All individual NEFA as well as the total fraction had weak to no correlations
+(r<0.3) with the basic characteristics (see `r cite_sf('corr')`).
 
-## GEE
+## GEE modeling
 
-As shown in {{supFigGeeUnadj}}, several associations were seen in the unadjusted
-GEE models, particularly for the concentration data. Full model adjustment
-attenuated most of these associations, with total NEFA, 16:0, and 18:2n-6 (all
-as nmol/mL) had negative associations between IGI/IR and ISSI-2 ({{figGee}}).
-The magnitude of association for each of these variables was fairly consistent
-for each beta-cell function measure. For every one SD decrease in any of these
-three NEFA variables, there was a predicted {{magnitude}}% lower IGI/IR and a
+Several associations were seen in the unadjusted GEE models, particularly for 
+the concentration data. Full model adjustment attenuated most of these 
+associations, with total NEFA, 16:0, and 18:2n-6 (all as nmol/mL) had negative 
+associations between IGI/IR and ISSI-2 (see `r cite_f('gee')`). The magnitude of
+association for each of these variables was fairly consistent for each beta-cell
+function measure. For every one SD decrease in any of these three NEFA 
+variables, there was a predicted {{magnitude}}% lower IGI/IR and a 
 {{magnitude}}% lower ISSI-2 at each clinic visit. None of the NEFA variables as 
-a mol% had any association with the outcomes.
-
-Adjusting for TAG attenuates all associations with IGI/IR and ISSI-2 (data not
-shown). This may be due to either or both of two factors, mediation of TAG on
-the pathway between NEFA and beta-cell function or the lower sample size at the
-6 year visit from the incomplete measurement of TAG due to cuts in funding.
-
-There were no significant interaction effects between time and the individual
-NEFA on any of the outcome measures (data not shown).
+a mol% had any association with the outcomes. Adjusting for TG attenuates all 
+associations with IGI/IR and ISSI-2 (data not shown), though sample size at the 
+6 year visit is restricted to n=128 in those who had TG measured. There were no
+significant interaction effects between time and the individual NEFA on any of
+the outcome measures (data not shown).
 
@@
 There was poor discriminatory ability in identifying patterns between tertiles 
-of ISSI-2, with only `r misclass_npct` of participants in the baseline
-visit who were classified correctly. In general, participants in the 3rd tertile
-(a high ISSI-2) had less of all NEFA, though the mean differences between
-tertiles were minor (`r cite_f('lda')`). The separation of tertiles of
-log(ISSI-2) and linear discriminants is shown in `r cite_sf('lda')`.
-
-These results confirm the findings from the longitudinal GEE modeling of
-individual NEFA variables.
-
-Results between visits were not substantially different.
+of ISSI-2, with only `r misclass_npct` of participants in the baseline visit who
+were classified correctly. In general, participants in the 3rd tertile (a high
+ISSI-2) had less of all NEFA, though the mean differences between tertiles were
+minor (`r cite_f('lda')`). The separation of tertiles of log(ISSI-2) by the two
+linear discriminants is shown in `r cite_sf('lda')`. Analyses on the later
+visits to confirm the baseline results were not substantially different from the
+baseline visit.
 
@@
 
-We found that higher palmitic acid (16:0) and linoleic acid (18:2n-6) associated
-with lower beta-cell function by {{number?}}. Expand on experimental studies,
-pathology/biology, mechanisms, etc.
+## Beta cell
 
-There is substantial experimental research on the role of increasing NEFA and
-increasing IR {{cite}}. Two major limitation of these studies are that they tend
-to be fairly short term ({{length of time}}) and they tend to increase NEFA to
-supraphysiological concentrations {{confirm}}. Epidemiological research has also
-shown a link between higher NEFA and higher IR {{cite}}. However, most of these
-studies are cross-sectional in nature and can not address causal direction.
-Adipose tissue can become insulin resistant and thus hormone-sensitive lipase
-may not respond to the inhibitory action of insulin {{cite}}. As such, IR may
-temporally occur sooner than elevations in blood NEFA. In this longitudinal
-analysis, we saw no association between NEFA composition and neither hepatic IR
-(HOMA-IR) nor whole-body insulin sensitivity (ISI). Possible explanations for
-this could be that over several years NEFA does not influence hepatic or muscle
-IR or that it is IR that elevates NEFA and as such NEFA would not predict IR
-over time.
+We found that higher palmitic acid (16:0), oleic acid (18:1n-9), and linoleic
+acid (18:2n-6) associated with lower beta-cell function by {{number?}}. Expand
+on experimental studies, pathology/biology, mechanisms, etc.
+[@Salgin2012a;@Giacca2011a;@Xiao2009a]
+
+Epidemiology [@Johns2014a]
+
+Experimental [@Carpentier2003a]
+Review [@Carpentier2000a;@Cnop2008]
+
+Most experimental studies on beta-cell function use only palmitic acid....
+
+First study to our knowledge to examine a broad spectrum of individual NEFA in a cohort setting on beta-cell function..
+
+We found no association of any individual NEFA as mol% on beta-cell function,
+but a very strong signal with total NEFA, palmitic acid, oleic acid, and
+linoleic acid as concentrations.
+
+## IR
+
+There is substantial experimental evidence highlighting the role of increased 
+NEFA and a subsequent increase in insulin resistance via impairment of insulin 
+signaling cascades (reviewed in multiple articles 
+[@Martins2012a;@Ebbert2013a;@Capurso2012a;@Surampudi2009]). However, in this
+longitudinal analysis, we saw no association between any individual or total
+NEFA and neither hepatic (1/HOMA-IR) nor whole-body insulin sensitivity (ISI). 
+There are some possible explanations for our findings. Previous studies done *in
+vivo* were mostly short term infusion 
+[@Xiao2009a;@Xiao2006;@Szendroedi2012a;@Kehlenbrink2012a] or pharmocologic
+[@Daniele2014a;@Liang2013a] trials and the observational research was generally
+cross-sectional [@Miller2012a;@Johns2014a;@Bush2012a]. The limitation of these
+studies were the short time periods examined and the experimentally-induced
+elevations in NEFA concentrations, making it difficult to assess the causal role
+in non-experimental environments over long time frames. Moreover, some studies
+have found null or weak associations of NEFA and IR 
+[@Gray2013a;@Magkos2012a;@Kehlenbrink2012a], suggesting that either NEFA only 
+have a minor role in IR or that NEFA is the effect, but not cause, of the 
+underlying IR. Adipose tissue, which release NEFA during fasting, can become 
+insulin resistant and thus lipolysis of intracellular TAG by hormone-sensitive 
+lipase may not respond to the inhibitory action of insulin. As such, IR may 
+develop before elevations in blood NEFA. There is also evidence to suggest that 
+insulin may be involved in NEFA update [@Ramos-Roman2012a], further supporting 
+the hypothesis that elevated NEFA is a consequence of higher IR.
+
+Another explanation for our null findings with insulin sensitivity may be
+differences in physiology between fasting and postprandial NEFA kinetics. For
+instance, some experimental studies using the clamp found that fasting NEFA was
+a weak predictor of insulin sensitivity compared to postprandial concentrations
+of NEFA [@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the
+adipose tissue following postprandial TAG lipolysis via lipoprotein lipase
+may result in NEFA spillover into the blood and a subsequent increase in circulating
+NEFA [@Almandoz2013a], which may be more metabolically active given postprandial
+activity.
+
+...Last possibility is that given these subjects were at-risk for diabetes already,
+the causal role of NEFA on IR may have long passed and it no longer influences
+IR....
 
@@
 in the production of TAG that is to be processed into very-low density 
-lipoproteins (VLDL) {{cite}}. Elevated NEFA may be mediating it's association
-through hypertriglyceridemia, a known risk factor for diabetes {{cite}}. A major
-caveat with the TAG-adjusted GEE models is the reduced sample size in the 6 year
-visit, as only a subset (n={{sample}}) of participants had TAG measured.
+lipoproteins (VLDL) [@Nielsen2012a]. Elevated NEFA may be mediating it's
+association through hypertriglyceridemia, a known risk factor for diabetes.
+However, a major limitation to the TAG-adjusted GEE models is the reduced sample
+size in the 6 year visit, as only a subset (n=128) of participants had TAG
+measured.
+
+## Composition
+
+This is the first study to our knowledge to examine the composition of the NEFA fraction and
+
+Composition of NEFA and metabolic syndrome components [@Dai2015a]
+
+Composition of NEFA and metabolic syndrome binary [@Novgorodtseva2011a]
+
+Composition of NEFA and diabetes status (x-sec) [@Liu2010a], found 16:0, 18:0,
+18:1, 18:2, 18:3 could be useful biomarkers for identifying healthy vs diabetes
 
@@
 
+These results confirm the findings from the longitudinal GEE modeling of
+individual NEFA variables.
+
 Limitations, etc...
@@
 ## Acknowledgements ##
-
-# References #
 
@@
 ```{r figGEE, dependson='geeData', fig.cap=fig_gee}
+gee_results[sort(order(
+    gee_results$Xterms)[
+        gee_df %>%
+    select(matches('ne\\d\\d')) %>%
+    gather(FA, Conc) %>%
+    transmute(FA = renaming_fa(FA)) %>%
+    unique() %>% 
+        .[[1]]
+    ])]
+order(gee_results$Xterms)
 plot_gee_results(gee_results)
+gee_results %>% filter(Xterms == 'Total')
+gee_results %>% str
 ```
@@
 
-```{r supFigGeeUnadj, dependson='geeData'}
+![`r sfig_consort`](img/flowDiagramSample.png)
+
+```{r supTabQIC}
+prep_qic <- gee_df %>%
+    select(
+        SID, VN, lISSI2, lISI, ne_Total, Sex, Ethnicity, MET, ALT, BaseAge,
+        Waist, BMI, AlcoholPerWk, TobaccoUse, FamHistDiab
+        )
+
+qic_is <- prep_qic %>% 
+    select(-lISSI2) %>% 
+    na.omit()
+
+qic_bcf <- prep_qic %>% 
+    select(-lISI) %>% 
+    na.omit()
+
+# QIC for IS
+M0 <- geepack::geeglm(
+    lISI ~ ne_Total + VN, data = qic_is,
+    id = SID, family = gaussian,
+    corstr = 'ar1'
+    )
+
+M1 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex)
+M2 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT)
+M3 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET)
+M4 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk) # Best
+M5 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab) # Best
+M6 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse) # Best
+M6.1 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse + ne_Total:VN) # Best
+
+qic_is_tab <- model.sel(M0, M1, M2, M3, M4, M5, M6, M6.1, rank = QIC) %>% 
+    add_rownames() %>%
+    select(Model = rowname, QIC, Delta = delta) %>%
+    mutate(id = seq_len(nrow(.))) %>%
+    bind_rows(., data.frame(Model = '**log(ISI)**', id = 0.5))
+
+# QIC data for BCF
+qic_bcf <- prep_qic %>% 
+    select(-lISI) %>% 
+    na.omit()
+
+M0 <- geepack::geeglm(
+    lISSI2 ~ ne_Total + VN, data = qic_bcf,
+    id = SID, family = gaussian,
+    corstr = 'ar1'
+    )
+M1 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex)
+M2 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT) 
+M3 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET)
+M4 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk)
+M5 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab) # best
+M6 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse) # best
+M7 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TobaccoUse) # best
+M8 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab) # best
+M8.1 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + ne_Total:VN) # best
+
+qic_bcf_tab <- model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M8.1, rank = QIC) %>% 
+    add_rownames() %>%
+    select(Model = rowname, QIC, Delta = delta) %>% 
+    mutate(id = seq_len(nrow(.))+8) %>%
+    bind_rows(., data.frame(Model = '**log(ISSI-2)**', id = 8.5))
+
+rbind(qic_is_tab, qic_bcf_tab) %>% 
+    arrange(id) %>%
+    select(-id) %>%
+    mutate(QIC = round(QIC, 1),
+           Delta = round(Delta, 1)) %>% 
+    pander(missing = '', caption = stab_qic)
+```
+
+Given the number of possible combinations of outcome and predictor variables,
+only ISI and ISSI-2 with total non-esterified fatty acids (nmol/mL) were used to
+compare various GEE models and to select a final model.  Baseline age was used
+as including both original age and the time variable would result in
+collinearity.  Column names: QIC is the quasi-likelihood information criteria
+(smaller values, eg. larger negative values, indicate a better fit compared to
+other models), Delta is the QIC minus the lowest QIC (models with delta <10 are 
+considered equivalent).  Models were:
+
+- M0: log(ISSI-2) or log(ISI) = total non-esterified fatty acids (nmol/mL) + time
+- M1: M0 + waist + sex + ethnicity + baseline age
+- M2: M1 + ALT
+- M3: M2 + physical activity
+- M4: M3 + alcohol intake
+- M5: M4 + family history of diabetes
+- M6: M5 + smoking status
+- M6.1: M6 + time by NEFA interaction
+- M7: M2 + family history of diabetes + smoking status
+- M8: M2 + family history of diabetes
+- M8.1: M8 + time by NEFA interaction
+
+```{r supFigHeatmap, fig.cap=sfig_corr}
+ds %>% 
+    filter(VN == 0) %>% 
+    design('cor', method = 'pearson') %>%
+    lay_base(c(outcomes, 'BMI', 'Waist', 'Age', 'lALT', 'lTAG', 'Chol', 'HDL', 'LDL'), ne_conc) %>%
+    build() %>%
+    polish(rename.vars.fun = function(x) renaming_list(x) %>%
+               gsub('^l(ALT|TAG)', 'log(\\1)', .)) %>% 
+    trance('heatmap') %>% 
+    visualize(colours = c('darkred', 'darkblue'), number.colours = 5) %>% 
+    vision_simple() +
+    ylab('Non-esterified fatty acids (nmol/mL)')
+```
+
+```{r supFigGeeUnadj, fig.cap=sfig_gee}
 plot_gee_results(gee_results_unadj)
@@
 
-```{r supTabQIC, eval=FALSE}
-
+```{r supTabGEE}
 ```
 
-```{r supTabGEE}
-
-```
+# References

### 2016-02-23 doc/lda.Rmd
Changed lda analysis over to lcmm + plsda

--- 2016-01-27 doc/lda.Rmd
@@
 ---
-title: "LDA exploration"
+title: "LCMM, PLS-DA, and LDA exploration"
 author: "Luke W. Johnston"
@@
 
-# Prepare the data for the LDA
+# Latent class mixed model
+
+```{r lccmPkg}
+ds.lcmm <- ds %>% 
+    select(SID, lISSI2, VN) %>% 
+    na.omit() %>% 
+    arrange(SID, VN)
+fit <- lcmm(lISSI2 ~ VN,
+            mixture = ~VN, random = ~VN, subject = 'SID',
+            ng = 3, idiag = TRUE, link = '3-equi-splines', 
+            data = ds.lcmm)
+as.data.frame(table(fit$pprob$class)) %>% 
+    rename(Group = Var1, N = Freq)
+
+ds.lcmm2 <- 
+    full_join(ds.lcmm,
+              as.data.frame(fit$pprob[,1:2]),
+              by = 'SID') %>%
+    mutate(class = factor(class, labels = c('Mid', 'High', 'Low')))
+ggplot(ds.lcmm2, aes(VN, lISSI2, group = SID, colour = class)) +
+    geom_line(aes(group = SID, colour = class), size = 0.1, alpha = 0.9) +
+    geom_smooth(aes(group = class), method = "loess", size = 2)  +
+    labs(x = "Time", y = "ISSI-2", colour = "Latent Class") +
+    theme_minimal()
+ds <- full_join(ds, ds.lcmm2)
+dim(ds)
+```
+
+# Partial least squares discriminant analysis
+
+```{r caretPkg}
+ds.plsda <- ds %>%
+    filter(VN == 0) %>% 
+    select(class, matches('^ne\\d\\d')) %>% 
+    na.omit()
+
+if (length(nearZeroVar(ds.plsda[, 2:23])) != 0) message('Near zero variance, check into it')
+ds.plsda2 <- predict(preProcess(ds.plsda), ds.plsda)
+y <- as.factor(ds.plsda2[[1]])
+x <- as.matrix(ds.plsda2[2:23])
+fit <- plsda(x, y, probMethod = 'Bayes')
+
+mismatch <- confusionMatrix(predict(fit, x), y) %>% 
+    .$table %>% 
+    as.data.frame %>% 
+    mutate(Match = ifelse(Prediction == Reference, 'yes', 'no')) %>% 
+    group_by(Match) %>% 
+    summarise(N = sum(Freq)) %>% 
+    ungroup() %>% 
+    mutate(Total = sum(N),
+           Percent = round((N / Total) * 100, 1))
+mismatch
+
+plot(fit, plottype = 'loadings')
+plot(fit, plottype = 'correlation')
+
+exp.prop <- round((fit$Xvar/fit$Xtotvar)*100, 2)
+exp.prop
+fit$loadings %>% 
+    as.data.frame.matrix %>% 
+    add_rownames %>% 
+    rename(Comp1 = `Comp 1`, Comp2 = `Comp 2`) %>% 
+    mutate(rowname = rowname %>% 
+               renaming_fa() %>% 
+               factor(., levels = unique(.))) %>% 
+    ggplot(aes(Comp1, Comp2)) +
+    geom_point(size = 0.5) +
+    geom_text(aes(label = rowname), hjust = 0, vjust = 0) +
+    labs(x = paste0('Component 1 (', round(exp.prop[1], 1), '%)'), 
+         y = paste0('Component 2 (', round(exp.prop[2], 1), '%)')) +
+    ggthemes::theme_tufte(11, 'sans') +
+    theme(panel.grid = element_line(),
+          panel.grid.minor = element_blank())
+
+fit$scores %>% 
+    as.data.frame.matrix %>% 
+    bind_cols(ds.plsda['class']) %>% 
+    rename(Comp1 = `Comp 1`, Comp2 = `Comp 2`) %>% 
+    ggplot(aes(Comp1, Comp2, colour = class)) +
+    stat_density2d(ggplot2::aes(alpha = ..level.., colour = class), size = 2) +
+    scale_alpha(guide = 'none') +
+    scale_color_discrete('LCMM Class') +
+    labs(x = paste0('Component 1 (', round(exp.prop[1], 1), '%)'), 
+         y = paste0('Component 2 (', round(exp.prop[2], 1), '%)')) +
+    ggthemes::theme_tufte(11, 'sans') +
+    theme(panel.grid = element_line(),
+          panel.grid.minor = element_blank())
+
+```
+
+# Linear discriminant analysis
 
 ```{r ldaData}
-lda_ds <- ds %>%
+ds.prep <- ds %>%
     filter(VN == 0) %>%
-    mutate(BCF = rstatsToolkit::aide.tertile(lISSI2) %>%
-               factor(., labels = c('1st tertile', '2nd tertile', '3rd tertile'), ordered = TRUE)) %>%
-    select(BCF, matches('^ne\\d\\d')) %>%
-    mutate_each(funs(as.numeric(scale(.))), -BCF) %>%
-    na.omit() %>% 
-    arrange(BCF)
+    select(class, matches('^ne\\d\\d')) %>% 
+    mutate_each(funs(as.numeric(scale(.))), -class) %>%
+    rename(Class = class)
+
+df_y1 <- ds.prep %>% 
+    na.omit()
 
 # include also the 3-yr and 6-yr data for training/testing.
-
-#fit <- qda(BCF ~ ., data = lda_ds)
-fit <- lda(BCF ~ ., data = lda_ds)
-fit.cv <- lda(BCF ~ ., data = lda_ds, CV = TRUE)
-prop_lda <- round(fit$svd^2/sum(fit$svd^2) * 100, 2)
-
-# proportion explained by each LD
-prop_lda
+fit_y1 <- MASS::lda(Class ~ ., data = df_y1)
 ```
@@
 ```{r testingLDA}
-fit.class <- predict(fit, newdata = lda_ds[, -1])$class %>% 
+fit.class_y1 <- MASS:::predict.lda(fit_y1, newdata = df_y1[, -1])$class %>% 
     factor(., ordered = TRUE)
-table(fit.class, lda_ds[[1]])
-table(fit.cv$class, lda_ds[[1]])
+
+table(fit.class_y1, df_y1[[1]])
+
+# Amount of participants who were misclassified
+table(fit.class_y1, df_y1[[1]]) %>% 
+    as.data.frame() %>% 
+    filter(fit.class_y1 != Var2) %>% 
+    summarize(count = sum(Freq))
+
+# Amount of participants who were correctly classified
+# table(fit.class_y1, df_y1[[1]]) %>% 
+#     as.data.frame() %>% 
+#     filter(fit.class_y1 == Var2) %>% 
+#     fit.class_y1 <- predict(fit_y1, newdata = df_y1[, -1])$class %>% 
+#     factor(., ordered = TRUE)summarize(count = sum(Freq))
+
 ```
@@
 ```{r FAperBCF}
-p_data <- fit$means %>%
-    as.data.frame() %>%
-    add_rownames('Group') %>%
-    gather(FA, Mean,-Group) %>%
-    mutate(Group = Group %>% factor(., c('1st tertile', '2nd tertile', '3rd tertile'), ordered = TRUE)) %>%
-    arrange(Group, Mean) %>%
-    mutate(FA = paste0(Group, FA) %>%
-               factor(., unique(.)))
-FA.orig <- p_data$FA %>% 
-    as.character()
-FA.label <- p_data$FA %>%
+seer::visualize(
+    fit_y1, 'means', label.rename.fun = function(x)
+        x %>%
         gsub('^.*(\\d\\d)(\\d)', '\\1:\\2', .) %>%
-    gsub('n(\\d)', 'n-\\1', .) %>%
-    as.character()
-p_data %>%
-    ggplot(aes(x = Mean, y = FA)) +
-    geom_point() +
-    geom_vline(xintercept = 0) +
-    facet_wrap(~ Group, scale = 'free_y') +
-    scale_y_discrete(breaks = FA.orig, label = FA.label) +
-    ggthemes::theme_tufte(12, 'sans') +
-    theme(panel.margin = unit(2, 'lines'),
-          panel.grid = element_line(colour = 'black'),
-          panel.grid.minor.x = element_blank(),
-          panel.grid.major.x = element_blank())
-
+        gsub('n(\\d)', 'n-\\1', .)
+    ) %>% 
+    seer::vision_simple()
 ```
@@
 ```{r FAperLD}
-p_data2 <- fit$scaling %>%
-    as.data.frame() %>%
-    add_rownames('Var') %>%
-    gather(LD, Value, -Var) %>%
-    arrange(LD, Value) %>% 
-    mutate(Var = factor(paste0(LD, Var), unique(paste0(LD, Var))))
-FA.orig <- p_data2$Var %>% 
-    as.character()
-FA.label <- p_data2$Var %>%
+seer::visualize(
+    fit_y1, 'scaling', label.rename.fun = function(x)
+        x %>%
         gsub('^.*(\\d\\d)(\\d)', '\\1:\\2', .) %>%
-    gsub('n(\\d)', 'n-\\1', .) %>%
-    as.character()
-p_data2 %>% 
-    ggplot(aes(x = Value, y = Var)) +
-    geom_point() +
-    geom_vline(xintercept = 0) +
-    facet_wrap(~LD, scale = 'free_y') +
-    scale_y_discrete(breaks = FA.orig, label = FA.label) +
-    ggthemes::theme_tufte(12, 'sans') +
-    theme(panel.margin = unit(2, 'lines'),
-          panel.grid = element_line(colour = 'black'),
-          panel.grid.minor.x = element_blank(),
-          panel.grid.major.x = element_blank())
+        gsub('n(\\d)', 'n-\\1', .)
+    ) %>% 
+    seer::vision_simple()
 ```
@@
 ```{r LDbyBCF}
-p_fit <- predict(object = fit, newdata = lda_ds)
-# predict on follow up visits too.
-p_ds <- data.frame(BCF = lda_ds[,"BCF"], lda = p_fit$x)
-ggplot(p_ds, aes(lda.LD1, lda.LD2, group = BCF)) +
-    geom_vline(xintercept = 0, linetype = 'dotted') +
-    geom_hline(yintercept = 0, linetype = 'dotted') +
-    stat_density2d(aes(alpha = ..level.., colour = BCF), size = 2) +
-    labs(x = paste0("LD1 (", prop_lda[1], "%)"),
-         y = paste0('LD2 (', prop_lda[2], '%)')) +
-    ggthemes::theme_tufte(12, 'sans') +
-    scale_alpha(guide = 'none') +
-    theme(legend.title = element_blank(),
-          legend.position = c(1,0),
-          legend.justification = c(1,0))
+seer::visualize(fit_y1, orig.data = df_y1, group.name = 'Class') %>% 
+    seer::vision_simple()
 ```

### 2016-02-23 R/functions.R
Major reorg and rewrite of many functions for manuscript

@@ -5,6 +5,9 @@
 ##' Custom functions used for analyses.
 ##'
 
+
+# Renaming ----------------------------------------------------------------
+
 renaming_table_rows <- function(x) {
     x %>%
         gsub('IGIIR', 'IGI/IR', .) %>%
@@ -48,14 +51,17 @@ renaming_list <- function(x) {
         renaming_outcomes()
 }
 
-analyze_gee <- function(data, y, x, covar, unit, adj.p = FALSE, adj.p.method = 'BH', int = FALSE) {
 
+# Analyze -----------------------------------------------------------------
+
+analyze_gee <- function(data, y, x, covar, unit,
+                        adj.p = FALSE, adj.p.method = 'BH', int = FALSE) {
     if (int) {
         extract_term <- ':'
     } else {
         extract_term <- 'Xterm$'
     }
-    out <- data %>%
+    gee_results <- data %>%
         mason::design('gee', family = gaussian, corstr = 'ar1') %>%
         {
             if (int) {
@@ -71,29 +77,82 @@ analyze_gee <- function(data, y, x, covar, unit, adj.p = FALSE, adj.p.method = '
             transform.beta.funs = function(x)
                 (exp(x) - 1) * 100,
             rename.vars.funs = renaming_list
-        )
+        ) %>%
+        dplyr::mutate(
+            unit = unit,
+            order1 = substr(Xterms, nchar(Xterms), nchar(Xterms)),
+            order2 = substr(Xterms, 1, 2),
+            Yterms = factor(Yterms, unique(Yterms))
+        ) %>%
+        dplyr::arrange(order1, order2, Yterms) %>%
+        dplyr::select(-order1,-order2)
 
     if (adj.p) {
-        out <- out %>%
+        gee_results <- gee_results %>%
             dplyr::mutate(p.value = p.adjust(p.value, method = adj.p.method))
     }
 
-    return(out)
-
+    return(gee_results)
 }
 
-get_gee_data <- function(ds) {
-    full_join(
-    ds %>%
-        select(-matches('pct_ne|^ne'),-ne_Total),
-    ## Scale all the fatty acids
-    ds %>%
-        filter(VN == 0) %>%
-        select(SID, ne_Total, matches('pct_ne|^ne')) %>%
-        mutate_each(funs(as.numeric(scale(.))),-SID),
-    by = 'SID') %>%
-    group_by(VN) %>%
-    mutate(
[…]

### 2016-02-23 doc/manuscript.Rmd
Added more code/analysis, used functions more, added results + disc re: lcmm+plsda

@@
 csl:
-    - /home/luke/Documents/bibliography/styles/american-medical-association-no-et-al.csl
+    - /home/luke/Documents/bibliography/styles/vancouver.csl
 output:
@@
 - fig:geeconc%
-- fig:lda
-
-supfigs/tabs:
-- tab:geeconc%
-- tab:distr
+- fig:plsda
 
 TODO:
-- Adjust for baseline prediabetes? SenAnalysis?
-- Do PLS-LDA instead of LDA, since it is apparently better for this type of
-analysis/data
+- Bootstrap for the longitudinal table data... like instead of an ANOVA/GEE for
+the p-value...?
 
@@
 P-values presented were adjusted for the false discovery rate.')
-fig_lda <- fig('lda', 'Linear discriminant analysis to identify patterns of NEFA composition between tertiles of log(ISSI-2) in 463 PROMISE participants at the baseline visit (2004-2006).')
 
@@
 P-values presented were adjusted for the false discovery rate.')
-sfig_lda <- sfig('lda', 'Linear discriminant factors with explained variance against tertiles of log(ISSI-2) in 463 PROMISE participants at the baseline visit (2004-2006).')
+stab_gee <- stab('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function using generalized estimating equations over the 6 years in the PROMISE cohort.
+Adjusted for {{covariates}}.
+Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
+P-values presented were adjusted for the BH false discovery rate.')
+stab_gee_unadj <- stab('gee_unadj', 'Unadjusted generalized estimating equations models of the longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
+Adjusted for time only.
+Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
+P-values presented were adjusted for the BH false discovery rate.')
+sfig_lcmm <- sfig('lcmm', 'Latent class mixed model analysis to identify individual classes of trajectories for log(ISSI-2) over the 6 years in the PROMISE cohort.')
+sfig_plsda_loadings <- sfig('plsda_loadings', 'Pattern loadings from partial least squares discriminant analysis to identify potential clusters of NEFA composition within the classes extracted from the latent class mixed model in 463 participants from the baseline PROMISE visit (2004-2006).
+The percent explained variance of each component is shown in brackets on each axis.')
+sfig_plsda_groups <- sfig('plsda_groups', 'Clustering of extracted components from the partial least squares discriminant analysis on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006).
+The percent explained variance of each component is shown in brackets on each axis.')
 ```
@@
 of the composition of the NEFA fraction on the pathogenesis of diabetes. Our
-research objective was to examine the longitudinal role of the NEFA composition
-on insulin sensitivity and beta-cell function. Some specific hypotheses included
-that higher palmitic acid while lower polyunsaturated fatty acids such as
-eicosapentaenoic acid would associate with lower insulin sensitivity and
-beta-cell function.
+research objective was to examine the role of the NEFA composition
+on insulin sensitivity and beta-cell function using longitudinal and clustering
+techniques. Some specific hypotheses included that higher palmitic acid while
+lower polyunsaturated fatty acids such as eicosapentaenoic acid would associate
+with lower insulin sensitivity and beta-cell function.
 
@@
 NEFA fraction. As such, after running the GEE models for the longitudinal 
-analysis, linear discriminant analysis (LDA) was used to identify the patterns
-of NEFA composition for tertiles of the metabolic disorders. Since LDA can't
-handle longitudinal data, only the baseline data was used and only the outcome
-and predictor variables could be included in the analysis. Data from the later
-visit dates were used to confirm the results of the LDA analysis.
+analysis, partial least squares discriminant analysis (PLS-DA) was used to
+identify the patterns of NEFA composition against beta-cell function (using ISSI-2,
+which has most of the associations in the GEE models). Since PLS-DA can't handle
+longitudinal data, the longitudinal data was maximized by extracting latent
+trajectories of beta-cell function over the 6 years using latent class mixed
+models (LCMM). A limitation of these methods is that they can not be adjusted
+for covariates.
 
 All analyses were performed using R `r getRversion()` [@Rbase], along with the R
-packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a] and 
-MASS `r packageVersion('MASS')` for LDA. The R code for this manuscript is
-available at {{link}}. Results were statistically significance at p<0.05, after
-adjusting for multiple testing using the False Discovery Rate [@Benjamini1995a].
+packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a], caret
+`r packageVersion('caret')` for PLS-DA, and lcmm `r packageVersion('lcmm')` for
+LCMM. The R code for this manuscript is available at {{link}}. Results were
+statistically significance at p<0.05, after adjusting for multiple testing using
+the False Discovery Rate [@Benjamini1995a].
 
@@
 
-```{r varSet}
+```{r variableSet}
 outcomes_is <- c('linvHOMA', 'lISI')
@@
     )
-gee_results %>% str
-# Use substr after this above to make the Xterms go in order.
 
@@
     analyze_gee(gee_df, outcomes, ne_conc, 'VN', 'nmol/mL', adj.p = TRUE)
-    ) %>% 
-    mutate(Yterms = factor(Yterms, unique(Yterms)))
+    )
 ```
 
-```{r ldaData}
+```{r lcmmPls}
 # include also the 3-yr and 6-yr data for training/testing.
-lda_df <- get_lda_data(ds)
-lda_results <- lda(BCF ~ ., data = lda_df)
+lcmm_results <- analyze_lcmm(ds)
+lcmm_data <- combine_lcmm_data(ds, lcmm_results)
+plsda_results <- analyze_plsda(lcmm_data)
 ```
@@
 pctne <- calculate_percent_nefa_contribution(ds) 
-misclass_npct <- calculate_misclass_npct(lda_results, lda_df)
-pctChg.o <- calculate_outcomes_pct_change(ds)
-
+misclass_npct <- calculate_plsda_misclass(plsda_results)
+pctChg_o <- calculate_outcomes_pct_change(ds)
+lcmm_grps <- calculate_ngroups_lcmm(lcmm_results)
 ```
@@
 
-The primary outcome variables declined between `r pctChg.o` over the 6 years.
+The primary outcome variables declined between `r pctChg_o` over the 6 years.
 
@@
 individual NEFA made up the vast majority (`r sum(pct_nefa$pct)`%) of the total
-NEFA fraction. The largest contributor was `r pctne$c[1]`, followed by
-`r pctne$c[2]`, `r pctne$c[3]`, and `r pctne$c[4]`.
-
-All individual NEFA as well as the total fraction had weak to no correlations
-(r<0.3) with the basic characteristics (see `r cite_sf('corr')`).
+NEFA fraction. The largest contributors were `r pctne$c[1]`, `r pctne$c[2]`, 
+`r pctne$c[3]`, and `r pctne$c[4]`. All individual NEFA as well as the total
+fraction had weak to no correlations (r<0.3) with the basic characteristics (see
+`r cite_sf('corr')`).
 
@@
 
-Several associations were seen in the unadjusted GEE models, particularly for 
-the concentration data. Full model adjustment attenuated most of these 
-associations, with total NEFA, 16:0, and 18:2n-6 (all as nmol/mL) had negative 
-associations between IGI/IR and ISSI-2 (see `r cite_f('gee')`). The magnitude of
-association for each of these variables was fairly consistent for each beta-cell
-function measure. For every one SD decrease in any of these three NEFA 
-variables, there was a predicted {{magnitude}}% lower IGI/IR and a 
-{{magnitude}}% lower ISSI-2 at each clinic visit. None of the NEFA variables as 
-a mol% had any association with the outcomes. Adjusting for TG attenuates all 
-associations with IGI/IR and ISSI-2 (data not shown), though sample size at the 
-6 year visit is restricted to n=128 in those who had TG measured. There were no
-significant interaction effects between time and the individual NEFA on any of
-the outcome measures (data not shown).
+Several associations were seen in the unadjusted GEE models 
+(`r cite_sf('gee')`), particularly for the concentration data. Full model
+adjustment attenuated most of these associations, with total NEFA, 16:0, and
+18:2n-6 (all as nmol/mL) had negative associations between IGI/IR and ISSI-2
+(see `r cite_f('gee')`). The magnitude of association for each of these
+variables was fairly consistent for each beta-cell function measure. For every
+one SD decrease in any of these three NEFA variables, there was a predicted
+{{magnitude}}% lower IGI/IR and a {{magnitude}}% lower ISSI-2 at each clinic
+visit. None of the NEFA variables as a mol% had any association with the
+outcomes. Adjusting for TG attenuates all associations with IGI/IR and ISSI-2
+(data not shown), though sample size at the 6 year visit is restricted to n=128
+in those who had TG measured. There were no significant interaction effects
+between time and the individual NEFA on any of the outcome measures (data not
+shown). Raw values from the GEE models are shown in `r cite_st('gee_unadj')` for
+unadjusted models and `r cite_st('gee')` for fully adjusted models.
 
-## Discriminant analysis
+## Latent trajectory and NEFA pattern recognition
 
-There was poor discriminatory ability in identifying patterns between tertiles 
-of ISSI-2, with only `r misclass_npct` of participants in the baseline visit who
-were classified correctly. In general, participants in the 3rd tertile (a high
-ISSI-2) had less of all NEFA, though the mean differences between tertiles were
-minor (`r cite_f('lda')`). The separation of tertiles of log(ISSI-2) by the two
-linear discriminants is shown in `r cite_sf('lda')`. Analyses on the later
-visits to confirm the baseline results were not substantially different from the
-baseline visit.
+Three latent classes were extracted from LCMM, with `r lcmm_grps[2,2]` 
+participants in the high, `r lcmm_grps[1,2]` in the middle, and `lcmm_grps[3,2]`
+in the low beta-cell function groups. The trajectories of the three groups are
+shown in `r cite_sf('lcmm')`. These groups were used as the response variables
+in the PLS-DA analysis. The PLS-DA results show a very poor discriminatory
+abililty of the NEFA composition in classifying participants into the correct
+beta-cell function latent class. Only `r misclass_npct[2,1]` 
+(`r misclass_npct[2,4]`) participants were correctly classified, primarily into the
+middle group. The specific NEFA composition was not able to accurately classify
+participants who had low or high beta-cell function.
+`r cite_sf('plsda_loadings')` shows the loadings of individual NEFA by the
+extracted component, while `r cite_sf('plsda_groups')` shows the clustering of the components by ISSI-2 latent class.
 
@@
 
-Epidemiology [@Johns2014a]
+In a cross-sectional analysis of the RISCK cohort, total NEFA had a negative
+association with insulin sensitivity and, particularly strong with beta-cell
+function [@Johns2014a]. The authors further analysis showed that greater
+adiposity was not a mediator between the total NEFA and metabolic function
+association.
 
@@
 linoleic acid as concentrations.
+
+Using cluster techniques, we found no evidence to suggest that any individual
+NEFA contributed uniquely to beta-cell function, confirming the results we saw
+in the GEE models. This is particularly seen with the poor predictive ability of
+the NEFA composition to classify between groups of beta-cell function.
+
+In the sensitivity analysis adjusting for TAG, all associations with beta-cell 
+function were attenuated. While there is strong experimental evidence on the 
+role of chronically elevated NEFA, particularly palmitic acid, on impaired 
+beta-cell function, within a free-living population NEFA may be mediating it's 
+association through TAG. In normal metabolism, NEFA enters the liver and assists
+in the production of TAG that is to be processed into very-low density 
+lipoproteins (VLDL) [@Nielsen2012a]. Elevated NEFA may be mediating it's
+association through hypertriglyceridemia, a known risk factor for diabetes.
+However, a major limitation to the TAG-adjusted GEE models is the reduced sample
+size in the 6 year visit, as only a subset (n=128) of participants had TAG
+measured.
 
@@
 have found null or weak associations of NEFA and IR 
-[@Gray2013a;@Magkos2012a;@Kehlenbrink2012a], suggesting that either NEFA only 
-have a minor role in IR or that NEFA is the effect, but not cause, of the 
-underlying IR. Adipose tissue, which release NEFA during fasting, can become 
+[@Gray2013a;@Magkos2012a;@Kehlenbrink2012a; @Johns2014a], suggesting that either
+NEFA only have a minor role in IR or that NEFA is the effect, but not cause, of
+the underlying IR. Adipose tissue, which release NEFA during fasting, can become
 insulin resistant and thus lipolysis of intracellular TAG by hormone-sensitive 
@@
 
-In the sensitivity analysis adjusting for TAG, all associations with beta-cell 
-function were attenuated. While there is strong experimental evidence on the 
-role of chronically elevated NEFA, particularly palmitic acid, on impaired 
-beta-cell function, within a free-living population NEFA may be mediating it's 
-association through TAG. In normal metabolism, NEFA enters the liver and assists
-in the production of TAG that is to be processed into very-low density 
-lipoproteins (VLDL) [@Nielsen2012a]. Elevated NEFA may be mediating it's
-association through hypertriglyceridemia, a known risk factor for diabetes.
-However, a major limitation to the TAG-adjusted GEE models is the reduced sample
-size in the 6 year visit, as only a subset (n=128) of participants had TAG
-measured.
-
 ## Composition
 
-This is the first study to our knowledge to examine the composition of the NEFA fraction and
+This is the first study to our knowledge to examine the composition of the NEFA
+fraction and longitudinal measures of metabolic functioning.
 
@@
 
-Composition of NEFA and diabetes status (x-sec) [@Liu2010a], found 16:0, 18:0,
-18:1, 18:2, 18:3 could be useful biomarkers for identifying healthy vs diabetes
-
-Comment on LDA analysis... ...LDA backs up some of the GEE model results,
-showing weak discriminatory ability of the NEFA pattern on identifying beta-cell
-function... ...While the overall discriminatory ability of the NEFA pattern was
-poor, there was a general pattern that most NEFA were lower in individuals with
-higher beta-cell function. However, the differences between groups were minor
-and confirms the findings from the GEE modeling.
-
-These results confirm the findings from the longitudinal GEE modeling of
-individual NEFA variables.
+Using complex measurement and statistical techniques that were rigorously
+applied {{? reword}} to the composition of the NEFA fraction, Liu *et al.*
+-[@Liu2010a] found that 16:0, 18:0, 18:1, 18:2, 18:3 may be useful biomarkers
+for identifying healthy compared to diabetic individuals. However, a limitation
+of this study was the smaller sample size {{confirm}} and the cross-sectional
+design.
 
@@
 ```{r figGEE, dependson='geeData', fig.cap=fig_gee}
-gee_results[sort(order(
-    gee_results$Xterms)[
-        gee_df %>%
-    select(matches('ne\\d\\d')) %>%
-    gather(FA, Conc) %>%
-    transmute(FA = renaming_fa(FA)) %>%
-    unique() %>% 
-        .[[1]]
-    ])]
-order(gee_results$Xterms)
 plot_gee_results(gee_results)
-gee_results %>% filter(Xterms == 'Total')
-gee_results %>% str
 ```
 
-```{r figLDA, fig.cap=fig_lda}
-visualize(lda_results, 'means', label.rename.fun = renaming_fa) %>% 
-    vision_simple() +
-    labs(y = 'Non-esterified fatty acids',
-         x = 'Standardized mean of the fatty acid (SD of the nmol/mL)')
-```
-
-# Supplemental Material #
+# Supplemental Material
 
@@
 compare various GEE models and to select a final model.  Baseline age was used
-as including both original age and the time variable would result in
+as including both the original age and the time variable would result in
 collinearity.  Column names: QIC is the quasi-likelihood information criteria
@@
 ```{r supFigHeatmap, fig.cap=sfig_corr}
-ds %>% 
-    filter(VN == 0) %>% 
-    design('cor', method = 'pearson') %>%
-    lay_base(c(outcomes, 'BMI', 'Waist', 'Age', 'lALT', 'lTAG', 'Chol', 'HDL', 'LDL'), ne_conc) %>%
-    build() %>%
-    polish(rename.vars.fun = function(x) renaming_list(x) %>%
-               gsub('^l(ALT|TAG)', 'log(\\1)', .)) %>% 
-    trance('heatmap') %>% 
-    visualize(colours = c('darkred', 'darkblue'), number.colours = 5) %>% 
-    vision_simple() +
-    ylab('Non-esterified fatty acids (nmol/mL)')
+plot_heatmap(ds)
 ```
@@
 ```{r supFigLDA}
-visualize(lda_results, orig.data = lda_df, group.name = 'BCF') %>% 
-    vision_simple()
+
 ```
@@
 ```{r supTabGEE}
+table_gee(gee_results, stab_gee)
+table_gee(gee_results_unadj, stab_gee_unadj)
+```
+
+```{r supFigLCMM, fig.cap=}
+plot_lcmm_results(lcmm_data)
+```
+
+```{r supFigPlsLoadings, fig.cap=}
+plot_plsda_loadings(plsda_results)
+```
+
+```{r supFigPlsGrp}
+plot_plsda_grouping(plsda_results)
 ```

### 2016-04-10 R/fetch_data.R
Updated fetching data as per changes to PROMISE dataset (aka as a package)

@@ -7,9 +7,7 @@
 #'
 .fetch_data <- function(master.dataset) {
     # Load the master dataset,
-    ds.prep <- read.table(master.dataset,
-                          header = TRUE, na = "",
-                          sep = ",") %>%
+    ds.prep <- master.dataset %>%
         filter(VN %in% c(1, 3, 6)) %>%
         ## Kick out Canoers
         filter(is.na(Canoe)) %>%
@@ -26,9 +24,11 @@
         select(
             SID, VN, BMI, Waist, HOMA, ISI, IGIIR, ISSI2, TAG, LDL, HDL, Chol,
             ALT, CRP, FamHistDiab, matches('meds'), Age, Sex, Ethnicity,
-            IFG, IGT, DM, MET, BaseAge, AlcoholPerWk, TobaccoUse, SelfEdu, Occupation
+            IFG, IGT, DM, MET, BaseAge, AlcoholPerWk, TobaccoUse, SelfEdu, Occupation,
+            TotalNE, matches('^ne\\d+')
         ) %>%
         mutate(
+            BaseTotalNE = TotalNE,
             FamHistDiab =
                 plyr::mapvalues(FamHistDiab, c(0, 1:12),
                                 c('No', rep('Yes', 12))) %>%
@@ -38,60 +38,70 @@
             lISI = log(ISI),
             lIGIIR = log(IGIIR),
             lISSI2 = log(ISSI2),
+            lALT = log(ALT),
+            lTAG = log(TAG),
             MedsLipidsChol = ifelse(is.na(MedsLipidsChol), 0, MedsLipidsChol)
         ) %>%
-        full_join(
-            .,
-            ## Merge in the FA dataset so it is time-independent.
-            full_join(
-                ds.prep %>%
-                    filter(VN == 1) %>%
-                    select(SID, totalNE, matches('^ne\\d+')) %>%
-                    filter(complete.cases(.)),
-                ## Create percent of total fatty acid values.
-                ds.prep %>%
-                    filter(VN == 1) %>%
-                    select(SID, totalNE, matches('^ne\\d+')) %>%
-                    gather(Total, TotConc, totalNE) %>%
-                    gather(Fat, FatConc,-SID,-Total,-TotConc) %>%
-                    mutate(
-                        Total = gsub('total', '', Total),
-                        FatPool = toupper(gsub('\\d.*$', '', Fat)) %>% factor(),
-                        Pct = (FatConc / TotConc) * 100,
[…]

### 2016-04-10 R/functions.R
Updated many functions to reflect dev of seer and mason

@@ -17,7 +17,7 @@ renaming_table_rows <- function(x) {
         gsub('Chol', 'Chol (mmol/L)', .) %>%
         gsub('LDL', 'LDL (mmol/L)', .) %>%
         gsub('HDL', 'HDL (mmol/L)', .) %>%
-        gsub('ne_Total', 'NEFA (nmol/mL)', .) %>%
+        gsub('BaseTotalNE', 'NEFA (nmol/mL)', .) %>%
         gsub('Age', 'Age (yrs)', .) %>%
         gsub('BMI', 'BMI (kg/m^2^)', .) %>%
         gsub('Waist', 'WC (cm)', .)
@@ -25,10 +25,13 @@ renaming_table_rows <- function(x) {
 
 renaming_outcomes <- function(x) {
     x %>%
-        gsub('linvHOMA', 'log(HOMA-IS)', .) %>%
+        gsub('linvHOMA', 'log(1/HOMA-IR)', .) %>%
         gsub('lISI', 'log(ISI)', .) %>%
         gsub('lIGIIR', 'log(IGI/IR)', .) %>%
-        gsub('lISSI2', 'log(ISSI-2)', .)
+        gsub('lISSI2', 'log(ISSI-2)', .) %>%
+        gsub('^invHOMA$', '1/HOMA-IR', .) %>%
+        gsub('^ISSI2$', 'ISSI-2', .) %>%
+        gsub('^IGIIR$', 'IGI/IR', .)
 }
 
 renaming_fa <- function(x) {
@@ -37,7 +40,7 @@ renaming_fa <- function(x) {
         gsub('n(\\d)$', 'n-\\1', .) %>%
         gsub('D(\\d\\d)$', 'D-\\1', .) %>%
         gsub('^pct_', '', .) %>%
-        gsub('ne_Total', 'Total', .)
+        gsub('TotalNE', 'Total', .)
 }
 
 renaming_fraction <- function(x) {
@@ -62,34 +65,34 @@ analyze_gee <- function(data, y, x, covar, unit,
         extract_term <- 'Xterm$'
     }
     gee_results <- data %>%
-        mason::design('gee', family = gaussian, corstr = 'ar1') %>%
-        {
+        mason::design_analysis('gee') %>%
+        mason::add_settings(family = gaussian('identity'), corstr = 'ar1', cluster.id = 'SID') %>%
+        mason::add_variables('yvars', y) %>%
+        mason::add_variables('xvars', x) %>%
[…]

### 2016-04-10 R/load_data.R
Updated load data to reflect changes to PROMISE dataset

@@ -16,11 +16,14 @@
 #'      'data/dsname.RData',
 #'      'R/fetchScript.R')
 #' }
-load_data <- function(master.dataset = '/home/luke/Documents/datasets/PROMISE/dataset/data/PROMISE_data.csv.gz',
-                      local.dataset = file.path('data', 'ds.RData'),
+load_data <- function(master.dataset = PROMISE::PROMISE_data,
+                      local.dataset = file.path('data', 'data.Rds'),
                       fetch.script = file.path('R', 'fetch_data.R'),
                       force = FALSE) {
-    if (force) .fetch_data(master.dataset)
+    source('.Rprofile')
+    if (force) {
+        .fetch_data(master.dataset)
+    }
     if (!file.exists(fetch.script))
         stop('Please create a file under the R folder that contains a function called fetch_data.')
     if (!file.exists(local.dataset)) {
@@ -34,5 +37,5 @@ load_data <- function(master.dataset = '/home/luke/Documents/datasets/PROMISE/da
         message('Dataset is up-to-date.')
     }
 
-    load(local.dataset, envir = .GlobalEnv)
+    readRDS(local.dataset)
 }

### 2016-04-10 doc/manuscript.Rmd
Added DC title page + added much to discussion (old edits)

@@
 output:
+    md_document:
+        variant: markdown
     word_document:
-      reference_docx: /home/luke/Templates/journal-templates/ajcn-template.docx
+        reference_docx: diabetes-care-template.docx
         fig_height: 6 
@@
 
-# Title page #
+# Title page
 
 Title: Total non-esterified fatty acids, but not composition, predict
-progression in beta-cell dysfunction
+progression in beta-cell dysfunction: The Prospective Metabolism and Islet Cell
+Evaluation (PROMISE) cohort
 
-Author: Luke W. Johnston, Richard P. Bazinet, Anthony J. Hanley
+<!-- 47 characters + spaces -->
+Running title: NEFA and the pathogenesis of diabetes
 
-Affiliations:
+<!-- First, middle initial, last, highest academic degree -->
+Author: Luke W. Johnston, Stewart Harris, Ravi Retnakaran, Bernard Zinman,
+Zhen Liu, Richard P. Bazinet, Anthony J. Hanley
 
-Disclaimers:
+<!-- During time of study. -->
+Affiliation:
 
-Correspondence:
+Corresponding author:
 
-Funding support:
+- Name: Anthony J. Hanley
+- Current address:
+- Phone number:
+- Fax number:
+- Email: anthony.hanley@utoronto.ca
 
-# Abstract #
+Word count: / 4000
+
+Table and figure count: / 4
+
+<!-- Reference limit / 40 -->
+
+<!-- 250 words -->
+# Abstract
+
+Objective:
+
+Research Design and Methods:
+
+Results:
+
+Conclusions:
+
+# Abstract
 
 <!-- outline
-
-Overall story: while nefa are known risk factors for diabetes, not sure about
-the composition. we find that the composition doesn't matter so much as the
-overall pool, specific to BCD. however, 16:0, etc, are the only two fa that have
-an assoc, suggesting that irrespective of other fa increases or decreases, these
-two fa are the fa that contribute to the harm seen in the overall nefa. using
-cluster analysis, we see that those with low bcd have a pattern of fa that skews
-to higher amounts of 16:0 etc and marginally less in other fa (maybe not enough
-to observe on an individual scale). while other studies have seen a role of nefa
-in ir, we do not. while this warrants further longitudinal research, this may
-suggest that given prev studies on ir have been acute in nature, over the long
-term the role of nefa on ir may not be substantial. this duality in nefa is also
-seen in bcd, where acutely it improves bc, but chronically no.
-
-- as GEE results suggest that % does nothing, but in conc 16:0 etc has assoc...
-is it that 16:0 is harmful or is it the pool? is it that as 16:0 inc, there is
-harm, irrespective of whether other FA increase or decrease as well?
-
-figs/tabs: 
-- tab:basic char
-- fig:distr
-- fig:nefaCorr (sup?)
-- fig:varCorr (sup?)
-- fig:geeconc%
-- fig:plsda
 
@@
 the p-value...?
+- What about adiponectin?
+- Precise followup time for each participant?
+- Interaction by sex or ethnicity
 
@@
 
-```{r knit_setup, collapse=TRUE}
+```{r knit_setup, echo=FALSE}
 knitr::opts_knit$set(root.dir = '../')
-knitr::opts_chunk$set(echo = FALSE)
+knitr::opts_chunk$set(echo = FALSE, dev = c('png', 'postscript'))
 ```
@@
 suppressMessages(run_setup())
-load_data()
-ds <- ds %>% 
-    mutate(lALT = log(ALT),
-           lTAG = log(TAG))
-dim(ds)
+ds <- load_data()
 ```
@@
 cite_f <- pryr::partial(fig, display = 'cite')
-sfig <- captioner(prefix = 'Supplementary Figure')
+sfig <- captioner(prefix = 'Supplemental Figure S')
 cite_sf <- pryr::partial(sfig, display = 'cite')
@@
 cite_t <- pryr::partial(tab, display = 'cite')
-stab <- captioner(prefix = 'Supplementary Table')
+stab <- captioner(prefix = 'Supplemental Table S')
 cite_st <- pryr::partial(stab, display = 'cite')
@@
 
-tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 time points.')
+tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 clinic visits.')
 fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006)')
@@
 the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that
-exposure of cell or animal to high concentrations of NEFA can induce insulin
-resistance in insulin-sensitive tissue such as the muscle or liver, impair
-beta-cell production of insulin, and cause hyperglycemia {{confirm...cite}}.
+exposure of cells or animals to high concentrations of NEFA can induce insulin
+resistance in insulin-sensitive tissue such as the muscle or liver and impair
+beta-cell production of insulin, and cause hyperglycemia [@Rachek2014a;@Cnop2008a].
 Epidemiological and clinical studies have observed similar findings, showing
-that elevated total plasma NEFA associates with risk for diabetes {{cite}}.
+that elevated total plasma NEFA associates with risk for diabetes [@Daniele2014a {{more}}].
 
@@
 ```{r variableSet}
+gee_df <- get_gee_data(ds)
+
 outcomes_is <- c('linvHOMA', 'lISI')
@@
 outcomes <- c(outcomes_is, outcomes_bcf)
-ne_pct <- grep('^pct_ne', names(ds), value = TRUE)
-ne_conc <- c('ne_Total', grep('^ne\\d\\d', names(ds), value = TRUE))
+ne_pct <- grep('^pct_ne', names(gee_df), value = TRUE)
+ne_conc <- c('TotalNE', grep('^ne\\d\\d', names(gee_df), value = TRUE))
 covariates_is <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'MET', 'AlcoholPerWk')
 covariates_bcf <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'FamHistDiab')
-
-gee_df <- get_gee_data(ds)
 ```
@@
 ```{r geeData, cache=TRUE}
-gee_results <- bind_rows(
+gee_results <- dplyr::bind_rows(
     analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', adj.p = TRUE),
@@
     analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', adj.p = TRUE)
-    )
+    ) %>% 
+    tidy_gee_results()
 
@@
     analyze_gee(gee_df, outcomes, ne_conc, 'VN', 'nmol/mL', adj.p = TRUE)
-    )
+    ) %>% 
+    tidy_gee_results()
 ```
@@
 ```{r inlineResults}
-pctne <- calculate_percent_nefa_contribution(ds) 
-misclass_npct <- calculate_plsda_misclass(plsda_results)
-pctChg_o <- calculate_outcomes_pct_change(ds)
-lcmm_grps <- calculate_ngroups_lcmm(lcmm_results)
+percent.nefa <- calculate_percent_nefa_contribution(ds) 
+misclass.n.percent <- calculate_plsda_misclass(plsda_results)
+percent.change.outcomes <- calculate_outcomes_pct_change(ds)
+lcmm.groups <- calculate_ngroups_lcmm(lcmm_results)
 ```
@@
 
-The primary outcome variables declined between `r pctChg_o` over the 6 years.
+The primary outcome variables had a median decline between 
+`r percent.change.outcomes` over the 6 years in this subset of the PROMISE cohort.
 
@@
 (`r cite_sf('gee')`), particularly for the concentration data. Full model
-adjustment attenuated most of these associations, with total NEFA, 16:0, and
+adjustment attenuated most of these associations, where total NEFA, 16:0, and
 18:2n-6 (all as nmol/mL) had negative associations between IGI/IR and ISSI-2
@@
 `r cite_sf('plsda_loadings')` shows the loadings of individual NEFA by the
-extracted component, while `r cite_sf('plsda_groups')` shows the clustering of the components by ISSI-2 latent class.
+extracted component and `r cite_sf('plsda_groups')` shows the clustering of the
+components by ISSI-2 latent class.
 
@@
 composition, influences the pathogenesis of diabetes.
-
 Given we found associations with the nmol/mL data, but none in the mol% data,
@@
 
-## Beta cell
+While future studies are needed to confirm these findings, our results suggest
+that measurement of the NEFA composition may not be warranted and that the
+simple, easy to measure total NEFA may be the measure most appropriate for
+further analysis. {{reword}}
 
-We found that higher palmitic acid (16:0), oleic acid (18:1n-9), and linoleic
-acid (18:2n-6) associated with lower beta-cell function by {{number?}}. Expand
-on experimental studies, pathology/biology, mechanisms, etc.
-[@Salgin2012a;@Giacca2011a;@Xiao2009a]
-
+The role of total NEFA on diabetes is a well-studied area of research.
+Epidemiology study showing that higher NEFA in both adults and children 
+associate with lower insulin secretion [@Salgin2012a]. Also show an association 
+with higher risk for developing IGT or diabetes after approximately 6 years in 
+adults [@Salgin2012a].
 In a cross-sectional analysis of the RISCK cohort, total NEFA had a negative
 association with insulin sensitivity and, particularly strong with beta-cell
-function [@Johns2014a]. The authors further analysis showed that greater
-adiposity was not a mediator between the total NEFA and metabolic function
-association.
-
-Experimental [@Carpentier2003a]
-Review [@Carpentier2000a;@Cnop2008]
-
-Most experimental studies on beta-cell function use only palmitic acid....
-
-First study to our knowledge to examine a broad spectrum of individual NEFA in a cohort setting on beta-cell function..
-
-We found no association of any individual NEFA as mol% on beta-cell function,
-but a very strong signal with total NEFA, palmitic acid, oleic acid, and
-linoleic acid as concentrations.
-
-Using cluster techniques, we found no evidence to suggest that any individual
-NEFA contributed uniquely to beta-cell function, confirming the results we saw
-in the GEE models. This is particularly seen with the poor predictive ability of
-the NEFA composition to classify between groups of beta-cell function.
+function [@Johns2014a]. The authors' further analysis suggested that greater
+adiposity was likely not a mediator between the total NEFA and metabolic
+function association.
+{{[@Giacca2011a;@Xiao2009a]}}.
+This is the first study to our knowledge to examine a broad spectrum of
+individual NEFA in a cohort setting on beta-cell function. We found that
+individual NEFA as mol% or of the NEFA composition overall had any association
+or predictive power on determining beta-cell function, but there was a very
+strong signal with total NEFA, palmitic acid, oleic acid, and linoleic acid as
+concentrations.
+Potential mechanisms for the role of NEFA on beta-cell function are well
+described experimentally, particularly for palmitic acid. Prolonged exposure to
+the elevated NEFA can induce apoptosis in the beta-cells, likely through 
+endoplasmic reticulum stress, formation of ceramides, and generation of nitric 
+oxide, as well as can impair proinsulin production and mitochrondial function 
+[@Cnop2008a; @Maris2013a; @Carpentier2003a].
 
@@
 measured.
-
-## IR
 
@@
 have found null or weak associations of NEFA and IR 
-[@Gray2013a;@Magkos2012a;@Kehlenbrink2012a; @Johns2014a], suggesting that either
+[@Gray2013a;@Magkos2012a;@Kehlenbrink2012a;@Johns2014a], suggesting that either
 NEFA only have a minor role in IR or that NEFA is the effect, but not cause, of
@@
 develop before elevations in blood NEFA. There is also evidence to suggest that 
-insulin may be involved in NEFA update [@Ramos-Roman2012a], further supporting 
+insulin may be involved in NEFA uptake [@Ramos-Roman2012a], further supporting 
 the hypothesis that elevated NEFA is a consequence of higher IR.
 
-Another explanation for our null findings with insulin sensitivity may be
-differences in physiology between fasting and postprandial NEFA kinetics. For
-instance, some experimental studies using the clamp found that fasting NEFA was
-a weak predictor of insulin sensitivity compared to postprandial concentrations
-of NEFA [@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the
-adipose tissue following postprandial TAG lipolysis via lipoprotein lipase
-may result in NEFA spillover into the blood and a subsequent increase in circulating
-NEFA [@Almandoz2013a], which may be more metabolically active given postprandial
-activity.
-
-...Last possibility is that given these subjects were at-risk for diabetes already,
-the causal role of NEFA on IR may have long passed and it no longer influences
-IR....
-
-## Composition
+There are two other possible explanations for our null findings for the insulin
+sensitivity measures. First, there may be differences in physiology between
+fasting and postprandial NEFA kinetics. For instance, some experimental studies
+using the clamp found that fasting NEFA was a weak predictor of insulin
+sensitivity compared to postprandial concentrations of NEFA
+[@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the adipose
+tissue following postprandial TAG lipolysis via lipoprotein lipase may result in
+NEFA spillover into the blood and a subsequent increase in circulating NEFA
+[@Almandoz2013a], which may be more metabolically active given postprandial 
+activity. Second, the null findings may be due to the high risk population
+examined. It may be that in this population, IR has become well established and
+the role of NEFA may no longer be evaluated in this context.
 
 This is the first study to our knowledge to examine the composition of the NEFA
-fraction and longitudinal measures of metabolic functioning.
+fraction and longitudinal measures of metabolic functioning. A recent, well
+analyzed study used a variety of advanced statistical techniques to explore the
+multivariate relationship between the NEFA composition and components of the
+metabolic syndrome (MetS) [@Dai2015a]. Specifically, using the statistical techniques,
+the authors identified the NEFA 16:1n-9, 20:1n-9, and 22:4n-6 to most correlate
+with components of the MetS. Another statistically similar studied diabetes
+specifically and found that 16:0, 18:0, 18:1, 18:2, 18:3 may be useful
+biomarkers for identifying healthy compared to diabetic individuals [@Liu2010a].
+However, a limitation of both these studies were the smaller sample sizes
+(approximately 100 subjects) and the cross-sectional design.
 
-Composition of NEFA and metabolic syndrome components [@Dai2015a]
-
-Composition of NEFA and metabolic syndrome binary [@Novgorodtseva2011a]
-
-Using complex measurement and statistical techniques that were rigorously
-applied {{? reword}} to the composition of the NEFA fraction, Liu *et al.*
--[@Liu2010a] found that 16:0, 18:0, 18:1, 18:2, 18:3 may be useful biomarkers
-for identifying healthy compared to diabetic individuals. However, a limitation
-of this study was the smaller sample size {{confirm}} and the cross-sectional
-design.
-
-Limitations, etc...
+There are a few important limitations to our study. NEFA were only quantified at
+the baseline visit and as such we can not investigate whether there were 
+concomitant changes in NEFA and the metabolic measures. However, we believe this
+is a strength for our specific objective, as the chance of reverse causality is 
+reduced given that fatty acid and glucose metabolism pathways are tightly 
+integrated. This is also an observational cohort, and there may be some residual
+confounding we haven't considered or that couldn't be measured. Nonetheless, we 
+were as thorough as possible in understanding and minimizing potential 
+confounding. Finally, our cohort consists of individuals at-risk for diabetes, 
+who are primarily female of European-ancestry and as such our results may not be
+generalizable to other populations. However, given these limitations, our study
+also has several strengths, including the longitudinal design and the rigorous
+statistical techniques and methods applied, which are specifically suited to
+investigating temporal relationships and to handling the multivariate nature of
+the data. In addition, our cohort contains highly detailed and comprehensive
+variable measurements at each collection visit, and has both concentration and
+mol% data for the fatty acids.
 
 Final, brief conclusion...
+In conclusion, we found that NEFA
+While future studies are needed to confirm these findings, our results suggest
+that measurement of the NEFA composition may not be warranted and that the
+simple, easy to measure total NEFA may be the measure most appropriate for
+further analysis. {{reword}}
 
-## Acknowledgements ##
+## Acknowledgements
 
-# Tables #
+The authors thank Jan Neuman, Paula Van Nostrand, Stella Kink, and Annette
+Barnie of the Leadership Sinai Centre for Diabetes, Mount Sinai Hospital,
+Toronto, Canada and Sheila Porter and Mauricio Marin of the Centre for Studies
+in Family Medicine, University of Western Ontario, London, Canada for their 
+expert technical assistance and dedication in their work for PROMISE. 
+The authors had the following responsibility: LWJ conducted research, analyzed
+data, and wrote the paper; RR, ZL, BZ, and SBH designed research, conducted
+research, and provided essential materials (infrastructure and clinical
+resources); RR, BZ, SBH, RPB, and AG provided intellectual feedback on the
+paper; RPB conducted research, provided essential reagents and materials; AJH
+designed research, assisted with interpretation, and provided intellectual
+feedback on all versions of the paper; LWJ and AJH had primary responsibility
+for final content. All authors read and approved the final manuscript.  The
+authors report no potential conflicts of interest relevant to this study.
+
+# Tables
 
 ```{r tabBasic}
-# table(ds$Ethnicity, ds$f.VN)
-# table(ds$f.VN)
-summary(ds$Sex)
-ds %>%
-    outline_table(c('ne_Total', 'HOMA', 'ISI', 'IGIIR', 'ISSI2',
-                    'TAG', 'Chol', 'BMI', 'Waist', 'HDL', 'FamHistDiab',
-                    'Age', 'Ethnicity', 'Sex', 'MET', 'IFG', 'IGT',
-                    'DM'), 'f.VN') %>% 
-#     add_rows(c('HOMA', 'ISI', 'IGIIR', 'ISSI2'), stat_medianIQR, digits = 1) %>% 
-#     add_rows(c('TAG', 'Chol', 'HDL', 'ne_Total', 'BMI', 'Waist', 'Age'), 
-#              stat_meanSD, digits = 1) %>% 
-    add_rows(c('Ethnicity', 'Sex')) %>% 
-    rename_rows(renaming_table_rows) %>%
-    rename_columns('Measure', 'Baseline', '3-yr', '6-yr') %>%
-    construct_table()
-
+table_basic(ds, caption = tab_basic)
 ```
 
-# Figures #
+Note: Values are in median (IQR), mean (SD), and n (%).
+
+# Figures
 
@@
 
-```{r supFigLDA}
-
-```
-
 ```{r supTabGEE}
@@
 
-```{r supFigLCMM, fig.cap=}
+```{r supFigLCMM, fig.cap=sfig_lcmm}
 plot_lcmm_results(lcmm_data)
@@
 
-```{r supFigPlsLoadings, fig.cap=}
+```{r supFigPlsLoadings, fig.cap=sfig_plsda_loadings}
 plot_plsda_loadings(plsda_results)
@@
 
-```{r supFigPlsGrp}
+```{r supFigPlsGrp, fig.cap=sfig_plsda_groups}
 plot_plsda_grouping(plsda_results)

### 2016-04-15 R/fetch_data.R
Added glucose0 and 120, plus default arg to fetch data

@@ -5,7 +5,7 @@
 #' the new dataset as an `.RData` file. The dot in front of the function hides
 #' it from the global environment.
 #'
-.fetch_data <- function(master.dataset) {
+.fetch_data <- function(master.dataset = PROMISE::PROMISE_data) {
     # Load the master dataset,
     ds.prep <- master.dataset %>%
         filter(VN %in% c(1, 3, 6)) %>%
@@ -25,7 +25,7 @@
             SID, VN, BMI, Waist, HOMA, ISI, IGIIR, ISSI2, TAG, LDL, HDL, Chol,
             ALT, CRP, FamHistDiab, matches('meds'), Age, Sex, Ethnicity,
             IFG, IGT, DM, MET, BaseAge, AlcoholPerWk, TobaccoUse, SelfEdu, Occupation,
-            TotalNE, matches('^ne\\d+')
+            TotalNE, matches('^ne\\d+'), Glucose0, Glucose120
         ) %>%
         mutate(
             BaseTotalNE = TotalNE,

### 2016-04-15 R/functions.R
Multiple changes to function, including getting dysgly data, and making a manuscript graph theme

@@ -5,7 +5,6 @@
 ##' Custom functions used for analyses.
 ##'
 
-
 # Renaming ----------------------------------------------------------------
 
 renaming_table_rows <- function(x) {
@@ -58,7 +57,7 @@ renaming_list <- function(x) {
 # Analyze -----------------------------------------------------------------
 
 analyze_gee <- function(data, y, x, covar, unit,
-                        adj.p = FALSE, adj.p.method = 'BH', int = FALSE) {
+                        adj.p = FALSE, adj.p.method = 'BH', int = FALSE, intvar = 'VN') {
     if (int) {
         extract_term <- ':'
     } else {
@@ -71,7 +70,7 @@ analyze_gee <- function(data, y, x, covar, unit,
         mason::add_variables('xvars', x) %>%
         mason::add_variables('covariates', covar) %>% {
             if (int) {
-                mason::add_variables(., 'interaction', 'VN')
+                mason::add_variables(., 'interaction', intvar)
             } else {
                 .
             }
@@ -113,10 +112,10 @@ analyze_lcmm <- function(data, lc.var = 'lISSI2') {
     return(lcmm_results)
 }
 
-analyze_plsda <- function(data.lcmm) {
-    data.prep <- data.lcmm %>%
+analyze_plsda <- function(data, variable = 'ClassLCMM') {
+    data.prep <- data %>%
         dplyr::filter(VN == 0) %>%
-        dplyr::select(ClassLCMM, matches('^ne\\d\\d')) %>%
+        dplyr::select(contains(variable), matches('^ne\\d\\d')) %>%
         na.omit()
 
     if (length(caret::nearZeroVar(data.prep[2:dim(data.prep)[2]])) != 0)
@@ -173,12 +172,43 @@ get_gee_data <- function(data) {
     return(gee_ready_data)
 }
 
+get_dysglycemia_data <- function(data) {
+    dysgly.data <-
+        dplyr::left_join(
+            data %>%
+                dplyr::filter(VN == 0),
+            data %>%
+                dplyr::filter(!is.na(TotalNE)) %>%
+                dplyr::mutate_each(funs(ifelse(is.na(.), 0, .)), IFG, IGT) %>%
+                dplyr::mutate(PreDM = as.numeric(rowSums(.[c('IFG', 'IGT')], na.rm = TRUE))) %>%
+                dplyr::mutate(FactorDysgly = ifelse(
+                    PreDM == 1, 'PreDM',
[…]

### 2016-04-15 R/setup.R
Added showtext to figure output in knitr options (setup)

@@ -24,7 +24,6 @@ run_setup <- function() {
     if (!require(caret)) install.packages('caret')
     if (!require(fdrtool)) install.packages('fdrtool')
     if (!require(captioner)) install.packages('captioner')
-    if (!require(dplyr)) install.packages('dplyr')
     if (!require(tidyr)) install.packages('tidyr')
     if (!require(ggplot2)) install.packages('ggplot2')
     if (!require(pander)) install.packages('pander')
@@ -33,6 +32,7 @@ run_setup <- function() {
     if (!require(MuMIn)) install.packages('MuMIn')
     if (!require(geepack)) install.packages('geepack')
     if (!require(magrittr)) install.packages('magrittr')
+    if (!require(dplyr)) install.packages('dplyr')
     # - For github packages:
     # if (!require(package)) packrat::install_github('username/package')
     if (!require(mason)) {
@@ -59,7 +59,8 @@ run_setup <- function() {
 
     # For the document (knitr)
     knitr::opts_chunk$set(
-        warning = FALSE, message = FALSE, collapse = TRUE
+        warning = FALSE, message = FALSE, collapse = TRUE,
+        fig.showtext = TRUE
     )
 }
 

### 2016-04-15 doc/gee.Rmd
Added gee analysis on interact of sex, ethn, + checking sample size of analysis

@@
 suppressMessages(run_setup())
-load_data()
-dim(ds)
+ds <- load_data()
 ds %>% 
@@
 ```{r geeData}
+gee_df <- get_gee_data(ds)
+
 outcomes <- c('linvHOMA', 'lISI', 'lIGIIR', 'lISSI2')
@@
 outcomes_bcf <- c('lIGIIR', 'lISSI2')
-ne_pct <- grep('^pct_ne', names(ds), value = TRUE)
-ne_conc <- c('ne_Total', grep('^ne\\d\\d', names(ds), value = TRUE))
+ne_pct <- grep('^pct_ne', names(gee_df), value = TRUE)
+ne_conc <- c('TotalNE', grep('^ne\\d\\d', names(gee_df), value = TRUE))
 covariates0 <- c('VN')
@@
                         'ALT', 'MET', 'AlcoholPerWk', 'TAG')
-
-gee_df <- get_gee_data(ds)
 ```
@@
 ```{r geeResults_Unadj}
-gee_results0 <- bind_rows(
-    analyze_gee(gee_df, outcomes, ne_pct, covariates0, 'mol%'),
-    analyze_gee(gee_df, outcomes, ne_conc, covariates0, 'nmol/mL')
-    )
+gee_results0 <- dplyr::bind_rows(
+    analyze_gee(gee_df, outcomes, ne_pct, covariates0, 'mol%', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes, ne_conc, covariates0, 'nmol/mL', adj.p = TRUE)
+    ) %>% 
+    tidy_gee_results()
 gee_results0 %>% 
-    mutate(p.value = p.adjust(p.value, 'BH')) %>% 
     filter(p.value < 0.1) %>% 
@@
 gee_results_is <- bind_rows(
-    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%'),
-    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL'),
-    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is_tag, 'mol% (w/ tag)'),
-    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is_tag, 'nmol/mL (w/ tag)')
-    )
+    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is_tag, 'mol%', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is_tag, 'nmol/mL', adj.p = TRUE)
+    ) %>% 
+    tidy_gee_results()
 gee_results_is %>%
-    group_by(unit) %>% 
-    mutate(p.value = p.adjust(p.value, 'BH')) %>% 
-    ungroup() %>% 
     filter(p.value < 0.1) %>% 
@@
     pander('Without or with TAG for IS')
+gee_results_is %>% 
+    select(Yterms, Xterms, sample.size) %>% 
+    group_by(Yterms) %>% 
+    summarise(MinSample = min(sample.size),
+              MaxSample = max(sample.size)) %>% 
+    pander('Sample size for IS')
 ```
@@
 gee_results_bcf <- bind_rows(
-    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%'),
-    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL'),
-    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf_tag, 'mol% (w/ tag)'),
-    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf_tag, 'nmol/mL (w/ tag)')
-    )
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf_tag, 'mol% (w/ tag)', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf_tag, 'nmol/mL (w/ tag)', adj.p = TRUE)
+    ) %>% 
+    tidy_gee_results()
 gee_results_bcf %>% 
-    group_by(unit) %>% 
-    mutate(p.value = p.adjust(p.value, 'BH')) %>% 
-    ungroup() %>% 
     filter(p.value < 0.05) %>% 
@@
     pander('Without or with TAG for BCF')
+gee_results_bcf %>% 
+    select(Yterms, Xterms, sample.size) %>% 
+    group_by(Yterms) %>% 
+    summarise(MinSample = min(sample.size),
+              MaxSample = max(sample.size)) %>% 
+    pander('Sample size for BCF')
 ```
@@
 gee_results_is_int <- bind_rows(
-    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', int = TRUE),
-    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', int = TRUE),
-    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is_tag, 'mol% (w/ tag)', int = TRUE),
-    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is_tag, 'nmol/mL (w/ tag)', int = TRUE)
-    )
+    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', adj.p = TRUE, int = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', adj.p = TRUE, int = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is_tag, 'mol% (w/ tag)', adj.p = TRUE, int = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is_tag, 'nmol/mL (w/ tag)', adj.p = TRUE, int = TRUE)
+    ) %>% 
+    tidy_gee_results()
 gee_results_is_int %>%
-    group_by(unit) %>% 
-    mutate(p.value = p.adjust(p.value, 'BH')) %>% 
-    ungroup() %>% 
+    select(p.value) %>% 
+    plot()
+```
+
+```{r geeResults_IS_int_sex}
+gee_results_is_int_sex <- bind_rows(
+    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', adj.p = TRUE, int = TRUE, intvar = 'Sex'),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', adj.p = TRUE, int = TRUE, intvar = 'Sex'),
+    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is_tag, 'mol% (w/ tag)', adj.p = TRUE, int = TRUE, intvar = 'Sex'),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is_tag, 'nmol/mL (w/ tag)', adj.p = TRUE, int = TRUE, intvar = 'Sex')
+    ) %>% 
+    tidy_gee_results()
+gee_results_is_int_sex %>%
+    select(p.value) %>% 
+    plot()
+gee_results_is_int_sex %>%
+    filter(p.value < 0.05) %>% 
+    select(Yterms, Xterms, term, estimate, conf.low, conf.high, unit) %>% 
+    pander('Interactions between male and female for IS.')
+```
+
+```{r geeResults_IS_int_ethn}
+gee_results_is_int_ethn <- bind_rows(
+    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', 
+                adj.p = TRUE, int = TRUE, intvar = 'Ethnicity'),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', 
+                adj.p = TRUE, int = TRUE, intvar = 'Ethnicity'),
+    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is_tag, 'mol% (w/ tag)', 
+                adj.p = TRUE, int = TRUE, intvar = 'Ethnicity'),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is_tag, 'nmol/mL (w/ tag)', 
+                adj.p = TRUE, int = TRUE, intvar = 'Ethnicity')
+    ) %>% 
+    tidy_gee_results()
+gee_results_is_int_ethn %>%
     select(p.value) %>% 
@@
 gee_results_bcf_int <- bind_rows(
-    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', int = TRUE),
-    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', int = TRUE),
-    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf_tag, 'mol% (w/ tag)', int = TRUE),
-    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf_tag, 'nmol/mL (w/ tag)', int = TRUE)
-    )
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', adj.p = TRUE, int = TRUE),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', adj.p = TRUE, int = TRUE),
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf_tag, 'mol% (w/ tag)', adj.p = TRUE, int = TRUE),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf_tag, 'nmol/mL (w/ tag)', adj.p = TRUE, int = TRUE)
+    ) %>% 
+    tidy_gee_results()
 gee_results_bcf_int %>% 
-    group_by(unit) %>% 
-    mutate(p.value = p.adjust(p.value, 'BH')) %>% 
-    ungroup() %>% 
     select(p.value) %>% 
@@
 ```
+
+```{r geeResults_BCF_int_sex}
+gee_results_bcf_int_sex <- bind_rows(
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', 
+                adj.p = TRUE, int = TRUE, intvar = 'Sex'),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', 
+                adj.p = TRUE, int = TRUE, intvar = 'Sex'),
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf_tag, 'mol% (w/ tag)', 
+                adj.p = TRUE, int = TRUE, intvar = 'Sex'),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf_tag, 'nmol/mL (w/ tag)', 
+                adj.p = TRUE, int = TRUE, intvar = 'Sex')
+    ) %>% 
+    tidy_gee_results()
+gee_results_bcf_int_sex %>% 
+    select(p.value) %>%
+    plot()
+gee_results_bcf_int_sex %>%
+    filter(p.value < 0.05) %>% 
+    select(Yterms, Xterms, term, estimate, conf.low, conf.high, unit) %>% 
+    pander('Interactions between male and female for BCF.')
+```
+
+```{r geeResults_BCF_int_ethn}
+gee_results_bcf_int_ethn <- bind_rows(
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', 
+                adj.p = TRUE, int = TRUE, intvar = 'Ethnicity'),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', 
+                adj.p = TRUE, int = TRUE, intvar = 'Ethnicity'),
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf_tag, 'mol% (w/ tag)', 
+                adj.p = TRUE, int = TRUE, intvar = 'Ethnicity'),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf_tag, 'nmol/mL (w/ tag)', 
+                adj.p = TRUE, int = TRUE, intvar = 'Ethnicity')
+    ) %>% 
+    tidy_gee_results()
+gee_results_bcf_int_ethn %>% 
+    select(p.value) %>% 
+    plot()
+```

### 2016-04-15 doc/manuscript.Rmd
Many changes to manuscript (extra analysis, figs, dysgly, edits) before sending to Tony

@@
     - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
-csl:
-    - /home/luke/Documents/bibliography/styles/vancouver.csl
+csl: /home/luke/Documents/bibliography/styles/dependent/diabetes-care.csl
 output:
-    md_document:
-        variant: markdown
     word_document:
@@
         fig_caption: true
+        keep_md: true
 ---
@@
 
-Word count: / 4000
+Word count: 3527 / 4000
 
-Table and figure count: / 4
+Table and figure count: 3 / 4
 
@@
 
-# Abstract
-
 <!-- outline
@@
 - Precise followup time for each participant?
-- Interaction by sex or ethnicity
+- [X] Interaction by sex or ethnicity
+- [X] Interaction by time
+- [X] Sensitivity analysis with TG
+- [X] PLS-DA with dysglycemia (for anyone who got it over the years).
 
@@
 
-```{r setup, message=FALSE}
+```{r setup, message=FALSE, echo=FALSE}
 source('.Rprofile')
@@
 ds <- load_data()
+ds0yr <- filter(ds, VN == 0)
 ```
@@
 tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 clinic visits.')
-fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006)')
+fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006).')
 fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function using generalized estimating equations over the 6 years in the PROMISE cohort.
-Adjusted for {{covariates}}.
+Adjusted for time, sex, ethnicity, baseline age, WC,
+ALT, and family history of diabetes (plus physical activity and alcohol intake for beta-cell function models).
 Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
-P-values presented were adjusted for the false discovery rate.')
+P-values were adjusted for the BH false discovery rate, presented as the dot size.')
 
@@
 stab_qic <- stab('qic', 'Comparing generalized estimating equation models adjusting for different covariates using Quasi-Likelihood Information Criterion.')
-sfig_corr <- sfig('corr', 'Pearson correlation heatmap of non-esterified fatty acids (nmol/mL) and basic PROMISE participant characteristics for the baseline visit (2004-2006)')
+sfig_corr <- sfig('corr', 'Pearson correlation heatmap of non-esterified fatty acids (nmol/mL) and basic PROMISE participant characteristics for the baseline visit (2004-2006).
+Darkness of the colour indicates the size of the correlation, with blue indicating positive and orange indicating negative correlations.')
 sfig_gee <- sfig('gee', 'Unadjusted generalized estimating equation modeling of the longitudinal association of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort.
@@
 Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
-P-values presented were adjusted for the false discovery rate.')
+P-values were adjusted for the BH false discovery rate, presented as the dot size.')
 stab_gee <- stab('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function using generalized estimating equations over the 6 years in the PROMISE cohort.
-Adjusted for {{covariates}}.
+Adjusted for time, sex, ethnicity, baseline age, WC,
+ALT, and family history of diabetes (plus physical activity and alcohol intake for beta-cell function models).
 Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
-P-values presented were adjusted for the BH false discovery rate.')
+P-values were adjusted for the BH false discovery rate, indicated by asterisk.')
 stab_gee_unadj <- stab('gee_unadj', 'Unadjusted generalized estimating equations models of the longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
@@
 Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
-P-values presented were adjusted for the BH false discovery rate.')
+P-values were adjusted for the BH false discovery rate, indicated by asterisk.')
 sfig_lcmm <- sfig('lcmm', 'Latent class mixed model analysis to identify individual classes of trajectories for log(ISSI-2) over the 6 years in the PROMISE cohort.')
+sfig_plsda_groups <- sfig('plsda_groups', 'Clustering of extracted components from the partial least squares discriminant analysis on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006).
+The percent explained variance of each component is shown in brackets on each axis.
+{{colour}}, {{colour}}, {{colour}} lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis.')
 sfig_plsda_loadings <- sfig('plsda_loadings', 'Pattern loadings from partial least squares discriminant analysis to identify potential clusters of NEFA composition within the classes extracted from the latent class mixed model in 463 participants from the baseline PROMISE visit (2004-2006).
 The percent explained variance of each component is shown in brackets on each axis.')
-sfig_plsda_groups <- sfig('plsda_groups', 'Clustering of extracted components from the partial least squares discriminant analysis on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006).
+sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', 'Clustering of extracted components from the partial least squares discriminant analysis for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).
+The percent explained variance of each component is shown in brackets on each axis.
+{{colour}} lines indicate dysglycemia conversion or maintanence and {{colour}} lines indicate no dysglycemia status.')
+sfig_plsda_loadings_dysgly <- sfig('plsda_loadings_dysgly', 'Pattern loadings from partial least squares discriminant analysis to identify potential clusters of NEFA composition for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).
 The percent explained variance of each component is shown in brackets on each axis.')
@@
 the pathogenesis of type 2 diabetes mellitus. Experimental work has shown that 
-exposure of cells or animals to high concentrations of NEFA can induce insulin
-resistance in insulin-sensitive tissue such as the muscle or liver and impair
-beta-cell production of insulin, and cause hyperglycemia [@Rachek2014a;@Cnop2008a].
-Epidemiological and clinical studies have observed similar findings, showing
-that elevated total plasma NEFA associates with risk for diabetes [@Daniele2014a {{more}}].
+exposure to high concentrations of NEFA can induce insulin resistance in
+insulin-sensitive tissue such as the muscle or liver and impair beta-cell
+production of insulin [@Rachek2014a;@Cnop2008a]. Epidemiological and clinical
+studies have observed similar findings, showing that elevated total plasma NEFA
+associates with risk for diabetes [@Daniele2014a;@Salgin2012a].
 
-Most previous experimental work uses palmitic acid (16:0), oleic acid (18:1n-9),
-or a standard oil such as soybean {{confirm, cite}} and generalize as total
-NEFA. However, fatty acids comprise multiple classes of molecules with diverse 
-physiological functions, with few studies comparing different fatty acids. For
-instance, one study on beta-cell {{plates? lines?}} suggests that
+Much of the previous experimental work frequently use fatty acids such as
+palmitic acid (16:0) or oleic acid (18:1n-9), or use a standard oil such as soybean
+oil as the exposure and generalize as total NEFA. However, fatty acids comprise
+multiple classes of molecules with diverse physiological functions, with few
+studies comparing different fatty acids. For instance, one study suggests that
 eicosapentaenoic acid (20:5n-3) can protect against the lipotoxic effect of
-palmitic acid {{xiao? cite}}.
+palmitic acid in the beta-cells [@Xiao2006].
 
-Prospective with diabetes [@Ilyasova2010a]
-
-Lifestyle interventions aimed at managing or preventing diabetes show a
-concomitant reduction in NEFA and insulin resistance {{cite}}.
-
-Lack of longitudinal data is particularly important given that NEFA release from
-adipocytes is inhibited by insulin, so if there is IR, more NEFA are released,
-making it difficult to disentangle which caused which.
-
-Despite the fairly extensive literature on NEFA and diabetes, there are
-substantial research gaps. Importantly, most of the experimental work using
-individual NEFA species has been in animal and cell lines or if in humans has
-been acute, infusion studies. For epidemiological studies, few have examined the
-role of total, but not individual, NEFA on the pathogenesis of diabetes and
-fewer still have been longitudinal, which is key to understanding causal
-associations and given the long timescale in diabetes pathophysiology.
-Presently, no longitudinal study has been conducted that has examined the role
-of the composition of the NEFA fraction on the pathogenesis of diabetes. Our
-research objective was to examine the role of the NEFA composition
-on insulin sensitivity and beta-cell function using longitudinal and clustering
-techniques. Some specific hypotheses included that higher palmitic acid while
-lower polyunsaturated fatty acids such as eicosapentaenoic acid would associate
-with lower insulin sensitivity and beta-cell function.
+Despite a sizable literature studying the role of NEFA in diabetes, there are 
+important gaps in the research. In particular, there have been few long term 
+studies in humans that have examined a range of NEFA species. Most studies have 
+been on animals and cells, have been short term human infusion trials, or have 
+been epidemiological studies that only looked at total NEFA. To date, there have
+been no longitudinal studies examining the role of the composition of the NEFA 
+blood lipid fraction on the pathogenesis of diabetes, which is key to 
+understanding causal associations given the chronic nature of diabetes 
+pathophysiology. Therefore, our research objective was to examine the role of
+the serum NEFA composition on insulin sensitivity and beta-cell function in a
+longitudinal cohort. Some specific hypotheses included that higher palmitic acid
+while lower polyunsaturated fatty acids such as eicosapentaenoic acid would 
+associate with lower insulin sensitivity and beta-cell function.
 
@@
 and/or a history of gestational diabetes or birth of a macrosomic infant. 
-Participants aged 30 years and older (n=726) conducted the baseline visit
+Participants aged 30 years and older (n=736) attended the baseline visit
 between 2004-2006. Follow-up examinations occur every three years, with three
@@
 standard prior to extracting total lipids according to the method of Folch 
-[@Folch1957a].  Each serum lipid fraction was isolated using thin layer 
-chromatography, which were visualized under UV light after lightly spraying with
-8-anilino-1-naphthalene sulfonic acid (0.1% wt/vol) and then converted to fatty
-acid methyl esters with 14% boron trifluoride in methanol at 100°C for 1 h. 
-Fatty acid methyl esters were separated and quantified using a Varian-430 gas
-chromatograph (Varian, Lake Forest, CA, USA) equipped with a Varian Factor Four
-capillary column and a flame ionization detector, which were injected in
-splitless mode. Fatty acid concentrations (nmol/ml) were calculated by
-proportional comparison of gas chromatography peak areas to that of the internal
-standards [@Nishi2014a; @Chen2011a].  There were 22 fatty acids measured in the
-NEFA fraction.
+[@Folch1957a]. Each serum lipid fraction (NEFA, cholesteryl ester, phospholipid,
+and triacylglycerol) was isolated using thin layer chromatography, which were
+visualized under UV light after lightly spraying with 8-anilino-1-naphthalene
+sulfonic acid (0.1% wt/vol) and then converted to fatty acid methyl esters with
+14% boron trifluoride in methanol at 100°C for 1 h. Fatty acid methyl esters
+were separated and quantified using a Varian-430 gas chromatograph (Varian, Lake
+Forest, CA, USA) equipped with a Varian Factor Four capillary column and a flame
+ionization detector, which were injected in splitless mode. Fatty acid
+concentrations (nmol/ml) were calculated by proportional comparison of gas
+chromatography peak areas to that of the internal standards [@Nishi2014a;
+@Chen2011a].  There were 22 fatty acids measured in the NEFA fraction. Given the
+diverse biology, as well as the complexity of the analyses, the other lipid
+fractions were split into separate reports (see {{pl and ce paper}} for the
+analysis of the phospholipid and cholesteryl ester fractions).
 
@@
 outcome variables and the predictor variables. Given the longitudinal design 
-and auto-regressive (AR1) correlation matrix was chosen for the GEE models, 
+an auto-regressive (AR1) correlation matrix was chosen for the GEE models, 
 though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
@@
 
-Covariates selection was based on previous literature, directed acyclic graph 
+Covariate selection was based on previous literature, directed acyclic graph 
 [@Greenland1999a] recommendations, and quasi-likelihood information criterion 
-(QIC). `r cite_st('qic')` shows the covariates compared using QIC and `r 
-cite_sf('dag')` shows the specified directed acyclic graph. The final GEE model 
+(QIC). `r cite_st('qic')` shows the covariates compared using QIC and 
+`r cite_sf('dag')` shows the specified directed acyclic graph. The final GEE model 
 we selected differed between insulin sensitivity and beta-cell function models. 
@@
 predictor variable given the covariates are held constant. While models with a
-time interaction did not perform better than others for the QIC (M6.1 and M8.1
+time interaction did not perform better than others for the QIC (M6.1 and M8.1)
 in `r cite_st('qic')`, we ran models with a time by predictor term for each of
-the NEFA to test if there was a time interaction. Given that TG is a risk factor
+the NEFA to test if there was an interaction by time. Given that TG is a risk factor
 for diabetes and that NEFA contribute to TG production, TG may act as a mediator
@@
 association between NEFA and the outcomes, TG was included in the GEE model as a
-sensitivity analysis.
+sensitivity analysis. Lastly for the GEE models, we tested for an interaction
+with sex or ethnicity by the predictor term for the outcome variables.
 
@@
 identify the patterns of NEFA composition against beta-cell function (using ISSI-2,
-which has most of the associations in the GEE models). Since PLS-DA can't handle
-longitudinal data, the longitudinal data was maximized by extracting latent
-trajectories of beta-cell function over the 6 years using latent class mixed
-models (LCMM). A limitation of these methods is that they can not be adjusted
-for covariates.
+which has most of the associations in the GEE models) and for those who converted or
+maintained dysglycemia status (either IFG, IGT, or DM) over the 6 years. Since
+PLS-DA can't handle longitudinal data, the longitudinal data was included by
+extracting latent trajectories of beta-cell function (ISSI-2) over the 6 years using
+latent class mixed models (LCMM). A limitation of these methods, excluding LCMM,
+is that they can not be adjusted for covariates.
 
@@
 statistically significance at p<0.05, after adjusting for multiple testing using
-the False Discovery Rate [@Benjamini1995a].
+the Benjamini-Hochberg False Discovery Rate [@Benjamini1995a].
 
@@
 
-```{r variableSet}
+```{r variableAndDataSet, message=FALSE, results='hide'}
 gee_df <- get_gee_data(ds)
+dysgly_data <- get_dysglycemia_data(ds)
 
@@
 
-```{r lcmmPls}
+```{r lcmmPls, cache=TRUE}
 # include also the 3-yr and 6-yr data for training/testing.
@@
 plsda_results <- analyze_plsda(lcmm_data)
+plsda_dysgly <- dysgly_data %>% 
+    mutate(ConvertDysgly = factor(ConvertDysgly, levels = c(0, 1), labels = c('No', 'Yes'))) %>% 
+    analyze_plsda('ConvertDysgly')
 ```
@@
 percent.nefa <- calculate_percent_nefa_contribution(ds) 
-misclass.n.percent <- calculate_plsda_misclass(plsda_results)
+misclass.n.percent.bcf <- calculate_plsda_misclass(plsda_results)
+misclass.n.percent.dysgly <- calculate_plsda_misclass(plsda_dysgly)
 percent.change.outcomes <- calculate_outcomes_pct_change(ds)
 lcmm.groups <- calculate_ngroups_lcmm(lcmm_results)
+dm.convert <- calculate_conversion_dysgly(dysgly_data, 'ConvertDM')
+predm.convert <- calculate_conversion_dysgly(dysgly_data, 'ConvertPreDM')
+sex.npct <- calculate_factor_npercent(ds0yr$Sex, 'Female')
+ethn.npct <- calculate_factor_npercent(ds0yr$Ethnicity, 'European')
+fhd.npct <- calculate_factor_npercent(ds0yr$FamHistDiab, 'Yes')
+est.gee <- calculate_gee_magnitude(gee_results)
 ```
@@
 
-The primary outcome variables had a median decline between 
-`r percent.change.outcomes` over the 6 years in this subset of the PROMISE cohort.
+Within this subset of the PROMISE cohort at the baseline visit, 
+`r sex.npct$npct` were females and `r ethn.npct$npct` had European-ancestry.
+The average age of the participants was `r aide::average(ds0yr$Age)` years 
+(`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
+`r aide::average(ds0yr$BMI)` (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
+subset of the cohort, `r fhd.npct$npct`, had a family history of diabetes.
+The primary outcome variables had a significant median decline between 
+`r percent.change.outcomes$chg` (p`r percent.change.outcomes$p`) over the 6
+years in this subset of the PROMISE cohort.
+Consistent with this decline, there were 42 (9%) <!--`r dm.convert`-->
+participants who developed diabetes while `r predm.convert` participants either
+converted to or maintained IFG or IGT status over the 6 years.
 
 `r cite_f('nefaDist')` shows the distribution of the NEFA composition. Four 
-individual NEFA made up the vast majority (`r sum(pct_nefa$pct)`%) of the total
-NEFA fraction. The largest contributors were `r pctne$c[1]`, `r pctne$c[2]`, 
-`r pctne$c[3]`, and `r pctne$c[4]`. All individual NEFA as well as the total
-fraction had weak to no correlations (r<0.3) with the basic characteristics (see
-`r cite_sf('corr')`).
+individual NEFA made up the vast majority (`r sum(percent.nefa$pct)`%) of the total
+NEFA fraction. The largest contributors were `r percent.nefa$c[1]`, 
+`r percent.nefa$c[2]`, `r percent.nefa$c[3]`, and `r percent.nefa$c[4]`. All
+individual NEFA as well as the total fraction had weak to no correlations
+(r<0.3) with the basic characteristics (see `r cite_sf('corr')`).
 
@@
 Several associations were seen in the unadjusted GEE models 
-(`r cite_sf('gee')`), particularly for the concentration data. Full model
-adjustment attenuated most of these associations, where total NEFA, 16:0, and
-18:2n-6 (all as nmol/mL) had negative associations between IGI/IR and ISSI-2
-(see `r cite_f('gee')`). The magnitude of association for each of these
-variables was fairly consistent for each beta-cell function measure. For every
-one SD decrease in any of these three NEFA variables, there was a predicted
-{{magnitude}}% lower IGI/IR and a {{magnitude}}% lower ISSI-2 at each clinic
-visit. None of the NEFA variables as a mol% had any association with the
+(see `r cite_sf('gee')`), particularly for the concentration data. Full model 
+adjustment (`r cite_f('gee')`) attenuated most of these associations, where
+total NEFA, 16:0, and 18:2n-6 (all as nmol/mL) had negative associations between
+IGI/IR and ISSI-2. The magnitude of association for each of these variables was
+fairly consistent for each beta-cell function measure. For every one SD decrease
+in any of these three NEFA variables, there was an average predicted
+`r est.gee$est[1]`% lower IGI/IR and `r est.gee$est[2]`% lower ISSI-2 at each
+clinic visit. None of the NEFA variables as a mol% had any association with the
 outcomes. Adjusting for TG attenuates all associations with IGI/IR and ISSI-2
@@
 in those who had TG measured. There were no significant interaction effects
-between time and the individual NEFA on any of the outcome measures (data not
-shown). Raw values from the GEE models are shown in `r cite_st('gee_unadj')` for
-unadjusted models and `r cite_st('gee')` for fully adjusted models.
+between time, sex, or ethnicity and the individual NEFA on any of the outcome
+measures (data not shown). Raw values from the GEE models are shown in 
+`r cite_st('gee_unadj')` for unadjusted models and `r cite_st('gee')` for fully
+adjusted models.
 
@@
 
-Three latent classes were extracted from LCMM, with `r lcmm_grps[2,2]` 
-participants in the high, `r lcmm_grps[1,2]` in the middle, and `lcmm_grps[3,2]`
+Three latent classes were extracted from LCMM, with `r lcmm.groups$Freq[2]` 
+participants in the high, `r lcmm.groups$Freq[1]` in the middle, and 
+`r lcmm.groups$Freq[3]`
 in the low beta-cell function groups. The trajectories of the three groups are
-shown in `r cite_sf('lcmm')`. These groups were used as the response variables
-in the PLS-DA analysis. The PLS-DA results show a very poor discriminatory
-abililty of the NEFA composition in classifying participants into the correct
-beta-cell function latent class. Only `r misclass_npct[2,1]` 
-(`r misclass_npct[2,4]`) participants were correctly classified, primarily into the
+shown in `r cite_sf('lcmm')`. These groups were used as the beta-cell function
+response variables in the PLS-DA analysis. The PLS-DA results show a very poor
+discriminatory ability of the NEFA composition in classifying participants into
+the correct beta-cell function latent class. Only 
+`r misclass.n.percent.bcf$NPct[2]` 
+participants were correctly classified, primarily into the
 middle group. The specific NEFA composition was not able to accurately classify
 participants who had low or high beta-cell function.
-`r cite_sf('plsda_loadings')` shows the loadings of individual NEFA by the
-extracted component and `r cite_sf('plsda_groups')` shows the clustering of the
-components by ISSI-2 latent class.
+`r cite_sf('plsda_groups')` shows the loadings of individual NEFA by the
+extracted component and `r cite_sf('plsda_loadings')` shows the clustering of
+the components by ISSI-2 latent class. The PLS-DA analysis using dysglycemia
+status conversion or maintanence as the response variable showed similar poor 
+discriminatory ability, though better than the beta-cell function trajectory
+results. While `r misclass.n.percent.dysgly$NPct[2]` participants were
+classified correctly based on dysglycemia status, examining the clustering of
+the NEFA composition showed poor discrimination between groups (see
+`r cite_sf('plsda_groups_dysgly')` and `r cite_sf('plsda_loadings_dysgly')`).
 
@@
 In a Canadian population of adults who are at-risk for diabetes, we found that
-higher total NEFA independently predicted a lower beta-cell function over 6
-years. In particular, the specific composition of the NEFA fraction had no
-association with any of the metabolic outcomes within this cohort. These
-findings suggest that the size of the total fraction, rather than the
-composition, influences the pathogenesis of diabetes.
-Given we found associations with the nmol/mL data, but none in the mol% data,
-this suggests to us that the total size of the NEFA fraction is more important
-in the pathogenesis of diabetes than is the specific FA composition. A
-limitation of the mol% data is the inherent dependency between the FA as an
-increase in one FA by definition means other FA must decrease. In the context of
-this dependency, the associations we found for 16:0 and 18:2n-6 as nmol/mL
-suggests that these NEFA may be contributing to the harmful association seen for
-the total NEFA fraction, irrespective of changes in the other NEFA.
-
-While future studies are needed to confirm these findings, our results suggest
-that measurement of the NEFA composition may not be warranted and that the
-simple, easy to measure total NEFA may be the measure most appropriate for
-further analysis. {{reword}}
+higher total NEFA independently predicted a lower beta-cell function over 6 years.
+While we found negative association with palmitic acid (16:0), oleic acid (18:1n-9), and
+linoleic acid (18:2n-6) as a concentration (nmol/mL), no associations were seen
+for the proportion of these fatty acids and further clustering analysis found no
+predictive ability of these NEFA on beta-cell function or dysglycemia. These
+observations suggest that the absolute size of the total fraction, rather than
+the specific composition of any individual NEFA, likely influences the
+pathogenesis of diabetes.
 
 The role of total NEFA on diabetes is a well-studied area of research. 
-Epidemiology study showing that higher NEFA in both adults and children 
-associate with lower insulin secretion [@Salgin2012a]. Also show an association 
-with higher risk for developing IGT or diabetes after approximately 6 years in 
-adults [@Salgin2012a].
-In a cross-sectional analysis of the RISCK cohort, total NEFA had a negative
-association with insulin sensitivity and, particularly strong with beta-cell
-function [@Johns2014a]. The authors' further analysis suggested that greater
-adiposity was likely not a mediator between the total NEFA and metabolic
-function association.
-{{[@Giacca2011a;@Xiao2009a]}}.
-This is the first study to our knowledge to examine a broad spectrum of
-individual NEFA in a cohort setting on beta-cell function. We found that
-individual NEFA as mol% or of the NEFA composition overall had any association
-or predictive power on determining beta-cell function, but there was a very
-strong signal with total NEFA, palmitic acid, oleic acid, and linoleic acid as
-concentrations.
-Potential mechanisms for the role of NEFA on beta-cell function are well
-described experimentally, particularly for palmitic acid. Prolonged exposure to
-the elevated NEFA can induce apoptosis in the beta-cells, likely through 
-endoplasmic reticulum stress, formation of ceramides, and generation of nitric 
-oxide, as well as can impair proinsulin production and mitochrondial function 
-[@Cnop2008a; @Maris2013a; @Carpentier2003a].
+Epidemiological studies have shown that higher NEFA associate with lower insulin
+secretion and a higher risk for developing diabetes 
+[@Salgin2012a;@Pankow2004a]. In a cross-sectional analysis of the RISCK cohort,
+total NEFA had a negative association with insulin sensitivity and a
+particularly strong negative association with beta-cell function [@Johns2014a].
+Experimentally, several potential mechanisms have been elucidated for the role
+of NEFA on beta-cell function, particularly for palmitic acid. Prolonged
+exposure to elevated NEFA can induce apoptosis in the beta-cells, possibly
+through endoplasmic reticulum stress, formation of ceramides, and generation of
+nitric oxide, as well as impairment of proinsulin production and mitochrondial
+function [@Cnop2008a; @Maris2013a; @Carpentier2003a; @Giacca2011a; @Xiao2009a].
+The present analysis is the first study to our knowledge to examine a broad
+spectrum of individual NEFA in a cohort setting on beta-cell function. We found
+that there was a strong signal of higher total NEFA, palmitic acid, oleic acid,
+and linoleic acid as concentrations with lower beta-cell function. However, as
+both a mol% and using clustering analyses, no individual or specific
+composition of the NEFA fraction predicted lower beta-cell function, insulin
+sensitivity, or dysglycemia status. Together, these results suggest that the
+absolute size of the NEFA fraction, irrespective of any specific composition of
+fatty acid species, exerts it's hypothesized lipotoxic effects on the
+beta-cells.
 
-In the sensitivity analysis adjusting for TAG, all associations with beta-cell 
-function were attenuated. While there is strong experimental evidence on the 
-role of chronically elevated NEFA, particularly palmitic acid, on impaired 
-beta-cell function, within a free-living population NEFA may be mediating it's 
-association through TAG. In normal metabolism, NEFA enters the liver and assists
-in the production of TAG that is to be processed into very-low density 
-lipoproteins (VLDL) [@Nielsen2012a]. Elevated NEFA may be mediating it's
-association through hypertriglyceridemia, a known risk factor for diabetes.
-However, a major limitation to the TAG-adjusted GEE models is the reduced sample
-size in the 6 year visit, as only a subset (n=128) of participants had TAG
-measured.
+Biologically, within a free-living population chronic elevation of NEFA may be 
+mediating it's association through TG. In normal metabolism, NEFA enters the 
+liver and assists in the production of TG that is to be processed into very-low
+density lipoproteins (VLDL) [@Nielsen2012a]. As such higher NEFA may contribute 
+to hypertriglyceridemia, which is a known risk factor for diabetes. In the
+sensitivity analysis adjusting for TG, all associations with beta-cell function
+were attenuated, suggesting that NEFA may in fact be mediating it's association
+through TG. However, a major limitation to the TG-adjusted GEE models is the
+reduced sample size in the 6 year visit, as only a subset (n=128) of
+participants had TG measured.
 
@@
 [@Daniele2014a;@Liang2013a] trials and the observational research was generally
-cross-sectional [@Miller2012a;@Johns2014a;@Bush2012a]. The limitation of these
-studies were the short time periods examined and the experimentally-induced
+cross-sectional or correlational
+[@Miller2012a;@Johns2014a;@Bush2012a;@Ilyasova2010a]. The limitation of these
+studies were the short time periods examined or the experimentally-induced
 elevations in NEFA concentrations, making it difficult to assess the causal role
@@
 [@Gray2013a;@Magkos2012a;@Kehlenbrink2012a;@Johns2014a], suggesting that either
-NEFA only have a minor role in IR or that NEFA is the effect, but not cause, of
+NEFA has only a minor role in IR or that NEFA is the effect, but not cause, of
 the underlying IR. Adipose tissue, which release NEFA during fasting, can become
-insulin resistant and thus lipolysis of intracellular TAG by hormone-sensitive 
+insulin resistant and thus lipolysis of intracellular TG by hormone-sensitive 
 lipase may not respond to the inhibitory action of insulin. As such, IR may 
@@
 insulin may be involved in NEFA uptake [@Ramos-Roman2012a], further supporting 
-the hypothesis that elevated NEFA is a consequence of higher IR.
+the hypothesis that elevated NEFA is a consequence of higher IR. However, we 
+could not examine this potential hypothesis as NEFA was only collected at the
+baseline visit.
 
@@
 [@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the adipose
-tissue following postprandial TAG lipolysis via lipoprotein lipase may result in
+tissue following postprandial TG lipolysis via lipoprotein lipase may result in
 NEFA spillover into the blood and a subsequent increase in circulating NEFA
@@
 activity. Second, the null findings may be due to the high risk population
-examined. It may be that in this population, IR has become well established and
-the role of NEFA may no longer be evaluated in this context.
+examined in PROMISE. It may be that in this population, IR has become well
+established and NEFA may not contribute to IR at this stage in the pathogenesis
+of diabetes.
 
-This is the first study to our knowledge to examine the composition of the NEFA
-fraction and longitudinal measures of metabolic functioning. A recent, well
-analyzed study used a variety of advanced statistical techniques to explore the
-multivariate relationship between the NEFA composition and components of the
-metabolic syndrome (MetS) [@Dai2015a]. Specifically, using the statistical techniques,
-the authors identified the NEFA 16:1n-9, 20:1n-9, and 22:4n-6 to most correlate
-with components of the MetS. Another statistically similar studied diabetes
-specifically and found that 16:0, 18:0, 18:1, 18:2, 18:3 may be useful
-biomarkers for identifying healthy compared to diabetic individuals [@Liu2010a].
-However, a limitation of both these studies were the smaller sample sizes
-(approximately 100 subjects) and the cross-sectional design.
+Few studies have examined the composition of NEFA on metabolic functioning. One
+recent, well analyzed study used a variety of advanced fatty acid measurement
+and statistical techniques to explore the multivariate relationship between the
+NEFA composition and components of the metabolic syndrome (MetS) [@Dai2015a].
+Specifically, using the statistical techniques the authors identified NEFA
+16:1n-9, 20:1n-9, and 22:4n-6 to most correlate with components of the MetS. 
+Another similar study examining diabetes found that 16:0, 18:0, 18:1, 18:2, 18:3
+may be useful biomarkers for identifying healthy compared to diabetic
+individuals [@Liu2010a]. However, in both studies, major limitations were the
+smaller sample sizes (approximately 100 subjects) and the cross-sectional
+design.
 
@@
 investigating temporal relationships and to handling the multivariate nature of
-the data. In addition, our cohort contains highly detailed and comprehensive
+the data. Lastly, our cohort contains highly detailed and comprehensive
 variable measurements at each collection visit, and has both concentration and
@@
 
-Final, brief conclusion...
-In conclusion, we found that NEFA
-While future studies are needed to confirm these findings, our results suggest
-that measurement of the NEFA composition may not be warranted and that the
-simple, easy to measure total NEFA may be the measure most appropriate for
-further analysis. {{reword}}
+In conclusion, we found that total NEFA was a strong predictor for lower
+beta-cell function over 6 years, irrespective of the specific composition of the
+NEFA fraction. While future studies are needed to confirm these findings, given
+the cost and labour involved in quantifying the NEFA composition, our results
+suggest that further research on the NEFA composition may not be as fruitful and
+cost-effective then if the simpler, easy to measure total NEFA concentration was
+used.
 
@@
 
-![`r sfig_consort`](img/flowDiagramSample.png)
+![`r sfig_consort`](../img/flowDiagramSample.png)
 
@@
 prep_qic <- gee_df %>%
-    select(
-        SID, VN, lISSI2, lISI, ne_Total, Sex, Ethnicity, MET, ALT, BaseAge,
+    dplyr::select(
+        SID, VN, lISSI2, lISI, TotalNE, Sex, Ethnicity, MET, ALT, BaseAge,
         Waist, BMI, AlcoholPerWk, TobaccoUse, FamHistDiab
@@
 qic_is <- prep_qic %>% 
-    select(-lISSI2) %>% 
+    dplyr::select(-lISSI2) %>% 
     na.omit()
@@
 qic_bcf <- prep_qic %>% 
-    select(-lISI) %>% 
+    dplyr::select(-lISI) %>% 
     na.omit()
@@
 M0 <- geepack::geeglm(
-    lISI ~ ne_Total + VN, data = qic_is,
+    lISI ~ TotalNE + VN, data = qic_is,
     id = SID, family = gaussian,
@@
 M6 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse) # Best
-M6.1 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse + ne_Total:VN) # Best
+M6.1 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse + TotalNE:VN) # Best
 
-qic_is_tab <- model.sel(M0, M1, M2, M3, M4, M5, M6, M6.1, rank = QIC) %>% 
-    add_rownames() %>%
-    select(Model = rowname, QIC, Delta = delta) %>%
-    mutate(id = seq_len(nrow(.))) %>%
-    bind_rows(., data.frame(Model = '**log(ISI)**', id = 0.5))
+qic_is_tab <- MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M6.1, rank = QIC) %>% 
+    dplyr::add_rownames() %>%
+    dplyr::select(Model = rowname, QIC, Delta = delta) %>%
+    dplyr::mutate(id = seq_len(nrow(.))) %>%
+    dplyr::bind_rows(., data.frame(Model = '**log(ISI)**', id = 0.5))
 
@@
 M0 <- geepack::geeglm(
-    lISSI2 ~ ne_Total + VN, data = qic_bcf,
+    lISSI2 ~ TotalNE + VN, data = qic_bcf,
     id = SID, family = gaussian,
@@
 M8 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab) # best
-M8.1 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + ne_Total:VN) # best
+M8.1 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TotalNE:VN) # best
 
-qic_bcf_tab <- model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M8.1, rank = QIC) %>% 
-    add_rownames() %>%
-    select(Model = rowname, QIC, Delta = delta) %>% 
-    mutate(id = seq_len(nrow(.))+8) %>%
-    bind_rows(., data.frame(Model = '**log(ISSI-2)**', id = 8.5))
+qic_bcf_tab <- MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M8.1, rank = QIC) %>% 
+    dplyr::add_rownames() %>%
+    dplyr::select(Model = rowname, QIC, Delta = delta) %>% 
+    dplyr::mutate(id = seq_len(nrow(.))+8) %>%
+    dplyr::bind_rows(., data.frame(Model = '**log(ISSI-2)**', id = 8.5))
 
 rbind(qic_is_tab, qic_bcf_tab) %>% 
-    arrange(id) %>%
-    select(-id) %>%
-    mutate(QIC = round(QIC, 1),
+    dplyr::arrange(id) %>%
+    dplyr::select(-id) %>%
+    dplyr::mutate(QIC = round(QIC, 1),
            Delta = round(Delta, 1)) %>% 
-    pander(missing = '', caption = stab_qic)
+    pander::pander(missing = '', caption = stab_qic)
 ```
@@
 
+```{r supFigPlsGrp, fig.cap=sfig_plsda_groups}
+plot_plsda_grouping(plsda_results)
+```
+
 ```{r supFigPlsLoadings, fig.cap=sfig_plsda_loadings}
@@
 
-```{r supFigPlsGrp, fig.cap=sfig_plsda_groups}
-plot_plsda_grouping(plsda_results)
+```{r supFigPlsGrpDysgly, fig.cap=sfig_plsda_groups_dysgly}
+plot_plsda_grouping(plsda_dysgly)
+```
+
+```{r supFigPlsLoadingsDysgly, fig.cap=sfig_plsda_loadings_dysgly}
+plot_plsda_loadings(plsda_dysgly)
 ```

### 2016-05-02 R/fetch_data.R
Minor edit to fetch_data + added BaseTAG for sensit analy

@@ -5,9 +5,9 @@
 #' the new dataset as an `.RData` file. The dot in front of the function hides
 #' it from the global environment.
 #'
-.fetch_data <- function(master.dataset = PROMISE::PROMISE_data) {
+.fetch_data <- function() {
     # Load the master dataset,
-    ds.prep <- master.dataset %>%
+    ds.prep <- PROMISE::PROMISE_data %>%
         filter(VN %in% c(1, 3, 6)) %>%
         ## Kick out Canoers
         filter(is.na(Canoe)) %>%
@@ -20,12 +20,12 @@
         full_join(ds.prep,
                   ds.prep %>%
                       filter(VN == 1) %>%
-                      select(SID, BaseAge = Age)) %>%
+                      select(SID, BaseAge = Age, BaseTAG = TAG)) %>%
         select(
             SID, VN, BMI, Waist, HOMA, ISI, IGIIR, ISSI2, TAG, LDL, HDL, Chol,
             ALT, CRP, FamHistDiab, matches('meds'), Age, Sex, Ethnicity,
             IFG, IGT, DM, MET, BaseAge, AlcoholPerWk, TobaccoUse, SelfEdu, Occupation,
-            TotalNE, matches('^ne\\d+'), Glucose0, Glucose120
+            TotalNE, matches('^ne\\d+'), Glucose0, Glucose120, BaseTAG
         ) %>%
         mutate(
             BaseTotalNE = TotalNE,

### 2016-05-02 R/load_data.R
Updated load_data script to match what I did in prodigenr

@@ -1,37 +1,38 @@
-# ---NO NEED TO EDIT---
+# ---(LIKELY) NO NEED TO EDIT---
 #
-#' Load or fetch the dataset for the project
-#'
-#' @param master.dataset File path to the master dataset to fetch and wrangle from.
-#' @param local.dataset File path to the local dataset (eg. \code{data/}).
-#' @param fetch.script File path to the \code{fetch_data} function.
-#'
-#' @return Either updates the dataset or loads it in the working environment.
-#' @export
-#'
-#' @examples
-#' \dontrun{
-#' load_data('/path/to/master/dataset/')
-#' load_data('/path/to/master/dataset/',
-#'      'data/dsname.RData',
-#'      'R/fetchScript.R')
-#' }
-load_data <- function(master.dataset = PROMISE::PROMISE_data,
-                      local.dataset = file.path('data', 'data.Rds'),
+# `load_data()` either loads the dataset inside the `data/` folder or, if the
+# `fetch_data.R` file has been changed, will re-run the `fetch_data.R` function
+# to update the dataset in `data/` and then loads the dataset.
+#
+# There are three arguments to this function:
+#
+# - `local.dataset`: Set the file path to the local dataset (eg. `data/`).
+# - `fetch.script`: Set the file path to the `fetch_data.R` function.
+# - `force`: Force updating the dataset in `data/` by running the `fetch_data.R`
+# function.
+#
+# Examples for usage:
+#
+#   ds <- load_data()
+#   ds <- load_data(
+#       'data/dsname.RData',
+#       'R/fetchScript.R')
+#
+load_data <- function(local.dataset = file.path('data', 'data.Rds'),
                       fetch.script = file.path('R', 'fetch_data.R'),
                       force = FALSE) {
     source('.Rprofile')
     if (force) {
-        .fetch_data(master.dataset)
+        .fetch_data()
     }
     if (!file.exists(fetch.script))
         stop('Please create a file under the R folder that contains a function called fetch_data.')
     if (!file.exists(local.dataset)) {
         dir.create('data')
-        .fetch_data(master.dataset)
+        .fetch_data()
         message('A dataset has been added and loaded.')
     } else if (file.mtime(fetch.script) > file.mtime(local.dataset)) {
-        .fetch_data(master.dataset)
+        .fetch_data()
         message('Dataset has been updated and loaded.')
     } else {
         message('Dataset is up-to-date.')

### 2016-05-02 R/functions.R
Added a table for distr function, plus minor edits/fixes for others

@@ -208,7 +208,8 @@ combine_lcmm_data <- function(data, results.lcmm) {
         dplyr::full_join(results.lcmm$pprob %>%
                              as.data.frame() %>%
                              dplyr::select(SID, class)) %>%
-        dplyr::mutate(class = factor(class, labels = c('High', 'Mid', 'Low'))) %>%
+        dplyr::mutate(class = factor(class, levels = c(2, 1, 3),
+                                     labels = c('High', 'Mid', 'Low'))) %>%
         dplyr::rename(ClassLCMM = class)
 
     return(data)
@@ -219,10 +220,11 @@ combine_lcmm_data <- function(data, results.lcmm) {
 
 plot_lcmm_results <- function(results.lcmm) {
     results.lcmm %>%
-        ggplot2::ggplot(ggplot2::aes(VN, lISSI2, group = SID, colour = ClassLCMM)) +
+        ggplot2::ggplot(ggplot2::aes(f.VN, lISSI2, group = SID, colour = ClassLCMM)) +
         ggplot2::geom_line(ggplot2::aes(group = SID, colour = ClassLCMM), size = 0.1, alpha = 0.9) +
         ggplot2::geom_smooth(aes(group = ClassLCMM), method = "loess", size = 2)  +
         ggplot2::labs(x = "Time", y = "log(ISSI-2)", colour = "Latent Class") +
+        ggplot2::scale_x_discrete(labels = c('Baseline', 'Year 3', 'Year 6'), expand = c(0.05, 0.05)) +
         graph_theme()
 }
 
@@ -409,18 +411,25 @@ calculate_plsda_misclass <- function(results.plsda) {
 }
 
 calculate_outcomes_pct_change <- function(data) {
-    change_over_time <- data %>%
+    prep.data <- data %>%
         dplyr::select(f.VN, HOMA, ISI, IGIIR, ISSI2) %>%
         tidyr::gather(Measure, Value,-f.VN) %>%
         na.omit() %>%
         dplyr::group_by(Measure, f.VN) %>%
-        dplyr::summarise(med = median(Value)) %>%
-        dplyr::ungroup() %>%
+        dplyr::summarise(med = median(Value),
+                         n = n()) %>%
+        dplyr::ungroup()
+
+    sample_size <- prep.data$n %>%
+        {paste0(min(.), '-', max(.))}
+
+    change_over_time <- prep.data %>%
+        dplyr::select(-n) %>%
[…]

### 2016-05-02 doc/gee.Rmd
Minor edit to YAML gee.Rmd
Included sensitivity analysis for TAG/BaseTAG

@@
     rmarkdown::html_vignette:
-        keep_md: true
         toc: true
@@
                     'MET', 'AlcoholPerWk')
-covariates_is_tag <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'TAG')
+covariates_is_tag <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'BaseTAG')
 covariates_bcf_tag <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 
-                        'ALT', 'MET', 'AlcoholPerWk', 'TAG')
+                        'ALT', 'MET', 'AlcoholPerWk', 'BaseTAG')
 ```
@@
     plot()
+gee_results_is_int %>%
+    .$p.value %>% 
+    min()
 ```
@@
     plot()
+gee_results_bcf_int %>%
+    .$p.value %>% 
+    min()
 ```

### 2016-05-02 doc/manuscript.Rmd
Added Tony's edits to the manuscript

@@
         fig_height: 6 
-        fig_width: 5
+        fig_width: 7
         fig_caption: true
@@
 
-Title: Total non-esterified fatty acids, but not composition, predict
+Title: Total non-esterified fatty acids, but not individual species, predict
 progression in beta-cell dysfunction: The Prospective Metabolism and Islet Cell
@@
 <!-- First, middle initial, last, highest academic degree -->
-Author: Luke W. Johnston, Stewart Harris, Ravi Retnakaran, Bernard Zinman,
-Zhen Liu, Richard P. Bazinet, Anthony J. Hanley
+Author: Luke W. Johnston (1), MSc; Stewart B. Harris (2), MD; Ravi Retnakaran
+(3,4), MD; Bernard Zinman (3,4), MD; Adria Giacca (5), PhD; Zhen Liu (1), PhD;
+Richard P. Bazinet (1), PhD; and Anthony J. Hanley (1,6), PhD
 
@@
 Affiliation:
+
+(1) Department of Nutritional Sciences, University of Toronto, Toronto, Ontario, Canada.
+(2) Centre for Studies in Family Medicine, University of Western Ontario, London, Ontario, Canada.
+(3) Division of Endocrinology, University of Toronto, Toronto, ON, Canada.
+(4) Lunenfeld Tanenbaum Research Institute, Mount Sinai Hospital, Toronto, Ontario, Canada.
+(5) Department of Physiology, University of Toronto, Toronto, Ontario, Canada.
+(6) Dalla Lana School of Public Health, University of Toronto, Toronto, Ontario, Canada.
 
@@
 - Current address:  
-- Phone number:
+    Department of Nutritional Sciences  
+    Faculty of Medicine  
+    University of Toronto  
+    FitzGerald Building, 150 College Street, Room 341  
+    Toronto, ON, Canada, M5S 3E2
+- Phone number: 416.978.3616
 - Fax number:
@@
 
-Word count: 3527 / 4000
+Word count: 3543 / 4000
 
@@
 
-<!-- 250 words -->
 # Abstract
+250 / 250 words
 
-Objective:
+Objective: Our aim was to determine the longitudinal associations of individual
+non-esterified fatty acid (NEFA) species on changes in insulin sensitivity and
+beta-cell function over 6 years in a cohort of individuals who are at-risk for
+diabetes.
 
-Research Design and Methods:
+Research Design and Methods: In the PROMISE cohort, 477 participants had serum
+NEFA measured at the baseline visit and who had completed an OGTT at 3 time
+points over 6 years. Outcome variables were calculated using the OGTT values.
+Insulin sensitivity was assessed using 1/HOMA-IR and the Matsuda index (ISI),
+while beta-cell function was assessed using the Insulinogenic index over HOMA-IR
+(IGI/IR) and the Insulin Secretion-Sensitivity Index-2 (ISSI-2). Generalized
+estimating equations was used, adjusting for time, waist, sex, ethnicity, age,
+ALT, alcohol intake, physical activity, and family history of diabetes. NEFA
+were analyzed as both a concentration and as a proportion of the total fraction.
 
-Results:
+Results: Insulin sensitivity and beta-cell function declined by 14-27% while
+BMI did not change over the 6-years. Total NEFA, 16:0, 18:1n-9, and 18:2n-6 as
+concentrations had negative associations with IGI/IR and ISSI-2. These four NEFA
+variables had similar magnitudes of association, predicting between 4-8% lower
+IGI/IR and ISSI-2 over the 6 years. No associations were seen for 1/HOMA-IR
+or ISI, nor were any associations seen for the mol% data.
 
-Conclusions:
+Conclusions: Total NEFA was a strong predictor for lower beta-cell function over
+6 years. Our results suggest that association with beta-cell function is due to
+the absolute size of the serum NEFA fraction, rather than the specific lipid
+composition.
 
@@
 - [X] PLS-DA with dysglycemia (for anyone who got it over the years).
+- [X] Replace TAG with BaseTAG in the QIC (or use both)
+- [X] Fix time axis labels on LCMM plot
+- Describe LCMM and PLS-DA better in plots
 
@@
 stab_qic <- stab('qic', 'Comparing generalized estimating equation models adjusting for different covariates using Quasi-Likelihood Information Criterion.')
+stab_dist <- stab('nefaDist', 'Raw concentration values (nmol/mL) of each non-esterified fatty acid in the PROMISE cohort at the baseline visit (2004-2006). Values are presented as mean (SD).')
 sfig_corr <- sfig('corr', 'Pearson correlation heatmap of non-esterified fatty acids (nmol/mL) and basic PROMISE participant characteristics for the baseline visit (2004-2006).
-Darkness of the colour indicates the size of the correlation, with blue indicating positive and orange indicating negative correlations.')
+Darkness of the colour indicates the magnitude of the correlation, with blue indicating positive and orange indicating negative correlations.')
 sfig_gee <- sfig('gee', 'Unadjusted generalized estimating equation modeling of the longitudinal association of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort.
-Only adjusted for time.
+Models are only adjusted for time.
 Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
@@
 stab_gee_unadj <- stab('gee_unadj', 'Unadjusted generalized estimating equations models of the longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
-Adjusted for time only.
+Models are only adjusted for time.
 Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
 P-values were adjusted for the BH false discovery rate, indicated by asterisk.')
-sfig_lcmm <- sfig('lcmm', 'Latent class mixed model analysis to identify individual classes of trajectories for log(ISSI-2) over the 6 years in the PROMISE cohort.')
-sfig_plsda_groups <- sfig('plsda_groups', 'Clustering of extracted components from the partial least squares discriminant analysis on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006).
-The percent explained variance of each component is shown in brackets on each axis.
-{{colour}}, {{colour}}, {{colour}} lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis.')
+sfig_lcmm <- sfig('lcmm', 'Latent class mixed model (LCMM) analysis to identify individual classes of trajectories for log(ISSI-2) over the 6 years in the PROMISE cohort.
+LCMM is a technique that identifies groups of participants that share a similar underlying trajectory in beta-cell function over the 6 years (e.g.
+no change compared to declines in ISSI-2).
+Red lines indicate individuals with a high beta-cell function who stayed high, green represents those in the middle, and blue represents those who had the lowest beta-cell function.')
+sfig_plsda_groups <- sfig('plsda_groups', 'Clustering of extracted components from the partial least squares discriminant analysis (PLS-DA) on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006).
+PLS-DA is a multivariate technique that The percent explained variance of each component is shown in brackets on each axis.
+Red, green, and blue lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis.')
 sfig_plsda_loadings <- sfig('plsda_loadings', 'Pattern loadings from partial least squares discriminant analysis to identify potential clusters of NEFA composition within the classes extracted from the latent class mixed model in 463 participants from the baseline PROMISE visit (2004-2006).
@@
 The percent explained variance of each component is shown in brackets on each axis.
-{{colour}} lines indicate dysglycemia conversion or maintanence and {{colour}} lines indicate no dysglycemia status.')
+Blue lines indicate dysglycemia conversion or maintanence and red lines indicate no dysglycemia status.')
 sfig_plsda_loadings_dysgly <- sfig('plsda_loadings_dysgly', 'Pattern loadings from partial least squares discriminant analysis to identify potential clusters of NEFA composition for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).
@@
 exposure to high concentrations of NEFA can induce insulin resistance in 
-insulin-sensitive tissue such as the muscle or liver and impair beta-cell
-production of insulin [@Rachek2014a;@Cnop2008a]. Epidemiological and clinical
-studies have observed similar findings, showing that elevated total plasma NEFA
-associates with risk for diabetes [@Daniele2014a;@Salgin2012a].
+insulin-sensitive tissues such as muscle and liver, and in addition can impair 
+pancreatic beta-cell production of insulin [@Rachek2014a;@Cnop2008a].
+Epidemiological and clinical studies have reported concordant findings, showing
+in particular that elevated total plasma NEFA associates with an increased risk
+for incident type 2 diabetes [@Daniele2014a;@Salgin2012a].
 
-Much of the previous experimental work frequently use fatty acids such as
-palmitic acid (16:0) or oleic acid (18:1n-9), or use a standard oil such as soybean
-oil as the exposure and generalize as total NEFA. However, fatty acids comprise
-multiple classes of molecules with diverse physiological functions, with few
-studies comparing different fatty acids. For instance, one study suggests that
-eicosapentaenoic acid (20:5n-3) can protect against the lipotoxic effect of
-palmitic acid in the beta-cells [@Xiao2006].
+Much of the previous experimental work on the role of NEFA in type 2 diabetes
+utilized individual lipid species such as palmitic acid (16:0) or oleic acid
+(18:1n-9), or alternatively used specific oils such as soybean oil, which is
+high in the polyunsaturated fatty acid (PUFA) linoleic acid (18:2n-6), as the
+exposure to characterize the impact of total NEFA. However, fatty acids comprise
+multiple classes of molecules with diverse physiological functions, and few 
+studies have compared the effects of a broader spectrum of fatty acids. Notably,
+one study suggests that eicosapentaenoic acid (20:5n-3) can protect against the
+lipotoxic effect of palmitic acid in the beta-cells [@Xiao2006].
 
-Despite a sizable literature studying the role of NEFA in diabetes, there are 
-important gaps in the research. In particular, there have been few long term 
-studies in humans that have examined a range of NEFA species. Most studies have 
-been on animals and cells, have been short term human infusion trials, or have 
-been epidemiological studies that only looked at total NEFA. To date, there have
+Despite a sizable literature studying the role of total NEFA concentration in 
+diabetes, there are important gaps in the research. The majority of previous 
+studies have used animal models or cell lines [@Szendroedi2012a;@Kehlenbrink2012a], 
+have been short term human infusion trials [@Daniele2014a;@Liang2013a], or have
+been epidemiological studies that only looked at total NEFA and not individual
+species [@Miller2012a;@Johns2014a]. To date, there have
 been no longitudinal studies examining the role of the composition of the NEFA
-blood lipid fraction on the pathogenesis of diabetes, which is key to 
-understanding causal associations given the chronic nature of diabetes 
-pathophysiology. Therefore, our research objective was to examine the role of
-the serum NEFA composition on insulin sensitivity and beta-cell function in a
-longitudinal cohort. Some specific hypotheses included that higher palmitic acid
-while lower polyunsaturated fatty acids such as eicosapentaenoic acid would 
-associate with lower insulin sensitivity and beta-cell function.
+blood lipid fraction on the pathogenesis of diabetes, which is critical to
+understanding associations at a more granular level given protracted natural
+history of diabetes and the growing appreciation of the divergent effects of
+individual lipid species. Therefore, our research objective was to examine the 
+association of serum NEFA composition on changes over time in insulin
+sensitivity and beta-cell function in a longitudinal cohort. We hypothesized
+that higher palmitic acid and lower polyunsaturated fatty acids such as
+eicosapentaenoic acid would associate with declining insulin sensitivity and
+beta-cell function over 6 years.
 
@@
 Prospective Metabolism and Islet cell Evaluation (PROMISE) cohort. Eligibility 
-for recruitment into PROMISE included having one or more risk factors for type 2
-diabetes mellitus, such as obesity, hypertension, family history of diabetes, 
+for recruitment into PROMISE required having one or more risk factors for type 2
+diabetes mellitus, including obesity, hypertension, family history of diabetes, 
 and/or a history of gestational diabetes or birth of a macrosomic infant. 
 Participants aged 30 years and older (n=736) attended the baseline visit
-between 2004-2006. Follow-up examinations occur every three years, with three
+between 2004-2006. Follow-up examinations in this cohort occur every three years, with three
 examination visits completed to date (2004-2006, 2007-2009, and 2010-2013).  The
@@
 and questionnaires on lifestyle and sociodemographics. Research ethics approval
-was obtained from Mount Sinai Hospital and the University of Western Ontario. 
+was obtained from Mount Sinai Hospital and the University of Western Ontario,
+and all participants provided written informed consent.
 Data collection methods were standardized across the 2 centres and research
@@
 procedures. Cholesterol, HDL, and triglycerides (TG) were measured using Roche
-Modular's enzymatic colometric tests (Mississauga, ON).  Both specific insulin
+Modular's enzymatic colorimetric tests (Mississauga, ON).  Both specific insulin
 and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
@@
 inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
-performed at the Banting and Best Diabetes Centre Core Lab at the University of
-Toronto. Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and
-diabetes were categorized using the 2006 WHO criteria [@WHO2006].
+performed at the Banting and Best Diabetes Centre Core Lab at Mt Sinai Hospital.
+Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and diabetes
+were categorized using the 2006 WHO criteria [@WHO2006].
 
@@
 stable at these temperatures for up to 10 years [@Matthan2010a]. A known amount 
-of heptadecanoic acid (17:0) {{confirm}} was added to the serum as an internal 
+of heptadecanoic acid (17:0) was added to the serum as an internal 
 standard prior to extracting total lipids according to the method of Folch 
 [@Folch1957a]. Each serum lipid fraction (NEFA, cholesteryl ester, phospholipid,
-and triacylglycerol) was isolated using thin layer chromatography, which were
+and triacylglycerol) was isolated using thin layer chromatography; each fraction was
 visualized under UV light after lightly spraying with 8-anilino-1-naphthalene
@@
 concentrations (nmol/ml) were calculated by proportional comparison of gas
-chromatography peak areas to that of the internal standards [@Nishi2014a;
-@Chen2011a].  There were 22 fatty acids measured in the NEFA fraction. Given the
-diverse biology, as well as the complexity of the analyses, the other lipid
-fractions were split into separate reports (see {{pl and ce paper}} for the
+chromatography peak areas to that of the internal standards [@Nishi2014a]. 
+There were 22 fatty acids measured in the NEFA fraction. Given the
+diverse biology, as well as the complexity of the analyses, findings for other lipid
+fractions in this cohort are reported separately (see {{pl and ce paper}} for the
 analysis of the phospholipid and cholesteryl ester fractions).
@@
 Insulin sensitivity and beta-cell function indices were computed using the OGTT 
-glucose and insulin. Insulin sensitivity was assessed using 1/HOMA-IR (1
+glucose and insulin data. Insulin sensitivity was assessed using 1/HOMA-IR (1
 divided by HOMA-IR) [@Matthews1985] and the Insulin Sensitivity Index (ISI)
@@
 is a measure of the first phase insulin secretion while ISSI-2 is analogous to
-the disposition index. Each index has been validated against gold standard
-measures [@Matthews1985; @Matsuda1999; @Retnakaran2009].
+the disposition index (but using OGTT values). Each index has been validated
+against gold standard measures [@Matthews1985; @Matsuda1999; @Retnakaran2009].
 
@@
 process. BMI was calculated by dividing weight (kg) by height (m) squared.
-Sociodemographic information, such as age, sex, and ethnicity, were determined
+Sociodemographic information, including age, sex, and ethnicity, were determined
 using questionnaires administered at each examination. In the lifestyle
@@
 
-The primary outcome variables for the PROMISE cohort and this analyses were 
+The primary outcome variables for this analysis were 
 1/HOMA-IR, ISI, IGI/IR, and ISSI-2; outcome variables were log-transformed for
-the statistical modeling. The primary predictor variables for this study were
-the 22 NEFA as a mole percent (mol%) of the total fraction and as a 
-concentration (nmol/mL). Pearson correlation coefficients were computed for the
-individual NEFA and with other continuous variables.
+the statistical modeling. The primary predictor variables for this analysis were
+22 individual NEFA species as mole percent (mol%) of the total fraction and as
+concentration (nmol/mL). Pearson correlation coefficients were computed to
+assess the relationships of individual NEFA with other continuous variables.
 
@@
 though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
-is well suited to cohort studies given it's capacity to handle missed visits.
-The predictor variables and continuous covariates were scaled (mean centered and
-standardized). The NEFA variables were classified as *time-independent* (held
-constant) as they were measured only at the baseline visit, while the outcome
-variables and covariates were set as *time-dependent*. No imputation was
-conducted on missing values.
+is well suited to longitudinal cohort studies given it's capacity to handle
+missed visits. The predictor variables and continuous covariates were scaled
+(mean centered and standardized). The NEFA variables were classified as
+*time-independent* (held constant) as they were measured only at the baseline
+visit, while the outcome variables and covariates were set as *time-dependent*.
+No imputation was conducted on missing values.
 
@@
 ALT, MET, alcohol intake, and family history of diabetes (M6). For beta-cell
-function, the last few models were comparable in performance, so we chose the
-simplest model (M8), adjusting for time, sex, ethnicity, baseline age, WC, ALT,
+function, the last few models (M5-M8) were comparable in performance, so we
+report results from the simplest model (M8), adjusting for time, sex, ethnicity,
+baseline age, WC, ALT,
 and family history of diabetes. After scaling, log-transforming, and
@@
 percent difference in the outcome variable for every SD increase in the
-predictor variable given the covariates are held constant. While models with a
-time interaction did not perform better than others for the QIC (M6.1 and M8.1)
-in `r cite_st('qic')`, we ran models with a time by predictor term for each of
-the NEFA to test if there was an interaction by time. Given that TG is a risk factor
+predictor variable given the covariates are held constant. While we a priori
+decided to test for and report on the time by predictor interaction results, we
+also assessed the overall fit of models that included a time interaction, which
+ultimately did not improve the overall fit of our models in terms of QIC
+(M6.1 and M8.1 as seen
+in `r cite_st('qic')`) and we therefore did not include them in the main effect
+analysis. Given that TG is a risk factor
 for diabetes and that NEFA contribute to TG production, TG may act as a mediator
-between NEFA and the outcomes. However, only a subset of the participants at the
-6-yr visit had clinically measured TG. To determine the role of TG in the
+between NEFA and the outcomes. To determine the role of TG in the
 association between NEFA and the outcomes, TG was included in the GEE model as a
 sensitivity analysis. Lastly for the GEE models, we tested for an interaction
-with sex or ethnicity by the predictor term for the outcome variables.
+with sex or ethnicity by the predictor term for each outcome variable.
 
-While GEE controls for the longitudinal design of the data, a limitation is that
-it can not analyze the inherent multivariate nature of the composition of the 
-NEFA fraction. As such, after running the GEE models for the longitudinal 
-analysis, partial least squares discriminant analysis (PLS-DA) was used to
-identify the patterns of NEFA composition against beta-cell function (using ISSI-2,
-which has most of the associations in the GEE models) and for those who converted or
-maintained dysglycemia status (either IFG, IGT, or DM) over the 6 years. Since
-PLS-DA can't handle longitudinal data, the longitudinal data was included by
-extracting latent trajectories of beta-cell function (ISSI-2) over the 6 years using
-latent class mixed models (LCMM). A limitation of these methods, excluding LCMM,
-is that they can not be adjusted for covariates.
+While GEE accounts for the longitudinal design of the data, this approach is
+limited in that it cannot analyze the inherent multivariate nature of the
+composition of the NEFA fraction. As such, after running the GEE models for the
+longitudinal analysis, partial least squares discriminant analysis (PLS-DA) was
+used to identify the patterns of NEFA composition against beta-cell function
+(using ISSI-2, which had most of the associations in the GEE models) and for
+those who converted to or maintained dysglycemia status (either IFG, IGT, or DM)
+over the 6 years. Since PLS-DA cannot handle longitudinal data, latent class
+mixed models (LCMM) were used to extract latent trajectories of beta-cell
+function (ISSI-2) over the 6 years.
 
 All analyses were performed using R `r getRversion()` [@Rbase], along with the R
-packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a], caret
+packages geepack `r packageVersion('geepack')` for GEE <!--[@Hoejsgaard2006a]-->, caret
 `r packageVersion('caret')` for PLS-DA, and lcmm `r packageVersion('lcmm')` for
@@
 
-gee_results_unadj <- bind_rows(
+gee_results_unadj <- dplyr::bind_rows(
     analyze_gee(gee_df, outcomes, ne_pct, 'VN', 'mol%', adj.p = TRUE),
@@
 
-```{r lcmmPls, cache=TRUE}
+```{r lcmmPls, cache=TRUE, results='hide'}
 # include also the 3-yr and 6-yr data for training/testing.
@@
 
-Within this subset of the PROMISE cohort at the baseline visit, 
-`r sex.npct$npct` were females and `r ethn.npct$npct` had European-ancestry.
+For this study, there were 
+`r sex.npct$npct` females and `r ethn.npct$npct` had European-ancestry.
 The average age of the participants was `r aide::average(ds0yr$Age)` years 
@@
 `r percent.change.outcomes$chg` (p`r percent.change.outcomes$p`) over the 6
-years in this subset of the PROMISE cohort.
-Consistent with this decline, there were 42 (9%) <!--`r dm.convert`-->
+years in this analysis of the PROMISE cohort (n=`r percent.change.outcomes$n`).
+Consistent with this decline, there were `r dm.convert`
 participants who developed diabetes while `r predm.convert` participants either
@@
 
-`r cite_f('nefaDist')` shows the distribution of the NEFA composition. Four 
-individual NEFA made up the vast majority (`r sum(percent.nefa$pct)`%) of the total
-NEFA fraction. The largest contributors were `r percent.nefa$c[1]`, 
-`r percent.nefa$c[2]`, `r percent.nefa$c[3]`, and `r percent.nefa$c[4]`. All
-individual NEFA as well as the total fraction had weak to no correlations
-(r<0.3) with the basic characteristics (see `r cite_sf('corr')`).
+`r cite_f('nefaDist')` shows the compositional distribution of NEFA in the study
+participants. Four individual NEFA made up the vast majority 
+(`r sum(percent.nefa$pct)`%) of the total NEFA fraction. The largest contributors
+were `r percent.nefa$c[1]`, `r percent.nefa$c[2]`, `r percent.nefa$c[3]`, and 
+`r percent.nefa$c[4]`. Raw concentration values are shown in `r cite_st('nefaDist')`. 
+All individual NEFA as well as the total fraction had
+correlations that ranged from weak to null (r<0.3) with the basic
+characteristics (see `r cite_sf('corr')`).
 
@@
 Several associations were seen in the unadjusted GEE models 
-(see `r cite_sf('gee')`), particularly for the concentration data. Full model 
-adjustment (`r cite_f('gee')`) attenuated most of these associations, where
-total NEFA, 16:0, and 18:2n-6 (all as nmol/mL) had negative associations between
+(see `r cite_sf('gee')`), particularly for the concentration compared to the
+mol% data. Full model adjustment (`r cite_f('gee')`) attenuated most of these
+associations, although total NEFA, 16:0, 18:1n-9, and 18:2n-6 (all as nmol/mL)
+had negative associations between
 IGI/IR and ISSI-2. The magnitude of association for each of these variables was
@@
 clinic visit. None of the NEFA variables as a mol% had any association with the
-outcomes. Adjusting for TG attenuates all associations with IGI/IR and ISSI-2
-(data not shown), though sample size at the 6 year visit is restricted to n=128
-in those who had TG measured. There were no significant interaction effects
+outcomes. Adjusting for TG attenuated all associations with IGI/IR and ISSI-2
+(data not shown).
+There were no significant interaction effects
 between time, sex, or ethnicity and the individual NEFA on any of the outcome
-measures (data not shown). Raw values from the GEE models are shown in 
+measures (all p>0.15). Raw values from the GEE models are shown in 
 `r cite_st('gee_unadj')` for unadjusted models and `r cite_st('gee')` for fully
@@
 extracted component and `r cite_sf('plsda_loadings')` shows the clustering of
-the components by ISSI-2 latent class. The PLS-DA analysis using dysglycemia
-status conversion or maintanence as the response variable showed similar poor 
-discriminatory ability, though better than the beta-cell function trajectory
-results. While `r misclass.n.percent.dysgly$NPct[2]` participants were
-classified correctly based on dysglycemia status, examining the clustering of
-the NEFA composition showed poor discrimination between groups (see
+the components by ISSI-2 latent class. The PLS-DA analysis using conversion to
+or maintenance of dysglycemia status as the response variable showed similar
+poor discriminatory ability, though better than the beta-cell function
+trajectory results. While `r misclass.n.percent.dysgly$NPct[2]` participants
+were classified correctly based on dysglycemia status, examining the clustering
+of the NEFA composition showed poor discrimination between groups (see 
 `r cite_sf('plsda_groups_dysgly')` and `r cite_sf('plsda_loadings_dysgly')`).
@@
 In a Canadian population of adults who are at-risk for diabetes, we found that
-higher total NEFA independently predicted a lower beta-cell function over 6 years.
-While we found negative association with palmitic acid (16:0), oleic acid (18:1n-9), and
-linoleic acid (18:2n-6) as a concentration (nmol/mL), no associations were seen
+higher total NEFA concentrations independently predicted
+lower beta-cell function after 6 years. While we found negative associations with
+palmitic acid (16:0), oleic acid (18:1n-9), and
+linoleic acid (18:2n-6) for the concentration (nmol/mL) modeling, no
+associations were seen
 for the proportion of these fatty acids and further clustering analysis found no
 predictive ability of these NEFA on beta-cell function or dysglycemia. These
-observations suggest that the absolute size of the total fraction, rather than
+observations suggest that the absolute size of the total NEFA fraction, rather than
 the specific composition of any individual NEFA, likely influences the
@@
 
-The role of total NEFA on diabetes is a well-studied area of research. 
+The role of total NEFA in the etiology of diabetes is well-documented.
 Epidemiological studies have shown that higher NEFA associate with lower insulin
@@
 through endoplasmic reticulum stress, formation of ceramides, and generation of
-nitric oxide, as well as impairment of proinsulin production and mitochrondial
-function [@Cnop2008a; @Maris2013a; @Carpentier2003a; @Giacca2011a; @Xiao2009a].
-The present analysis is the first study to our knowledge to examine a broad
-spectrum of individual NEFA in a cohort setting on beta-cell function. We found
-that there was a strong signal of higher total NEFA, palmitic acid, oleic acid,
-and linoleic acid as concentrations with lower beta-cell function. However, as
-both a mol% and using clustering analyses, no individual or specific
-composition of the NEFA fraction predicted lower beta-cell function, insulin
-sensitivity, or dysglycemia status. Together, these results suggest that the
-absolute size of the NEFA fraction, irrespective of any specific composition of
-fatty acid species, exerts it's hypothesized lipotoxic effects on the
-beta-cells.
+nitric oxide, as well as impairment of proinsulin production and mitochrondrial
+function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a].
+The present analysis is the first study to our knowledge to examine the 
+longitudinal association of a broad spectrum of individual NEFA species on
+beta-cell function in a large cohort. We found that there was a strong signal of
+higher total NEFA, palmitic acid, oleic acid, and linoleic acid as
+concentrations with lower beta-cell function. However, in modeling these fatty
+acids as a mol% and using novel clustering analysis approaches, no individual or
+specific fatty acid species in the NEFA fraction predicted lower beta-cell
+function, insulin sensitivity, or dysglycemia status. Taken together, these
+results suggest that it is the absolute size of the circulating NEFA fraction,
+irrespective of any specific composition of fatty acid species, that is
+responsible for the hypothesized lipotoxic effects of NEFA on the beta-cells.
 
@@
 were attenuated, suggesting that NEFA may in fact be mediating it's association
-through TG. However, a major limitation to the TG-adjusted GEE models is the
-reduced sample size in the 6 year visit, as only a subset (n=128) of
-participants had TG measured.
+through TG.
 
 There is substantial experimental evidence highlighting the role of increased 
-NEFA and a subsequent increase in insulin resistance via impairment of insulin 
-signaling cascades (reviewed in multiple articles 
-[@Martins2012a;@Ebbert2013a;@Capurso2012a;@Surampudi2009]). However, in this
-longitudinal analysis, we saw no association between any individual or total
-NEFA and neither hepatic (1/HOMA-IR) nor whole-body insulin sensitivity (ISI). 
-There are some possible explanations for our findings. Previous studies done *in
-vivo* were mostly short term infusion 
+NEFA and the subsequent increase in insulin resistance via impairment of insulin 
+signaling cascades, as reviewed in previously published articles 
+[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this
+longitudinal analysis, we saw no association of any individual or total
+NEFA with hepatic (1/HOMA-IR) or whole-body insulin sensitivity (ISI). 
+There are some possible explanations for these findings. Previous studies done *in
+vivo* were mostly short term infusion protocols
 [@Xiao2009a;@Xiao2006;@Szendroedi2012a;@Kehlenbrink2012a] or pharmocologic
@@
 cross-sectional or correlational
-[@Miller2012a;@Johns2014a;@Bush2012a;@Ilyasova2010a]. The limitation of these
-studies were the short time periods examined or the experimentally-induced
-elevations in NEFA concentrations, making it difficult to assess the causal role
+[@Miller2012a;@Johns2014a;@Bush2012a;@Ilyasova2010a]. The short time periods of
+these studies and the experimentally-induced
+elevations in NEFA concentrations make it difficult to assess the causal role
 in non-experimental environments over long time frames. Moreover, some studies
@@
 fasting and postprandial NEFA kinetics. For instance, some experimental studies
-using the clamp found that fasting NEFA was a weak predictor of insulin
+using clamp protocols found that fasting NEFA was a weak predictor of insulin
 sensitivity compared to postprandial concentrations of NEFA
@@
 examined in PROMISE. It may be that in this population, IR has become well
-established and NEFA may not contribute to IR at this stage in the pathogenesis
-of diabetes.
+established and NEFA may not contribute to IR at this somewhat mode advanced
+stage in the pathogenesis of diabetes.
 
 Few studies have examined the composition of NEFA on metabolic functioning. One
-recent, well analyzed study used a variety of advanced fatty acid measurement
+recent, well-analyzed study used a variety of advanced fatty acid measurement
 and statistical techniques to explore the multivariate relationship between the
 NEFA composition and components of the metabolic syndrome (MetS) [@Dai2015a].
-Specifically, using the statistical techniques the authors identified NEFA
-16:1n-9, 20:1n-9, and 22:4n-6 to most correlate with components of the MetS. 
+Specifically, the authors identified NEFA
+16:1n-9, 20:1n-9, and 22:4n-6 to correlate with components of the MetS. 
 Another similar study examining diabetes found that 16:0, 18:0, 18:1, 18:2, 18:3
@@
 There are a few important limitations to our study. NEFA were only quantified at
-the baseline visit and as such we can not investigate whether there were 
+the baseline visit and as such we cannot investigate whether there were 
 concomitant changes in NEFA and the metabolic measures. However, we believe this
@@
 integrated. This is also an observational cohort, and there may be some residual
-confounding we haven't considered or that couldn't be measured. Nonetheless, we 
-were as thorough as possible in understanding and minimizing potential 
+confounding we haven't considered or that couldn't be measured. Nonetheless,
+potential covariates were empirically analyzed prior to inclusion into the
+GEE models to best understand and minimizing potential
 confounding. Finally, our cohort consists of individuals at-risk for diabetes, 
@@
 
+<!--
+A limitation of these methods, excluding LCMM,
+is that they can not be adjusted for covariates.
+-->
+
 In conclusion, we found that total NEFA was a strong predictor for lower
 beta-cell function over 6 years, irrespective of the specific composition of the
-NEFA fraction. While future studies are needed to confirm these findings, given
-the cost and labour involved in quantifying the NEFA composition, our results
-suggest that further research on the NEFA composition may not be as fruitful and
-cost-effective then if the simpler, easy to measure total NEFA concentration was
-used.
+NEFA fraction. While future studies are needed to confirm these findings, our
+results reinforce the importance of continuing to investigate the role of
+circulating NEFA concentration on the natural history of diabetes.
 
@@
 authors report no potential conflicts of interest relevant to this study. 
+This study was supported by grants from the Canadian Diabetes Association (CDA),
+the Canadian Institutes for Health Research (CIHR), and the University of
+Toronto Banting and Best Diabetes Centre (BBDC); LWJ is supported by a CDA
+Doctoral Student Research Award; RR is supported by a Heart and Stroke
+Foundation of Ontario Mid-Career Investigator Award; SBH holds the CDA Chair in
+National Diabetes Management and the Ian McWhinney Chair of Family Medicine
+Studies at the University of Western Ontario; BZ holds the Sam and Judy Pencer
+Family Chair in Diabetes Research at Mount Sinai Hospital and University of
+Toronto; RBP holds a Tier II Canada Research Chair in Brain Lipid Metabolism;
+AJH holds a Tier II Canada Research Chair in Diabetes Epidemiology.
 
@@
 
-Note: Values are in median (IQR), mean (SD), and n (%).
+Note: Values are in median (IQR), mean (SD), and n (%). The proportion of ethnic
+and sex groups did not change over the 6 years.
 
@@
 
+```{r supTabDist}
+table_distribution(ds, stab_dist)
+```
+
 ```{r supFigHeatmap, fig.cap=sfig_corr}

### 2016-05-04 R/functions.R
Minor fixes to table func, plus added example plsda plot func

@@ -115,7 +115,7 @@ analyze_lcmm <- function(data, lc.var = 'lISSI2') {
 analyze_plsda <- function(data, variable = 'ClassLCMM') {
     data.prep <- data %>%
         dplyr::filter(VN == 0) %>%
-        dplyr::select(contains(variable), matches('^ne\\d\\d')) %>%
+        dplyr::select(contains(variable), matches('^pct_ne\\d\\d')) %>%
         na.omit()
 
     if (length(caret::nearZeroVar(data.prep[2:dim(data.prep)[2]])) != 0)
@@ -306,7 +306,7 @@ plot_plsda_grouping <- function(results.plsda, legend = 'LCMM Class') {
 
     plot_data %>%
         ggplot2::ggplot(ggplot2::aes(Comp1, Comp2, colour = class.lcmm)) +
-        ggplot2::stat_density2d(ggplot2::aes(alpha = ..level.., colour = class.lcmm), size = 2) +
+        ggplot2::stat_density2d(ggplot2::aes(alpha = ..level.., colour = class.lcmm), size = 1) +
         ggplot2::scale_alpha(guide = 'none') +
         ggplot2::scale_color_discrete(legend) +
         ggplot2::labs(
@@ -350,7 +350,7 @@ plot_heatmap <- function(data, x = c(outcomes, 'BMI', 'Waist', 'Age', 'lALT',
             ylab = 'Non-esterified fatty acids (nmol/mL)',
             number.colours = 5,
             values.size = 4) +
-        graph_theme(ticks = FALSE)
+        graph_theme(ticks = FALSE, legend.pos = 'right')
 }
 
 # Calculate or extract for inline -----------------------------------------
@@ -507,9 +507,9 @@ table_gee <- function(results, caption, digits = 1) {
 
     gee_table <-
         dplyr::bind_rows(
-            data.frame(Xterms = paste0('**', unique(gee_table_prep$unit)[1], '**')),
-            dplyr::filter(gee_table_prep, unit == 'mol%'),
             data.frame(Xterms = paste0('**', unique(gee_table_prep$unit)[2], '**')),
+            dplyr::filter(gee_table_prep, unit == 'mol%'),
+            data.frame(Xterms = paste0('**', unique(gee_table_prep$unit)[1], '**')),
             dplyr::filter(gee_table_prep, unit == 'nmol/mL')
         ) %>%
         dplyr::select(-unit) %>%
[…]

### 2016-05-04 doc/manuscript.Rmd
Added Richard's edits/suggestions, sent to Tony

@@
 
+<!--
+
+TODO:
+- Use pct for PLSDA
+
+-->
+
 # Title page
 
-Title: Total non-esterified fatty acids, but not individual species, predict
+Title: Total non-esterified fatty acids, but not individual fatty acids, predict
 progression in beta-cell dysfunction: The Prospective Metabolism and Islet Cell
@@
 
-Word count: 3543 / 4000
+Word count: 3569 / 4000
 
@@
 # Abstract
-250 / 250 words
+249 / 250 words
 
 Objective: Our aim was to determine the longitudinal associations of individual
-non-esterified fatty acid (NEFA) species on changes in insulin sensitivity and
+non-esterified fatty acids (NEFA) on changes in insulin sensitivity and
 beta-cell function over 6 years in a cohort of individuals who are at-risk for
@@
 6 years. Our results suggest that association with beta-cell function is due to
-the absolute size of the serum NEFA fraction, rather than the specific lipid
+the absolute size of the serum NEFA fraction, rather than the specific fatty acid
 composition.
@@
 - [X] Interaction by time
-- [X] Sensitivity analysis with TG
+- [X] Sensitivity analysis with TAG
 - [X] PLS-DA with dysglycemia (for anyone who got it over the years).
@@
 Red lines indicate individuals with a high beta-cell function who stayed high, green represents those in the middle, and blue represents those who had the lowest beta-cell function.')
-sfig_plsda_groups <- sfig('plsda_groups', 'Clustering of extracted components from the partial least squares discriminant analysis (PLS-DA) on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006).
-PLS-DA is a multivariate technique that The percent explained variance of each component is shown in brackets on each axis.
-Red, green, and blue lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis.')
+sfig_plsda_groups <- sfig('plsda_groups', '')
 sfig_plsda_loadings <- sfig('plsda_loadings', 'Pattern loadings from partial least squares discriminant analysis to identify potential clusters of NEFA composition within the classes extracted from the latent class mixed model in 463 participants from the baseline PROMISE visit (2004-2006).
 The percent explained variance of each component is shown in brackets on each axis.')
-sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', 'Clustering of extracted components from the partial least squares discriminant analysis for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).
-The percent explained variance of each component is shown in brackets on each axis.
-Blue lines indicate dysglycemia conversion or maintanence and red lines indicate no dysglycemia status.')
+sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', '')
 sfig_plsda_loadings_dysgly <- sfig('plsda_loadings_dysgly', 'Pattern loadings from partial least squares discriminant analysis to identify potential clusters of NEFA composition for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).
 The percent explained variance of each component is shown in brackets on each axis.')
+sfig_plsda_example <- sfig('plsda_example', 'Example results of a high discriminatory ability to classify groups accurately when using partial least squares discriminatory analysis (PLS-DA).
+Discriminatory ability is evident from the amount of separation between the two groups.
+In this case, PLS-DA was 82% accurate at correctly classifying groups.')
+sfig_plsda_groups <- sfig('plsda_groups', paste0('Clustering of extracted components from the partial least squares discriminant analysis (PLS-DA) on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006).
+PLS-DA is a multivariate technique that The percent explained variance of each component is shown in brackets on each axis.
+Red, green, and blue lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis.
+See ', cite_sf('plsda_example'), ' for an example plot showing good discriminatory ability between groups.'))
+sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', paste0('Clustering of extracted components from the partial least squares discriminant analysis for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).
+The percent explained variance of each component is shown in brackets on each axis.
+Blue lines indicate dysglycemia conversion or maintanence and red lines indicate no dysglycemia status.
+See ', cite_sf('plsda_example'), ' for an example plot showing good discriminatory ability between groups.'))
 ```
@@
 pancreatic beta-cell production of insulin [@Rachek2014a;@Cnop2008a].
-Epidemiological and clinical studies have reported concordant findings, showing
+Observational and clinical studies have reported concordant findings, showing
 in particular that elevated total plasma NEFA associates with an increased risk
@@
 Much of the previous experimental work on the role of NEFA in type 2 diabetes
-utilized individual lipid species such as palmitic acid (16:0) or oleic acid
+utilized individual fatty acids such as palmitic acid (16:0) or oleic acid
 (18:1n-9), or alternatively used specific oils such as soybean oil, which is
@@
 exposure to characterize the impact of total NEFA. However, fatty acids comprise
-multiple classes of molecules with diverse physiological functions, and few 
+multiple molecules with diverse physiological functions, and few 
 studies have compared the effects of a broader spectrum of fatty acids. Notably,
@@
 been epidemiological studies that only looked at total NEFA and not individual
-species [@Miller2012a;@Johns2014a]. To date, there have
-been no longitudinal studies examining the role of the composition of the NEFA
-blood lipid fraction on the pathogenesis of diabetes, which is critical to
-understanding associations at a more granular level given protracted natural
+fatty acids [@Miller2012a;@Johns2014a]. To date, there have
+been no longitudinal studies examining the role of the composition of the serum NEFA
+fraction on the pathogenesis of diabetes, which is critical to
+understanding associations at a more granular level given the protracted natural
 history of diabetes and the growing appreciation of the divergent effects of
-individual lipid species. Therefore, our research objective was to examine the 
+individual fatty acids. Therefore, our research objective was to examine the 
 association of serum NEFA composition on changes over time in insulin
@@
 -70°C. Alanine aminotransferase (ALT) was measured using standard laboratory
-procedures. Cholesterol, HDL, and triglycerides (TG) were measured using Roche
+procedures. Cholesterol, HDL, and triacylglycerides (TAG) were measured using Roche
 Modular's enzymatic colorimetric tests (Mississauga, ON).  Both specific insulin
@@
 [@Folch1957a]. Each serum lipid fraction (NEFA, cholesteryl ester, phospholipid,
-and triacylglycerol) was isolated using thin layer chromatography; each fraction was
+and TAG) was isolated using thin layer chromatography; each fraction was
 visualized under UV light after lightly spraying with 8-anilino-1-naphthalene
@@
 the statistical modeling. The primary predictor variables for this analysis were
-22 individual NEFA species as mole percent (mol%) of the total fraction and as
+22 individual NEFA as mole percent (mol%) of the total fraction and as
 concentration (nmol/mL). Pearson correlation coefficients were computed to
@@
 outcome variables and the predictor variables. Given the longitudinal design 
-an auto-regressive (AR1) correlation matrix was chosen for the GEE models, 
+an auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE models, 
 though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
@@
 in `r cite_st('qic')`) and we therefore did not include them in the main effect
-analysis. Given that TG is a risk factor
-for diabetes and that NEFA contribute to TG production, TG may act as a mediator
-between NEFA and the outcomes. To determine the role of TG in the
-association between NEFA and the outcomes, TG was included in the GEE model as a
+analysis. Given that TAG is a risk factor
+for diabetes and that NEFA contribute to TAG production, TAG may act as a mediator
+between NEFA and the outcomes. To determine the role of TAG in the
+association between NEFA and the outcomes, TAG was included in the GEE model as a
 sensitivity analysis. Lastly for the GEE models, we tested for an interaction
@@
 clinic visit. None of the NEFA variables as a mol% had any association with the
-outcomes. Adjusting for TG attenuated all associations with IGI/IR and ISSI-2
+outcomes. Adjusting for TAG attenuated all associations with IGI/IR and ISSI-2
 (data not shown).
@@
 measures (all p>0.15). Raw values from the GEE models are shown in 
-`r cite_st('gee_unadj')` for unadjusted models and `r cite_st('gee')` for fully
-adjusted models.
+`r cite_st('gee')` for fully adjusted models and `r cite_st('gee_unadj')` for unadjusted models.
 
@@
 middle group. The specific NEFA composition was not able to accurately classify
-participants who had low or high beta-cell function.
-`r cite_sf('plsda_groups')` shows the loadings of individual NEFA by the
-extracted component and `r cite_sf('plsda_loadings')` shows the clustering of
+participants who had low or high beta-cell function, however, those individuals 
+classified into the low beta-cell function group tended to have patterns of NEFA 
+composition higher in 14:0, 16:0, 14:1n-7, and 16:1n-7.
+`r cite_sf('plsda_loadings')` shows the loadings of individual NEFA by the
+extracted component and `r cite_sf('plsda_groups')` shows the clustering of
 the components by ISSI-2 latent class. The PLS-DA analysis using conversion to
-or maintenance of dysglycemia status as the response variable showed similar
+or maintenance of dysglycemia status over the 6 years as the response variable showed similar
 poor discriminatory ability, though better than the beta-cell function
@@
 The present analysis is the first study to our knowledge to examine the 
-longitudinal association of a broad spectrum of individual NEFA species on
+longitudinal association of a broad spectrum of individual NEFA on
 beta-cell function in a large cohort. We found that there was a strong signal of
@@
 acids as a mol% and using novel clustering analysis approaches, no individual or
-specific fatty acid species in the NEFA fraction predicted lower beta-cell
+specific fatty acids in the NEFA fraction strongly predicted lower beta-cell
 function, insulin sensitivity, or dysglycemia status. Taken together, these
 results suggest that it is the absolute size of the circulating NEFA fraction,
-irrespective of any specific composition of fatty acid species, that is
+irrespective of any specific composition of fatty acids, that is
 responsible for the hypothesized lipotoxic effects of NEFA on the beta-cells.
@@
 Biologically, within a free-living population chronic elevation of NEFA may be 
-mediating it's association through TG. In normal metabolism, NEFA enters the 
-liver and assists in the production of TG that is to be processed into very-low
+mediating it's association through TAG. In normal metabolism, NEFA enters the 
+liver and assists in the production of TAG that is to be processed into very-low
 density lipoproteins (VLDL) [@Nielsen2012a]. As such higher NEFA may contribute 
 to hypertriglyceridemia, which is a known risk factor for diabetes. In the
-sensitivity analysis adjusting for TG, all associations with beta-cell function
+sensitivity analysis adjusting for TAG, all associations with beta-cell function
 were attenuated, suggesting that NEFA may in fact be mediating it's association
-through TG.
+through TAG.
 
@@
 the underlying IR. Adipose tissue, which release NEFA during fasting, can become
-insulin resistant and thus lipolysis of intracellular TG by hormone-sensitive 
+insulin resistant and thus lipolysis of intracellular TAG by hormone-sensitive 
 lipase may not respond to the inhibitory action of insulin. As such, IR may 
@@
 [@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the adipose
-tissue following postprandial TG lipolysis via lipoprotein lipase may result in
+tissue following postprandial TAG lipolysis via lipoprotein lipase may result in
 NEFA spillover into the blood and a subsequent increase in circulating NEFA
@@
 may be useful biomarkers for identifying healthy compared to diabetic
-individuals [@Liu2010a]. However, in both studies, major limitations were the
+individuals [@Liu2010a]. However, both studies were limited by
 smaller sample sizes (approximately 100 subjects) and the cross-sectional
@@
 ```{r supFigPlsGrp, fig.cap=sfig_plsda_groups}
-plot_plsda_grouping(plsda_results)
+plot_plsda_grouping(plsda_results, 'Beta-cell function')
 ```
@@
 ```{r supFigPlsGrpDysgly, fig.cap=sfig_plsda_groups_dysgly}
-plot_plsda_grouping(plsda_dysgly)
+plot_plsda_grouping(plsda_dysgly, 'Dysglycemia')
 ```
@@
 
+```{r supFigPlsExample, fig.cap=sfig_plsda_example}
+plot_plsda_grouping(example_plsda_results(), 'Example groups')
+```
+
 # References

### 2016-05-17 R/setup.R
Moved mason and seer down to local for setup

@@ -9,6 +9,7 @@ run_setup <- function() {
     .load_packages()
     .set_options()
     .load_packages()
+    .load_packages()
 }
 
 .load_packages <- function() {
@@ -35,16 +36,11 @@ run_setup <- function() {
     if (!require(dplyr)) install.packages('dplyr')
     # - For github packages:
     # if (!require(package)) packrat::install_github('username/package')
-    if (!require(mason)) {
-        if (!require(curl)) install.packages('curl')
-        packrat::install_github('lwjohnst86/mason')
-    }
-    if (!require(seer)) {
-        if (!require(curl)) install.packages('curl')
-        packrat::install_github('lwjohnst86/seer')
-    }
+
     # How to manage local? Not sure.
     library(carpenter)
+    library(mason)
+    library(seer)
 }
 
 .set_options <- function() {

### 2016-05-17 R/functions.R
Minor fixes to functions based on changes to mason

@@ -322,7 +322,7 @@ plot_heatmap <- function(data, x = c(outcomes, 'BMI', 'Waist', 'Age', 'lALT',
     data %>%
         dplyr::filter(VN == 0) %>%
         mason::design_analysis('cor') %>%
-        mason::add_settings(method = 'pearson', obs.usage = 'complete.obs') %>%
+        mason::add_settings(method = 'pearson', use = 'complete.obs') %>%
         mason::add_variables('yvars', y) %>%
         mason::add_variables('xvars', x) %>%
         mason::construct_analysis() %>%
@@ -332,7 +332,6 @@ plot_heatmap <- function(data, x = c(outcomes, 'BMI', 'Waist', 'Age', 'lALT',
         mason::polish_renaming(function(x)
             gsub('l(ALT|TAG|IGIIR|invHOMA|ISI|ISSI2)', '\\1', x) %>%
                 renaming_outcomes(), 'Vars1') %>%
-        mason::polish_round(2) %>%
         dplyr::mutate(
             order1 = substr(Vars2, nchar(Vars2), nchar(Vars2)),
             order1 = ifelse(order1 == 0, 10, order1),

### 2016-05-17 doc/manuscript.Rmd
Updated manuscript as per Tony's edits

@@
 csl: /home/luke/Documents/bibliography/styles/dependent/diabetes-care.csl
-output:
-    word_document:
-        reference_docx: diabetes-care-template.docx
-        fig_height: 6 
-        fig_width: 7
-        fig_caption: true
-        keep_md: true
+output: fost::diabetes_care_manuscript
 ---
-
-<!--
-
-TODO:
-- Use pct for PLSDA
-
--->
 
@@
 
-Word count: 3569 / 4000
+Word count: 3577 / 4000
 
@@
 
-<!-- Reference limit / 40 -->
+<!-- Reference limit 41 / 40 -->
 
 # Abstract
-249 / 250 words
+248 / 250 words
 
 Objective: Our aim was to determine the longitudinal associations of individual
-non-esterified fatty acids (NEFA) on changes in insulin sensitivity and
+non-esterified fatty acids (NEFA) on changes in insulin sensitivity (IS) and
 beta-cell function over 6 years in a cohort of individuals who are at-risk for
@@
 points over 6 years. Outcome variables were calculated using the OGTT values.
-Insulin sensitivity was assessed using 1/HOMA-IR and the Matsuda index (ISI),
-while beta-cell function was assessed using the Insulinogenic index over HOMA-IR
-(IGI/IR) and the Insulin Secretion-Sensitivity Index-2 (ISSI-2). Generalized
-estimating equations was used, adjusting for time, waist, sex, ethnicity, age,
-ALT, alcohol intake, physical activity, and family history of diabetes. NEFA
-were analyzed as both a concentration and as a proportion of the total fraction.
+IS was assessed using 1/HOMA-IR and the Matsuda index, while beta-cell function
+was assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin
+Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were
+used, adjusting for time, waist, sex, ethnicity, age, ALT, alcohol intake,
+physical activity, and family history of diabetes. NEFA were analyzed as both
+concentrations and proportions (mol%) of the total fraction.
 
-Results: Insulin sensitivity and beta-cell function declined by 14-27% while
-BMI did not change over the 6-years. Total NEFA, 16:0, 18:1n-9, and 18:2n-6 as
-concentrations had negative associations with IGI/IR and ISSI-2. These four NEFA
-variables had similar magnitudes of association, predicting between 4-8% lower
-IGI/IR and ISSI-2 over the 6 years. No associations were seen for 1/HOMA-IR
-or ISI, nor were any associations seen for the mol% data.
+Results: IS and beta-cell function declined by 14-27% while BMI
+did not significantly change over 6 years of follow-up. Total NEFA, 16:0,
+18:1n-9, and 18:2n-6 as concentrations had significant negative associations
+with IGI/IR and ISSI-2, predicting between 4-8% lower IGI/IR and ISSI-2 over the
+6 years. However, individual NEFA modeled as mol% showed no associations with
+these outcomes. Total and individual NEFA were not associated with IS measures.
 
-Conclusions: Total NEFA was a strong predictor for lower beta-cell function over
-6 years. Our results suggest that association with beta-cell function is due to
-the absolute size of the serum NEFA fraction, rather than the specific fatty acid
-composition.
+Conclusions: Total NEFA concentration was a strong predictor of lower beta-cell
+function over 6 years. Our results suggest that the association with beta-cell
+function is due to the absolute size of the serum NEFA fraction, rather than the
+specific fatty acid composition.
 
@@
 - Describe LCMM and PLS-DA better in plots
+- [X] Use pct for PLSDA
+- [X] Explained variation in X and in Y for PLS-DA
+- Supplemental only file (separate from manuscript).. maybe?
 
@@
 knitr::opts_knit$set(root.dir = '../')
-knitr::opts_chunk$set(echo = FALSE, dev = c('png', 'postscript'))
+knitr::opts_chunk$set(echo = FALSE, dev = c('png', 'postscript'), dpi = 180)
 ```
@@
 multiple molecules with diverse physiological functions, and few 
-studies have compared the effects of a broader spectrum of fatty acids. Notably,
+studies have analyzed the effects of a broader spectrum of fatty acids. Notably,
 one study suggests that eicosapentaenoic acid (20:5n-3) can protect against the
@@
 Despite a sizable literature studying the role of total NEFA concentration in 
-diabetes, there are important gaps in the research. The majority of previous 
+diabetes, there are important gaps in this research. The majority of previous 
 studies have used animal models or cell lines [@Szendroedi2012a;@Kehlenbrink2012a], 
@@
 history of diabetes and the growing appreciation of the divergent effects of
-individual fatty acids. Therefore, our research objective was to examine the 
+individual fatty acids. Therefore, our objective was to examine the 
 association of serum NEFA composition on changes over time in insulin
@@
 chromatography peak areas to that of the internal standards [@Nishi2014a]. 
-There were 22 fatty acids measured in the NEFA fraction. Given the
+There were 22 fatty acids measured in the NEFA fraction. Given their
 diverse biology, as well as the complexity of the analyses, findings for other lipid
-fractions in this cohort are reported separately (see {{pl and ce paper}} for the
+fractions in this cohort are reported separately (see [@Johnston2016a] for the
 analysis of the phospholipid and cholesteryl ester fractions).
@@
 ALT, MET, alcohol intake, and family history of diabetes (M6). For beta-cell
-function, the last few models (M5-M8) were comparable in performance, so we
+function, the last four models (M5-M8) were comparable in performance, therefore we
 report results from the simplest model (M8), adjusting for time, sex, ethnicity,
@@
 percent difference in the outcome variable for every SD increase in the
-predictor variable given the covariates are held constant. While we a priori
-decided to test for and report on the time by predictor interaction results, we
+predictor variable given the covariates are held constant. While we decided *a priori*
+to test for and report on the time by predictor interaction results, we
 also assessed the overall fit of models that included a time interaction, which
@@
 used to identify the patterns of NEFA composition against beta-cell function
-(using ISSI-2, which had most of the associations in the GEE models) and for
+(using ISSI-2, which had the majority of the significant associations in the GEE models) and for
 those who converted to or maintained dysglycemia status (either IFG, IGT, or DM)
@@
 `r packageVersion('caret')` for PLS-DA, and lcmm `r packageVersion('lcmm')` for
-LCMM. The R code for this manuscript is available at {{link}}. Results were
-statistically significance at p<0.05, after adjusting for multiple testing using
-the Benjamini-Hochberg False Discovery Rate [@Benjamini1995a].
+LCMM. The R code for this manuscript is available at {{add link before
+submission}}. Results were considered statistically significance at p<0.05,
+after adjusting for multiple testing using the Benjamini-Hochberg False
+Discovery Rate [@Benjamini1995a].
 
@@
 `r aide::average(ds0yr$BMI)` (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
-subset of the cohort, `r fhd.npct$npct`, had a family history of diabetes.
-The primary outcome variables had a significant median decline between 
+participants, `r fhd.npct$npct`, had a family history of diabetes.
+The primary outcome variables showed a significant median decline between 
 `r percent.change.outcomes$chg` (p`r percent.change.outcomes$p`) over the 6
-years in this analysis of the PROMISE cohort (n=`r percent.change.outcomes$n`).
+years in this analysis of the PROMISE cohort.
 Consistent with this decline, there were `r dm.convert`
@@
 All individual NEFA as well as the total fraction had
-correlations that ranged from weak to null (r<0.3) with the basic
+correlations that ranged from weak to null (r<0.3) with participant
 characteristics (see `r cite_sf('corr')`).
@@
 
-Several associations were seen in the unadjusted GEE models 
-(see `r cite_sf('gee')`), particularly for the concentration compared to the
-mol% data. Full model adjustment (`r cite_f('gee')`) attenuated most of these
+A number of associations were seen in the unadjusted GEE models 
+(see `r cite_sf('gee')`), particularly for NEFA modeled as a concentration.
+Full model adjustment (`r cite_f('gee')`) attenuated most of these
 associations, although total NEFA, 16:0, 18:1n-9, and 18:2n-6 (all as nmol/mL)
-had negative associations between
+had negative associations with
 IGI/IR and ISSI-2. The magnitude of association for each of these variables was
-fairly consistent for each beta-cell function measure. For every one SD decrease
+fairly consistent for each beta-cell function measure. For every one SD increase
 in any of these three NEFA variables, there was an average predicted
 `r est.gee$est[1]`% lower IGI/IR and `r est.gee$est[2]`% lower ISSI-2 at each
-clinic visit. None of the NEFA variables as a mol% had any association with the
+clinic visit. However, in contrast to these results using NEFA modeled as concentrations, none of the NEFA variables modeled as mol% were associated with the
 outcomes. Adjusting for TAG attenuated all associations with IGI/IR and ISSI-2
@@
 `r lcmm.groups$Freq[3]`
-in the low beta-cell function groups. The trajectories of the three groups are
+in the low beta-cell function groups. The trajectories of the these groups are
 shown in `r cite_sf('lcmm')`. These groups were used as the beta-cell function
@@
 trajectory results. While `r misclass.n.percent.dysgly$NPct[2]` participants 
-were classified correctly based on dysglycemia status, examining the clustering
-of the NEFA composition showed poor discrimination between groups (see 
-`r cite_sf('plsda_groups_dysgly')` and `r cite_sf('plsda_loadings_dysgly')`).
+were classified correctly based on dysglycemia status, this was almost entirely
+due to classifying correctly into the non-dysglycemia status. Examining the
+clustering of the NEFA composition showed poor discrimination between groups
+(see `r cite_sf('plsda_groups_dysgly')` and 
+`r cite_sf('plsda_loadings_dysgly')`).
 
@@
 linoleic acid (18:2n-6) for the concentration (nmol/mL) modeling, no
-associations were seen
-for the proportion of these fatty acids and further clustering analysis found no
-predictive ability of these NEFA on beta-cell function or dysglycemia. These
-observations suggest that the absolute size of the total NEFA fraction, rather than
-the specific composition of any individual NEFA, likely influences the
-pathogenesis of diabetes.
+associations were seen for the proportion of these fatty acids and further
+clustering analysis found no predictive ability of these individual NEFA on
+beta-cell function or dysglycemia. These observations suggest that the absolute
+size of the total NEFA fraction, rather than its specific composition, likely
+influences the pathogenesis of diabetes.
 
@@
 through endoplasmic reticulum stress, formation of ceramides, and generation of
-nitric oxide, as well as impairment of proinsulin production and mitochrondrial
+nitric oxide, as well as impairment of proinsulin production and mitochondrial
 function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a].
@@
 beta-cell function in a large cohort. We found that there was a strong signal of
-higher total NEFA, palmitic acid, oleic acid, and linoleic acid as
+higher total NEFA, palmitic acid, oleic acid, and linoleic acid modeled as
 concentrations with lower beta-cell function. However, in modeling these fatty
-acids as a mol% and using novel clustering analysis approaches, no individual or
-specific fatty acids in the NEFA fraction strongly predicted lower beta-cell
+acids as a mol% and using novel clustering analysis approaches, no specific
+fatty acids in the NEFA fraction strongly predicted lower beta-cell
 function, insulin sensitivity, or dysglycemia status. Taken together, these
@@
 
-Biologically, within a free-living population chronic elevation of NEFA may be 
-mediating it's association through TAG. In normal metabolism, NEFA enters the 
+Biologically, in free-living populations chronic elevation of NEFA may be 
+mediating it's association with metabolic outcomes through TAG. In normal
+metabolism, NEFA enters the
 liver and assists in the production of TAG that is to be processed into very-low
-density lipoproteins (VLDL) [@Nielsen2012a]. As such higher NEFA may contribute 
+density lipoproteins (VLDL) [@Nielsen2012a]. As such, higher NEFA may contribute 
 to hypertriglyceridemia, which is a known risk factor for diabetes. In the
@@
 were attenuated, suggesting that NEFA may in fact be mediating it's association
-through TAG.
+with beta-cell dysfunction through higher TAG.
 
@@
 NEFA with hepatic (1/HOMA-IR) or whole-body insulin sensitivity (ISI). 
-There are some possible explanations for these findings. Previous studies done *in
-vivo* were mostly short term infusion protocols
+There are several possible explanations for these findings. Previous studies
+done *in vivo* were mostly short term infusion protocols
 [@Xiao2009a;@Xiao2006;@Szendroedi2012a;@Kehlenbrink2012a] or pharmocologic
-[@Daniele2014a;@Liang2013a] trials and the observational research was generally
+[@Daniele2014a;@Liang2013a] trials and previous observational research was generally
 cross-sectional or correlational
@@
 elevations in NEFA concentrations make it difficult to assess the causal role
-in non-experimental environments over long time frames. Moreover, some studies
+in free-living environments over long time frames. Moreover, some studies
 have found null or weak associations of NEFA and IR 
@@
 insulin may be involved in NEFA uptake [@Ramos-Roman2012a], further supporting 
-the hypothesis that elevated NEFA is a consequence of higher IR. However, we 
+the hypothesis that elevated NEFA may be a consequence of higher IR. However, we 
 could not examine this potential hypothesis as NEFA was only collected at the
@@
 examined in PROMISE. It may be that in this population, IR has become well
-established and NEFA may not contribute to IR at this somewhat mode advanced
+established and NEFA may not contribute to IR at this somewhat more advanced
 stage in the pathogenesis of diabetes.

### 2016-06-17 doc/manuscript.Rmd
Removed Zinman (as he requested) from author list
Made edits as per Ravi's edits

@@
 Author: Luke W. Johnston (1), MSc; Stewart B. Harris (2), MD; Ravi Retnakaran
-(3,4), MD; Bernard Zinman (3,4), MD; Adria Giacca (5), PhD; Zhen Liu (1), PhD;
+(3,4), MD; Adria Giacca (5), PhD; Zhen Liu (1), PhD;
 Richard P. Bazinet (1), PhD; and Anthony J. Hanley (1,6), PhD
@@
 
-At each examination, an 8-12 hour fasting blood sample was draw from each
+At each examination, an 8-12 hour fasting blood sample was drawn from each
 participant, followed by a 75g oral glucose tolerance test (OGTT) with a 30
@@
 though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
-is well suited to longitudinal cohort studies given it's capacity to handle
+is well suited to longitudinal cohort studies given its capacity to handle
 missed visits. The predictor variables and continuous covariates were scaled
@@
 `r lcmm.groups$Freq[3]`
-in the low beta-cell function groups. The trajectories of the these groups are
+in the low beta-cell function groups. The trajectories of these groups are
 shown in `r cite_sf('lcmm')`. These groups were used as the beta-cell function
@@
 Biologically, in free-living populations chronic elevation of NEFA may be 
-mediating it's association with metabolic outcomes through TAG. In normal
+mediating its association with metabolic outcomes through TAG. In normal
 metabolism, NEFA enters the
@@
 sensitivity analysis adjusting for TAG, all associations with beta-cell function
-were attenuated, suggesting that NEFA may in fact be mediating it's association
+were attenuated, suggesting that NEFA may in fact be mediating its association
 with beta-cell dysfunction through higher TAG.
@@
 done *in vivo* were mostly short term infusion protocols
-[@Xiao2009a;@Xiao2006;@Szendroedi2012a;@Kehlenbrink2012a] or pharmocologic
+[@Xiao2009a;@Xiao2006;@Szendroedi2012a;@Kehlenbrink2012a] or pharmacologic
 [@Daniele2014a;@Liang2013a] trials and previous observational research was generally

### 2016-06-28 doc/manuscript.Rmd
Updated analysis and manuscript as per Adria's suggestions

@@
     - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
-csl: /home/luke/Documents/bibliography/styles/dependent/diabetes-care.csl
+#csl: /home/luke/Documents/bibliography/styles/dependent/diabetes-care.csl
+csl: /home/luke/Documents/bibliography/styles/vancouver-brackets-no-et-al.csl
 output: fost::diabetes_care_manuscript
@@
 
-Word count: 3577 / 4000
+Word count: 3543 / 4000
 
@@
 # Abstract
-248 / 250 words
+243 / 250 words
 
@@
 Research Design and Methods: In the PROMISE cohort, 477 participants had serum
-NEFA measured at the baseline visit and who had completed an OGTT at 3 time
+NEFA measured at the baseline visit and completed an OGTT at 3 time
 points over 6 years. Outcome variables were calculated using the OGTT values.
@@
 Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were
-used, adjusting for time, waist, sex, ethnicity, age, ALT, alcohol intake,
-physical activity, and family history of diabetes. NEFA were analyzed as both
-concentrations and proportions (mol%) of the total fraction.
+used, adjusting for time, waist, sex, ethnicity, baseline age, ALT, and family
+history of diabetes. NEFA were analyzed as both concentrations and proportions
+(mol%) of the total fraction.
 
@@
 knitr::opts_knit$set(root.dir = '../')
-knitr::opts_chunk$set(echo = FALSE, dev = c('png', 'postscript'), dpi = 180)
 ```
@@
 fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006).')
-fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function using generalized estimating equations over the 6 years in the PROMISE cohort.
-Adjusted for time, sex, ethnicity, baseline age, WC,
-ALT, and family history of diabetes (plus physical activity and alcohol intake for beta-cell function models).
+fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
+Generalized estimating equation models were adjusted for time, sex, ethnicity, baseline age, WC,
+ALT, and family history of diabetes.
 Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
@@
 sfig_corr <- sfig('corr', 'Pearson correlation heatmap of non-esterified fatty acids (nmol/mL) and basic PROMISE participant characteristics for the baseline visit (2004-2006).
-Darkness of the colour indicates the magnitude of the correlation, with blue indicating positive and orange indicating negative correlations.')
+Darkness of the colour indicates the magnitude of the correlation, with orange indicating positive and blue indicating negative correlations.')
 sfig_gee <- sfig('gee', 'Unadjusted generalized estimating equation modeling of the longitudinal association of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort.
-Models are only adjusted for time.
+GEE models are only adjusted for time.
 Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
 P-values were adjusted for the BH false discovery rate, presented as the dot size.')
-stab_gee <- stab('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function using generalized estimating equations over the 6 years in the PROMISE cohort.
-Adjusted for time, sex, ethnicity, baseline age, WC,
-ALT, and family history of diabetes (plus physical activity and alcohol intake for beta-cell function models).
+stab_gee <- stab('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
+GEE models were adjusted for time, sex, ethnicity, baseline age, WC,
+ALT, and family history of diabetes.
 Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
-P-values were adjusted for the BH false discovery rate, indicated by asterisk.')
-stab_gee_unadj <- stab('gee_unadj', 'Unadjusted generalized estimating equations models of the longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
-Models are only adjusted for time.
+P-values were adjusted for the BH false discovery rate, with significant (p<0.05) associations indicated by asterisk.')
+stab_gee_unadj <- stab('gee_unadj', 'Unadjusted generalized estimating equation models of the longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
+GEE models are only adjusted for time.
 Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
-P-values were adjusted for the BH false discovery rate, indicated by asterisk.')
+P-values were adjusted for the BH false discovery rate, with significant (p<0.05) associations indicated by asterisk.')
 sfig_lcmm <- sfig('lcmm', 'Latent class mixed model (LCMM) analysis to identify individual classes of trajectories for log(ISSI-2) over the 6 years in the PROMISE cohort.
@@
 sfig_plsda_groups <- sfig('plsda_groups', '')
-sfig_plsda_loadings <- sfig('plsda_loadings', 'Pattern loadings from partial least squares discriminant analysis to identify potential clusters of NEFA composition within the classes extracted from the latent class mixed model in 463 participants from the baseline PROMISE visit (2004-2006).
-The percent explained variance of each component is shown in brackets on each axis.')
+sfig_plsda_loadings <- sfig('plsda_loadings', 'Pattern loadings from partial least squares discriminant analysis (PLS-DA) to identify potential clusters of NEFA composition within the classes extracted from the latent class mixed model in 463 participants from the baseline PROMISE visit (2004-2006).
+The percent explained variance of each component is shown in brackets on each axis.
+The solid line represents an explained variance of 100% while the dashed line represents an explained variance of 50%.
+See the Supplemental Methods for a description of PLS-DA and an explanation of interpreting this plot.')
 sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', '')
-sfig_plsda_loadings_dysgly <- sfig('plsda_loadings_dysgly', 'Pattern loadings from partial least squares discriminant analysis to identify potential clusters of NEFA composition for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).
-The percent explained variance of each component is shown in brackets on each axis.')
+sfig_plsda_loadings_dysgly <- sfig('plsda_loadings_dysgly', 'Pattern loadings from partial least squares discriminant analysis (PLS-DA) to identify potential clusters of NEFA composition for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).
+See the Supplemental Methods for a description of PLS-DA and an explanation of interpreting this plot.
+The percent explained variance of each component is shown in brackets on each axis.
+The solid line represents an explained variance of 100% while the dashed line represents an explained variance of 50%.')
 sfig_plsda_example <- sfig('plsda_example', 'Example results of a high discriminatory ability to classify groups accurately when using partial least squares discriminatory analysis (PLS-DA).
 Discriminatory ability is evident from the amount of separation between the two groups.
-In this case, PLS-DA was 82% accurate at correctly classifying groups.')
+In this case, PLS-DA was 82% accurate at correctly classifying groups.
+See the Supplemental Methods for a description of PLS-DA and for interpreting this plot.')
 sfig_plsda_groups <- sfig('plsda_groups', paste0('Clustering of extracted components from the partial least squares discriminant analysis (PLS-DA) on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006).
-PLS-DA is a multivariate technique that The percent explained variance of each component is shown in brackets on each axis.
+See the Supplemental Methods for a description of PLS-DA and interpreting this plot.
+The percent explained variance of each component is shown in brackets on each axis.
 Red, green, and blue lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis.
@@
 sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', paste0('Clustering of extracted components from the partial least squares discriminant analysis for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).
+See the Supplemental Methods for a description of PLS-DA and interpreting this plot.
 The percent explained variance of each component is shown in brackets on each axis.
@@
 in particular that elevated total plasma NEFA associates with an increased risk
-for incident type 2 diabetes [@Daniele2014a;@Salgin2012a].
+for incident type 2 diabetes [@Paolisso1995a;@Salgin2012a].
 
@@
 one study suggests that eicosapentaenoic acid (20:5n-3) can protect against the
-lipotoxic effect of palmitic acid in the beta-cells [@Xiao2006].
+lipotoxic effect of palmitic acid in the beta-cells [@Kato2008].
 
@@
 diabetes, there are important gaps in this research. The majority of previous 
-studies have used animal models or cell lines [@Szendroedi2012a;@Kehlenbrink2012a], 
-have been short term human infusion trials [@Daniele2014a;@Liang2013a], or have
+studies have used animal models or cell lines [@Maris2013a;@Deguil2011a], 
+have been short term human trials 
+[@Szendroedi2012a;@Daniele2014a;@Liang2013a;@Kehlenbrink2012a], or have
 been epidemiological studies that only looked at total NEFA and not individual
@@
 been no longitudinal studies examining the role of the composition of the serum NEFA
-fraction on the pathogenesis of diabetes, which is critical to
-understanding associations at a more granular level given the protracted natural
+fraction on the pathogenesis of diabetes, which is critical given the protracted natural
 history of diabetes and the growing appreciation of the divergent effects of
@@
 sensitivity and beta-cell function in a longitudinal cohort. We hypothesized
-that higher palmitic acid and lower polyunsaturated fatty acids such as
+that higher palmitic acid and lower PUFA such as
 eicosapentaenoic acid would associate with declining insulin sensitivity and
@@
 procedures. Cholesterol, HDL, and triacylglycerides (TAG) were measured using Roche
-Modular's enzymatic colorimetric tests (Mississauga, ON).  Both specific insulin
+Modular's enzymatic colorimetric tests (Mississauga, ON).  Both insulin
 and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
@@
 immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
-and the Des 31,32 circulating split form (Linco Res. Inc), and has a CV of 9.3%.
+and the Des 31,32 circulating split form (Linco Res. Inc), and has a coefficient of variation (CV) of 9.3%.
 Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
@@
 diverse biology, as well as the complexity of the analyses, findings for other lipid
-fractions in this cohort are reported separately (see [@Johnston2016a] for the
+fractions in this cohort are reported separately (see ref @Johnston2016a for the
 analysis of the phospholipid and cholesteryl ester fractions).
@@
 though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
-is well suited to longitudinal cohort studies given its capacity to handle
+is well suited to longitudinal cohort studies given it's capacity to handle
 missed visits. The predictor variables and continuous covariates were scaled
@@
 [@Greenland1999a] recommendations, and quasi-likelihood information criterion 
-(QIC). `r cite_st('qic')` shows the covariates compared using QIC and 
-`r cite_sf('dag')` shows the specified directed acyclic graph. The final GEE model 
-we selected differed between insulin sensitivity and beta-cell function models. 
-For insulin sensitivity, we adjusted for time, sex, ethnicity, baseline age, WC,
-ALT, MET, alcohol intake, and family history of diabetes (M6). For beta-cell
-function, the last four models (M5-M8) were comparable in performance, therefore we
-report results from the simplest model (M8), adjusting for time, sex, ethnicity,
-baseline age, WC, ALT,
-and family history of diabetes. After scaling, log-transforming, and
-exponentiating the GEE estimates, the GEE results are interpreted as an expected
-percent difference in the outcome variable for every SD increase in the
-predictor variable given the covariates are held constant. While we decided *a priori*
-to test for and report on the time by predictor interaction results, we
-also assessed the overall fit of models that included a time interaction, which
-ultimately did not improve the overall fit of our models in terms of QIC
-(M6.1 and M8.1 as seen
-in `r cite_st('qic')`) and we therefore did not include them in the main effect
-analysis. Given that TAG is a risk factor
-for diabetes and that NEFA contribute to TAG production, TAG may act as a mediator
-between NEFA and the outcomes. To determine the role of TAG in the
-association between NEFA and the outcomes, TAG was included in the GEE model as a
-sensitivity analysis. Lastly for the GEE models, we tested for an interaction
-with sex or ethnicity by the predictor term for each outcome variable.
+(QIC). `r cite_st('qic')` shows the covariates compared using QIC. While the
+final GEE model selected as best fitting differed between insulin sensitivity
+and beta-cell function measures, the differences in QIC values were less than
+10 between the models, suggesting similar fit. As such, we selected the model
+that had the fewest covariates and that had similar fit between the outcome
+measures (M8). The final GEE model was adjusted for time, sex, ethnicity,
+baseline age, WC, ALT, and family history of diabetes. After scaling,
+log-transforming, and exponentiating, the GEE estimates are
+interpreted as an expected percent difference in the outcome variable for every
+SD increase in the predictor variable given the covariates are held constant.
+Since TAG is a risk factor for diabetes and since NEFA contribute to TAG
+production, TAG may act as a mediator between NEFA and the outcomes. To
+determine the role of TAG in the association between NEFA and the outcomes, TAG
+was included in the GEE model as a sensitivity analysis. Lastly for the GEE
+models, we tested for an interaction with sex, ethnicity, or time by the
+predictor term for each outcome variable.
 
@@
 limited in that it cannot analyze the inherent multivariate nature of the
-composition of the NEFA fraction. As such, after running the GEE models for the
-longitudinal analysis, partial least squares discriminant analysis (PLS-DA) was
-used to identify the patterns of NEFA composition against beta-cell function
-(using ISSI-2, which had the majority of the significant associations in the GEE models) and for
-those who converted to or maintained dysglycemia status (either IFG, IGT, or DM)
-over the 6 years. Since PLS-DA cannot handle longitudinal data, latent class
-mixed models (LCMM) were used to extract latent trajectories of beta-cell
-function (ISSI-2) over the 6 years.
+composition of the NEFA fraction. Therefore, to confirm the results of the GEE
+analyses in a multivariate environment (all NEFA variables in a single model),
+partial least squares discriminant analysis (PLS-DA) was used to identify the
+patterns of NEFA composition against beta-cell function (using ISSI-2, which had
+the majority of the associations in the GEE models) and for those
+who converted to or maintained dysglycemia status (either IFG, IGT, or DM) over
+the 6 years. Latent class mixed models (LCMM) were used to extract the underlying
+trajectories of beta-cell function (ISSI-2) over the 6 years. For more detailed
+explanation of GEE, PLS-DA, and LCMM, please see the Supplemental Methods.
 
@@
 ne_conc <- c('TotalNE', grep('^ne\\d\\d', names(gee_df), value = TRUE))
-covariates_is <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'MET', 'AlcoholPerWk')
-covariates_bcf <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'FamHistDiab')
+covariates <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'FamHistDiab')
 ```
@@
 gee_results <- dplyr::bind_rows(
-    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', adj.p = TRUE),
-    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', adj.p = TRUE),
-    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', adj.p = TRUE),
-    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', adj.p = TRUE)
+    analyze_gee(gee_df, outcomes_is, ne_pct, covariates, 'mol%', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covariates, 'nmol/mL', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates, 'mol%', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates, 'nmol/mL', adj.p = TRUE)
     ) %>% 
@@
 gee_results_unadj <- dplyr::bind_rows(
-    analyze_gee(gee_df, outcomes, ne_pct, 'VN', 'mol%', adj.p = TRUE),
-    analyze_gee(gee_df, outcomes, ne_conc, 'VN', 'nmol/mL', adj.p = TRUE)
+    analyze_gee(gee_df, outcomes_is, ne_pct, 'VN', 'mol%', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_conc, 'VN', 'nmol/mL', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, 'VN', 'mol%', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, 'VN', 'nmol/mL', adj.p = TRUE)
     ) %>% 
@@
 participants, `r fhd.npct$npct`, had a family history of diabetes.
-The primary outcome variables showed a significant median decline between 
+Insulin sensitivity and beta-cell function measures showed a significant median
+decline between
 `r percent.change.outcomes$chg` (p`r percent.change.outcomes$p`) over the 6
@@
 converted to or maintained IFG or IGT status over the 6 years.
+<!-- maintain vs convert -->
 
@@
 (see `r cite_sf('gee')`), particularly for NEFA modeled as a concentration.
-Full model adjustment (`r cite_f('gee')`) attenuated most of these
-associations, although total NEFA, 16:0, 18:1n-9, and 18:2n-6 (all as nmol/mL)
-had negative associations with
-IGI/IR and ISSI-2. The magnitude of association for each of these variables was
+While full model adjustment (`r cite_f('gee')`) attenuated most of these
+associations, total NEFA, 16:0, 18:1n-9, and 18:2n-6 (all as nmol/mL)
+continued to show negative associations with IGI/IR and ISSI-2. Sensitivity
+analyses revealed that adjustment for WC was responsible for attenuating the
+unadjusted results. The magnitude of association for each of these variables was
 fairly consistent for each beta-cell function measure. For every one SD increase
@@
 `r est.gee$est[1]`% lower IGI/IR and `r est.gee$est[2]`% lower ISSI-2 at each
-clinic visit. However, in contrast to these results using NEFA modeled as concentrations, none of the NEFA variables modeled as mol% were associated with the
-outcomes. Adjusting for TAG attenuated all associations with IGI/IR and ISSI-2
-(data not shown).
-There were no significant interaction effects
-between time, sex, or ethnicity and the individual NEFA on any of the outcome
-measures (all p>0.15). Raw values from the GEE models are shown in 
-`r cite_st('gee')` for fully adjusted models and `r cite_st('gee_unadj')` for unadjusted models.
+clinic visit. However, in contrast to these results using NEFA modeled as
+concentrations, none of the NEFA variables modeled as mol% were associated with
+the outcomes. Adjusting for TAG attenuated all associations with IGI/IR and
+ISSI-2 (data not shown). Sensitivity analyses to examine the association with
+the sum of fatty acid classes (saturated vs unsaturated) revealed similar
+findings as seen with the adjusted GEE model for total NEFA (data not shown).
+There were no significant interaction effects between time, sex, or ethnicity
+and the individual NEFA on any of the outcome measures, except for a single
+marginally significant (p=0.04) interaction between sex, 20:3n-6 (mol%), and
+ISI, which showed that males had a 12% higher ISI for every SD increase in
+20:3n-6. Raw values from the GEE models are shown in `r cite_st('gee')` for
+fully adjusted models and `r cite_st('gee_unadj')` for unadjusted models.
 
-## Latent trajectory and NEFA pattern recognition
+## Latent trajectories and differences in NEFA composition
 
@@
 shown in `r cite_sf('lcmm')`. These groups were used as the beta-cell function
-response variables in the PLS-DA analysis. The PLS-DA results show a very poor
-discriminatory ability of the NEFA composition in classifying participants into
-the correct beta-cell function latent class. Only 
-`r misclass.n.percent.bcf$NPct[2]` 
-participants were correctly classified, primarily into the
-middle group. The specific NEFA composition was not able to accurately classify
-participants who had low or high beta-cell function, however, those individuals 
-classified into the low beta-cell function group tended to have patterns of NEFA 
-composition higher in 14:0, 16:0, 14:1n-7, and 16:1n-7.
-`r cite_sf('plsda_loadings')` shows the loadings of individual NEFA by the
-extracted component and `r cite_sf('plsda_groups')` shows the clustering of
-the components by ISSI-2 latent class. The PLS-DA analysis using conversion to
-or maintenance of dysglycemia status over the 6 years as the response variable showed similar
-poor discriminatory ability, though better than the beta-cell function
-trajectory results. While `r misclass.n.percent.dysgly$NPct[2]` participants 
-were classified correctly based on dysglycemia status, this was almost entirely
-due to classifying correctly into the non-dysglycemia status. Examining the
-clustering of the NEFA composition showed poor discrimination between groups
-(see `r cite_sf('plsda_groups_dysgly')` and 
-`r cite_sf('plsda_loadings_dysgly')`).
+response variables in the PLS-DA analysis, which was used to confirm the results
+of the GEE modeling, but in a multivariate environment (all NEFA in a single
+model). The PLS-DA results show a very poor discriminatory ability of the
+NEFA composition in classifying participants into the correct beta-cell function
+latent class. The percent of participants correctly predicted to belonging to
+their group (i.e. those predicted in the low group and actually belonging to the
+low group) was `r misclass.n.percent.bcf$Percent[9]` in the low group,
+`r misclass.n.percent.bcf$Percent[5]` in the middle group, and 
+`r misclass.n.percent.bcf$Percent[1]` in the high group (the distribution of the
+groupings based on the components from PLS-DA are shown in `r cite_sf('plsda_groups')`). 
+The loadings plot of the NEFA composition against the extracted PLS-DA
+components is shown in `r cite_sf('plsda_loadings')` and is used to identify
+fatty acids that contribute the most explained variance to the beta-cell
+function groupings. However, because of the high rate of misclassification, no
+meaningful predictive information can be extracted even if any NEFA does
+contribute to the explained variance (e.g. 16:0). The PLS-DA analysis using
+conversion to or maintenance of dysglycemia status over the 6 years as the
+response variable showed similar poor discriminatory ability, though better than
+the beta-cell function trajectory results. While 
+`r misclass.n.percent.dysgly$Percent[1]`
+of participants were correctly predicted to not have dysglycemia, only 
+`r misclass.n.percent.dysgly$Percent[4]` were correctly predicted to have
+dysglycemia (see `r cite_sf('plsda_groups_dysgly')` for the distribution of 
+dysglycemia status based on the components from the PLS-DA and 
+`r cite_sf('plsda_loadings_dysgly')` for the loadings plot of the NEFA composition
+by extracted PLS-DA components).
 
@@
 secretion and a higher risk for developing diabetes 
-[@Salgin2012a;@Pankow2004a]. In a cross-sectional analysis of the RISCK cohort,
-total NEFA had a negative association with insulin sensitivity and a
-particularly strong negative association with beta-cell function [@Johns2014a].
-Experimentally, several potential mechanisms have been elucidated for the role
-of NEFA on beta-cell function, particularly for palmitic acid. Prolonged
-exposure to elevated NEFA can induce apoptosis in the beta-cells, possibly
-through endoplasmic reticulum stress, formation of ceramides, and generation of
-nitric oxide, as well as impairment of proinsulin production and mitochondrial
-function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a].
-The present analysis is the first study to our knowledge to examine the 
-longitudinal association of a broad spectrum of individual NEFA on
-beta-cell function in a large cohort. We found that there was a strong signal of
-higher total NEFA, palmitic acid, oleic acid, and linoleic acid modeled as
-concentrations with lower beta-cell function. However, in modeling these fatty
-acids as a mol% and using novel clustering analysis approaches, no specific
-fatty acids in the NEFA fraction strongly predicted lower beta-cell
-function, insulin sensitivity, or dysglycemia status. Taken together, these
-results suggest that it is the absolute size of the circulating NEFA fraction,
-irrespective of any specific composition of fatty acids, that is
-responsible for the hypothesized lipotoxic effects of NEFA on the beta-cells.
+[@Paolisso1995a;@Salgin2012a;@Pankow2004a]. In a cross-sectional analysis of the
+RISCK cohort, total NEFA had a negative association with insulin sensitivity and
+a particularly strong negative association with beta-cell function
+[@Johns2014a]. Experimentally, several potential mechanisms have been elucidated
+for the role of NEFA on beta-cell function, particularly for palmitic acid.
+Prolonged exposure to elevated NEFA can induce apoptosis in the beta-cells,
+possibly through endoplasmic reticulum stress, formation of ceramides, and
+generation of nitric oxide, as well as impairment of proinsulin production and
+mitochondrial function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a]. The
+present analysis is the first study to our knowledge to examine the longitudinal
+association of a broad spectrum of individual NEFA on beta-cell function in a
+large cohort. We found that there was a strong signal of higher total NEFA,
+palmitic acid, oleic acid, and linoleic acid modeled as concentrations with
+lower beta-cell function. However, in modeling these fatty acids as a mol% and
+using novel clustering analysis approaches, no specific fatty acids in the NEFA
+fraction strongly predicted lower beta-cell function, insulin sensitivity, or
+dysglycemia status. Taken together, these results suggest that it is the
+absolute size of the circulating NEFA fraction, irrespective of any specific
+composition of fatty acids, that is responsible for the hypothesized lipotoxic
+effects of NEFA on the beta-cells.
 
@@
 mediating its association with metabolic outcomes through TAG. In normal 
-metabolism, NEFA enters the
-liver and assists in the production of TAG that is to be processed into very-low
-density lipoproteins (VLDL) [@Nielsen2012a]. As such, higher NEFA may contribute 
-to hypertriglyceridemia, which is a known risk factor for diabetes. In the
-sensitivity analysis adjusting for TAG, all associations with beta-cell function
-were attenuated, suggesting that NEFA may in fact be mediating its association
-with beta-cell dysfunction through higher TAG.
+metabolism, NEFA enters the liver and assists in the production of TAG that is
+to be processed into very-low density lipoproteins (VLDL) [@Nielsen2012a]. As
+such, higher NEFA may contribute to hypertriglyceridemia, which is a known risk
+factor for diabetes. In the sensitivity analysis adjusting for TAG, all
+associations with beta-cell function were attenuated, suggesting that NEFA may
+in fact be mediating its association with beta-cell dysfunction through higher
+TAG.
 
@@
 signaling cascades, as reviewed in previously published articles 
-[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this
-longitudinal analysis, we saw no association of any individual or total
-NEFA with hepatic (1/HOMA-IR) or whole-body insulin sensitivity (ISI). 
-There are several possible explanations for these findings. Previous studies
-done *in vivo* were mostly short term infusion protocols
-[@Xiao2009a;@Xiao2006;@Szendroedi2012a;@Kehlenbrink2012a] or pharmacologic
-[@Daniele2014a;@Liang2013a] trials and previous observational research was generally
-cross-sectional or correlational
-[@Miller2012a;@Johns2014a;@Bush2012a;@Ilyasova2010a]. The short time periods of
-these studies and the experimentally-induced
-elevations in NEFA concentrations make it difficult to assess the causal role
-in free-living environments over long time frames. Moreover, some studies
-have found null or weak associations of NEFA and IR 
-[@Gray2013a;@Magkos2012a;@Kehlenbrink2012a;@Johns2014a], suggesting that either
-NEFA has only a minor role in IR or that NEFA is the effect, but not cause, of
-the underlying IR. Adipose tissue, which release NEFA during fasting, can become
-insulin resistant and thus lipolysis of intracellular TAG by hormone-sensitive 
-lipase may not respond to the inhibitory action of insulin. As such, IR may 
-develop before elevations in blood NEFA. There is also evidence to suggest that 
-insulin may be involved in NEFA uptake [@Ramos-Roman2012a], further supporting 
-the hypothesis that elevated NEFA may be a consequence of higher IR. However, we 
-could not examine this potential hypothesis as NEFA was only collected at the
-baseline visit.
-
-There are two other possible explanations for our null findings for the insulin
-sensitivity measures. First, there may be differences in physiology between
-fasting and postprandial NEFA kinetics. For instance, some experimental studies
-using clamp protocols found that fasting NEFA was a weak predictor of insulin
-sensitivity compared to postprandial concentrations of NEFA
-[@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the adipose
-tissue following postprandial TAG lipolysis via lipoprotein lipase may result in
-NEFA spillover into the blood and a subsequent increase in circulating NEFA
-[@Almandoz2013a], which may be more metabolically active given postprandial 
-activity. Second, the null findings may be due to the high risk population
-examined in PROMISE. It may be that in this population, IR has become well
-established and NEFA may not contribute to IR at this somewhat more advanced
-stage in the pathogenesis of diabetes.
+[@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this longitudinal
+analysis, we saw no association of any individual or total NEFA with hepatic
+(1/HOMA-IR) or whole-body insulin sensitivity (ISI). We identified that
+adjustment for WC was likely the cause for the attenuation in results between
+unadjusted and adjusted models. There are some possible explanations for our
+null findings for the insulin sensitivity measures and the influence of WC on
+the results. First, there may be differences in physiology between fasting and 
+postprandial NEFA kinetics that we are not able to investigate but that may 
+explain our null findings for fasting NEFA and insulin sensitivity. For
+instance, some experimental studies using clamp protocols found that fasting
+NEFA was a weak predictor of insulin sensitivity compared to postprandial
+concentrations of NEFA [@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA
+uptake into the adipose tissue (e.g. as seen in obesity and larger WC) following
+postprandial TAG lipolysis via lipoprotein lipase may result in NEFA spillover
+into the blood and a subsequent increase in circulating NEFA [@Almandoz2013a],
+which may be more metabolically active given postprandial activity. Second, the
+null findings may be due to the high risk population examined in PROMISE,
+including having a higher WC. It may be that in this population, IR has become
+well established and NEFA may not contribute to IR at this somewhat more
+advanced stage in the pathogenesis of diabetes.
 
@@
 integrated. This is also an observational cohort, and there may be some residual
-confounding we haven't considered or that couldn't be measured. Nonetheless,
+confounding we have not considered or that could not be measured. Nonetheless,
 potential covariates were empirically analyzed prior to inclusion into the
@@
 The authors had the following responsibility: LWJ conducted research, analyzed
-data, and wrote the paper; RR, ZL, BZ, and SBH designed research, conducted
+data, and wrote the paper; RR, ZL, and SBH designed research, conducted
 research, and provided essential materials (infrastructure and clinical
-resources); RR, BZ, SBH, RPB, and AG provided intellectual feedback on the
+resources); RR, SBH, RPB, and AG provided intellectual feedback on the
 paper; RPB conducted research, provided essential reagents and materials; AJH
@@
 authors report no potential conflicts of interest relevant to this study.
+
 This study was supported by grants from the Canadian Diabetes Association (CDA),
@@
 National Diabetes Management and the Ian McWhinney Chair of Family Medicine
-Studies at the University of Western Ontario; BZ holds the Sam and Judy Pencer
-Family Chair in Diabetes Research at Mount Sinai Hospital and University of
-Toronto; RBP holds a Tier II Canada Research Chair in Brain Lipid Metabolism;
-AJH holds a Tier II Canada Research Chair in Diabetes Epidemiology.
+Studies at the University of Western Ontario; RBP holds a Tier II Canada
+Research Chair in Brain Lipid Metabolism; AJH holds a Tier II Canada Research
+Chair in Diabetes Epidemiology.
 
@@
 # Supplemental Material
+
+## Supplemental Material
+
+Generalized estimating equations (GEE) is analogous to mixed effects models,
+except it emphasizes population-level trends over individual-level trends that 
+mixed models use. Because it can handle autocorrelations in the outcome (as 
+would be seen in variables measured over time), GEE modeling is well suited to longitudinal 
+data.
+Latent class mixed models (LCMM) is a form of mixed effects modeling that tries
+to categorize individuals into groups that have similar clusters; in the case of
+the present analysis are the trajectories in beta-cell function over time. Because
+mixed models use the participant as the random effect, LCMM can identify
+individuals who similarly change over time.
+
+Partial least squares (PLS) regression is a technique similar to principal
+component analysis (PCA) that is designed to extract meaningful information from
+multivariate, high dimensionality data (e.g. as in metabolomic and other -omic type 
+analyses). These types of methods try to extract as much of the variance in the 
+data into a smaller number of components or factors. The difference between PCA 
+and PLS is that PLS is a supervised method, while PCA is not. This means that
+PLS uses a response variable(s), i.e. an outcome or **Y**, to describe the
+variation in the predictors (**X**) while PCA only describes the variation 
+inherent in **X**. Because of this, PLS can be better at predicting the 
+contributions of predictor variables against an outcome variable. Partial least
+squares discriminant analysis (PLS-DA) is a form of PLS, except the **Y** is 
+a discrete (i.e. categorical) variable and not a continuous one. Therefore, 
+PLS-DA is able to predict who will belong to which group based on the values of
+the predictors.
+
+The two plots generated from the PLS-DA analysis show the results from both
+the response (i.e. LCMM beta-cell function groups and dyslgycemia status) and
+the predictors (all 22 NEFA). The response result plots as seen in 
+`r cite_sf('plsda_groups')` and `r cite_sf('plsda_groups_dysgly')` show the
+discriminatory ability of the PLS-DA model. When the predictors represent an
+underlying pattern that may exist between groups, the plots will show greater
+separation between the groups in addition to a low misclassification rate. The
+predictor result plots as seen in `r cite_sf('plsda_loadings')` and
+`r cite_sf('plsda_loadings_dysgly')` show which of the predictor variables
+contribute the most to the explained variance. Variables between the solid and 
+dashed lines explain between 50-100% of the explained variance. Whether the 
+variables have positive or negative values on either of the components axes 
+determines how they contribute to each group. For instance, if those individuals
+with dysglycemia have a positive value for component 1 and the fatty acid 16:0 also
+has a positive value for component 1, this indicates that 16:0 is predictive of
+having dysglycemia.
+
+## Supplemental Data
 
@@
 M1 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex)
-M2 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT)
-M3 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET)
-M4 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk) # Best
-M5 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab) # Best
-M6 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse) # Best
-M6.1 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse + TotalNE:VN) # Best
+M2 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT) # best
+M3 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET) # best
+M4 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk) # best
+M5 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab) # best
+M6 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse) # best
+M7 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TobaccoUse) # best
+M8 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab) # best
+M9 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TotalNE:VN)
 
-qic_is_tab <- MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M6.1, rank = QIC) %>% 
+qic_is_tab <- MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M9, rank = QIC) %>% 
     dplyr::add_rownames() %>%
@@
 M8 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab) # best
-M8.1 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TotalNE:VN) # best
+M9 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TotalNE:VN) # best
 
-qic_bcf_tab <- MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M8.1, rank = QIC) %>% 
+qic_bcf_tab <- MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M9, rank = QIC) %>% 
     dplyr::add_rownames() %>%
     dplyr::select(Model = rowname, QIC, Delta = delta) %>% 
-    dplyr::mutate(id = seq_len(nrow(.))+8) %>%
-    dplyr::bind_rows(., data.frame(Model = '**log(ISSI-2)**', id = 8.5))
+    dplyr::mutate(id = seq_len(nrow(.))+10) %>%
+    dplyr::bind_rows(., data.frame(Model = '**log(ISSI-2)**', id = 10.5))
 
@@
 as including both the original age and the time variable would result in
-collinearity.  Column names: QIC is the quasi-likelihood information criteria
+collinearity.  Column names are: QIC is the quasi-likelihood information criteria
 (smaller values, eg. larger negative values, indicate a better fit compared to
@@
 - M6: M5 + smoking status
-- M6.1: M6 + time by NEFA interaction
-- M7: M2 + family history of diabetes + smoking status
-- M8: M2 + family history of diabetes
-- M8.1: M8 + time by NEFA interaction
+- M7: M2 + ALT + family history of diabetes + smoking status
+- M8: M2 + ALT + family history of diabetes
+- M9: M8 + time by NEFA interaction
 

### 2016-06-28 R/functions.R

@@ -327,7 +327,7 @@ plot_heatmap <- function(data, x = c(outcomes, 'BMI', 'Waist', 'Age', 'lALT',
         dplyr::mutate(Vars2 = factor(Vars2, unique(Vars2)),
                       Vars1 = factor(Vars1, unique(Vars1)),
                       Correlations = round(Correlations, 2)) %>%
-        seer::visualize_corr_heatmap(
+        seer::view_heatmap(
             y = 'Vars2',
             x = 'Vars1',
             ylab = 'Non-esterified fatty acids (nmol/mL)',

### 2016-06-28 R/setup.R

@@ -56,7 +56,7 @@ run_setup <- function() {
     # For the document (knitr)
     knitr::opts_chunk$set(
         warning = FALSE, message = FALSE, collapse = TRUE,
-        fig.showtext = TRUE
+        fig.showtext = TRUE, echo = FALSE, dev = c('png', 'postscript')
     )
 }
 

### 2016-06-29 R/functions.R
Added arg to plot gee to remove legend

@@ -228,7 +228,7 @@ plot_lcmm_results <- function(results.lcmm) {
         graph_theme()
 }
 
-plot_gee_results <- function(results.gee) {
+plot_gee_results <- function(results.gee, leg = 'bottom') {
     results.gee %>%
         seer::view_main_effect(
             graph.options = 'dot.size',
@@ -237,7 +237,7 @@ plot_gee_results <- function(results.gee) {
             xlab = 'Percent difference with 95% CI in the outcomes\nfor each SD increase in fatty acid',
             ylab = 'Non-esterified fatty acids'
             ) +
-        graph_theme(ticks = FALSE)
+        graph_theme(ticks = FALSE, legend.pos = leg)
 }
 
 plot_nefa_distribution <- function(data) {

### 2016-06-29 R/setup.R
Changed dev to cairo_ps instead of png/postscript

@@ -56,7 +56,7 @@ run_setup <- function() {
     # For the document (knitr)
     knitr::opts_chunk$set(
         warning = FALSE, message = FALSE, collapse = TRUE,
-        fig.showtext = TRUE, echo = FALSE, dev = c('png', 'postscript')
+        fig.showtext = TRUE, echo = FALSE, dev = c('cairo_ps')
     )
 }
 

### 2016-06-29 doc/manuscript.Rmd
Updated manuscript to reflect DC reqs plus Tony's edits

@@
 
-Word count: 3543 / 4000
+Word count: 3602 / 4000
 
@@
 Research Design and Methods: In the PROMISE cohort, 477 participants had serum
-NEFA measured at the baseline visit and completed an OGTT at 3 time
-points over 6 years. Outcome variables were calculated using the OGTT values.
-IS was assessed using 1/HOMA-IR and the Matsuda index, while beta-cell function
-was assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin
+NEFA measured at the baseline visit and completed an OGTT at 3 time points over
+6 years. Outcome variables were calculated using the OGTT values. IS was
+assessed using 1/HOMA-IR and the Matsuda index, while beta-cell function was
+assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin
 Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were
@@
 
-Results: IS and beta-cell function declined by 14-27% while BMI
-did not significantly change over 6 years of follow-up. Total NEFA, 16:0,
-18:1n-9, and 18:2n-6 as concentrations had significant negative associations
-with IGI/IR and ISSI-2, predicting between 4-8% lower IGI/IR and ISSI-2 over the
-6 years. However, individual NEFA modeled as mol% showed no associations with
-these outcomes. Total and individual NEFA were not associated with IS measures.
+Results: IS and beta-cell function declined by 14-27% while BMI did not
+significantly change over 6 years of follow-up. Total NEFA, 16:0, 18:1n-9, and
+18:2n-6 as concentrations had significant negative associations with IGI/IR and
+ISSI-2, predicting between 4-8% lower IGI/IR and ISSI-2 over the 6 years.
+However, individual NEFA modeled as mol% showed no associations with these
+outcomes. Total and individual NEFA were not associated with IS measures.
 
@@
 Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
-P-values were adjusted for the BH false discovery rate, presented as the dot size.')
+P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association.')
 
@@
 Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
-P-values were adjusted for the BH false discovery rate, presented as the dot size.')
+P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association (*).')
 stab_gee <- stab('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
@@
 Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
-P-values were adjusted for the BH false discovery rate, with significant (p<0.05) associations indicated by asterisk.')
+P-values were adjusted for the BH false discovery rate, with significant (p<0.05) associations indicated by asterisk (*).')
 stab_gee_unadj <- stab('gee_unadj', 'Unadjusted generalized estimating equation models of the longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
@@
 
-# Subjects and Methods
+# Research Design and Methods
 
@@
 ISI, which showed that males had a 12% higher ISI for every SD increase in
-20:3n-6. Raw values from the GEE models are shown in `r cite_st('gee')` for
+20:3n-6 compared to females. Raw values from the GEE models are shown in `r cite_st('gee')` for
 fully adjusted models and `r cite_st('gee_unadj')` for unadjusted models.
@@
 `r cite_sf('plsda_loadings_dysgly')` for the loadings plot of the NEFA composition
-by extracted PLS-DA components).
+by extracted PLS-DA components). The results from the PLS-DA analysis are consistent
+with what the associations seen in the GEE model.
 
-# Discussion
+# Conclusions
 
 In a Canadian population of adults who are at-risk for diabetes, we found that 
-higher total NEFA concentrations independently predicted
-lower beta-cell function after 6 years. While we found negative associations with
-palmitic acid (16:0), oleic acid (18:1n-9), and
-linoleic acid (18:2n-6) for the concentration (nmol/mL) modeling, no
-associations were seen for the proportion of these fatty acids and further
-clustering analysis found no predictive ability of these individual NEFA on
-beta-cell function or dysglycemia. These observations suggest that the absolute
-size of the total NEFA fraction, rather than its specific composition, likely
-influences the pathogenesis of diabetes.
+higher total NEFA concentrations independently predicted lower beta-cell
+function after 6 years. While we found negative associations with palmitic acid
+(16:0), oleic acid (18:1n-9), and linoleic acid (18:2n-6) for the concentration
+(nmol/mL) modeling, no associations were seen for the proportion of these fatty
+acids and further confirmed by clustering analysis found no predictive ability of these
+individual NEFA on beta-cell function or dysglycemia. These observations suggest
+that the absolute size of the total NEFA fraction, rather than its specific
+composition, likely influences the pathogenesis of diabetes, at least within a
+population at-risk for diabetes.
 
@@
 analysis, we saw no association of any individual or total NEFA with hepatic
-(1/HOMA-IR) or whole-body insulin sensitivity (ISI). We identified that
-adjustment for WC was likely the cause for the attenuation in results between
+(1/HOMA-IR) or whole-body insulin sensitivity (ISI). We identified
+WC as having a powerful attenuating effect on the associations between
 unadjusted and adjusted models. There are some possible explanations for our
 null findings for the insulin sensitivity measures and the influence of WC on
-the results. First, there may be differences in physiology between fasting and 
-postprandial NEFA kinetics that we are not able to investigate but that may 
-explain our null findings for fasting NEFA and insulin sensitivity. For
-instance, some experimental studies using clamp protocols found that fasting
-NEFA was a weak predictor of insulin sensitivity compared to postprandial
-concentrations of NEFA [@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA
-uptake into the adipose tissue (e.g. as seen in obesity and larger WC) following
-postprandial TAG lipolysis via lipoprotein lipase may result in NEFA spillover
-into the blood and a subsequent increase in circulating NEFA [@Almandoz2013a],
-which may be more metabolically active given postprandial activity. Second, the
-null findings may be due to the high risk population examined in PROMISE,
-including having a higher WC. It may be that in this population, IR has become
-well established and NEFA may not contribute to IR at this somewhat more
-advanced stage in the pathogenesis of diabetes.
+the results. First, WC may be a strong causal link in previously reported
+associations between NEFA and insulin sensitivity, given the role adipocytes
+play in metabolism (e.g. leptin, adiponectin). Second, there may be differences
+in physiology between fasting and postprandial NEFA kinetics that we are not
+able to investigate but that may explain our null findings for fasting NEFA and
+insulin sensitivity. For instance, some experimental studies using clamp
+protocols found that fasting NEFA was a weak predictor of insulin sensitivity
+compared to postprandial concentrations of NEFA
+[@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the adipose
+tissue (e.g. as seen in obesity and larger WC) following postprandial TAG
+lipolysis via lipoprotein lipase may result in NEFA spillover into the blood and
+a subsequent increase in circulating NEFA [@Almandoz2013a], which may be more
+metabolically active given postprandial activity. Third, the null findings seen
+may be due to the high risk population examined in PROMISE, including having a 
+high WC and BMI. It may be that in this population, IR has become well
+established and NEFA may not contribute to IR at this somewhat more advanced
+stage in the pathogenesis of diabetes.
 
@@
 NEFA composition and components of the metabolic syndrome (MetS) [@Dai2015a]. 
-Specifically, the authors identified NEFA
-16:1n-9, 20:1n-9, and 22:4n-6 to correlate with components of the MetS. 
-Another similar study examining diabetes found that 16:0, 18:0, 18:1, 18:2, 18:3
-may be useful biomarkers for identifying healthy compared to diabetic
-individuals [@Liu2010a]. However, both studies were limited by
-smaller sample sizes (approximately 100 subjects) and the cross-sectional
-design.
+Specifically, the authors identified NEFA 16:1n-9, 20:1n-9, and 22:4n-6 to
+correlate with components of the MetS. Another similar study examining diabetes
+found that 16:0, 18:0, 18:1, 18:2, 18:3 may be useful biomarkers for identifying
+healthy compared to diabetic individuals [@Liu2010a]. However, both studies were
+limited by smaller sample sizes (approximately 100 subjects) and the
+cross-sectional design.
 
@@
 feedback on all versions of the paper; LWJ and AJH had primary responsibility
-for final content. All authors read and approved the final manuscript.  The
-authors report no potential conflicts of interest relevant to this study.
+for final content. All authors read and approved the final manuscript. AJH is
+the guarantor of this work and, as such, had full access to all the data in the
+study and takes responsibility for the integrity of the data and the accuracy of
+the data analysis.
 
@@
 Research Chair in Brain Lipid Metabolism; AJH holds a Tier II Canada Research
-Chair in Diabetes Epidemiology.
+Chair in Diabetes Epidemiology. The authors report no potential conflicts of
+interest relevant to this study.
+
+Portions of the results in the present analysis have been presented in
+abstract form at the following conferences: International Symposium on Diabetes
+& Nutrition (2015); Biennial Congress of the International Society for the Study
+of Fatty Acids and Lipids (2014); International Diabetes Federation World
+Diabetes Congress (2015); American Diabetes Association Scientific Sessions
+(2014, 2015).
 
@@
 ```{r figGEE, dependson='geeData', fig.cap=fig_gee}
-plot_gee_results(gee_results)
+plot_gee_results(gee_results, leg = 'none')
 ```
@@
 
-## Supplemental Material
+## Supplemental Methods
 
@@
 mixed models use. Because it can handle autocorrelations in the outcome (as 
-would be seen in variables measured over time), GEE modeling is well suited to longitudinal 
-data.
-Latent class mixed models (LCMM) is a form of mixed effects modeling that tries
-to categorize individuals into groups that have similar clusters; in the case of
-the present analysis are the trajectories in beta-cell function over time. Because
-mixed models use the participant as the random effect, LCMM can identify
-individuals who similarly change over time.
+would be seen in variables measured over time), GEE modeling is well suited to
+longitudinal data. Latent class mixed models (LCMM) is a form of mixed effects
+modeling that tries to categorize individuals into groups that have similar
+clusters; in the case of the present analysis are the trajectories in beta-cell
+function over time. Because mixed models use the participant as the random
+effect, LCMM can identify individuals who similarly change over time.
 
@@
 `r cite_sf('plsda_groups')` and `r cite_sf('plsda_groups_dysgly')` show the
-discriminatory ability of the PLS-DA model. When the predictors represent an
+discriminatory ability of the PLS-DA model, with `r cite_sf('plsda_example')` 
+showing an example of good discriminatory ability. When the predictors represent an
 underlying pattern that may exist between groups, the plots will show greater
@@
 
-## Supplemental Data
+## Supplemental Tables and Figures
 
@@
 
-```{r supFigPlsLoadings, fig.cap=sfig_plsda_loadings}
+```{r supFigPlsLoadings, fig.cap=sfig_plsda_loadings, fig.width=7, fig.height=7}
 plot_plsda_loadings(plsda_results)
@@
 
-```{r supFigPlsLoadingsDysgly, fig.cap=sfig_plsda_loadings_dysgly}
+```{r supFigPlsLoadingsDysgly, fig.cap=sfig_plsda_loadings_dysgly, fig.width=7, fig.height=7}
 plot_plsda_loadings(plsda_dysgly)

### 2016-07-06 R/functions.R
Fixed changes to mason packages functions

@@ -64,7 +64,7 @@ analyze_gee <- function(data, y, x, covar, unit,
         extract_term <- 'Xterm$'
     }
     gee_results <- data %>%
-        mason::design_analysis('gee') %>%
+        mason::design('gee') %>%
         mason::add_settings(family = gaussian('identity'), corstr = 'ar1', cluster.id = 'SID') %>%
         mason::add_variables('yvars', y) %>%
         mason::add_variables('xvars', x) %>%
@@ -75,7 +75,7 @@ analyze_gee <- function(data, y, x, covar, unit,
                 .
             }
         } %>%
-        mason::construct_analysis() %>%
+        mason::construct() %>%
         mason::scrub() %>%
         mason::polish_filter(extract_term, 'term') %>%
         mason::polish_transform_estimates(function(x) (exp(x) - 1) * 100) %>%
@@ -305,11 +305,11 @@ plot_heatmap <- function(data, x = c(outcomes, 'BMI', 'Waist', 'Age', 'lALT',
                          y = ne_conc) {
     data %>%
         dplyr::filter(VN == 0) %>%
-        mason::design_analysis('cor') %>%
+        mason::design('cor') %>%
         mason::add_settings(method = 'pearson', use = 'complete.obs') %>%
         mason::add_variables('yvars', y) %>%
         mason::add_variables('xvars', x) %>%
-        mason::construct_analysis() %>%
+        mason::construct() %>%
         mason::scrub() %>%
         mason::polish_renaming(renaming_fa, 'Vars2') %>%
         #mason::polish_renaming(renaming_outcomes, 'Vars1') %>%
@@ -417,11 +417,11 @@ calculate_outcomes_pct_change <- function(data) {
         round(0) %>%
         {paste0(min(.), '% to ', max(.), '%')}
 
-    pval <- design_analysis(data, 'gee') %>%
+    pval <- design(data, 'gee') %>%
         add_settings(family = gaussian, corstr = 'ar1', cluster.id = 'SID') %>%
         add_variables('yvars', c('linvHOMA', 'lISI', 'lIGIIR', 'lISSI2')) %>%
         add_variables('xvars', 'VN') %>%
-        construct_analysis() %>%
+        construct() %>%
         scrub() %>%
         polish_filter('Xterm$', 'term') %>%
         dplyr::summarise(p.value = mean(p.value)) %>%

### 2016-07-06 doc/manuscript.Rmd
Made changes to manuscript as per tony's edits + prep for sub

@@
 (3,4), MD; Adria Giacca (5), PhD; Zhen Liu (1), PhD;
-Richard P. Bazinet (1), PhD; and Anthony J. Hanley (1,6), PhD
+Richard P. Bazinet (1), PhD; and Anthony J. Hanley (1,3,6), PhD
 
@@
 
-Word count: 3602 / 4000
+Word count: 3620 / 4000
 
@@
 
-<!-- Reference limit 41 / 40 -->
+<!-- Reference limit 39 / 40 -->
 
 # Abstract
-243 / 250 words
+<!--243 / 250 words-->
 
@@
 fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
-Generalized estimating equation models were adjusted for time, sex, ethnicity, baseline age, WC,
-ALT, and family history of diabetes.
+Generalized estimating equation models were adjusted for time, sex, ethnicity, baseline age, WC, ALT, and family history of diabetes.
 Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
@@
 Red, green, and blue lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis.
-See ', cite_sf('plsda_example'), ' for an example plot showing good discriminatory ability between groups.'))
+See ', cite_sf('plsda_example'), ' for a hypothetical example plot showing good discriminatory ability between groups.'))
 sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', paste0('Clustering of extracted components from the partial least squares discriminant analysis for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).
@@
 Blue lines indicate dysglycemia conversion or maintanence and red lines indicate no dysglycemia status.
-See ', cite_sf('plsda_example'), ' for an example plot showing good discriminatory ability between groups.'))
+See ', cite_sf('plsda_example'), ' for a hypothetical example plot showing good discriminatory ability between groups.'))
 ```
@@
 high in the polyunsaturated fatty acid (PUFA) linoleic acid (18:2n-6), as the
-exposure to characterize the impact of total NEFA. However, fatty acids comprise
+exposure to characterize the metabolic impact of total NEFA. However, fatty acids comprise
 multiple molecules with diverse physiological functions, and few 
 studies have analyzed the effects of a broader spectrum of fatty acids. Notably,
-one study suggests that eicosapentaenoic acid (20:5n-3) can protect against the
+one study suggested that eicosapentaenoic acid (20:5n-3) can protect against the
 lipotoxic effect of palmitic acid in the beta-cells [@Kato2008].
@@
 that higher palmitic acid and lower PUFA such as
-eicosapentaenoic acid would associate with declining insulin sensitivity and
+eicosapentaenoic acid in the NEFA fraction would associate with declining insulin sensitivity and
 beta-cell function over 6 years.
@@
 current study used data on participants who did not have diabetes at baseline,
-who returned for one or more follow-up examinations, and who had samples
+who returned for one or more of the follow-up examinations, and who had samples
 available for fatty acid measurements (n=477). A diagram of the sample size at
@@
 and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR
-is a measure of the first phase insulin secretion while ISSI-2 is analogous to
+is a measure of the first phase of insulin secretion while ISSI-2 is analogous to
 the disposition index (but using OGTT values). Each index has been validated
@@
 interpreted as an expected percent difference in the outcome variable for every
-SD increase in the predictor variable given the covariates are held constant.
+standard deviation (SD) increase in the predictor variable given the covariates are held constant.
 Since TAG is a risk factor for diabetes and since NEFA contribute to TAG
@@
 by extracted PLS-DA components). The results from the PLS-DA analysis are consistent
-with what the associations seen in the GEE model.
+with the associations seen in the GEE models.
 
@@
 (16:0), oleic acid (18:1n-9), and linoleic acid (18:2n-6) for the concentration
-(nmol/mL) modeling, no associations were seen for the proportion of these fatty
-acids and further confirmed by clustering analysis found no predictive ability of these
+(nmol/mL) modeling, no associations were seen for these fatty
+acids as a proportion, an observation that was further confirmed by multivariate
+cluster techniques, which found no predictive ability of these
 individual NEFA on beta-cell function or dysglycemia. These observations suggest
@@
 metabolically active given postprandial activity. Third, the null findings seen
-may be due to the high risk population examined in PROMISE, including having a 
+may be due to the high risk population examined in PROMISE, as the majority of subjects had a 
 high WC and BMI. It may be that in this population, IR has become well
@@
 found that 16:0, 18:0, 18:1, 18:2, 18:3 may be useful biomarkers for identifying
-healthy compared to diabetic individuals [@Liu2010a]. However, both studies were
+healthy compared to individuals with diabetes [@Liu2010a]. However, both studies were
 limited by smaller sample sizes (approximately 100 subjects) and the
@@
 potential covariates were empirically analyzed prior to inclusion into the
-GEE models to best understand and minimizing potential
+GEE models to best understand and minimize potential
 confounding. Finally, our cohort consists of individuals at-risk for diabetes, 
-who are primarily female of European-ancestry and as such our results may not be
+who are primarily female and European-ancestry and as such our results may not be
 generalizable to other populations. However, given these limitations, our study
@@
 
+# Figures
+
+```{r figure1, fig.cap=fig_dist}
+plot_nefa_distribution(ds)
+```
+
+```{r figure2, dependson='geeData', fig.cap=fig_gee}
+plot_gee_results(gee_results, leg = 'none')
+```
+
 # Tables
 
-```{r tabBasic}
+```{r table1}
 table_basic(ds, caption = tab_basic)
@@
 and sex groups did not change over the 6 years.
-
-# Figures
-
-```{r figDist, fig.cap=fig_dist}
-plot_nefa_distribution(ds)
-```
-
-```{r figGEE, dependson='geeData', fig.cap=fig_gee}
-plot_gee_results(gee_results, leg = 'none')
-```
 
@@
 `r cite_sf('plsda_groups')` and `r cite_sf('plsda_groups_dysgly')` show the
-discriminatory ability of the PLS-DA model, with `r cite_sf('plsda_example')` 
-showing an example of good discriminatory ability. When the predictors represent an
+poor discriminatory ability of the PLS-DA model, with `r cite_sf('plsda_example')` 
+showing a hypothetical example of good discriminatory ability. When the predictors represent an
 underlying pattern that may exist between groups, the plots will show greater
@@
 
-```{r supTabQIC}
+```{r onlineSupplementalTableS1}
 prep_qic <- gee_df %>%
@@
 
-```{r supTabDist}
+```{r onlineSupplementalTableS2}
 table_distribution(ds, stab_dist)
@@
 
-```{r supFigHeatmap, fig.cap=sfig_corr}
+```{r onlineSupplementalFigureS2, fig.cap=sfig_corr}
 plot_heatmap(ds)
@@
 
-```{r supFigGeeUnadj, fig.cap=sfig_gee}
+```{r onlineSupplementalFigureS3, fig.cap=sfig_gee}
 plot_gee_results(gee_results_unadj)
@@
 
-```{r supTabGEE}
+```{r onlineSupplementalTableS3_4}
 table_gee(gee_results, stab_gee)
@@
 
-```{r supFigLCMM, fig.cap=sfig_lcmm}
+```{r onlineSupplementalFigureS4, fig.cap=sfig_lcmm}
 plot_lcmm_results(lcmm_data)
@@
 
-```{r supFigPlsGrp, fig.cap=sfig_plsda_groups}
+```{r onlineSupplementalFigureS5, fig.cap=sfig_plsda_groups}
 plot_plsda_grouping(plsda_results, 'Beta-cell function')
@@
 
-```{r supFigPlsLoadings, fig.cap=sfig_plsda_loadings, fig.width=7, fig.height=7}
+```{r onlineSupplementalFigureS6, fig.cap=sfig_plsda_loadings, fig.width=7, fig.height=7}
 plot_plsda_loadings(plsda_results)
@@
 
-```{r supFigPlsGrpDysgly, fig.cap=sfig_plsda_groups_dysgly}
+```{r onlineSupplementalFigureS7, fig.cap=sfig_plsda_groups_dysgly}
 plot_plsda_grouping(plsda_dysgly, 'Dysglycemia')
@@
 
-```{r supFigPlsLoadingsDysgly, fig.cap=sfig_plsda_loadings_dysgly, fig.width=7, fig.height=7}
+```{r onlineSupplementalFigureS8, fig.cap=sfig_plsda_loadings_dysgly, fig.width=7, fig.height=7}
 plot_plsda_loadings(plsda_dysgly)
@@
 
-```{r supFigPlsExample, fig.cap=sfig_plsda_example}
+```{r onlineSupplementalFigureS9, fig.cap=sfig_plsda_example}
 plot_plsda_grouping(example_plsda_results(), 'Example groups')

### 2016-07-06 doc/gee.Rmd
First commit for submission code

--- 2016-01-27 doc/gee.Rmd
@@
     summarize(n = n())
+
+gee_table <- function(data, caption, p.val = 0.05) {
+    data %>%
+        filter(p.value < p.val) %>%
+        select(Yterms, Xterms, term, unit, estimate, p.value) %>%
+        mutate_each(funs(round(., 2)), estimate, p.value) %>%
+        arrange(unit, Yterms) %>% 
+        pander(caption)
+}
 ```
@@
 outcomes_bcf <- c('lIGIIR', 'lISSI2')
-ne_pct <- grep('^pct_ne', names(ds), value = TRUE)
-ne_conc <- c('ne_Total', grep('^ne\\d\\d', names(ds), value = TRUE))
+ne_pct <- grep('^pct_ne', names(gee_df), value = TRUE)
+ne_conc <- c('TotalNE', grep('^ne\\d\\d', names(gee_df), value = TRUE))
 covariates0 <- c('VN')
-covariates_is <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT')
-covariates_bcf <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 
-                    'MET', 'AlcoholPerWk')
-covariates_is_tag <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'TAG')
-covariates_bcf_tag <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 
-                        'ALT', 'MET', 'AlcoholPerWk', 'TAG')
-
-gee_df <- get_gee_data(ds)
+covariates_bcf <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'FamHistDiab')
+covariates <- covariates_is <- covariates_bcf
+covariates_is_tag <- c(covariates_is, 'BaseTAG')
+covariates_bcf_tag <- c(covariates_bcf, 'BaseTAG')
 ```
@@
 ```{r geeResults_Unadj}
-gee_results0 <- bind_rows(
-    analyze_gee(gee_df, outcomes, ne_pct, covariates0, 'mol%'),
-    analyze_gee(gee_df, outcomes, ne_conc, covariates0, 'nmol/mL')
-    )
-gee_results0 %>% 
-    mutate(p.value = p.adjust(p.value, 'BH')) %>% 
-    filter(p.value < 0.1) %>% 
-    select(Yterms, Xterms, unit, estimate, p.value) %>% 
-    mutate_each(funs(round(., 2)), estimate, p.value) %>% 
-    pander('Unadjusted')
+gee_results0 <- dplyr::bind_rows(
+    analyze_gee(gee_df, outcomes, ne_pct, covariates0, 'mol%', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes, ne_conc, covariates0, 'nmol/mL', adj.p = TRUE)
+    ) %>% 
+    tidy_gee_results()
+gee_table(gee_results0, 'Unadjusted')
 ```
@@
 gee_results_is <- bind_rows(
-    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%'),
-    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL'),
-    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is_tag, 'mol% (w/ tag)'),
-    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is_tag, 'nmol/mL (w/ tag)')
-    )
+    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is_tag, 'mol%', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is_tag, 'nmol/mL', adj.p = TRUE)
+    ) %>% 
+    tidy_gee_results()
+
+gee_table(gee_results_is, 'Without or with TAG for IS')
 gee_results_is %>% 
-    group_by(unit) %>% 
-    mutate(p.value = p.adjust(p.value, 'BH')) %>% 
-    ungroup() %>% 
-    filter(p.value < 0.1) %>% 
-    select(Yterms, Xterms, unit, estimate, p.value) %>% 
-    mutate_each(funs(round(., 2)), estimate, p.value) %>% 
-    pander('Without or with TAG for IS')
+    select(Yterms, Xterms, sample.total) %>% 
+    group_by(Yterms) %>% 
+    summarise(MinSample = min(sample.total),
+              MaxSample = max(sample.total)) %>% 
+    pander('Sample size for IS')
 ```
@@
 gee_results_bcf <- bind_rows(
-    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%'),
-    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL'),
-    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf_tag, 'mol% (w/ tag)'),
-    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf_tag, 'nmol/mL (w/ tag)')
-    )
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf_tag, 'mol% (w/ tag)', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf_tag, 'nmol/mL (w/ tag)', adj.p = TRUE)
+    ) %>% 
+    tidy_gee_results()
+
+gee_table(gee_results_bcf, 'Without or with TAG for BCF')
 gee_results_bcf %>% 
-    group_by(unit) %>% 
-    mutate(p.value = p.adjust(p.value, 'BH')) %>% 
-    ungroup() %>% 
-    filter(p.value < 0.05) %>% 
-    select(Yterms, Xterms, unit, estimate, p.value) %>% 
-    mutate_each(funs(round(., 2)), estimate, p.value) %>% 
-    pander('Without or with TAG for BCF')
+    select(Yterms, Xterms, sample.total) %>% 
+    group_by(Yterms) %>% 
+    summarise(MinSample = min(sample.total),
+              MaxSample = max(sample.total)) %>% 
+    pander('Sample size for BCF')
+```
+
+# Adjusting using a single model for both IS and BCF (cancelled)
+
+Decided on using only one model. Not running the below code but keeping it for 
+now for record-keeping.
+
+```{r geeResults_SingleModel, eval = FALSE}
+gee_results <- bind_rows(
+    analyze_gee(gee_df, outcomes, ne_pct, covariates_is, 'mol% (IS model)', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes, ne_conc, covariates_is, 'nmol/mL (IS model)', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes, ne_pct, covariates_bcf, 'mol% (BCF model)', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes, ne_conc, covariates_bcf, 'nmol/mL (BCF model)', adj.p = TRUE)
+    ) %>% 
+    tidy_gee_results()
+
+gee_table(gee_results, 'Comparing IS vs BCF models.')
+```
+
+# Identifying which variable changes association for IS
+
+```{r geeResults_VariableInfluence}
+covar_base <- c('VN', 'Sex', 'Ethnicity', 'BaseAge')
+covar_wc <- c(covar_base, 'Waist')
+covar_alt <- c(covar_base, 'ALT')
+covar_fh <- c(covar_base, 'FamHistDiab')
+gee_results <- bind_rows(
+    analyze_gee(gee_df, outcomes_is, ne_pct, covar_base, 'mol% (base)', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covar_base, 'nmol/mL (base)', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_pct, covar_wc, 'mol% (wc)', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covar_wc, 'nmol/mL (wc)', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_pct, covar_fh, 'mol% (fh)', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covar_fh, 'nmol/mL (fh)', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_pct, covar_alt, 'mol% (alt)', adj.p = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covar_alt, 'nmol/mL (alt)', adj.p = TRUE)
+    ) %>% 
+    tidy_gee_results()
+
+gee_table(gee_results, 'Covariate influencing results.')
+```
+
+# Grouping NEFA into common classes (SFA, MUFA, PUFA)
+
+```{r geeResults_FAGroupings}
+sfa <- grep('0$', c(ne_conc), value = TRUE)
+mufa <- grep('1n.*$', c(ne_conc), value = TRUE)
+pufa <- grep('[2-6]n.*$', c(ne_conc), value = TRUE)
+n3pufa <- grep('[2-6]n3$', c(ne_conc), value = TRUE)
+n6pufa <- grep('[2-6]n6$', c(ne_conc), value = TRUE)
+
+gee_df2 <- ds %>% 
+    mutate(
+        SFA = rowSums(.[sfa]),
+        pct_SFA = (SFA / TotalNE) * 100,
+        MUFA = rowSums(.[mufa]),
+        pct_MUFA = (MUFA / TotalNE) * 100,
+        PUFA = rowSums(.[pufa]),
+        pct_PUFA = (PUFA / TotalNE) * 100,
+        n3PUFA = rowSums(.[n3pufa]),
+        pct_n3PUFA = (n3PUFA / TotalNE) * 100,
+        n6PUFA = rowSums(.[n6pufa]),
+        pct_n6PUFA = (n6PUFA / TotalNE) * 100
+        ) %>% 
+    get_gee_data(fa = '(^|^pct_).*(SFA|MUFA|PUFA)')
+
+gee_results <- bind_rows(
+    analyze_gee(gee_df2, outcomes, c('SFA', 'MUFA', 'PUFA', 'n3PUFA', 'n6PUFA'), 
+                covariates, 'nmol/mL', adj.p = TRUE), 
+    analyze_gee(gee_df2, outcomes, c('pct_SFA', 'pct_MUFA', 'pct_PUFA', 
+                                     'pct_n3PUFA', 'pct_n6PUFA'), 
+                covariates, 'mol%', adj.p = TRUE)
+    ) %>% 
+    tidy_gee_results()
+
+gee_table(gee_results, 'Results using fatty acid classes (saturated, etc).')
 ```
@@
 gee_results_is_int <- bind_rows(
-    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', int = TRUE),
-    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', int = TRUE),
-    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is_tag, 'mol% (w/ tag)', int = TRUE),
-    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is_tag, 'nmol/mL (w/ tag)', int = TRUE)
-    )
-gee_results_is_int %>%
-    group_by(unit) %>% 
-    mutate(p.value = p.adjust(p.value, 'BH')) %>% 
-    ungroup() %>% 
-    select(p.value) %>% 
-    plot()
+    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', adj.p = TRUE, int = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', adj.p = TRUE, int = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is_tag, 'mol% (w/ tag)', adj.p = TRUE, int = TRUE),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is_tag, 'nmol/mL (w/ tag)', adj.p = TRUE, int = TRUE)
+    ) %>% 
+    tidy_gee_results()
+gee_table(gee_results_is_int, 'Interaction for time for IS')
+```
+
+```{r geeResults_IS_int_sex}
+gee_results_is_int_sex <- bind_rows(
+    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', adj.p = TRUE, int = TRUE, intvar = 'Sex'),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', adj.p = TRUE, int = TRUE, intvar = 'Sex'),
+    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is_tag, 'mol% (w/ tag)', adj.p = TRUE, int = TRUE, intvar = 'Sex'),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is_tag, 'nmol/mL (w/ tag)', adj.p = TRUE, int = TRUE, intvar = 'Sex')
+    ) %>% 
+    tidy_gee_results()
+
+gee_table(gee_results_is_int_sex, 'Interactions between male and female for IS.')
+```
+
+```{r geeResults_IS_int_ethn}
+gee_results_is_int_ethn <- bind_rows(
+    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is, 'mol%', 
+                adj.p = TRUE, int = TRUE, intvar = 'Ethnicity'),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is, 'nmol/mL', 
+                adj.p = TRUE, int = TRUE, intvar = 'Ethnicity'),
+    analyze_gee(gee_df, outcomes_is, ne_pct, covariates_is_tag, 'mol% (w/ tag)', 
+                adj.p = TRUE, int = TRUE, intvar = 'Ethnicity'),
+    analyze_gee(gee_df, outcomes_is, ne_conc, covariates_is_tag, 'nmol/mL (w/ tag)', 
+                adj.p = TRUE, int = TRUE, intvar = 'Ethnicity')
+    ) %>% 
+    tidy_gee_results()
+
+gee_table(gee_results_is_int_ethn, 'Interactions between ethnicities for IS.')
 ```
@@
 gee_results_bcf_int <- bind_rows(
-    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', int = TRUE),
-    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', int = TRUE),
-    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf_tag, 'mol% (w/ tag)', int = TRUE),
-    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf_tag, 'nmol/mL (w/ tag)', int = TRUE)
-    )
-gee_results_bcf_int %>% 
-    group_by(unit) %>% 
-    mutate(p.value = p.adjust(p.value, 'BH')) %>% 
-    ungroup() %>% 
-    select(p.value) %>% 
-    plot()
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', adj.p = TRUE, int = TRUE),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', adj.p = TRUE, int = TRUE),
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf_tag, 'mol% (w/ tag)', adj.p = TRUE, int = TRUE),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf_tag, 'nmol/mL (w/ tag)', adj.p = TRUE, int = TRUE)
+    ) %>% 
+    tidy_gee_results()
+
+gee_table(gee_results_bcf_int, 'Interactions for time for BCF.')
 ```
+
+```{r geeResults_BCF_int_sex}
+gee_results_bcf_int_sex <- bind_rows(
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', 
+                adj.p = TRUE, int = TRUE, intvar = 'Sex'),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', 
+                adj.p = TRUE, int = TRUE, intvar = 'Sex'),
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf_tag, 'mol% (w/ tag)', 
+                adj.p = TRUE, int = TRUE, intvar = 'Sex'),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf_tag, 'nmol/mL (w/ tag)', 
+                adj.p = TRUE, int = TRUE, intvar = 'Sex')
+    ) %>% 
+    tidy_gee_results()
+
+gee_table(gee_results_bcf_int_sex, 'Interactions between male and female for BCF.')
+```
+
+```{r geeResults_BCF_int_ethn}
+gee_results_bcf_int_ethn <- bind_rows(
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf, 'mol%', 
+                adj.p = TRUE, int = TRUE, intvar = 'Ethnicity'),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf, 'nmol/mL', 
+                adj.p = TRUE, int = TRUE, intvar = 'Ethnicity'),
+    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates_bcf_tag, 'mol% (w/ tag)', 
+                adj.p = TRUE, int = TRUE, intvar = 'Ethnicity'),
+    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates_bcf_tag, 'nmol/mL (w/ tag)', 
+                adj.p = TRUE, int = TRUE, intvar = 'Ethnicity')
+    ) %>% 
+    tidy_gee_results()
+
+gee_table(gee_results_bcf_int_ethn, 'Interactions between ethnicities for BCF.')
+```

### 2016-07-06 R/fetch_data.R

@@ -1,0 +1,115 @@
+#' Fetch data from the master dataset.
+#'
+#' This function fetchs the main dataset, keeps variables relevant to
+#' the analysis, restrict the sample size as needed, and lastly save
+#' the new dataset as an `.RData` file. The dot in front of the function hides
+#' it from the global environment.
+#'
+.fetch_data <- function() {
+    # Load the master dataset,
+    ds.prep <- PROMISE::PROMISE_data %>%
+        filter(VN %in% c(1, 3, 6)) %>%
+        ## Kick out Canoers
+        filter(is.na(Canoe)) %>%
+        tbl_df()
+
+    print(paste0('Original dataset rows are ', dim(ds.prep)[1], ' and columns are ', dim(ds.prep)[2]))
+
+    ##' Munge and wrangle the data into the final version.
+    ds <-
+        full_join(ds.prep,
+                  ds.prep %>%
+                      filter(VN == 1) %>%
+                      select(SID, BaseAge = Age, BaseTAG = TAG)) %>%
+        select(
+            SID, VN, BMI, Waist, HOMA, ISI, IGIIR, ISSI2, TAG, LDL, HDL, Chol,
+            ALT, CRP, FamHistDiab, matches('meds'), Age, Sex, Ethnicity,
+            IFG, IGT, DM, MET, BaseAge, AlcoholPerWk, TobaccoUse, SelfEdu, Occupation,
+            TotalNE, matches('^ne\\d+'), Glucose0, Glucose120, BaseTAG
+        ) %>%
+        mutate(
+            BaseTotalNE = TotalNE,
+            FamHistDiab =
+                plyr::mapvalues(FamHistDiab, c(0, 1:12),
+                                c('No', rep('Yes', 12))) %>%
+                as.factor(),
+            invHOMA = (1 / HOMA),
+            linvHOMA = log(invHOMA),
+            lISI = log(ISI),
+            lIGIIR = log(IGIIR),
+            lISSI2 = log(ISSI2),
+            lALT = log(ALT),
+            lTAG = log(TAG),
+            MedsLipidsChol = ifelse(is.na(MedsLipidsChol), 0, MedsLipidsChol)
+        ) %>%
+        arrange(SID, VN) %>%
+        group_by(SID) %>%
+        fill(TotalNE, matches('^ne\\d+')) %>%
+        ungroup()
+
+    ds <- ds %>%
+        full_join(ds %>%
+                      filter(VN == 1) %>%
+                      select(SID, TotalNE, matches('^ne\\d+')) %>%
[…]

### 2016-07-06 R/functions.R

@@ -1,0 +1,648 @@
+##'
+##' Functions
+##' =========
+##'
+##' Custom functions used for analyses.
+##'
+
+# Renaming ----------------------------------------------------------------
+
+renaming_table_rows <- function(x) {
+    x %>%
+        gsub('IGIIR', 'IGI/IR', .) %>%
+        gsub('ISSI2', 'ISSI-2', .) %>%
+        gsub('HOMA', 'HOMA-IR', .) %>%
+        gsub('TAG', 'TAG (mmol/L)', .) %>%
+        gsub('Chol', 'Chol (mmol/L)', .) %>%
+        gsub('LDL', 'LDL (mmol/L)', .) %>%
+        gsub('HDL', 'HDL (mmol/L)', .) %>%
+        gsub('BaseTotalNE', 'NEFA (nmol/mL)', .) %>%
+        gsub('Age', 'Age (yrs)', .) %>%
+        gsub('BMI', 'BMI (kg/m^2^)', .) %>%
+        gsub('Waist', 'WC (cm)', .)
+}
+
+renaming_outcomes <- function(x) {
+    x %>%
+        gsub('linvHOMA', 'log(1/HOMA-IR)', .) %>%
+        gsub('lISI', 'log(ISI)', .) %>%
+        gsub('lIGIIR', 'log(IGI/IR)', .) %>%
+        gsub('lISSI2', 'log(ISSI-2)', .) %>%
+        gsub('^invHOMA$', '1/HOMA-IR', .) %>%
+        gsub('^ISSI2$', 'ISSI-2', .) %>%
+        gsub('^IGIIR$', 'IGI/IR', .)
+}
+
+renaming_fa <- function(x) {
+    x %>%
+        gsub('.*(\\d\\d)(\\d)', '\\1:\\2', .) %>%
+        gsub('n(\\d)$', 'n-\\1', .) %>%
+        gsub('D(\\d\\d)$', 'D-\\1', .) %>%
+        gsub('^pct_', '', .) %>%
+        gsub('TotalNE', 'Total', .)
+}
+
+renaming_fraction <- function(x) {
+    x %>%
+        gsub('ne', 'Non-esterified', .)
+}
+
+renaming_list <- function(x) {
+    x %>%
+        renaming_fa() %>%
+        renaming_outcomes()
+}
+
+
+# Analyze -----------------------------------------------------------------
+
+analyze_gee <- function(data, y, x, covar, unit,
[…]

### 2016-07-06 R/load_data.R

@@ -1,0 +1,42 @@
+# ---(LIKELY) NO NEED TO EDIT---
+#
+# `load_data()` either loads the dataset inside the `data/` folder or, if the
+# `fetch_data.R` file has been changed, will re-run the `fetch_data.R` function
+# to update the dataset in `data/` and then loads the dataset.
+#
+# There are three arguments to this function:
+#
+# - `local.dataset`: Set the file path to the local dataset (eg. `data/`).
+# - `fetch.script`: Set the file path to the `fetch_data.R` function.
+# - `force`: Force updating the dataset in `data/` by running the `fetch_data.R`
+# function.
+#
+# Examples for usage:
+#
+#   ds <- load_data()
+#   ds <- load_data(
+#       'data/dsname.RData',
+#       'R/fetchScript.R')
+#
+load_data <- function(local.dataset = file.path('data', 'data.Rds'),
+                      fetch.script = file.path('R', 'fetch_data.R'),
+                      force = FALSE) {
+    source('.Rprofile')
+    if (force) {
+        .fetch_data()
+    }
+    if (!file.exists(fetch.script))
+        stop('Please create a file under the R folder that contains a function called fetch_data.')
+    if (!file.exists(local.dataset)) {
+        dir.create('data')
+        .fetch_data()
+        message('A dataset has been added and loaded.')
+    } else if (file.mtime(fetch.script) > file.mtime(local.dataset)) {
+        .fetch_data()
+        message('Dataset has been updated and loaded.')
+    } else {
+        message('Dataset is up-to-date.')
+    }
+
+    readRDS(local.dataset)
+}

### 2016-07-06 R/setup.R

@@ -1,0 +1,62 @@
+#' Code preamble
+#' =============
+#'
+#' Install packages if they are not installed and load them. Set options for
+#' the packages.
+#'
+run_setup <- function() {
+    # Run the following packages for setup
+    .load_packages()
+    .set_options()
+    .load_packages()
+    .load_packages()
+}
+
+.load_packages <- function() {
+    # May need to run this function another time if a package needs to be
+    # installed.
+    #
+    # Add your packages you want to use here, using the following format:
+    # - For pre-bundled packages
+
+    # - For CRAN packages:
+    # if (!require(package)) install.packages('package')
+    if (!require(lcmm)) install.packages('lcmm')
+    if (!require(caret)) install.packages('caret')
+    if (!require(fdrtool)) install.packages('fdrtool')
+    if (!require(captioner)) install.packages('captioner')
+    if (!require(tidyr)) install.packages('tidyr')
+    if (!require(ggplot2)) install.packages('ggplot2')
+    if (!require(pander)) install.packages('pander')
+    if (!require(knitr)) install.packages('knitr')
+    if (!require(DiagrammeR)) install.packages('DiagrammeR')
+    if (!require(MuMIn)) install.packages('MuMIn')
+    if (!require(geepack)) install.packages('geepack')
+    if (!require(magrittr)) install.packages('magrittr')
+    if (!require(dplyr)) install.packages('dplyr')
+    # - For github packages:
+    # if (!require(package)) packrat::install_github('username/package')
+
+    # How to manage local? Not sure.
+    library(carpenter)
+    library(mason)
+    library(seer)
+}
+
+.set_options <- function() {
+    # Set the options here for individual packages
+
+    # For tables (pander)
+    panderOptions('table.split.table', Inf)
+    panderOptions('table.style', 'rmarkdown')
+    panderOptions('table.alignment.default',
+                  function(df)
+                      ifelse(sapply(df, is.numeric), 'center', 'left'))
+
+    # For the document (knitr)
+    knitr::opts_chunk$set(
+        warning = FALSE, message = FALSE, collapse = TRUE,
+        fig.showtext = TRUE, echo = FALSE, dev = c('cairo_ps')
+    )
+}
+

### 2016-07-06 doc/manuscript.R

@@ -1,0 +1,204 @@
+## ----knit_setup, echo=FALSE----------------------------------------------
+knitr::opts_knit$set(root.dir = '../')
+
+## ----setup, message=FALSE, echo=FALSE------------------------------------
+source('.Rprofile')
+suppressMessages(run_setup())
+ds <- load_data()
+ds0yr <- filter(ds, VN == 0)
+
+## ----caption_setup-------------------------------------------------------
+## Include captions below using `captioner` package
+fig <- captioner(prefix = 'Figure')
+cite_f <- pryr::partial(fig, display = 'cite')
+sfig <- captioner(prefix = 'Supplemental Figure S')
+cite_sf <- pryr::partial(sfig, display = 'cite')
+tab <- captioner(prefix = 'Table')
+cite_t <- pryr::partial(tab, display = 'cite')
+stab <- captioner(prefix = 'Supplemental Table S')
+cite_st <- pryr::partial(stab, display = 'cite')
+# usage: cite_st('basicChar')
+
+tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 clinic visits.')
+fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006).')
+fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.  Generalized estimating equation models were adjusted for time, sex, ethnicity, baseline age, WC, ALT, and family history of diabetes. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association.')
+
+sfig_consort <- sfig('consort', 'CONSORT diagram of sample size at each examination visit.')
+stab_qic <- stab('qic', 'Comparing generalized estimating equation models adjusting for different covariates using Quasi-Likelihood Information Criterion.')
+stab_dist <- stab('nefaDist', 'Raw concentration values (nmol/mL) of each non-esterified fatty acid in the PROMISE cohort at the baseline visit (2004-2006). Values are presented as mean (SD).')
[…]

### 2016-07-08 figshare.R
Added the code and steps to upload R analysis code to figshare

@@ -1,0 +1,22 @@
+library(rfigshare)
+
+# Steps to take:
+# 1. make new orphan branch
+# 2. keep only R files and purl(manuscript.Rmd)
+# 3. git commit the files
+# 4. git archive --format=zip --output=git-nefa-code.zip manuscript-code #(branch name)
+# 5. then run the below code
+id <- fs_new_article("Analysis code for study on non-esterified fatty acids and the pathogenesis of diabetes",
+                     "R code used for the analysis in the NEFA manuscript.",
+                     type = "fileset",
+                     # there was a problem with the authors...
+                     # authors = c("Luke Johnston", "Stewart Harris", "Ravi Retnakaran",
+                     #             "Adria Giacca", "Zhen Liu", "Richard P. Bazinet",
+                     #             "Anthony J. Hanley"),
+                     tags = c("insulin sensitivity", "beta-cell function", "fatty acids",
+                              "non-esterified fatty acids", "free fatty acids",
+                              "cohort", "longitudinal"),
+                     categories = c("Diseases", "Pathogenesis", "Epidemiology"),
+                     #links = "link to github",
+                     files = "git-nefa-code.zip",
+                     visibility = "draft")

### 2016-07-27 R/functions.R
Updated code for carpenter, fix plots

@@ -8,12 +8,14 @@
 # Renaming ----------------------------------------------------------------
 
 renaming_table_rows <- function(x) {
+    `%>%` <- magrittr::`%>%`
     x %>%
         gsub('IGIIR', 'IGI/IR', .) %>%
         gsub('ISSI2', 'ISSI-2', .) %>%
         gsub('HOMA', 'HOMA-IR', .) %>%
         gsub('TAG', 'TAG (mmol/L)', .) %>%
         gsub('Chol', 'Chol (mmol/L)', .) %>%
+        gsub('ALT', 'ALT (U/L)', .) %>%
         gsub('LDL', 'LDL (mmol/L)', .) %>%
         gsub('HDL', 'HDL (mmol/L)', .) %>%
         gsub('BaseTotalNE', 'NEFA (nmol/mL)', .) %>%
@@ -23,6 +25,7 @@ renaming_table_rows <- function(x) {
 }
 
 renaming_outcomes <- function(x) {
+    `%>%` <- magrittr::`%>%`
     x %>%
         gsub('linvHOMA', 'log(1/HOMA-IR)', .) %>%
         gsub('lISI', 'log(ISI)', .) %>%
@@ -34,6 +37,7 @@ renaming_outcomes <- function(x) {
 }
 
 renaming_fa <- function(x) {
+    `%>%` <- magrittr::`%>%`
     x %>%
         gsub('.*(\\d\\d)(\\d)', '\\1:\\2', .) %>%
         gsub('n(\\d)$', 'n-\\1', .) %>%
@@ -43,11 +47,13 @@ renaming_fa <- function(x) {
 }
 
 renaming_fraction <- function(x) {
+    `%>%` <- magrittr::`%>%`
     x %>%
         gsub('ne', 'Non-esterified', .)
 }
 
 renaming_list <- function(x) {
+    `%>%` <- magrittr::`%>%`
     x %>%
         renaming_fa() %>%
         renaming_outcomes()
@@ -338,6 +344,19 @@ plot_heatmap <- function(data, x = c(outcomes, 'BMI', 'Waist', 'Age', 'lALT',
 
 # Calculate or extract for inline -----------------------------------------
 
+calculate_sample <- function(data) {
+    data %>%
+        tbl_df() %>%
+        select(VN, HOMA, ISI, IGIIR, ISSI2) %>%
+        gather(Measure, Value, -VN) %>%
+        na.omit() %>%
+        group_by(Measure, VN) %>%
[…]

### 2016-07-27 R/setup.R
Rearranged library statements in setup

@@ -34,13 +34,13 @@ run_setup <- function() {
     if (!require(geepack)) install.packages('geepack')
     if (!require(magrittr)) install.packages('magrittr')
     if (!require(dplyr)) install.packages('dplyr')
+    if (!require(mason)) install.packages('mason')
     # - For github packages:
     # if (!require(package)) packrat::install_github('username/package')
 
     # How to manage local? Not sure.
-    library(carpenter)
-    library(mason)
-    library(seer)
+    library(carpenter) #github: lwjohnst86/carpenter
+    library(seer) #github: lwjohnst86/seer
 }
 
 .set_options <- function() {
@@ -56,7 +56,8 @@ run_setup <- function() {
     # For the document (knitr)
     knitr::opts_chunk$set(
         warning = FALSE, message = FALSE, collapse = TRUE,
-        fig.showtext = TRUE, echo = FALSE, dev = c('cairo_ps')
+        #fig.showtext = TRUE,
+        echo = FALSE, dev = c('cairo_ps')
     )
 }
 

### 2016-07-27 doc/manuscript.Rmd
Updated R code link in manuscript
Updated manuscript to fit with Diabetologia reqs

@@
     - /home/luke/Documents/bibliography/bib/bib-db/myrefs.bib
-#csl: /home/luke/Documents/bibliography/styles/dependent/diabetes-care.csl
-csl: /home/luke/Documents/bibliography/styles/vancouver-brackets-no-et-al.csl
-output: fost::diabetes_care_manuscript
+    - /home/luke/Documents/bibliography/bib/bib-db/methods.bib
+output: fost::diabetologia_manuscript
 ---
@@
 <!-- 47 characters + spaces -->
-Running title: NEFA and the pathogenesis of diabetes
+<!--Running title: NEFA and the pathogenesis of diabetes-->
 
-<!-- First, middle initial, last, highest academic degree -->
-Author: Luke W. Johnston (1), MSc; Stewart B. Harris (2), MD; Ravi Retnakaran
-(3,4), MD; Adria Giacca (5), PhD; Zhen Liu (1), PhD;
-Richard P. Bazinet (1), PhD; and Anthony J. Hanley (1,3,6), PhD
+<!-- First, middle initial, last -->
+Author: Luke W. Johnston (1); Stewart B. Harris (2); Ravi Retnakaran
+(3,4); Adria Giacca (5); Zhen Liu (1);
+Richard P. Bazinet (1); and Anthony J. Hanley (1,3,6)
 
-<!-- During time of study. -->
-Affiliation:
+Institutions of origin:
 
@@
 
-Word count: 3620 / 4000
+Word count - Abstract: 242 / 250
+
+Word count - Main text: 3726 / 4000
 
@@
 
-<!-- Reference limit 39 / 40 -->
+<!-- Reference limit 39 / 50 -->
 
@@
 
-Objective: Our aim was to determine the longitudinal associations of individual
+Aim: Our aim was to determine the longitudinal associations of individual
 non-esterified fatty acids (NEFA) on changes in insulin sensitivity (IS) and
@@
 
-Research Design and Methods: In the PROMISE cohort, 477 participants had serum
+Methods: In the PROMISE cohort, 477 participants had serum
 NEFA measured at the baseline visit and completed an OGTT at 3 time points over
@@
 Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were
-used, adjusting for time, waist, sex, ethnicity, baseline age, ALT, and family
-history of diabetes. NEFA were analyzed as both concentrations and proportions
-(mol%) of the total fraction.
+used, adjusting for time, waist, sex, ethnicity, baseline age, alanine
+aminotransferase, and family history of diabetes. NEFA were analyzed as both
+concentrations (nmol/mL) and proportions (mol%) of the total fraction.
 
@@
 specific fatty acid composition.
+
+## Keywords:
+
+insulin sensitivity; beta-cell function; longitudinal cohort; non-esterified
+fatty acids; diabetes pathogenesis; fatty acid composition
+
+## Abbreviations:
+
+- GEE: Generalized estimating equations
+- LCMM: Latent curve mixed models
+- PLS-DA: Partial least squares regression discriminant analysis
+- IGI/IR: Insulinogenic index over HOMA-IR
+- ISI: Insulin Sensitivity Index (Matsuda index)
+- ISSI-2: Insulin Secretion-Sensitivity Index-2
+- ALT: Alanine aminotransferase
+- TAG: Triacylglyceride
+- PROMISE: Prospective Metabolism and Islet Cell Evaluation Cohort
+- QIC: Quasi-Likelihood Information Criterion
+- MAQ: Modified Activity Questionnaire
+- WC: Waist circumference
 
@@
 ## Include captions below using `captioner` package
-fig <- captioner(prefix = 'Figure')
+fig <- captioner(prefix = 'Fig.')
 cite_f <- pryr::partial(fig, display = 'cite')
-sfig <- captioner(prefix = 'Supplemental Figure S')
+sfig <- captioner(prefix = 'ESM Fig.')
 cite_sf <- pryr::partial(sfig, display = 'cite')
@@
 cite_t <- pryr::partial(tab, display = 'cite')
-stab <- captioner(prefix = 'Supplemental Table S')
+stab <- captioner(prefix = 'ESM Table')
 cite_st <- pryr::partial(stab, display = 'cite')
@@
 tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 clinic visits.')
-fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006).')
-fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
-Generalized estimating equation models were adjusted for time, sex, ethnicity, baseline age, WC, ALT, and family history of diabetes.
-Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
+fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each NEFA in PROMISE participants at the baseline visit (2004-2006).')
+fig_gee <- fig('gee', 'Longitudinal associations of individual NEFA (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
+Fully adjusted generalized estimating equation models.
+X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.
 P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association.')
@@
 The solid line represents an explained variance of 100% while the dashed line represents an explained variance of 50%.
-See the Supplemental Methods for a description of PLS-DA and an explanation of interpreting this plot.')
+See the ESM Methods for a description of PLS-DA and an explanation of interpreting this plot.')
 sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', '')
 sfig_plsda_loadings_dysgly <- sfig('plsda_loadings_dysgly', 'Pattern loadings from partial least squares discriminant analysis (PLS-DA) to identify potential clusters of NEFA composition for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).
-See the Supplemental Methods for a description of PLS-DA and an explanation of interpreting this plot.
+See the ESM Methods for a description of PLS-DA and an explanation of interpreting this plot.
 The percent explained variance of each component is shown in brackets on each axis.
@@
 In this case, PLS-DA was 82% accurate at correctly classifying groups.
-See the Supplemental Methods for a description of PLS-DA and for interpreting this plot.')
+See the ESM Methods for a description of PLS-DA and for interpreting this plot.')
 sfig_plsda_groups <- sfig('plsda_groups', paste0('Clustering of extracted components from the partial least squares discriminant analysis (PLS-DA) on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006).
-See the Supplemental Methods for a description of PLS-DA and interpreting this plot.
+See the ESM Methods for a description of PLS-DA and interpreting this plot.
 The percent explained variance of each component is shown in brackets on each axis.
@@
 sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', paste0('Clustering of extracted components from the partial least squares discriminant analysis for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).
-See the Supplemental Methods for a description of PLS-DA and interpreting this plot.
+See the ESM Methods for a description of PLS-DA and interpreting this plot.
 The percent explained variance of each component is shown in brackets on each axis.
@@
 
-# Background
+# Introduction
 
@@
 
-# Research Design and Methods
+# Methods
 
@@
 between 2004-2006. Follow-up examinations in this cohort occur every three years, with three
-examination visits completed to date (2004-2006, 2007-2009, and 2010-2013).  The
+examination visits completed to date (2004-2006, 2007-2009, and 2010-2013). 
+Participants are contacted annually by telephone. The
 current study used data on participants who did not have diabetes at baseline,
@@
 trajectories of beta-cell function (ISSI-2) over the 6 years. For more detailed
-explanation of GEE, PLS-DA, and LCMM, please see the Supplemental Methods.
+explanation of GEE, PLS-DA, and LCMM, please see the ESM Methods.
 
 All analyses were performed using R `r getRversion()` [@Rbase], along with the R
-packages geepack `r packageVersion('geepack')` for GEE <!--[@Hoejsgaard2006a]-->, caret
+packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a], caret
 `r packageVersion('caret')` for PLS-DA, and lcmm `r packageVersion('lcmm')` for
-LCMM. The R code for this manuscript is available at {{add link before
-submission}}. Results were considered statistically significance at p<0.05,
+LCMM. The R code for this manuscript is available at https://figshare.com/s/a45a1e1eec5d9421eba3. Results were considered statistically significance at p<0.05,
 after adjusting for multiple testing using the Benjamini-Hochberg False
-Discovery Rate [@Benjamini1995a].
+Discovery Rate [@Benjamini1995a]. STROBE was used as a guideline for reporting [@Vandenbroucke2007b].
 
@@
 est.gee <- calculate_gee_magnitude(gee_results)
+sample.outcomes <- calculate_sample(ds)
 ```
@@
 
+There are a few important limitations to our study. NEFA were only quantified at
+the baseline visit and as such we cannot investigate whether there were 
+concomitant changes in NEFA and the metabolic measures. However, we believe this
+is a strength for our specific objective, as the chance of reverse causality is 
+reduced given that fatty acid and glucose metabolism pathways are tightly 
+integrated. This is also an observational cohort, and there may be some residual
+confounding we have not considered or that could not be measured. Nonetheless,
+potential covariates were empirically analyzed prior to inclusion into the
+GEE models to best understand and minimize potential
+confounding. Finally, our cohort consists of individuals at-risk for diabetes, 
+who are primarily female and European-ancestry and as such our results may not be
+generalizable to other populations. However, given these limitations, our study
+also has several strengths, including the longitudinal design and the rigorous
+statistical techniques and methods applied, which are specifically suited to
+investigating temporal relationships and to handling the multivariate nature of
+the data. Lastly, our cohort contains highly detailed and comprehensive
+variable measurements at each collection visit, and has both concentration and
+mol% data for the fatty acids.
+
 The role of total NEFA in the etiology of diabetes is well-documented.
@@
 
-There are a few important limitations to our study. NEFA were only quantified at
-the baseline visit and as such we cannot investigate whether there were 
-concomitant changes in NEFA and the metabolic measures. However, we believe this
-is a strength for our specific objective, as the chance of reverse causality is 
-reduced given that fatty acid and glucose metabolism pathways are tightly 
-integrated. This is also an observational cohort, and there may be some residual
-confounding we have not considered or that could not be measured. Nonetheless,
-potential covariates were empirically analyzed prior to inclusion into the
-GEE models to best understand and minimize potential
-confounding. Finally, our cohort consists of individuals at-risk for diabetes, 
-who are primarily female and European-ancestry and as such our results may not be
-generalizable to other populations. However, given these limitations, our study
-also has several strengths, including the longitudinal design and the rigorous
-statistical techniques and methods applied, which are specifically suited to
-investigating temporal relationships and to handling the multivariate nature of
-the data. Lastly, our cohort contains highly detailed and comprehensive
-variable measurements at each collection visit, and has both concentration and
-mol% data for the fatty acids.
-
-<!--
-A limitation of these methods, excluding LCMM,
-is that they can not be adjusted for covariates.
--->
-
 In conclusion, we found that total NEFA was a strong predictor for lower
 beta-cell function over 6 years, irrespective of the specific composition of the
-NEFA fraction. While future studies are needed to confirm these findings, our
-results reinforce the importance of continuing to investigate the role of
-circulating NEFA concentration on the natural history of diabetes.
+NEFA fraction, suggesting that efforts at lower circulating NEFA
+indiscriminately to reduce the risk of diabetes are well warranted. While future
+studies are needed to confirm these findings, our results reinforce the
+importance of continuing to investigate the role of circulating NEFA
+concentration on the natural history of diabetes.
 
-## Acknowledgements
+# Acknowledgements
 
@@
 expert technical assistance and dedication in their work for PROMISE.
+
+## Funding
+
+This study was supported by grants from the Canadian Diabetes Association (CDA),
+the Canadian Institutes for Health Research, and the University of
+Toronto Banting and Best Diabetes Centre; LWJ is supported by a CDA
+Doctoral Student Research Award; RR is supported by a Heart and Stroke
+Foundation of Ontario Mid-Career Investigator Award; SBH holds the CDA Chair in
+National Diabetes Management and the Ian McWhinney Chair of Family Medicine
+Studies at the University of Western Ontario; RBP holds a Tier II Canada
+Research Chair in Brain Lipid Metabolism; AJH holds a Tier II Canada Research
+Chair in Diabetes Epidemiology.
+
+## Duality of interest
+
+The authors report no potential conflicts of interest relevant to this study.
+
+## Contribution statement
+
 The authors had the following responsibility: LWJ conducted research, analyzed
@@
 feedback on all versions of the paper; LWJ and AJH had primary responsibility
-for final content. All authors read and approved the final manuscript. AJH is
-the guarantor of this work and, as such, had full access to all the data in the
-study and takes responsibility for the integrity of the data and the accuracy of
-the data analysis.
+for final content. All authors read and approved the final manuscript.
 
-This study was supported by grants from the Canadian Diabetes Association (CDA),
-the Canadian Institutes for Health Research (CIHR), and the University of
-Toronto Banting and Best Diabetes Centre (BBDC); LWJ is supported by a CDA
-Doctoral Student Research Award; RR is supported by a Heart and Stroke
-Foundation of Ontario Mid-Career Investigator Award; SBH holds the CDA Chair in
-National Diabetes Management and the Ian McWhinney Chair of Family Medicine
-Studies at the University of Western Ontario; RBP holds a Tier II Canada
-Research Chair in Brain Lipid Metabolism; AJH holds a Tier II Canada Research
-Chair in Diabetes Epidemiology. The authors report no potential conflicts of
-interest relevant to this study.
+# References
 
-Portions of the results in the present analysis have been presented in
-abstract form at the following conferences: International Symposium on Diabetes
-& Nutrition (2015); Biennial Congress of the International Society for the Study
-of Fatty Acids and Lipids (2014); International Diabetes Federation World
-Diabetes Congress (2015); American Diabetes Association Scientific Sessions
-(2014, 2015).
+# Tables
 
-# Figures
+```{r table1}
+table_basic(ds, caption = tab_basic)
+```
+
+Note: Values are in median (IQR), mean (SD), and n (%). The proportion of ethnic
+and sex groups did not change over the 6 years. Outcome variables sample size
+ranges from `r sample.outcomes` over the 3 visits.
+
+# Figure legends
 
@@
 
-# Tables
+# ESM Material
 
-```{r table1}
-table_basic(ds, caption = tab_basic)
-```
-
-Note: Values are in median (IQR), mean (SD), and n (%). The proportion of ethnic
-and sex groups did not change over the 6 years.
-
-# Supplemental Material
-
-## Supplemental Methods
+## ESM Methods
 
@@
 
-## Supplemental Tables and Figures
+## ESM Tables and Figures
 
@@
 
-```{r onlineSupplementalTableS1}
+```{r ESM_Table1}
 prep_qic <- gee_df %>%
@@
 
-```{r onlineSupplementalTableS2}
+```{r ESM_Table2}
 table_distribution(ds, stab_dist)
@@
 
-```{r onlineSupplementalFigureS2, fig.cap=sfig_corr}
+```{r ESM_Fig2, fig.cap=sfig_corr}
 plot_heatmap(ds)
@@
 
-```{r onlineSupplementalFigureS3, fig.cap=sfig_gee}
+```{r ESM_Fig3, fig.cap=sfig_gee}
 plot_gee_results(gee_results_unadj)
@@
 
-```{r onlineSupplementalTableS3_4}
+```{r ESM_Table3_4}
 table_gee(gee_results, stab_gee)
@@
 
-```{r onlineSupplementalFigureS4, fig.cap=sfig_lcmm}
+```{r ESM_Fig4, fig.cap=sfig_lcmm}
 plot_lcmm_results(lcmm_data)
@@
 
-```{r onlineSupplementalFigureS5, fig.cap=sfig_plsda_groups}
+```{r ESM_Fig5, fig.cap=sfig_plsda_groups}
 plot_plsda_grouping(plsda_results, 'Beta-cell function')
@@
 
-```{r onlineSupplementalFigureS6, fig.cap=sfig_plsda_loadings, fig.width=7, fig.height=7}
+```{r ESM_Fig6, fig.cap=sfig_plsda_loadings, fig.width=7, fig.height=7}
 plot_plsda_loadings(plsda_results)
@@
 
-```{r onlineSupplementalFigureS7, fig.cap=sfig_plsda_groups_dysgly}
+```{r ESM_Fig7, fig.cap=sfig_plsda_groups_dysgly}
 plot_plsda_grouping(plsda_dysgly, 'Dysglycemia')
@@
 
-```{r onlineSupplementalFigureS8, fig.cap=sfig_plsda_loadings_dysgly, fig.width=7, fig.height=7}
+```{r ESM_Fig8, fig.cap=sfig_plsda_loadings_dysgly, fig.width=7, fig.height=7}
 plot_plsda_loadings(plsda_dysgly)
@@
 
-```{r onlineSupplementalFigureS9, fig.cap=sfig_plsda_example}
+```{r ESM_Fig9, fig.cap=sfig_plsda_example}
 plot_plsda_grouping(example_plsda_results(), 'Example groups')
 ```
-
-# References

### 2016-07-27 R/fetch_data.R
Updated scripts and data to reflect new proj style

@@ -1,33 +1,39 @@
-#' Fetch data from the master dataset.
+#' Fetch data from the original source
 #'
 #' This function fetchs the main dataset, keeps variables relevant to
 #' the analysis, restrict the sample size as needed, and lastly save
-#' the new dataset as an `.RData` file. The dot in front of the function hides
-#' it from the global environment.
+#' the new dataset as an `.RData` file.
 #'
-.fetch_data <- function() {
+#' @return Saves the wrangled data into the data/ folder.
+#' @export
+#'
+#' @examples
+#' \dontrun{
+#' fetch_data()
+#' }
+fetch_data <- function() {
     # Load the master dataset,
     ds.prep <- PROMISE::PROMISE_data %>%
-        filter(VN %in% c(1, 3, 6)) %>%
+        dplyr::filter(VN %in% c(1, 3, 6)) %>%
         ## Kick out Canoers
-        filter(is.na(Canoe)) %>%
-        tbl_df()
+        dplyr::filter(is.na(Canoe)) %>%
+        dplyr::tbl_df()
 
     print(paste0('Original dataset rows are ', dim(ds.prep)[1], ' and columns are ', dim(ds.prep)[2]))
 
     ##' Munge and wrangle the data into the final version.
     ds <-
-        full_join(ds.prep,
+        dplyr::full_join(ds.prep,
                   ds.prep %>%
-                      filter(VN == 1) %>%
-                      select(SID, BaseAge = Age, BaseTAG = TAG)) %>%
-        select(
+                      dplyr::filter(VN == 1) %>%
+                      dplyr::select(SID, BaseAge = Age, BaseTAG = TAG)) %>%
+        dplyr::select(
             SID, VN, BMI, Waist, HOMA, ISI, IGIIR, ISSI2, TAG, LDL, HDL, Chol,
-            ALT, CRP, FamHistDiab, matches('meds'), Age, Sex, Ethnicity,
+            ALT, CRP, FamHistDiab, dplyr::matches('meds'), Age, Sex, Ethnicity,
             IFG, IGT, DM, MET, BaseAge, AlcoholPerWk, TobaccoUse, SelfEdu, Occupation,
-            TotalNE, matches('^ne\\d+'), Glucose0, Glucose120, BaseTAG
+            TotalNE, dplyr::matches('^ne\\d+'), Glucose0, Glucose120, BaseTAG
         ) %>%
-        mutate(
+        dplyr::mutate(
[…]

### 2016-07-27 R/functions.R
Updated r functions to reflect new style

@@ -71,7 +71,7 @@ analyze_gee <- function(data, y, x, covar, unit,
     }
     gee_results <- data %>%
         mason::design('gee') %>%
-        mason::add_settings(family = gaussian('identity'), corstr = 'ar1', cluster.id = 'SID') %>%
+        mason::add_settings(family = stats::gaussian('identity'), corstr = 'ar1', cluster.id = 'SID') %>%
         mason::add_variables('yvars', y) %>%
         mason::add_variables('xvars', x) %>%
         mason::add_variables('covariates', covar) %>% {
@@ -106,7 +106,7 @@ analyze_gee <- function(data, y, x, covar, unit,
 analyze_lcmm <- function(data, lc.var = 'lISSI2') {
     data.prep <- data %>%
         dplyr::select_('SID', lc.var, 'VN') %>%
-        na.omit() %>%
+        stats::na.omit() %>%
         dplyr::arrange(SID, VN)
     lcmm_results <- lcmm::lcmm(
         lISSI2 ~ VN,
@@ -121,7 +121,7 @@ analyze_lcmm <- function(data, lc.var = 'lISSI2') {
 analyze_plsda <- function(data, variable = 'ClassLCMM') {
     data.prep <- data %>%
         dplyr::filter(VN == 0) %>%
-        dplyr::select(contains(variable), matches('^pct_ne\\d\\d')) %>%
+        dplyr::select(dplyr::contains(variable), dplyr::matches('^pct_ne\\d\\d')) %>%
         na.omit()
 
     if (length(caret::nearZeroVar(data.prep[2:dim(data.prep)[2]])) != 0)
@@ -149,12 +149,12 @@ analyze_plsda <- function(data, variable = 'ClassLCMM') {
 get_gee_data <- function(data, fa = 'pct_ne|^ne') {
     gee_ready_data <- dplyr::full_join(
         data %>%
-            dplyr::select(-matches(fa), -TotalNE),
+            dplyr::select(-dplyr::matches(fa), -TotalNE),
         ## Scale all the fatty acids
         data %>%
             dplyr::filter(VN == 0) %>%
-            dplyr::select(SID, TotalNE, matches(fa)) %>%
-            dplyr::mutate_each(funs(as.numeric(scale(
+            dplyr::select(SID, TotalNE, dplyr::matches(fa)) %>%
+            dplyr::mutate_each(dplyr::funs(as.numeric(scale(
[…]

### 2016-07-27 R/setup.R

@@ -1,55 +1,27 @@
-#' Code preamble
-#' =============
+# Setup description -------------------------------------------------------
+#
+# These functions are used to setup your files so that all options are in one
+# location, and references via the function call.
+#
+# These functions can be accessed using either `library(projectname)` or
+# `devtools::load_all()` (Ctrl-Shift-L in RStudio).
+
+#' Set options for all documents and scripts.
 #'
-#' Install packages if they are not installed and load them. Set options for
-#' the packages.
+#' @export
+#' @importFrom magrittr %>%
 #'
-run_setup <- function() {
-    # Run the following packages for setup
-    .load_packages()
-    .set_options()
-    .load_packages()
-    .load_packages()
-}
-
-.load_packages <- function() {
-    # May need to run this function another time if a package needs to be
-    # installed.
-    #
-    # Add your packages you want to use here, using the following format:
-    # - For pre-bundled packages
-
-    # - For CRAN packages:
-    # if (!require(package)) install.packages('package')
-    if (!require(lcmm)) install.packages('lcmm')
-    if (!require(caret)) install.packages('caret')
-    if (!require(fdrtool)) install.packages('fdrtool')
-    if (!require(captioner)) install.packages('captioner')
-    if (!require(tidyr)) install.packages('tidyr')
-    if (!require(ggplot2)) install.packages('ggplot2')
-    if (!require(pander)) install.packages('pander')
-    if (!require(knitr)) install.packages('knitr')
-    if (!require(DiagrammeR)) install.packages('DiagrammeR')
-    if (!require(MuMIn)) install.packages('MuMIn')
-    if (!require(geepack)) install.packages('geepack')
-    if (!require(magrittr)) install.packages('magrittr')
-    if (!require(dplyr)) install.packages('dplyr')
-    if (!require(mason)) install.packages('mason')
-    # - For github packages:
-    # if (!require(package)) packrat::install_github('username/package')
-
-    # How to manage local? Not sure.
-    library(carpenter) #github: lwjohnst86/carpenter
-    library(seer) #github: lwjohnst86/seer
-}
-
-.set_options <- function() {
+#' @examples
+#'
+#' set_options()
+#'
+set_options <- function() {
     # Set the options here for individual packages
 
     # For tables (pander)
-    panderOptions('table.split.table', Inf)
-    panderOptions('table.style', 'rmarkdown')
-    panderOptions('table.alignment.default',
+    pander::panderOptions('table.split.table', Inf)
[…]

### 2016-07-27 doc/manuscript.Rmd
Updated manuscript to reflect new proj style

@@
 ```{r knit_setup, echo=FALSE}
-knitr::opts_knit$set(root.dir = '../')
+#knitr::opts_knit$set(root.dir = '../')
 ```
@@
 ```{r setup, message=FALSE, echo=FALSE}
-source('.Rprofile')
-suppressMessages(run_setup())
-ds <- load_data()
-ds0yr <- filter(ds, VN == 0)
+#source('.Rprofile')
+devtools::load_all()
+set_options()
+ds <- project_data
+ds0yr <- dplyr::filter(ds, VN == 0)
 ```
@@
 ## Include captions below using `captioner` package
+library(captioner)
 fig <- captioner(prefix = 'Fig.')
@@
 plsda_dysgly <- dysgly_data %>% 
-    mutate(ConvertDysgly = factor(ConvertDysgly, levels = c(0, 1), labels = c('No', 'Yes'))) %>% 
+    dplyr::mutate(ConvertDysgly = factor(ConvertDysgly, levels = c(0, 1), labels = c('No', 'Yes'))) %>% 
     analyze_plsda('ConvertDysgly')
@@
 ```{r ESM_Table1}
+library(magrittr)
 prep_qic <- gee_df %>%
@@
     dplyr::select(-lISSI2) %>% 
-    na.omit()
+    stats::na.omit()
 
@@
     dplyr::select(-lISI) %>% 
-    na.omit()
+    stats::na.omit()
 
@@
     lISI ~ TotalNE + VN, data = qic_is,
-    id = SID, family = gaussian,
+    id = SID, family = stats::gaussian,
     corstr = 'ar1'
@@
 
-M1 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex)
-M2 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT) # best
-M3 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET) # best
-M4 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk) # best
-M5 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab) # best
-M6 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse) # best
-M7 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TobaccoUse) # best
-M8 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab) # best
-M9 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TotalNE:VN)
+M1 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex)
+M2 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT) # best
+M3 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET) # best
+M4 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk) # best
+M5 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab) # best
+M6 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse) # best
+M7 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TobaccoUse) # best
+M8 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab) # best
+M9 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TotalNE:VN)
 
-qic_is_tab <- MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M9, rank = QIC) %>% 
-    dplyr::add_rownames() %>%
-    dplyr::select(Model = rowname, QIC, Delta = delta) %>%
+qic_is_tab <- MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M9, rank = MuMIn::QIC) %>% 
+    tibble::rownames_to_column() %>%
+    dplyr::select(Model = rowname, QIC = IC, Delta = delta) %>%
     dplyr::mutate(id = seq_len(nrow(.))) %>%
@@
     lISSI2 ~ TotalNE + VN, data = qic_bcf,
-    id = SID, family = gaussian,
+    id = SID, family = stats::gaussian,
     corstr = 'ar1'
     )
-M1 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex)
-M2 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT) 
-M3 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET)
-M4 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk)
-M5 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab) # best
-M6 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse) # best
-M7 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TobaccoUse) # best
-M8 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab) # best
-M9 <- update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TotalNE:VN) # best
+M1 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex)
+M2 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT) 
+M3 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET)
+M4 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk)
+M5 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab) # best
+M6 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse) # best
+M7 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TobaccoUse) # best
+M8 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab) # best
+M9 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TotalNE:VN) # best
 
-qic_bcf_tab <- MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M9, rank = QIC) %>% 
-    dplyr::add_rownames() %>%
-    dplyr::select(Model = rowname, QIC, Delta = delta) %>% 
+qic_bcf_tab <- MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M9, rank = MuMIn::QIC) %>% 
+    tibble::rownames_to_column() %>%
+    dplyr::select(Model = rowname, QIC = IC, Delta = delta) %>% 
     dplyr::mutate(id = seq_len(nrow(.))+10) %>%

### 2016-07-27 doc/manuscript.Rmd
Added tony's edits

@@
 
-Table and figure count: 3 / 4
+<!-- Table and figure count: 3 / 4 -->
 
@@
 Participants are contacted annually by telephone. The
-current study used data on participants who did not have diabetes at baseline,
+current study sample used data on participants who did not have diabetes at baseline,
 who returned for one or more of the follow-up examinations, and who had samples
@@
 
-In a Canadian population of adults who are at-risk for diabetes, we found that 
+In a Canadian population of adults who were at-risk for diabetes, we found that 
 higher total NEFA concentrations independently predicted lower beta-cell
@@
 acids as a proportion, an observation that was further confirmed by multivariate
-cluster techniques, which found no predictive ability of these
-individual NEFA on beta-cell function or dysglycemia. These observations suggest
+cluster techniques, which found no ability of these
+individual NEFA to predict beta-cell function or dysglycemia. These observations suggest
 that the absolute size of the total NEFA fraction, rather than its specific
@@
 the baseline visit and as such we cannot investigate whether there were 
-concomitant changes in NEFA and the metabolic measures. However, we believe this
+concomitant changes in NEFA and the metabolic measures over time. However, we believe this
 is a strength for our specific objective, as the chance of reverse causality is 
@@
 beta-cell function over 6 years, irrespective of the specific composition of the
-NEFA fraction, suggesting that efforts at lower circulating NEFA
-indiscriminately to reduce the risk of diabetes are well warranted. While future
+NEFA fraction, suggesting that efforts at lowering circulating total NEFA
+through medication and/or lifestyle strategies to reduce the risk of diabetes
+are well warranted. While future
 studies are needed to confirm these findings, our results reinforce the

### 2016-07-27 R/functions.R
Updated code from master branch

@@ -8,12 +8,14 @@
 # Renaming ----------------------------------------------------------------
 
 renaming_table_rows <- function(x) {
+    `%>%` <- magrittr::`%>%`
     x %>%
         gsub('IGIIR', 'IGI/IR', .) %>%
         gsub('ISSI2', 'ISSI-2', .) %>%
         gsub('HOMA', 'HOMA-IR', .) %>%
         gsub('TAG', 'TAG (mmol/L)', .) %>%
         gsub('Chol', 'Chol (mmol/L)', .) %>%
+        gsub('ALT', 'ALT (U/L)', .) %>%
         gsub('LDL', 'LDL (mmol/L)', .) %>%
         gsub('HDL', 'HDL (mmol/L)', .) %>%
         gsub('BaseTotalNE', 'NEFA (nmol/mL)', .) %>%
@@ -23,6 +25,7 @@ renaming_table_rows <- function(x) {
 }
 
 renaming_outcomes <- function(x) {
+    `%>%` <- magrittr::`%>%`
     x %>%
         gsub('linvHOMA', 'log(1/HOMA-IR)', .) %>%
         gsub('lISI', 'log(ISI)', .) %>%
@@ -34,6 +37,7 @@ renaming_outcomes <- function(x) {
 }
 
 renaming_fa <- function(x) {
+    `%>%` <- magrittr::`%>%`
     x %>%
         gsub('.*(\\d\\d)(\\d)', '\\1:\\2', .) %>%
         gsub('n(\\d)$', 'n-\\1', .) %>%
@@ -43,11 +47,13 @@ renaming_fa <- function(x) {
 }
 
 renaming_fraction <- function(x) {
+    `%>%` <- magrittr::`%>%`
     x %>%
         gsub('ne', 'Non-esterified', .)
 }
 
 renaming_list <- function(x) {
+    `%>%` <- magrittr::`%>%`
     x %>%
         renaming_fa() %>%
         renaming_outcomes()
@@ -65,7 +71,7 @@ analyze_gee <- function(data, y, x, covar, unit,
     }
     gee_results <- data %>%
         mason::design('gee') %>%
-        mason::add_settings(family = gaussian('identity'), corstr = 'ar1', cluster.id = 'SID') %>%
+        mason::add_settings(family = stats::gaussian('identity'), corstr = 'ar1', cluster.id = 'SID') %>%
         mason::add_variables('yvars', y) %>%
         mason::add_variables('xvars', x) %>%
[…]

### 2016-07-27 R/load_data.R

@@ -1,42 +1,26 @@
-# ---(LIKELY) NO NEED TO EDIT---
-#
-# `load_data()` either loads the dataset inside the `data/` folder or, if the
-# `fetch_data.R` file has been changed, will re-run the `fetch_data.R` function
-# to update the dataset in `data/` and then loads the dataset.
-#
-# There are three arguments to this function:
-#
-# - `local.dataset`: Set the file path to the local dataset (eg. `data/`).
-# - `fetch.script`: Set the file path to the `fetch_data.R` function.
-# - `force`: Force updating the dataset in `data/` by running the `fetch_data.R`
-# function.
-#
-# Examples for usage:
-#
-#   ds <- load_data()
-#   ds <- load_data(
-#       'data/dsname.RData',
-#       'R/fetchScript.R')
-#
-load_data <- function(local.dataset = file.path('data', 'data.Rds'),
-                      fetch.script = file.path('R', 'fetch_data.R'),
-                      force = FALSE) {
-    source('.Rprofile')
-    if (force) {
-        .fetch_data()
-    }
-    if (!file.exists(fetch.script))
-        stop('Please create a file under the R folder that contains a function called fetch_data.')
-    if (!file.exists(local.dataset)) {
-        dir.create('data')
-        .fetch_data()
-        message('A dataset has been added and loaded.')
-    } else if (file.mtime(fetch.script) > file.mtime(local.dataset)) {
-        .fetch_data()
-        message('Dataset has been updated and loaded.')
-    } else {
-        message('Dataset is up-to-date.')
-    }
-
-    readRDS(local.dataset)
+#' Load the data found in data/.
+#'
+#' @param update Whether you want to update the dataset using the fetch_data
+#'   function.
+#' @param project.name Name of your project. Set the default to keep it always
+#'   working.
+#'
+#' @return Outputs the datasets in the data/ folder.
+#' @export
+#'
+#' @examples
+#'
+#' \dontrun{
+#' load_data()
+#' load_data(update = TRUE)
+#' }
+#'
+load_data <- function(update = FALSE, project.name = 'nefaDiabetes') {
+    if (is.null(project.name))
+        stop('Please change the project.name default to the name of the project.')
+    if (!dir.exists(system.file('data', package = project.name)))
+        fetch_data()
+    if (update)
+        fetch_data()
+    devtools::load_data()
 }

### 2016-07-27 R/setup.R

@@ -1,62 +1,35 @@
-#' Code preamble
-#' =============
+# Setup description -------------------------------------------------------
+#
+# These functions are used to setup your files so that all options are in one
+# location, and references via the function call.
+#
+# These functions can be accessed using either `library(projectname)` or
+# `devtools::load_all()` (Ctrl-Shift-L in RStudio).
+
+#' Set options for all documents and scripts.
 #'
-#' Install packages if they are not installed and load them. Set options for
-#' the packages.
+#' @export
+#' @importFrom magrittr %>%
 #'
-run_setup <- function() {
-    # Run the following packages for setup
-    .load_packages()
-    .set_options()
-    .load_packages()
-    .load_packages()
-}
-
-.load_packages <- function() {
-    # May need to run this function another time if a package needs to be
-    # installed.
-    #
-    # Add your packages you want to use here, using the following format:
-    # - For pre-bundled packages
-
-    # - For CRAN packages:
-    # if (!require(package)) install.packages('package')
-    if (!require(lcmm)) install.packages('lcmm')
-    if (!require(caret)) install.packages('caret')
-    if (!require(fdrtool)) install.packages('fdrtool')
-    if (!require(captioner)) install.packages('captioner')
-    if (!require(tidyr)) install.packages('tidyr')
-    if (!require(ggplot2)) install.packages('ggplot2')
-    if (!require(pander)) install.packages('pander')
-    if (!require(knitr)) install.packages('knitr')
-    if (!require(DiagrammeR)) install.packages('DiagrammeR')
-    if (!require(MuMIn)) install.packages('MuMIn')
-    if (!require(geepack)) install.packages('geepack')
-    if (!require(magrittr)) install.packages('magrittr')
-    if (!require(dplyr)) install.packages('dplyr')
-    # - For github packages:
-    # if (!require(package)) packrat::install_github('username/package')
-
-    # How to manage local? Not sure.
-    library(carpenter)
-    library(mason)
-    library(seer)
-}
-
-.set_options <- function() {
+#' @examples
+#'
+#' set_options()
+#'
+set_options <- function() {
     # Set the options here for individual packages
 
     # For tables (pander)
-    panderOptions('table.split.table', Inf)
-    panderOptions('table.style', 'rmarkdown')
-    panderOptions('table.alignment.default',
+    pander::panderOptions('table.split.table', Inf)
[…]

### 2016-07-27 doc/manuscript.R

@@ -1,27 +1,29 @@
 ## ----knit_setup, echo=FALSE----------------------------------------------
-knitr::opts_knit$set(root.dir = '../')
+#knitr::opts_knit$set(root.dir = '../')
 
 ## ----setup, message=FALSE, echo=FALSE------------------------------------
-source('.Rprofile')
-suppressMessages(run_setup())
-ds <- load_data()
-ds0yr <- filter(ds, VN == 0)
+#source('.Rprofile')
+devtools::load_all()
+set_options()
+ds <- project_data
+ds0yr <- dplyr::filter(ds, VN == 0)
 
 ## ----caption_setup-------------------------------------------------------
 ## Include captions below using `captioner` package
-fig <- captioner(prefix = 'Figure')
+library(captioner)
+fig <- captioner(prefix = 'Fig.')
 cite_f <- pryr::partial(fig, display = 'cite')
-sfig <- captioner(prefix = 'Supplemental Figure S')
+sfig <- captioner(prefix = 'ESM Fig.')
 cite_sf <- pryr::partial(sfig, display = 'cite')
 tab <- captioner(prefix = 'Table')
 cite_t <- pryr::partial(tab, display = 'cite')
-stab <- captioner(prefix = 'Supplemental Table S')
+stab <- captioner(prefix = 'ESM Table')
 cite_st <- pryr::partial(stab, display = 'cite')
 # usage: cite_st('basicChar')
 
 tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 clinic visits.')
-fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each non-esterified fatty acid in PROMISE participants at the baseline visit (2004-2006).')
-fig_gee <- fig('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.  Generalized estimating equation models were adjusted for time, sex, ethnicity, baseline age, WC, ALT, and family history of diabetes. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association.')
+fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each NEFA in PROMISE participants at the baseline visit (2004-2006).')
[…]

### 2017-02-20 doc/submission/reviewer-response.md
Pulled out reviewer comments from decision letter into a response file

# Responses to reviewers

## Referee: 1

First, I would like to congratulate you for the exhaustive and complete
research work that has been done for this paper. I think the statistical study
is very successful and useful for researchers who have accumulated a large
database and need a method with the ability to handle a large number of
variables and their interactions. I really liked read in the discussion the
negative points of the study, with which I fully agree, and not an attempt to
hide the limitations of any scientific work. In retrospect, perhaps it will miss
data that clarify the pre-diabetic state, or not, of the patients (HbA1c or
C-peptide). I think it would be good for the study know if some patients were
taking any medication that could affect the results. Although it is a
complicated scientific work to understand for researchers who do not usually use
this kind of statistical studies, I believe that is described in a very
appropriate and clear manner.

## Referee: 2

The present paper addresses an important and yet unresolved research question
regarding the role of non-esterified fatty acids on the trajectories of
metabolic traits of diabetes (insulin sensitivity and beta-cell function). The
analysis is based on a cohort study with repeated measures (up to 3 times per
individual) over 6 years of follow-up using generalized estimating equations
(GEE). This is a well-established method to analyse longitudinal data. The
authors in general did a thorough analysis however some major questions related
to methods and interpretation remains. After the trajectory analysis they run
partial least-squares discriminant analyses (PLS-DA) with groups defined by
beta-cell trajectories (latent class analysis) and defined by glycemic status as
outcomes. While the primary analysis is well-described and easy to follow, **these
additional analyses are described mostly in the online appendix and much harder
to interpret for the general reader. In my opinion, these do not add substantial
information to the paper and the conclusion, and I would move these to the
online appendix or drop them from the paper.** I think that the paper needs a
major revision however both the dataset and the study questions are adequate and
important thus conditional acceptance is reasonable.

1. Title of the paper: it is a statement of the results. I would prefer to leave
it open ended without an interpretation of the results, as with different levels
of adjustments the results and conclusion will change substantially. See also
comments on the results section.

2. Methods in the abstract:
    - Please give some information on the cohort characteristics:
        - age, sex
        - high risk of diabetes but no prevalent disease at baseline
    - incident diabetes cases (how were these handled? what about treated
    patients? – these should go in the main methods section)
    - adjustment for time – I would prefer to state that time since baseline was
    used as the underlying time variable instead of adjusting for time

3. Results in the abstract:
    - a major problem in my view with the interpretation of the results is that
    all characteristics are taken as confounders although a considerable
    proportion of time-varying covariates are taken after the NEFA measurements.
    Furthermore, it is not clear whether these measures are confounders,
    mediators or NEFAs are mediators of these variables. I think that unadjusted
    results are as interesting and require discussion as the fully adjusted and
    selected models. Regarding the abstract, I would show both the unadjusted
    and the fully adjusted results here.

4. Conclusions in abstract: I have a different interpretation of the results,
particularly because adjustment waist may be an over-adjustment.

5. INTRODUCTION
    - a clear and well-written introduction with a clear objective and
    hypothesis

6. MATERIALS & METHODS:
    - ESM Figure 1 is not clear to me.
        - Prevalent diabetes cases were excluded at baseline (n=XX)
        - the lines suggest that everyone had to have a year 6 visit, ehile in
        the main text, the authors state that any follow-up visit was enough for
        being included (this corresponds well with the GEE requirements).
        - what happened with incident diabetes cases? Were treated diabetes
        cases included in the analysis? How does that effect measures of insulin
        sensitivity and beta-cell function? I suggest to exclude those visits
        where diabetes was treated.
    - I would prefer to see the equations for the outcome measures. Why not use
    HOMA2-IS instead of 1/HOMA-IR?
    - In the statistical analysis you state that co-variates were
    time-dependent. This is not true for age at baseline, family history, sex,
    ethnicity. Please clarify in the text.
    - misspelling: “statistically significance”
    - I am not an expert of PLS-DA but it is not clear to me, why does this
    analysis reflect more the role of relative NEFA (mole%) analysis than the
    concentration analysis? As I can see, it depends on what variables you feed
    in the model.

7. RESULTS:
    - Please give ns in Table 1.
    - It would be also nice to see, how many participants had 1, 2 or 3 time
    points in the analysis.
    - How were the changes calculated (GEE?) in the basic characteristics of
    the PROMISE cohort section?
    - I suggest to include the unadjusted data to the main paper (ESM Figure 3)
    and describe these findings as we do not know the exact model where
    over-adjustment start to happen.
    - In a trajectory analysis, the interaction between time and the main
    predictor (NEFA in this case) is of major interest. The authors check for
    this interaction only as the last step of their model building (see ESM). I
    would suggest to perform this as the first step of the analysis (M0), and
    drop this term, if it is non-significant or does not improve model fit.
    Furthermore, given that this interaction was dropped from the model, it is
    hard to except the major conclusion that NEFA predicts progression of
    beat-cell dysfunction. Actually what is found here is that NEFA are related
    to beta-cell function in repeated cross-sectional analyses and NEFA related
    differences in beta-cell function remain constant over time.
    - The model building is based on literature data which is the right way to
    build the models but a more detailed discussion on over-adjustment is
    required before interpretation.
    - Latent classes. Although this is a very attractive method, based on ESM
    figure 4, there are 3 parallel declining group trajectories that is mostly
    affected by the baseline values. Groups based on baseline (last follow-up)
    tertiles would be easier to interpret.
    - The PLS-DA analysis needs further description. I would expect variables
    with the highest loadings to be the variables in the unadjusted GEE
    analysis. I am probably wrong given the results.
    - Do we really expect good discrimination? The associations are week to nul
    between NEFAs and the outcomes. I do not really see the point the extensive
    supplementary analysis on PLS-DA. Furthermore dysglycemia is an outcome that
    is not directly related to beta-cell function only. Dysglycemia requires
    both insulin resistance and beta-cell dysfunction.
    - The authors mention in the Discussion the external validity is mostly
    limited to white females. I suggest to run a sensitivity analysis on this
    group.

8. DISCUSSION
    - I would like to see more on the DAGs that underlies the selected
    variables in the models. That would either confirm the way the analysis is
    currently performed, or would show that some of the adjustments are
    over-adjustment.
    - In the time-varying analysis some covariates are measured after NEFAs,
    thus adjustment for them could be over-adjustment just because they are
    measured closer to the outcome of interest.

# Associate Editor

Many thanks for submitting this potentially interesting paper. The second
reviewer comes with some very helpful comments that will need to be addressed in
detail. There is need to consider the results from the Paris Prospective study:
Charles MA et al.  The role of non-esterified fatty acids in the deterioration
of glucose tolerance in Caucasian subjects: results of the Paris Prospective
Study. Diabetologia 1997;40:1101-6

### 2017-02-20 doc/submission/reviewer-response.md
First draft response to Reviewer 1

@@
 
-First, I would like to congratulate you for the exhaustive and complete
+1. *First, I would like to congratulate you for the exhaustive and complete
 research work that has been done for this paper. I think the statistical study
@@
 hide the limitations of any scientific work. In retrospect, perhaps it will miss
-data that clarify the pre-diabetic state, or not, of the patients (HbA1c or
-C-peptide). I think it would be good for the study know if some patients were
-taking any medication that could affect the results. Although it is a
+data that **clarify the pre-diabetic state, or not, of the patients (HbA1c or
+C-peptide)**. I think it would be good for the study know if some patients were
+taking **any medication that could affect the results**. Although it is a
 complicated scientific work to understand for researchers who do not usually use
 this kind of statistical studies, I believe that is described in a very
-appropriate and clear manner.
+appropriate and clear manner.*
+
+    We thank the reviewer for these comments. {{not sure about comment on
+    pre-diabetes}}. In the PROMISE cohort, we collected medication use data from
+    all participants. While most participants took some form of medication over
+    the 6 years, during the times of data collection, few ({{get number}}) were
+    on medications. However, for those who were taking medications that could
+    influence the results, sensitivity analyses revealed they did not
+    substantially influence NEFA concentrations, the outcome measures, nor the
+    GEE model results {{need to run/confirm this}}. These results have been
+    appended to the ESM {{yes?}}.
 

### 2017-02-22 doc/submission/reviewer-response.md
First draft response to Reviewer 2

@@
 
-1. *First, I would like to congratulate you for the exhaustive and complete
+- *"First, I would like to congratulate you for the exhaustive and complete
 research work that has been done for this paper. I think the statistical study
@@
 this kind of statistical studies, I believe that is described in a very
-appropriate and clear manner.*
+appropriate and clear manner."*
 
@@
 
-The present paper addresses an important and yet unresolved research question
+- *"The present paper addresses an important and yet unresolved research question
 regarding the role of non-esterified fatty acids on the trajectories of
@@
 authors in general did a thorough analysis however some major questions related
-to methods and interpretation remains. After the trajectory analysis they run
-partial least-squares discriminant analyses (PLS-DA) with groups defined by
-beta-cell trajectories (latent class analysis) and defined by glycemic status as
-outcomes. While the primary analysis is well-described and easy to follow, **these
-additional analyses are described mostly in the online appendix and much harder
-to interpret for the general reader. In my opinion, these do not add substantial
-information to the paper and the conclusion, and I would move these to the
-online appendix or drop them from the paper.** I think that the paper needs a
-major revision however both the dataset and the study questions are adequate and
-important thus conditional acceptance is reasonable.
+to methods and interpretation remains."*
 
-1. Title of the paper: it is a statement of the results. I would prefer to leave
+    We appreciated the reviewer's positive comments and detailed suggestions. In
+    the responses below we try to address each of the comments raised by the
+    reviewer.
+
+- *"After the trajectory analysis they run partial least-squares discriminant
+analyses (PLS-DA) with groups defined by beta-cell trajectories (latent class
+analysis) and defined by glycemic status as outcomes. While the primary analysis
+is well-described and easy to follow, **these additional analyses are described
+mostly in the online appendix and much harder to interpret for the general
+reader. In my opinion, these do not add substantial information to the paper and
+the conclusion, and I would move these to the online appendix or drop them from
+the paper.** I think that the paper needs a major revision however both the
+dataset and the study questions are adequate and important thus conditional
+acceptance is reasonable."*
+
+    Given that GEE models do not take into account the high dimensionality of
+    the NEFA composition data, we ran the PLS analyses in order to confirm the
+    GEE results when using a multivariate approach. Use of the LCMM analysis
+    allowed us to take include the longitudinal data in the PLS analysis.
+    However, we agree with the reviewer that these analyses, while interesting
+    in their confirmation of the GEE findings, do not contribute to the overall
+    manuscript. In light of this reviewer's comments, we decided to move {{or
+    drop?}} the analyses into the ESM, including the description of the methods
+    and results.
+
+1. *"Title of the paper: it is a statement of the results. I would prefer to leave
 it open ended without an interpretation of the results, as with different levels
 of adjustments the results and conclusion will change substantially. See also
-comments on the results section.
+comments on the results section."*
 
-2. Methods in the abstract:
-    - Please give some information on the cohort characteristics:
-        - age, sex
-        - high risk of diabetes but no prevalent disease at baseline
-    - incident diabetes cases (how were these handled? what about treated
-    patients? – these should go in the main methods section)
-    - adjustment for time – I would prefer to state that time since baseline was
-    used as the underlying time variable instead of adjusting for time
+    As suggested, we've made the title results agnostic.
 
-3. Results in the abstract:
-    - a major problem in my view with the interpretation of the results is that
-    all characteristics are taken as confounders although a considerable
-    proportion of time-varying covariates are taken after the NEFA measurements.
-    Furthermore, it is not clear whether these measures are confounders,
-    mediators or NEFAs are mediators of these variables. I think that unadjusted
-    results are as interesting and require discussion as the fully adjusted and
-    selected models. Regarding the abstract, I would show both the unadjusted
-    and the fully adjusted results here.
+2. *"Methods in the abstract: a) Please give some information on the cohort
+characteristics: age, sex, high risk of diabetes but no prevalent disease at
+baseline; b) incident diabetes cases (how were these handled? what about treated
+patients? - these should go in the main methods section); c) adjustment for time
+– I would prefer to state that time since baseline was used as the underlying
+time variable instead of adjusting for time"*
 
-4. Conclusions in abstract: I have a different interpretation of the results,
-particularly because adjustment waist may be an over-adjustment.
+    We have made the requested changes to the methods and stated the exact
+    analyses used more explicitly. Incident diabetes cases are presented only
+    for descriptive purposes and were not included in any further analysis. We
+    thank the reviewer for the suggestion and agree that using time since
+    baseline is a better variable to adjust for; they have been replaced in GEE
+    models, though the change does not substantially impact the final results.
 
-5. INTRODUCTION
-    - a clear and well-written introduction with a clear objective and
-    hypothesis
+3. *"Results in the abstract: a major problem in my view with the
+interpretation of the results is that all characteristics are taken as
+confounders although a considerable proportion of time-varying covariates are
+taken after the NEFA measurements. Furthermore, it is not clear whether these
+measures are confounders, mediators or NEFAs are mediators of these variables. I
+think that unadjusted results are as interesting and require discussion as the
+fully adjusted and selected models. Regarding the abstract, I would show both
+the unadjusted and the fully adjusted results here."*
 
-6. MATERIALS & METHODS:
-    - ESM Figure 1 is not clear to me.
-        - Prevalent diabetes cases were excluded at baseline (n=XX)
-        - the lines suggest that everyone had to have a year 6 visit, ehile in
-        the main text, the authors state that any follow-up visit was enough for
-        being included (this corresponds well with the GEE requirements).
-        - what happened with incident diabetes cases? Were treated diabetes
-        cases included in the analysis? How does that effect measures of insulin
-        sensitivity and beta-cell function? I suggest to exclude those visits
-        where diabetes was treated.
-    - I would prefer to see the equations for the outcome measures. Why not use
-    HOMA2-IS instead of 1/HOMA-IR?
-    - In the statistical analysis you state that co-variates were
+    The specifics of the covariates are expanded upon in the methods section. We
+    have included the unadjusted results in the main manuscript and have 
+    described it in the results and discussion sections, in addition to making
+    the changes to the abstract.
+
+4. *"Conclusions in abstract: I have a different interpretation of the results,
+particularly because adjustment waist may be an over-adjustment."*
+
+    We have included a sensitivity analysis identifying which exact covariate
+    attenuates the results between the unadjusted and adjusted models. {{briefly
+    describe the results here}}.
+
+5. *"INTRODUCTION: a clear and well-written introduction with a clear objective
+and hypothesis"*
+
+6. *"MATERIALS & METHODS:"*
+    - *"ESM Figure 1 is not clear to me: a) Prevalent diabetes cases were excluded
+    at baseline (n=XX); b) the lines suggest that everyone had to have a year 6
+    visit, ehile in the main text, the authors state that any follow-up visit
+    was enough for being included (this corresponds well with the GEE
+    requirements); c) what happened with incident diabetes cases? Were treated
+    diabetes cases included in the analysis? How does that effect measures of
+    insulin sensitivity and beta-cell function? I suggest to exclude those
+    visits where diabetes was treated."*
+
+        We have clarified how we dealt with incident diabetes cases. Briefly, we
+        only report incident diabetes and dysglycemia cases in the text, but in
+        the modeling these were excluded. {{not sure how to address a) and b)...}}
+
+    - *"I would prefer to see the equations for the outcome measures. Why not use
+    HOMA2-IS instead of 1/HOMA-IR?"*
+
+        We have included the equations for the outcome measures as requested. In
+        addition, we replaced HOMA-IR with HOMA2-%S from the HOMA2 Calculator.
+
+    - *"In the statistical analysis you state that co-variates were
     time-dependent. This is not true for age at baseline, family history, sex,
-    ethnicity. Please clarify in the text.
-    - misspelling: “statistically significance”
-    - I am not an expert of PLS-DA but it is not clear to me, why does this
+    ethnicity. Please clarify in the text."* ... *"misspelling: 'statistically
+    significance'"*
+
+        These have been corrected.
+
+    - *"I am not an expert of PLS-DA but it is not clear to me, why does this
     analysis reflect more the role of relative NEFA (mole%) analysis than the
     concentration analysis? As I can see, it depends on what variables you feed
-    in the model.
+    in the model."*
+
+        {{bit confused about this one. Will need to look into the wording in the
+        manuscript.}}
 
 7. RESULTS:
-    - Please give ns in Table 1.
-    - It would be also nice to see, how many participants had 1, 2 or 3 time
-    points in the analysis.
-    - How were the changes calculated (GEE?) in the basic characteristics of
-    the PROMISE cohort section?
-    - I suggest to include the unadjusted data to the main paper (ESM Figure 3)
+    - *"Please give ns in Table 1."* ... *"It would be also nice to see, how
+    many participants had 1, 2 or 3 time points in the analysis."*
+
+        These have been added as requested.
+
+    - *"How were the changes calculated (GEE?) in the basic characteristics of
+    the PROMISE cohort section?"*
+
+        {{not completely sure what is meant here}} The changes as reported in
+        the results of the outcome variables are simple median differences
+        between the baseline and 6-year visit. The p-value {{confirm this is
+        added}} representing the difference was computed using an unadjusted GEE
+        model.
+
+    - *"I suggest to include the unadjusted data to the main paper (ESM Figure 3)
     and describe these findings as we do not know the exact model where
-    over-adjustment start to happen.
-    - In a trajectory analysis, the interaction between time and the main
+    over-adjustment start to happen."*
+
+        The unadjusted results and figure has been added to the main paper and
+        described in the results section. We've also added a sensitivity
+        analysis confirming which covariate attenuates the associations.
+
+    - *"In a trajectory analysis, the interaction between time and the main
     predictor (NEFA in this case) is of major interest. The authors check for
@@
     to beta-cell function in repeated cross-sectional analyses and NEFA related
-    differences in beta-cell function remain constant over time.
-    - The model building is based on literature data which is the right way to
+    differences in beta-cell function remain constant over time."*
+
+        The reviewer makes an excellent point regarding the interaction check in
+        model building. We have incorporated this change in the analysis and
+        report on the results in the paper. In simple terms, the reviewer is
+        correct that without an interaction term the model is essentially
+        "repeated cross-sectional analyses", however, GEE handles the analysis
+        in a slightly more nuanced method. GEE calculates the estimates of the
+        association of NEFA on the outcomes at any given time point, taking into
+        consideration the inherent multivariate structure of the repeated
+        measurements of the outcome variables. This nuance provides
+        substantially more power to the model and the proceeding results.
+
+    - *"The model building is based on literature data which is the right way to
     build the models but a more detailed discussion on over-adjustment is
-    required before interpretation.
-    - Latent classes. Although this is a very attractive method, based on ESM
+    required before interpretation."*
+
+        We have included further mention of potential over-adjustment to the
+        paper. {{where would this go exactly? throughout the results and
+        discussion?}}
+
+    - *"Latent classes. Although this is a very attractive method, based on ESM
     figure 4, there are 3 parallel declining group trajectories that is mostly
     affected by the baseline values. Groups based on baseline (last follow-up)
-    tertiles would be easier to interpret.
-    - The PLS-DA analysis needs further description. I would expect variables
-    with the highest loadings to be the variables in the unadjusted GEE
-    analysis. I am probably wrong given the results.
-    - Do we really expect good discrimination? The associations are week to nul
-    between NEFAs and the outcomes. I do not really see the point the extensive
-    supplementary analysis on PLS-DA. Furthermore dysglycemia is an outcome that
-    is not directly related to beta-cell function only. Dysglycemia requires
-    both insulin resistance and beta-cell dysfunction.
-    - The authors mention in the Discussion the external validity is mostly
+    tertiles would be easier to interpret."* ... *"The PLS-DA analysis needs
+    further description. I would expect variables with the highest loadings to
+    be the variables in the unadjusted GEE analysis. I am probably wrong given
+    the results."* ... *"Do we really expect good discrimination? The
+    associations are week to nul between NEFAs and the outcomes. I do not really
+    see the point the extensive supplementary analysis on PLS-DA. Furthermore
+    dysglycemia is an outcome that is not directly related to beta-cell function
+    only. Dysglycemia requires both insulin resistance and beta-cell
+    dysfunction."*
+
+        Given this reviewer's previous comments and suggestions, we have decided
+        to focus on the GEE analysis in the main paper and move the PLS-DA to
+        the ESM and expand on {{or remove? it seems this reviewer prefers if we
+        remove it. I could simplify it by doing what I did in the TAG as a
+        confirmation, and include it in the ESM}} it there.
+
+    - *"The authors mention in the Discussion the external validity is mostly
     limited to white females. I suggest to run a sensitivity analysis on this
-    group.
+    group."*
+
+        We agree with the reviewer's suggestion and ran additional sensitivity
+        analyses stratified by sex and ethnicity. We mention the results of this
+        analysis in the results.
 
 8. DISCUSSION
-    - I would like to see more on the DAGs that underlies the selected
+    - *"I would like to see more on the DAGs that underlies the selected
     variables in the models. That would either confirm the way the analysis is
     currently performed, or would show that some of the adjustments are
-    over-adjustment.
-    - In the time-varying analysis some covariates are measured after NEFAs,
+    over-adjustment."*
+
+        As requested, we have included the DAG underlying that hypothesized
+        associations in the ESM.
+
+    - *"In the time-varying analysis some covariates are measured after NEFAs,
     thus adjustment for them could be over-adjustment just because they are
-    measured closer to the outcome of interest.
+    measured closer to the outcome of interest."*
+
+        In order to confirm the impact the comment made by the reviewer has on
+        the reported associations, we ran sensitivity analyses where we had all
+        covariates as time-independent, to match the NEFA variables in the GEE
+        model. We found {{state findings}}.
 
@@
 
-Many thanks for submitting this potentially interesting paper. The second
+- *"Many thanks for submitting this potentially interesting paper. The second
 reviewer comes with some very helpful comments that will need to be addressed in
@@
 of glucose tolerance in Caucasian subjects: results of the Paris Prospective
-Study. Diabetologia 1997;40:1101-6
+Study. Diabetologia 1997;40:1101-6"*
+
+    {{do I need to include this?}} We thank the associate editor for the kind
+    comments. We have included and discussed the cited paper in the discussion
+    section.

### 2017-02-22 doc/manuscript.Rmd
Updated figshare link
Reorganized code and analysis flow to fit new prodigenr setup

@@
 ```{r knit_setup, echo=FALSE}
-#knitr::opts_knit$set(root.dir = '../')
+knitr::opts_chunk$set(echo = FALSE, dev = c('cairo_ps'))
 ```
@@
 ```{r setup, message=FALSE, echo=FALSE}
-#source('.Rprofile')
-devtools::load_all()
-set_options()
+devtools::load_all('.')
 ds <- project_data
@@
 `r packageVersion('caret')` for PLS-DA, and lcmm `r packageVersion('lcmm')` for
-LCMM. The R code for this manuscript is available at https://figshare.com/s/a45a1e1eec5d9421eba3. Results were considered statistically significance at p<0.05,
+LCMM. The R code for this manuscript is available at 
+https://dx.doi.org/10.6084/m9.figshare.3472433
+. Results were considered statistically significance at p<0.05,
 after adjusting for multiple testing using the Benjamini-Hochberg False

### 2017-02-22 R/load_data.R

@@ -1,26 +1,0 @@
-#' Load the data found in data/.
-#'
-#' @param update Whether you want to update the dataset using the fetch_data
-#'   function.
-#' @param project.name Name of your project. Set the default to keep it always
-#'   working.
-#'
-#' @return Outputs the datasets in the data/ folder.
-#' @export
-#'
-#' @examples
-#'
-#' \dontrun{
-#' load_data()
-#' load_data(update = TRUE)
-#' }
-#'
-load_data <- function(update = FALSE, project.name = 'nefaDiabetes') {
-    if (is.null(project.name))
-        stop('Please change the project.name default to the name of the project.')
-    if (!dir.exists(system.file('data', package = project.name)))
-        fetch_data()
-    if (update)
-        fetch_data()
-    devtools::load_data()
-}

### 2017-02-22 R/setup.R

@@ -27,9 +27,7 @@ set_options <- function() {
 
     # For the document (knitr)
     knitr::opts_chunk$set(
-        warning = FALSE, message = FALSE, collapse = TRUE,
-        #fig.showtext = TRUE,
-        echo = FALSE, dev = c('cairo_ps')
+        warning = FALSE, message = FALSE, collapse = TRUE
     )
 }
 

### 2017-02-22 R/zzz.R

@@ -1,0 +1,37 @@
+#' Load the data found in data/.
+#'
+#' @param update Whether you want to update the dataset using the fetch_data
+#'   function.
+#' @param project.name Name of your project. Set the default to keep it always
+#'   working.
+#'
+#' @return Outputs the datasets in the data/ folder.
+#' @export
+#'
+#' @examples
+#' load_data()
+#' load_data(update = TRUE)
+#'
+attach_data <- function(update = FALSE) {
+    pkg <- devtools::as.package('.')$path
+    rfile <- normalizePath(file.path(pkg, 'R', 'fetch_data.R'))
+    datadir <- file.path(pkg, 'data')
+    if (length(rfile) == 0)
+        stop("Oops, fetch_data.R file must have been deleted, please keep it.", .call = FALSE)
+    if (!dir.exists(datadir))
+        fetch_data()
+    datafile <- list.files(datadir, pattern = 'rda$', full.names = TRUE)
+    if (file.mtime(rfile) > file.mtime(datafile))
+        fetch_data()
+    if (update)
+        fetch_data()
+    devtools::load_data()
+}
+
+# This function loads everything after using `devtools::load_all()` so you don't
+# need to include all these functions in each code chunk.
+#
+.onLoad <- function(libname, pkgname) {
+    set_options()
+    attach_data()
+}

### 2017-02-22 doc/submission/reviewer-response.md
Second draft of response to reviewers (sent to Tony) # create branch

@@
+---
+output: fost::default_manuscript
+---
+
 # Responses to reviewers
@@
     We thank the reviewer for these comments. {{not sure about comment on
-    pre-diabetes}}. In the PROMISE cohort, we collected medication use data from
+    pre-diabetes}}. In the PROMISE cohort, we collected medication data from
     all participants. While most participants took some form of medication over
@@
     thank the reviewer for the suggestion and agree that using time since
-    baseline is a better variable to adjust for; they have been replaced in GEE
-    models, though the change does not substantially impact the final results.
+    baseline is a better variable to adjust for; the time variable has been
+    replaced in GEE models, though the change does not substantially impact the
+    final results. {{confirm}}
 
@@
         addition, we replaced HOMA-IR with HOMA2-%S from the HOMA2 Calculator.
+        The results were not substantially different after the replacement of
+        HOMA-IR. {{confirm}}
 
@@
 
-7. RESULTS:
+7. *"RESULTS:"*
     - *"Please give ns in Table 1."* ... *"It would be also nice to see, how
@@
 
-        {{not completely sure what is meant here}} The changes as reported in
-        the results of the outcome variables are simple median differences
-        between the baseline and 6-year visit. The p-value {{confirm this is
-        added}} representing the difference was computed using an unadjusted GEE
-        model.
+        {{not completely sure what is meant here}} The changes over time as
+        reported in the results of the outcome variables are simple median
+        differences between the baseline and 6-year visit. The p-value {{confirm
+        this is added}} representing the difference was computed using an
+        unadjusted GEE model.
 
@@
         described in the results section. We've also added a sensitivity
-        analysis confirming which covariate attenuates the associations.
+        analysis confirming which covariate(s) attenuates the associations.
 
@@
         "repeated cross-sectional analyses", however, GEE handles the analysis
-        in a slightly more nuanced method. GEE calculates the estimates of the
+        in a slightly more nuanced approach. GEE calculates the estimates of the
         association of NEFA on the outcomes at any given time point, taking into
         consideration the inherent multivariate structure of the repeated
-        measurements of the outcome variables. This nuance provides
-        substantially more power to the model and the proceeding results.
+        measurements of the outcome variables. As such, inherent similarities in
+        the outcome measurements over time within a subject are included in the
+        computation of the GEE estimates. This nuance provides substantially
+        more power to the model and the proceeding results.
 
@@
 
-8. DISCUSSION
+8. *"DISCUSSION"*
     - *"I would like to see more on the DAGs that underlies the selected

### 2017-06-11 doc/submission/reviewer-response.md
Minor edit to reviewer response

@@
 hide the limitations of any scientific work. In retrospect, perhaps it will miss
-data that **clarify the pre-diabetic state, or not, of the patients (HbA1c or
-C-peptide)**. I think it would be good for the study know if some patients were
-taking **any medication that could affect the results**. Although it is a
+data that clarify the pre-diabetic state, or not, of the patients (HbA1c or
+C-peptide). I think it would be good for the study know if some patients were
+taking any medication that could affect the results. Although it is a
 complicated scientific work to understand for researchers who do not usually use

### 2017-07-01 R/renaming.R
added homa2 renaming

@@ -3,6 +3,7 @@
 
 renaming_table_rows <- function(x) {
     x %>%
+        gsub('HOMA2_S', 'HOMA2-%S', .) %>%
         gsub('IGIIR', 'IGI/IR', .) %>%
         gsub('ISSI2', 'ISSI-2', .) %>%
         gsub('HOMA', 'HOMA-IR', .) %>%
@@ -19,6 +20,7 @@ renaming_table_rows <- function(x) {
 
 renaming_outcomes <- function(x) {
     x %>%
+        gsub('lHOMA2_S', 'log(HOMA2-%S)', .) %>%
         gsub('linvHOMA', 'log(1/HOMA-IR)', .) %>%
         gsub('lHOMA2_S', 'log(HOMA2-%S)', .) %>%
         gsub('lISI', 'log(ISI)', .) %>%

### 2017-07-01 R/calculate.R
added/updated/moved calc fun into own file

@@ -1,0 +1,122 @@
+
+#' Calculate the conversion of participants to DM or pre DM (incidence).
+#'
+#' @param variable which dysgly variable to use
+#'
+calc_conversion_dysgly <- function(variable = c('ConvertDM', 'ConvertPreDM')) {
+    variable <- match.arg(variable)
+
+    conversion <-
+        table(dysgly_data[variable])[2] %>%
+        {
+            paste0(., ' (', round((. / 477) * 100, 0), '%)')
+        }
+
+    return(conversion)
+}
+
+#' Calculate the percentage of contribution for NEFA that are >10 percent.
+#'
+#' @param data project data
+#'
+calc_pct_nefa <- function(data) {
+    over_10pct_nefa <- data %>%
+        dplyr::filter(VN == 0) %>%
+        dplyr::select(dplyr::matches('pct_ne')) %>%
+        tidyr::gather(fat, value) %>%
+        dplyr::group_by(fat) %>%
+        dplyr::summarise(pct = round(mean(value, na.rm = TRUE), 1)) %>%
+        dplyr::mutate(fat = renaming_fa(fat),
+               c = paste0(fat, ' (', pct, '%)')) %>%
+        dplyr::arrange(dplyr::desc(pct)) %>%
+        dplyr::filter(pct >= 10)
+
+    return(over_10pct_nefa)
+}
+
+#' Calculate the percent change in the outcomes (plus sample size and p-value).
+#'
+#' @param data project data
+#'
+calc_outcomes_chg_n <- function(data) {
+    prep_data <- data %>%
+        dplyr::select(f.VN, HOMA2_S, ISI, IGIIR, ISSI2) %>%
+        tidyr::gather(Measure, Value,-f.VN) %>%
+        stats::na.omit() %>%
+        dplyr::group_by(Measure, f.VN) %>%
+        dplyr::summarise(med = stats::median(Value),
+                         n = n()) %>%
+        dplyr::ungroup()
+
+    sample_size <- prep_data$n %>%
+        {paste0(min(.), '-', max(.))}
+
+    change_over_time <- prep_data %>%
+        dplyr::select(-n) %>%
+        tidyr::spread(f.VN, med) %>%
+        dplyr::mutate(pctChg = ((yr6 - yr0) / yr0) * 100) %>%
+        dplyr::select(pctChg) %>%
+        abs() %>%
+        round(0) %>%
+        {paste0(min(.), '% to ', max(.), '%')}
+
+    pval <- mason::design(data, 'gee') %>%
[…]

### 2017-07-01 R/lcmm.R
moved untested fun into lcmm

@@ -59,3 +59,14 @@ plot_lcmm_results <- function(data, results, lc_var = "lISSI2") {
         graph_theme()
 }
 
+# Untested right now. Worked previously.
+calculate_ngroups_lcmm <- function(results.lcmm) {
+    n_by_group <- results.lcmm$pprob$class %>%
+        table %>%
+        as.data.frame
+    names(n_by_group) <- c('Group', 'Freq')
+    n_by_group <-
+        dplyr::mutate(n_by_group, Group = factor(Group, labels = c('Mid', 'High', 'Low')))
+
+    return(n_by_group)
+}

### 2017-07-01 doc/manuscript.Rmd
Cleaned up first setup chunks manuscript

@@
 
-```{r knit_setup, echo=FALSE}
-knitr::opts_chunk$set(echo = FALSE, dev = c('cairo_ps'))
+```{r setup, echo=FALSE}
+devtools::load_all('.')
 ```
 
-```{r setup, message=FALSE, echo=FALSE}
-devtools::load_all('.')
-ds <- project_data
-ds0yr <- dplyr::filter(ds, VN == 0)
-```
-
-```{r caption_setup}
+```{r captions}
 ## Include captions below using `captioner` package

### 2017-07-01 doc/manuscript.Rmd
moved lcmm and plsda captions into vignettes

@@
 P-values were adjusted for the BH false discovery rate, with significant (p<0.05) associations indicated by asterisk.')
-sfig_lcmm <- sfig('lcmm', 'Latent class mixed model (LCMM) analysis to identify individual classes of trajectories for log(ISSI-2) over the 6 years in the PROMISE cohort.
-LCMM is a technique that identifies groups of participants that share a similar underlying trajectory in beta-cell function over the 6 years (e.g.
-no change compared to declines in ISSI-2).
-Red lines indicate individuals with a high beta-cell function who stayed high, green represents those in the middle, and blue represents those who had the lowest beta-cell function.')
-sfig_plsda_groups <- sfig('plsda_groups', '')
-sfig_plsda_loadings <- sfig('plsda_loadings', 'Pattern loadings from partial least squares discriminant analysis (PLS-DA) to identify potential clusters of NEFA composition within the classes extracted from the latent class mixed model in 463 participants from the baseline PROMISE visit (2004-2006).
-The percent explained variance of each component is shown in brackets on each axis.
-The solid line represents an explained variance of 100% while the dashed line represents an explained variance of 50%.
-See the ESM Methods for a description of PLS-DA and an explanation of interpreting this plot.')
-sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', '')
-sfig_plsda_loadings_dysgly <- sfig('plsda_loadings_dysgly', 'Pattern loadings from partial least squares discriminant analysis (PLS-DA) to identify potential clusters of NEFA composition for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).
-See the ESM Methods for a description of PLS-DA and an explanation of interpreting this plot.
-The percent explained variance of each component is shown in brackets on each axis.
-The solid line represents an explained variance of 100% while the dashed line represents an explained variance of 50%.')
-sfig_plsda_example <- sfig('plsda_example', 'Example results of a high discriminatory ability to classify groups accurately when using partial least squares discriminatory analysis (PLS-DA).
-Discriminatory ability is evident from the amount of separation between the two groups.
-In this case, PLS-DA was 82% accurate at correctly classifying groups.
-See the ESM Methods for a description of PLS-DA and for interpreting this plot.')
-sfig_plsda_groups <- sfig('plsda_groups', paste0('Clustering of extracted components from the partial least squares discriminant analysis (PLS-DA) on the classes extracted from the latent class mixed model (LCMM) in 463 participants from the baseline PROMISE visit (2004-2006).
-See the ESM Methods for a description of PLS-DA and interpreting this plot.
-The percent explained variance of each component is shown in brackets on each axis.
-Red, green, and blue lines indicate participants classified as high, middle, and low for beta-cell function, respectively, from the LCMM analysis.
-See ', cite_sf('plsda_example'), ' for a hypothetical example plot showing good discriminatory ability between groups.'))
-sfig_plsda_groups_dysgly <- sfig('plsda_groups_dysgly', paste0('Clustering of extracted components from the partial least squares discriminant analysis for dysglycemia (IFG, IGT, DM) conversion status over the 6-years in the participants from the baseline PROMISE visit (2004-2006).
-See the ESM Methods for a description of PLS-DA and interpreting this plot.
-The percent explained variance of each component is shown in brackets on each axis.
-Blue lines indicate dysglycemia conversion or maintanence and red lines indicate no dysglycemia status.
-See ', cite_sf('plsda_example'), ' for a hypothetical example plot showing good discriminatory ability between groups.'))
 ```

### 2017-07-01 doc/manuscript.Rmd
moved methods for plsda/lcmm to vig, updated inline code

@@
 analyses in a multivariate environment (all NEFA variables in a single model),
-partial least squares discriminant analysis (PLS-DA) was used to identify the
-patterns of NEFA composition against beta-cell function (using ISSI-2, which had
-the majority of the associations in the GEE models) and for those
-who converted to or maintained dysglycemia status (either IFG, IGT, or DM) over
-the 6 years. Latent class mixed models (LCMM) were used to extract the underlying
-trajectories of beta-cell function (ISSI-2) over the 6 years. For more detailed
-explanation of GEE, PLS-DA, and LCMM, please see the ESM Methods.
+partial least squares
 
 All analyses were performed using R `r getRversion()` [@Rbase], along with the R
-packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a], caret
-`r packageVersion('caret')` for PLS-DA, and lcmm `r packageVersion('lcmm')` for
-LCMM. The R code for this manuscript is available at 
-https://dx.doi.org/10.6084/m9.figshare.3472433
-. Results were considered statistically significance at p<0.05,
+packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a].
+The R code for this manuscript is available at 
+https://dx.doi.org/10.6084/m9.figshare.3472433. Results were considered statistically significance at p<0.05,
 after adjusting for multiple testing using the Benjamini-Hochberg False
@@
 
-```{r variableAndDataSet, message=FALSE, results='hide'}
-gee_df <- get_gee_data(ds)
-dysgly_data <- get_dysglycemia_data(ds)
+```{r inline_results}
+pct_nefa <- calc_pct_nefa(project_data)
+chg_outcomes <- calc_outcomes_chg_n(project_data)
+dm_incid <- calc_conversion_dysgly('ConvertDM')
+predm_incid <- calc_conversion_dysgly('ConvertPreDM')
 
-outcomes_is <- c('linvHOMA', 'lISI')
-outcomes_bcf <- c('lIGIIR', 'lISSI2')
-outcomes <- c(outcomes_is, outcomes_bcf)
-ne_pct <- grep('^pct_ne', names(gee_df), value = TRUE)
-ne_conc <- c('TotalNE', grep('^ne\\d\\d', names(gee_df), value = TRUE))
-covariates <- c('VN', 'Waist', 'Sex', 'Ethnicity', 'BaseAge', 'ALT', 'FamHistDiab')
-```
-
-```{r geeData, cache=TRUE}
-gee_results <- dplyr::bind_rows(
-    analyze_gee(gee_df, outcomes_is, ne_pct, covariates, 'mol%', adj.p = TRUE),
-    analyze_gee(gee_df, outcomes_is, ne_conc, covariates, 'nmol/mL', adj.p = TRUE),
-    analyze_gee(gee_df, outcomes_bcf, ne_pct, covariates, 'mol%', adj.p = TRUE),
-    analyze_gee(gee_df, outcomes_bcf, ne_conc, covariates, 'nmol/mL', adj.p = TRUE)
-    ) %>% 
-    tidy_gee_results()
-
-gee_results_unadj <- dplyr::bind_rows(
-    analyze_gee(gee_df, outcomes_is, ne_pct, 'VN', 'mol%', adj.p = TRUE),
-    analyze_gee(gee_df, outcomes_is, ne_conc, 'VN', 'nmol/mL', adj.p = TRUE),
-    analyze_gee(gee_df, outcomes_bcf, ne_pct, 'VN', 'mol%', adj.p = TRUE),
-    analyze_gee(gee_df, outcomes_bcf, ne_conc, 'VN', 'nmol/mL', adj.p = TRUE)
-    ) %>% 
-    tidy_gee_results()
-```
-
-```{r lcmmPls, cache=TRUE, results='hide'}
-# include also the 3-yr and 6-yr data for training/testing.
-lcmm_results <- analyze_lcmm(ds)
-lcmm_data <- combine_lcmm_data(ds, lcmm_results)
-plsda_results <- analyze_plsda(lcmm_data)
-plsda_dysgly <- dysgly_data %>% 
-    dplyr::mutate(ConvertDysgly = factor(ConvertDysgly, levels = c(0, 1), labels = c('No', 'Yes'))) %>% 
-    analyze_plsda('ConvertDysgly')
-```
-
-```{r inlineResults}
-percent.nefa <- calculate_percent_nefa_contribution(ds) 
-misclass.n.percent.bcf <- calculate_plsda_misclass(plsda_results)
-misclass.n.percent.dysgly <- calculate_plsda_misclass(plsda_dysgly)
-percent.change.outcomes <- calculate_outcomes_pct_change(ds)
-lcmm.groups <- calculate_ngroups_lcmm(lcmm_results)
-dm.convert <- calculate_conversion_dysgly(dysgly_data, 'ConvertDM')
-predm.convert <- calculate_conversion_dysgly(dysgly_data, 'ConvertPreDM')
-sex.npct <- calculate_factor_npercent(ds0yr$Sex, 'Female')
-ethn.npct <- calculate_factor_npercent(ds0yr$Ethnicity, 'European')
-fhd.npct <- calculate_factor_npercent(ds0yr$FamHistDiab, 'Yes')
-est.gee <- calculate_gee_magnitude(gee_results)
-sample.outcomes <- calculate_sample(ds)
+ds0yr <- dplyr::filter(project_data, VN == 0)
+female <- calc_npercent(ds0yr$Sex, 'Female')
+european <- calc_npercent(ds0yr$Ethnicity, 'European')
+famhist <- calc_npercent(ds0yr$FamHistDiab, 'Yes')
+est_gee <- calculate_gee_magnitude(gee_results)
 ```

### 2017-07-01 doc/manuscript.Rmd
moved plsda/lcmm results to vig

@@
 
-## Latent trajectories and differences in NEFA composition
-
-Three latent classes were extracted from LCMM, with `r lcmm.groups$Freq[2]` 
-participants in the high, `r lcmm.groups$Freq[1]` in the middle, and 
-`r lcmm.groups$Freq[3]`
-in the low beta-cell function groups. The trajectories of these groups are
-shown in `r cite_sf('lcmm')`. These groups were used as the beta-cell function
-response variables in the PLS-DA analysis, which was used to confirm the results
-of the GEE modeling, but in a multivariate environment (all NEFA in a single
-model). The PLS-DA results show a very poor discriminatory ability of the
-NEFA composition in classifying participants into the correct beta-cell function
-latent class. The percent of participants correctly predicted to belonging to
-their group (i.e. those predicted in the low group and actually belonging to the
-low group) was `r misclass.n.percent.bcf$Percent[9]` in the low group,
-`r misclass.n.percent.bcf$Percent[5]` in the middle group, and 
-`r misclass.n.percent.bcf$Percent[1]` in the high group (the distribution of the
-groupings based on the components from PLS-DA are shown in `r cite_sf('plsda_groups')`). 
-The loadings plot of the NEFA composition against the extracted PLS-DA
-components is shown in `r cite_sf('plsda_loadings')` and is used to identify
-fatty acids that contribute the most explained variance to the beta-cell
-function groupings. However, because of the high rate of misclassification, no
-meaningful predictive information can be extracted even if any NEFA does
-contribute to the explained variance (e.g. 16:0). The PLS-DA analysis using
-conversion to or maintenance of dysglycemia status over the 6 years as the
-response variable showed similar poor discriminatory ability, though better than
-the beta-cell function trajectory results. While 
-`r misclass.n.percent.dysgly$Percent[1]`
-of participants were correctly predicted to not have dysglycemia, only 
-`r misclass.n.percent.dysgly$Percent[4]` were correctly predicted to have
-dysglycemia (see `r cite_sf('plsda_groups_dysgly')` for the distribution of 
-dysglycemia status based on the components from the PLS-DA and 
-`r cite_sf('plsda_loadings_dysgly')` for the loadings plot of the NEFA composition
-by extracted PLS-DA components). The results from the PLS-DA analysis are consistent
-with the associations seen in the GEE models.
+```{r figure2, dependson='geeData', fig.cap=fig_gee}
+gee_results %>% 
+    dplyr::filter(model == "Adjusted") %>% 
+    plot_gee_main()
+```
 

### 2017-07-01 doc/manuscript.Rmd
Moved figs/tabs to end, moved esm to own file

@@
 
-```{r figure2, dependson='geeData', fig.cap=fig_gee}
-gee_results %>% 
-    dplyr::filter(model == "Adjusted") %>% 
-    plot_gee_main()
-```
-
 # Conclusions
@@
 
-# References
+# Tables and figures
 
-# Tables
-
-```{r table1}
-table_basic(ds, caption = tab_basic)
+```{r table1_basic}
+table_basic(project_data, caption = tab_basic)
 ```
@@
 
-# Figure legends
-
-```{r figure1, fig.cap=fig_dist}
-plot_nefa_distribution(ds)
+```{r figure1_nefa, fig.cap=fig_dist}
+plot_nefa_distribution()
 ```
 
-```{r figure2, dependson='geeData', fig.cap=fig_gee}
-plot_gee_results(gee_results, leg = 'none')
+```{r figure2_gee, dependson='geeData', fig.cap=fig_gee}
+gee_results %>% 
+    dplyr::filter(model == "Adjusted") %>% 
+    plot_gee_main()
 ```
 
-# ESM Material
-
-## ESM Methods
-
-Generalized estimating equations (GEE) is analogous to mixed effects models, 
-except it emphasizes population-level trends over individual-level trends that 
-mixed models use. Because it can handle autocorrelations in the outcome (as 
-would be seen in variables measured over time), GEE modeling is well suited to
-longitudinal data. Latent class mixed models (LCMM) is a form of mixed effects
-modeling that tries to categorize individuals into groups that have similar
-clusters; in the case of the present analysis are the trajectories in beta-cell
-function over time. Because mixed models use the participant as the random
-effect, LCMM can identify individuals who similarly change over time.
-
-Partial least squares (PLS) regression is a technique similar to principal
-component analysis (PCA) that is designed to extract meaningful information from
-multivariate, high dimensionality data (e.g. as in metabolomic and other -omic type 
-analyses). These types of methods try to extract as much of the variance in the 
-data into a smaller number of components or factors. The difference between PCA 
-and PLS is that PLS is a supervised method, while PCA is not. This means that
-PLS uses a response variable(s), i.e. an outcome or **Y**, to describe the
-variation in the predictors (**X**) while PCA only describes the variation 
-inherent in **X**. Because of this, PLS can be better at predicting the 
-contributions of predictor variables against an outcome variable. Partial least
-squares discriminant analysis (PLS-DA) is a form of PLS, except the **Y** is 
-a discrete (i.e. categorical) variable and not a continuous one. Therefore, 
-PLS-DA is able to predict who will belong to which group based on the values of
-the predictors.
-
-The two plots generated from the PLS-DA analysis show the results from both
-the response (i.e. LCMM beta-cell function groups and dyslgycemia status) and
-the predictors (all 22 NEFA). The response result plots as seen in 
-`r cite_sf('plsda_groups')` and `r cite_sf('plsda_groups_dysgly')` show the
-poor discriminatory ability of the PLS-DA model, with `r cite_sf('plsda_example')` 
-showing a hypothetical example of good discriminatory ability. When the predictors represent an
-underlying pattern that may exist between groups, the plots will show greater
-separation between the groups in addition to a low misclassification rate. The
-predictor result plots as seen in `r cite_sf('plsda_loadings')` and
-`r cite_sf('plsda_loadings_dysgly')` show which of the predictor variables
-contribute the most to the explained variance. Variables between the solid and 
-dashed lines explain between 50-100% of the explained variance. Whether the 
-variables have positive or negative values on either of the components axes 
-determines how they contribute to each group. For instance, if those individuals
-with dysglycemia have a positive value for component 1 and the fatty acid 16:0 also
-has a positive value for component 1, this indicates that 16:0 is predictive of
-having dysglycemia.
-
-## ESM Tables and Figures
-
-![`r sfig_consort`](../img/flowDiagramSample.png)
-
-```{r ESM_Table1}
-library(magrittr)
-prep_qic <- gee_df %>%
-    dplyr::select(
-        SID, VN, lISSI2, lISI, TotalNE, Sex, Ethnicity, MET, ALT, BaseAge,
-        Waist, BMI, AlcoholPerWk, TobaccoUse, FamHistDiab
-        )
-
-qic_is <- prep_qic %>% 
-    dplyr::select(-lISSI2) %>% 
-    stats::na.omit()
-
-qic_bcf <- prep_qic %>% 
-    dplyr::select(-lISI) %>% 
-    stats::na.omit()
-
-# QIC for IS
-M0 <- geepack::geeglm(
-    lISI ~ TotalNE + VN, data = qic_is,
-    id = SID, family = stats::gaussian,
-    corstr = 'ar1'
-    )
-
-M1 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex)
-M2 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT) # best
-M3 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET) # best
-M4 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk) # best
-M5 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab) # best
-M6 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse) # best
-M7 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TobaccoUse) # best
-M8 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab) # best
-M9 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TotalNE:VN)
-
-qic_is_tab <- MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M9, rank = MuMIn::QIC) %>% 
-    tibble::rownames_to_column() %>%
-    dplyr::select(Model = rowname, QIC = IC, Delta = delta) %>%
-    dplyr::mutate(id = seq_len(nrow(.))) %>%
-    dplyr::bind_rows(., data.frame(Model = '**log(ISI)**', id = 0.5))
-
-# QIC data for BCF
-qic_bcf <- prep_qic %>% 
-    dplyr::select(-lISI) %>% 
-    stats::na.omit()
-
-M0 <- geepack::geeglm(
-    lISSI2 ~ TotalNE + VN, data = qic_bcf,
-    id = SID, family = stats::gaussian,
-    corstr = 'ar1'
-    )
-M1 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex)
-M2 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT) 
-M3 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET)
-M4 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk)
-M5 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab) # best
-M6 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + MET + AlcoholPerWk + FamHistDiab + TobaccoUse) # best
-M7 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TobaccoUse) # best
-M8 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab) # best
-M9 <- stats::update(M0, . ~ . + Waist + BaseAge + Ethnicity + Sex + ALT + FamHistDiab + TotalNE:VN) # best
-
-qic_bcf_tab <- MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M9, rank = MuMIn::QIC) %>% 
-    tibble::rownames_to_column() %>%
-    dplyr::select(Model = rowname, QIC = IC, Delta = delta) %>% 
-    dplyr::mutate(id = seq_len(nrow(.))+10) %>%
-    dplyr::bind_rows(., data.frame(Model = '**log(ISSI-2)**', id = 10.5))
-
-rbind(qic_is_tab, qic_bcf_tab) %>% 
-    dplyr::arrange(id) %>%
-    dplyr::select(-id) %>%
-    dplyr::mutate(QIC = round(QIC, 1),
-           Delta = round(Delta, 1)) %>% 
-    pander::pander(missing = '', caption = stab_qic)
-```
-
-Given the number of possible combinations of outcome and predictor variables,
-only ISI and ISSI-2 with total non-esterified fatty acids (nmol/mL) were used to
-compare various GEE models and to select a final model.  Baseline age was used
-as including both the original age and the time variable would result in
-collinearity.  Column names are: QIC is the quasi-likelihood information criteria
-(smaller values, eg. larger negative values, indicate a better fit compared to
-other models), Delta is the QIC minus the lowest QIC (models with delta <10 are 
-considered equivalent).  Models were:
-
-- M0: log(ISSI-2) or log(ISI) = total non-esterified fatty acids (nmol/mL) + time
-- M1: M0 + waist + sex + ethnicity + baseline age
-- M2: M1 + ALT
-- M3: M2 + physical activity
-- M4: M3 + alcohol intake
-- M5: M4 + family history of diabetes
-- M6: M5 + smoking status
-- M7: M2 + ALT + family history of diabetes + smoking status
-- M8: M2 + ALT + family history of diabetes
-- M9: M8 + time by NEFA interaction
-
-```{r ESM_Table2}
-table_distribution(ds, stab_dist)
-```
-
-```{r ESM_Fig2, fig.cap=sfig_corr}
-plot_heatmap(ds)
-```
-
-```{r ESM_Fig3, fig.cap=sfig_gee}
-plot_gee_results(gee_results_unadj)
-```
-
-```{r ESM_Table3_4}
-table_gee(gee_results, stab_gee)
-table_gee(gee_results_unadj, stab_gee_unadj)
-```
-
-```{r ESM_Fig4, fig.cap=sfig_lcmm}
-plot_lcmm_results(lcmm_data)
-```
-
-```{r ESM_Fig5, fig.cap=sfig_plsda_groups}
-plot_plsda_grouping(plsda_results, 'Beta-cell function')
-```
-
-```{r ESM_Fig6, fig.cap=sfig_plsda_loadings, fig.width=7, fig.height=7}
-plot_plsda_loadings(plsda_results)
-```
-
-```{r ESM_Fig7, fig.cap=sfig_plsda_groups_dysgly}
-plot_plsda_grouping(plsda_dysgly, 'Dysglycemia')
-```
-
-```{r ESM_Fig8, fig.cap=sfig_plsda_loadings_dysgly, fig.width=7, fig.height=7}
-plot_plsda_loadings(plsda_dysgly)
-```
-
-```{r ESM_Fig9, fig.cap=sfig_plsda_example}
-plot_plsda_grouping(example_plsda_results(), 'Example groups')
-```
+# References

### 2017-07-01 doc/manuscript.Rmd
results agnostic title

@@
 
-Title: Total non-esterified fatty acids, but not individual fatty acids, predict
-progression in beta-cell dysfunction: The Prospective Metabolism and Islet Cell
+Title: Association of non-esterified fatty acid composition with insulin
+sensitivity and beta-cell dysfunction in the Prospective Metabolism and Islet Cell
 Evaluation (PROMISE) cohort

### 2017-07-01 doc/manuscript.Rmd
homa-ir to homa2

@@
 6 years. Outcome variables were calculated using the OGTT values. IS was
-assessed using 1/HOMA-IR and the Matsuda index, while beta-cell function was
+assessed using HOMA2-%S and the Matsuda index, while beta-cell function was
 assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin
@@
 Insulin sensitivity and beta-cell function indices were computed using the OGTT 
-glucose and insulin data. Insulin sensitivity was assessed using 1/HOMA-IR (1
-divided by HOMA-IR) [@Matthews1985] and the Insulin Sensitivity Index (ISI)
-[@Matsuda1999]. HOMA-IR largely reflects hepatic insulin resistance, while ISI
+glucose and insulin data. Insulin sensitivity was assessed using HOMA2-%S [@Levy1998a]
+and the Insulin Sensitivity Index (ISI)
+[@Matsuda1999]. HOMA2-%S largely reflects hepatic insulin sensitivity, while ISI
 reflects whole-body insulin sensitivity [@AbdulGhani2007].  Beta-cell function
@@
 The primary outcome variables for this analysis were 
-1/HOMA-IR, ISI, IGI/IR, and ISSI-2; outcome variables were log-transformed for
+HOMA2-%S, ISI, IGI/IR, and ISSI-2; outcome variables were log-transformed for
 the statistical modeling. The primary predictor variables for this analysis were
@@
 analysis, we saw no association of any individual or total NEFA with hepatic
-(1/HOMA-IR) or whole-body insulin sensitivity (ISI). We identified
+(HOMA2-%S) or whole-body insulin sensitivity (ISI). We identified
 WC as having a powerful attenuating effect on the associations between

### 2017-07-01 doc/manuscript.Rmd
removed lcmm and plsda wording from keywords

@@
 - GEE: Generalized estimating equations
-- LCMM: Latent curve mixed models
-- PLS-DA: Partial least squares regression discriminant analysis
 - IGI/IR: Insulinogenic index over HOMA-IR

### 2017-07-01 doc/manuscript.Rmd
moved captions for esm into esm file

@@
 
-tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 clinic visits.')
-fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each NEFA in PROMISE participants at the baseline visit (2004-2006).')
+tab_basic <- tab('basic', 'Basic characteristics of the PROMISE participants at each of the three clinic visits.')
+fig_dist <- fig('nefa', 'Concentrations (nmol/mL) of each NEFA in PROMISE participants at the baseline visit (2004-2006).')
 fig_gee <- fig('gee', 'Longitudinal associations of individual NEFA (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
-Fully adjusted generalized estimating equation models.
+Fully adjusted generalized estimating equation models, adjusting for time, WC, sex, ethnicity, baseline age, ALT, and family history of diabetes.
 X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.
-P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association.')
-
-sfig_consort <- sfig('consort', 'CONSORT diagram of sample size at each examination visit.')
-stab_qic <- stab('qic', 'Comparing generalized estimating equation models adjusting for different covariates using Quasi-Likelihood Information Criterion.')
-stab_dist <- stab('nefaDist', 'Raw concentration values (nmol/mL) of each non-esterified fatty acid in the PROMISE cohort at the baseline visit (2004-2006). Values are presented as mean (SD).')
-sfig_corr <- sfig('corr', 'Pearson correlation heatmap of non-esterified fatty acids (nmol/mL) and basic PROMISE participant characteristics for the baseline visit (2004-2006).
-Darkness of the colour indicates the magnitude of the correlation, with orange indicating positive and blue indicating negative correlations.')
-sfig_gee <- sfig('gee', 'Unadjusted generalized estimating equation modeling of the longitudinal association of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort.
-GEE models are only adjusted for time.
-Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
-P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association (*).')
-stab_gee <- stab('gee', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
-GEE models were adjusted for time, sex, ethnicity, baseline age, WC,
-ALT, and family history of diabetes.
-Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
-P-values were adjusted for the BH false discovery rate, with significant (p<0.05) associations indicated by asterisk (*).')
-stab_gee_unadj <- stab('gee_unadj', 'Unadjusted generalized estimating equation models of the longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
-GEE models are only adjusted for time.
-Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.
-P-values were adjusted for the BH false discovery rate, with significant (p<0.05) associations indicated by asterisk.')
+P-values were adjusted for the BH false discovery rate, with the {{largest dot}} representing a significant (p<0.05) association.')
 ```

### 2017-07-01 doc/manuscript.Rmd
removed activity questionnaire description in methods

@@
 Sociodemographic information, including age, sex, and ethnicity, were determined
-using questionnaires administered at each examination. In the lifestyle
-questionnaire, physical activity was determined using a version of the
-Modifiable Activity Questionnaire (MAQ) [@Kriska1990].  The MAQ collects 
-information on leisure and occupational activity, including intensity, 
-frequency, and duration, over the past year.  Each reported activity from the 
-MAQ was weighted by its metabolic intensity allowing for the estimation of 
-MET-hours per week.
+using questionnaires administered at each examination.
 

### 2017-07-02 R/calculate.R
added fun to calc followup time

@@ -120,3 +120,16 @@ extract_gee_estimateCI <- function(data) {
       dplyr::arrange(dep, dplyr::desc(estimate)) %>%
       dplyr::select(dep, indep, estCI)
 }
+
+#' Mean and SD followup time for those who attended the 6 year visit.
+#'
+#' @param data Project data
+#'
+calc_followup_time <- function(data = project_data) {
+    data %>%
+        dplyr::arrange(SID, VN) %>%
+        dplyr::group_by(SID) %>%
+        dplyr::slice(n()) %>%
+        dplyr::ungroup() %>%
+        dplyr::summarise(MeanFollowup = aide::ave_sd(YearsFromBaseline)) %>%
+        .[[1]]

### 2017-07-02 R/calculate.R
default data arg as project_data in calc funs

@@ -19,7 +19,7 @@ calc_conversion_dysgly <- function(variable = c('ConvertDM', 'ConvertPreDM')) {
 #'
 #' @param data project data
 #'
-calc_pct_nefa <- function(data) {
+calc_pct_nefa <- function(data = project_data) {
     over_10pct_nefa <- data %>%
         dplyr::filter(VN == 0) %>%
         dplyr::select(dplyr::matches('pct_ne')) %>%
@@ -38,7 +38,7 @@ calc_pct_nefa <- function(data) {
 #'
 #' @param data project data
 #'
-calc_outcomes_chg_n <- function(data) {
+calc_outcomes_chg_n <- function(data = project_data) {
     prep_data <- data %>%
         dplyr::select(f.VN, HOMA2_S, ISI, IGIIR, ISSI2) %>%
         tidyr::gather(Measure, Value,-f.VN) %>%

### 2017-07-02 R/calculate.R
for dysgly s/conversion/incid

@@ -3,7 +3,7 @@
 #'
 #' @param variable which dysgly variable to use
 #'
-calc_conversion_dysgly <- function(variable = c('ConvertDM', 'ConvertPreDM')) {
+calc_incid_dysgly <- function(variable = c('ConvertDM', 'ConvertPreDM')) {
     variable <- match.arg(variable)
 
     conversion <-
@@ -133,3 +133,4 @@ calc_followup_time <- function(data = project_data) {
         dplyr::ungroup() %>%
         dplyr::summarise(MeanFollowup = aide::ave_sd(YearsFromBaseline)) %>%
         .[[1]]
+}

### 2017-07-02 R/setup.R
add text in figs as vector lines (not font) via showtext

@@ -27,7 +27,7 @@ set_options <- function() {
     # For the document (knitr)
     knitr::opts_chunk$set(
         warning = FALSE, message = FALSE, collapse = TRUE, echo = FALSE,
-        dev = c("cairo_ps", "pdf", "png")
+        dev = c("cairo_ps", "pdf", "png"), fig.showtext = TRUE
     )
 }
 

### 2017-07-02 doc/esm-methods.Rmd
added esm-methods file

---
output: fost::diabetologia_manuscript
---

```{r setup, include=FALSE}
devtools::load_all()
set_options()
knitr::opts_chunk$set(results = "asis")
```

# ESM Methods

## Equations for outcomes

The equations for calculating the outcome variables are listed below. HOMA2-%S was 
calculated using the Excel found at https://www.dtu.ox.ac.uk/homacalculator/.

```{r}
PROMISE.methods::insert_formula("ISI")
```

```{r}
PROMISE.methods::insert_formula("HOMA")
```

```{r}
PROMISE.methods::insert_formula("IGIIR")
```

```{r}
PROMISE.methods::insert_formula("ISSI2")
```

Area under the curve (AUC) for both glucose (G) and insulin (I) was computed
using the trapezoidal rule. ISI~OGTT~ (ISI) uses all the glucose and insulin
values from the oral glucose tolerance (the overall mean).

## Further description of statistical methods

Generalized estimating equations (GEE) is analogous to mixed effects models, 
except it emphasizes population-level trends over individual-level trends that 
mixed models use. Because it can handle autocorrelations in the outcome (as 
would be seen in variables measured over time), GEE modeling is well suited to
longitudinal data. Latent class mixed models (LCMM) is a form of mixed effects
modeling that tries to categorize individuals into groups that have similar
clusters; in the case of the present analysis are the trajectories in beta-cell
function over time. Because mixed models use the participant as the random
effect, LCMM can identify individuals who similarly change over time.

Partial least squares (PLS) regression is a technique similar to principal
component analysis (PCA) that is designed to extract meaningful information from
multivariate, high dimensionality data (e.g. as in metabolomic and other -omic type 
analyses). These types of methods try to extract as much of the variance in the 
data into a smaller number of components or factors. The difference between PCA 
and PLS is that PLS is a supervised method, while PCA is not. This means that
PLS uses a response variable(s), i.e. an outcome or **Y**, to describe the
variation in the predictors (**X**) while PCA only describes the variation 
inherent in **X**. Because of this, PLS can be better at predicting the 
contributions of predictor variables against an outcome variable.

### 2017-07-02 doc/manuscript.Rmd
added caption for unadjusted gee fig

@@
 fig_dist <- fig('nefa', 'Concentrations (nmol/mL) of each NEFA in PROMISE participants at the baseline visit (2004-2006).')
-fig_gee <- fig('gee', 'Longitudinal associations of individual NEFA (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
-Fully adjusted generalized estimating equation models, adjusting for time, WC, sex, ethnicity, baseline age, ALT, and family history of diabetes.
+fig_gee_unadj <- fig('gee_unadj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort.
+Models are only adjusted for time.
+X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.
+P-values were adjusted for the BH false discovery rate, with the {{largest dot}} representing a significant (p<0.05) association.')
+fig_gee_adj <- fig('gee_adj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
+Generalized estimating equation models were fully adjusted.
 X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.

### 2017-07-02 doc/manuscript.Rmd
added to methods on reference to equations

@@
 the disposition index (but using OGTT values). Each index has been validated
-against gold standard measures [@Matthews1985; @Matsuda1999; @Retnakaran2009].
+against gold standard measures [@Levy1998a; @Matsuda1999; @Retnakaran2009]. {{confirm re homa2}}
+Formulas used to calculate these measures can be found in the ESM Methods.
 

### 2017-07-02 doc/manuscript.Rmd
stated other vars that were time independent
added unadj gee fig

@@
 measures (M8). The final GEE model was adjusted for time, sex, ethnicity,
-baseline age, WC, ALT, and family history of diabetes. After scaling,
+baseline age, WC, ALT, and family history of diabetes. Sex, ethnicity, and baseline 
+age were time-*independent* variables. After scaling,
 log-transforming, and exponentiating, the GEE estimates are
@@
 
-```{r figure2_gee, dependson='geeData', fig.cap=fig_gee}
+```{r figure2_gee_unadj, fig.cap=fig_gee_unadj}
+gee_results %>% 
+    dplyr::filter(model == "Unadjusted") %>% 
+    plot_gee_main()
+```
+
+```{r figure2_gee_adj, fig.cap=fig_gee_adj}
 gee_results %>% 

### 2017-07-03 R/plots.R
axis labels reversed

@@ -13,7 +13,7 @@ plot_nefa_distribution <- function() {
 
     nefa_dist %>%
         seer::view_boxplots(dots = TRUE, "Measure", "Value") +
-        ggplot2::labs(x = 'Concentration (nmol/mL)',
-                      y = 'Non-esterified fatty acid') +
+        ggplot2::labs(y = 'Concentration (nmol/mL)',
+                      x = 'Non-esterified fatty acid') +
         graph_theme(ticks = FALSE)
 }

### 2017-07-03 R/gee.R
formatted gee plot to have more spacing

@@ -100,10 +100,11 @@ plot_gee_main <- function(results) {
             xlab = 'Percent difference with 95% CI in the outcomes\nfor each SD increase in fatty acid',
             ylab = 'Non-esterified fatty acids'
             ) +
-        graph_theme(ticks = FALSE) +
+        graph_theme(ticks = FALSE, legend.pos = "right") +
         ggplot2::facet_grid(unit~Yterms, switch = 'y',
                             scales = 'free_y',
-                            space = 'free_y')
+                            space = 'free_y') +
+        ggplot2::theme(panel.spacing = grid::unit(0.6, "lines"))
 }
 
 #' Create a table from the GEE results.

### 2017-07-03 R/calculate.R
s/data.frame/list calc fun

@@ -83,7 +83,7 @@ calc_outcomes_chg_n <- function(data = project_data) {
 calc_npercent <- function(x, group = c('Yes', 'Female', 'European')) {
     nums <- table(x)
     pct <- (nums[group]/sum(nums)) * 100
-    data.frame(
+    list(
         n = nums[group],
         pct = paste0(round(pct, 1), '%'),
         npct = paste0(nums[group], ' (', paste0(round(pct, 1), '%)'))

### 2017-07-03 R/corr.R
removed default for y for corr

@@ -4,7 +4,7 @@
 #' @param x Variables on the X-axis
 #' @param y Variables on the Y-axis
 analyze_corr <- function(x = c(outcomes, 'BMI', 'Waist', 'Age', 'lALT', 'lTAG', 'Chol', 'HDL', 'LDL'),
-                         y = ne_conc) {
+                         y) {
 
     baseline_data <- project_data %>%
         dplyr::filter(VN == 0)

### 2017-07-03 R/gee.R
Added exclusion of diabetes cases

@@ -1,6 +1,9 @@
 #' Get the dataset organized and wrangled into a format for the GEE analysis.
 #'
 prep_gee_data <- function(data) {
+    data <- data %>%
+        dplyr::filter(DM != 1)
+
     data_without_fa <- dplyr::select(data, -dplyr::matches('pct_ne|^ne'), -TotalNE)
 
     data_with_fa_scaled <- data %>%

### 2017-07-03 R/generate_results.R
added corr analysis for both conc and pct

@@ -54,9 +54,13 @@ generate_results_gee <- function() {
 #' Generate and save the results of the correlation analysis as a dataset.
 #'
 generate_results_corr <- function() {
-    corr_results <- analyze_corr()
+    corr_results_conc <- analyze_corr(y = ne_conc)
     # Save output of results into dataset
-    devtools::use_data(corr_results, overwrite = TRUE)
+    devtools::use_data(corr_results_conc, overwrite = TRUE)
+
+    corr_results_pct <- analyze_corr(y = ne_pct)
+    # Save output of results into dataset
+    devtools::use_data(corr_results_pct, overwrite = TRUE)
 }
 
 #' Generate results for LCMM modeling as a dataset.

### 2017-07-03 R/corr.R
corr plot arg for unit of FA

@@ -31,12 +31,12 @@ analyze_corr <- function(x = c(outcomes, 'BMI', 'Waist', 'Age', 'lALT', 'lTAG', 
 #' Plot the correlations into a heatmap.
 #'
 #' @param results Correlation results
-plot_corr_heatmap <- function(results) {
+plot_corr_heatmap <- function(results, unit) {
     results %>%
         seer::view_heatmap(
             y = 'Vars2',
             x = 'Vars1',
-            ylab = 'Non-esterified fatty acids (nmol/mL)',
+            ylab = paste0('Non-esterified fatty acids (', unit, ")"),
             number.colours = 5,
             values.text = FALSE
         ) +

### 2017-07-03 doc/esm-figures.Rmd
esm figures file done

---
output: fost::diabetologia_manuscript
---

```{r setup, include=FALSE}
devtools::load_all()
set_options()

# Include captions below using `captioner` package via library(captioner)
sfig <- captioner::captioner(prefix = 'ESM Fig.')
cite_sf <- pryr::partial(sfig, display = 'cite')

sfig_consort <- sfig('consort', 'CONSORT diagram of sample size at each examination visit.')
sfig_dag <- sfig("dagitty", "Directed acyclic graph taken from DAGitty (DAGitty.net).")
sfig_corr_conc <- sfig('corr_conc', 'Pearson correlation heatmap of non-esterified fatty acids (*nmol/mL*) and basic PROMISE participant characteristics for the baseline visit (2004-2006). Darkness of the colour indicates the magnitude of the correlation, with orange indicating positive and blue indicating negative correlations.')
sfig_corr_pct <- sfig('corr_pct', 'Pearson correlation heatmap of non-esterified fatty acids (*mol%*) and basic PROMISE participant characteristics for the baseline visit (2004-2006). Darkness of the colour indicates the magnitude of the correlation, with orange indicating positive and blue indicating negative correlations.')
```

# ESM Figures

![`r sfig_consort`](../img/flowDiagramSample.png)

![`r sfig_dag`](../img/dagitty.png)

```{r ESM_Fig3, fig.cap=sfig_corr_conc}
plot_corr_heatmap(corr_results_conc, unit = "nmol/mL")
```

```{r ESM_Fig4, fig.cap=sfig_corr_pct}
plot_corr_heatmap(corr_results_pct, unit = "mol%")
```

### 2017-07-03 R/qic.R
moved qic code into own file and funs

@@ -1,0 +1,133 @@
+
+#' Prepare the data for using in the QIC model comparisons.
+#'
+#' @param data Project data
+#'
+prep_qic_data <- function(data) {
+    data %>%
+        prep_gee_data() %>%
+        dplyr::select(
+            SID,
+            YearsFromBaseline,
+            lISSI2,
+            lISI,
+            TotalNE,
+            Sex,
+            BiEthnicity,
+            MET,
+            ALT,
+            BaseAge,
+            Waist,
+            BMI,
+            AlcoholPerWk,
+            TobaccoUse,
+            FamHistDiab,
+            BaseTAG
+        )
+}
+
+#' Clean data output from the QIC model selection functions.
+#'
+#' @param data QIC output data.
+#'
+clean_qic <- function(data) {
+    data %>%
+        tibble::rownames_to_column() %>%
+        dplyr::select(Model = rowname, QIC = IC, Delta = delta) %>%
+        dplyr::mutate(id = seq_len(nrow(.)))
+}
+
+qic_is <- function(data) {
+    qic_data <<- data %>%
+        prep_qic_data() %>%
+        dplyr::select(-lISSI2) %>%
+        stats::na.omit()
+
+    M0 <- geepack::geeglm(
+        lISI ~ TotalNE + YearsFromBaseline,
+        data = qic_data,
+        id = SID,
+        family = stats::gaussian,
+        corstr = 'ar1'
+    )
+
+    M1 <- stats::update(M0, . ~ . + TotalNE:YearsFromBaseline)
+    M2 <- stats::update(M0, . ~ . + BaseAge + BiEthnicity + Sex)
+    M3 <- stats::update(M2, . ~ . + Waist)
+    M4 <- stats::update(M3, . ~ . + ALT)
+    M5 <- stats::update(M3, . ~ . + MET)
+    M6 <- stats::update(M3, . ~ . + AlcoholPerWk)
+    M7 <- stats::update(M3, . ~ . + FamHistDiab)
+    M8 <- stats::update(M3, . ~ . + TobaccoUse)
+    M9 <- stats::update(M3, . ~ . + ALT + MET)
+    M10 <- stats::update(M3, . ~ . + TotalNE:YearsFromBaseline)
+    M11 <- stats::update(M9, . ~ . + TotalNE:YearsFromBaseline)
+
[…]

### 2017-07-03 doc/esm-tables.Rmd
moved esm tables into own file and updated

---
output: fost::diabetologia_manuscript
---

```{r setup, include=FALSE}
devtools::load_all()
set_options()
```

```{r captions}
# Include captions below using `captioner` package via library(captioner)
stab <- captioner::captioner(prefix = 'ESM Table')
cite_st <- pryr::partial(stab, display = 'cite')

stab_qic <- stab('qic', 'Comparing generalized estimating equation models adjusting for different covariates using Quasi-Likelihood Information Criterion.')
stab_dist <- stab('nefa', 'Raw concentration values of each non-esterified fatty acid in the PROMISE cohort at the baseline visit (2004-2006). Values are presented as mean (SD).')
stab_gee_unadj <- stab('gee_unadj', 'Unadjusted generalized estimating equation models of the longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.  GEE models are only adjusted for time. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values were adjusted for the BH false discovery rate, with significant (p<0.05) associations indicated by asterisk.')
stab_gee_adj <- stab('gee_adj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort. GEE models were adjusted for time, sex, ethnicity, baseline age, WC, ALT, and family history of diabetes. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values were adjusted for the BH false discovery rate, with significant (p<0.05) associations indicated by asterisk (*).')
```

# ESM Tables

```{r ESM_Table1, results="asis"}
analyze_qic() %>% 
    table_qic(stab_qic)
```

Given the number of possible combinations of outcome and predictor variables,
only ISI and ISSI-2 with total non-esterified fatty acids (nmol/mL) were used to
compare various GEE models and to select a final model.  Baseline age was used
as including both the original age and the time variable would result in
collinearity.  Column names are: QIC is the quasi-likelihood information criteria
(smaller values, eg. larger negative values, indicate a better fit compared to
other models), Delta is the QIC minus the lowest QIC (models that have a delta <10 
between them are considered equivalent).  Models were:

- M0: log(ISSI-2) or log(ISI) = total non-esterified fatty acids (nmol/mL) + years from baseline
- M1: M0 + NEFA by time interaction
- M2: M0 + sex + ethnicity + baseline age
- M3: M2 + waist
- M4: M3 + ALT
- M5: M3 + physical activity
- M6: M3 + alcohol intake
- M7: M3 + family history of diabetes
- M8: M3 + smoking status
- M9: M3 + ALT + physical activity
- M10: M3 + NEFA by time interaction
- M11: M9 + NEFA by time interaction

```{r ESM_Table2}
table_distribution(project_data, stab_dist)
```

```{r ESM_Table3_4}
gee_results %>% 
    dplyr::filter(model == "Unadjusted") %>% 
    table_gee_main(stab_gee_unadj)
gee_results %>% 
    dplyr::filter(model == "Adjusted") %>% 
    table_gee_main(stab_gee_adj)
```

### 2017-07-03 R/variables.R
Updated model variables based on qic

@@ -2,4 +2,4 @@ outcomes <- c('lHOMA2_S', 'lISI', 'lIGIIR', 'lISSI2')
 ne_pct <- grep('^pct_ne', vars, value = TRUE)
 ne_conc <- grep('^ne\\d\\d', vars, value = TRUE)
 ne_total <- "TotalNE"
-covariates <- c("YearsFromBaseline", 'Waist', 'Sex', 'BiEthnicity', 'BaseAge', 'ALT', 'FamHistDiab')
+covariates <- c("YearsFromBaseline", 'Waist', 'Sex', 'BiEthnicity', 'BaseAge', 'ALT', 'MET')

### 2017-07-03 R/fetch_data.R
roxygen comment in wrong place

@@ -11,13 +11,13 @@ fetch_data <- function() {
     # Load the master dataset,
     ds.prep <- PROMISE::PROMISE %>%
         dplyr::filter(VN %in% c(1, 3, 6)) %>%
-        ## Kick out Canoers
+        # Kick out Canoers
         dplyr::filter(is.na(Canoe)) %>%
         dplyr::tbl_df()
 
     print(paste0('Original dataset rows are ', dim(ds.prep)[1], ' and columns are ', dim(ds.prep)[2]))
 
-    ##' Munge and wrangle the data into the final version.
+    # Munge and wrangle the data into the final version.
     ds <- ds.prep %>%
         dplyr::select(
             SID, VN, BMI, Waist, HOMA, ISI, IGIIR, ISSI2, TAG, LDL, HDL, Chol,

### 2017-07-03 R/pls.R
removed export from pls

@@ -8,7 +8,6 @@
 #' @param y Y or outcome variable.
 #' @param x X or exposure variables of interest.
 #'
-#' @export
 prep_pls_data <- function(data, y, x) {
     data %>%
         dplyr::filter(VN == 0) %>%
@@ -26,7 +25,6 @@ prep_pls_data <- function(data, y, x) {
 #' @param ncomp Number of components.
 #' @param cv Whether to use CV.
 #'
-#' @export
 analyze_pls <- function(data, y, x = ne_pct, ncomp = NULL, cv = TRUE) {
     data %>%
         prep_pls_data(y = y, x = x) %>%
@@ -44,7 +42,6 @@ analyze_pls <- function(data, y, x = ne_pct, ncomp = NULL, cv = TRUE) {
 #'
 #' @param data PLS output results.
 #'
-#' @export
 plot_pls <- function(data) {
     seer::view_pls_xloadings(data, renaming.x = renaming_fa, dot.colour = "black") +
         graph_theme(minor.grid.lines = FALSE)
@@ -58,7 +55,6 @@ plot_pls <- function(data) {
 #' @param test The test data to predict from.
 #' @param ncomps The component to predict on.
 #'
-#' @export
 calc_pred_corr <- function(model, test, ncomps = 1) {
     predicted <- stats::predict(model, ncomp = ncomps, newdata = test)
     measured <- as.matrix(stats::model.response(

### 2017-07-03 R/setup.R
removed export and example from set_options

@@ -8,12 +8,6 @@
 
 #' Set options for all documents and scripts.
 #'
-#' @export
-#'
-#' @examples
-#'
-#' set_options()
-#'
 set_options <- function() {
     # Set the options here for individual packages
 

### 2017-07-05 doc/manuscript.Rmd
updated wording and covariates of gee model in methods

@@
 Covariate selection was based on previous literature, directed acyclic graph 
-[@Greenland1999a] recommendations, and quasi-likelihood information criterion 
-(QIC). `r cite_st('qic')` shows the covariates compared using QIC. While the
-final GEE model selected as best fitting differed between insulin sensitivity
-and beta-cell function measures, the differences in QIC values were less than
-10 between the models, suggesting similar fit. As such, we selected the model
+[@Greenland1999a] recommendations (see `r cite_sf("dag")`), and quasi-likelihood information criterion 
+(QIC). `r cite_st('qic')` shows the covariates compared using QIC. To reduce
+complexity between models using either insulin sensitivity or beta-cell function
+measures, only the model that had similar good fit for both was used. 
+As such, we selected the model
 that had the fewest covariates and that had similar fit between the outcome
-measures (M8). The final GEE model was adjusted for time, sex, ethnicity,
-baseline age, WC, ALT, and family history of diabetes. Sex, ethnicity, and baseline 
+measures (M9). The final GEE model was adjusted for years from baseline (time),
+sex (female or male), ethnicity (European ancestry or non-European ancestry),
+baseline age, WC, ALT, and MET (physical activity). Sex, ethnicity, and baseline 
 age were time-*independent* variables. After scaling,
@@
 interpreted as an expected percent difference in the outcome variable for every
-standard deviation (SD) increase in the predictor variable given the covariates are held constant.
+standard deviation (SD) increase in the predictor variable given the covariates
+are held constant.
 Since TAG is a risk factor for diabetes and since NEFA contribute to TAG

### 2017-07-05 doc/manuscript.Rmd
finished pls sentence in methods

@@
 analyses in a multivariate environment (all NEFA variables in a single model),
-partial least squares
+partial least squares (PLS) was used with the outcomes as the independent variable.
 

### 2017-07-05 doc/manuscript.Rmd
updated inline results funs and updated basic char results

@@
 ```{r inline_results}
-pct_nefa <- calc_pct_nefa(project_data)
+nefa <- calc_pct_nefa(project_data)
 chg_outcomes <- calc_outcomes_chg_n(project_data)
-dm_incid <- calc_conversion_dysgly('ConvertDM')
-predm_incid <- calc_conversion_dysgly('ConvertPreDM')
 
@@
 famhist <- calc_npercent(ds0yr$FamHistDiab, 'Yes')
-est_gee <- calculate_gee_magnitude(gee_results)
+est_gee <- gee_results %>% 
+    dplyr::filter(model == "Adjusted") %>% 
+    calc_gee_magnitude()
 ```
@@
 For this study, there were 
-`r sex.npct$npct` females and `r ethn.npct$npct` had European-ancestry.
+`r female$npct` females and `r european$npct` had European-ancestry (see 
+`r cite_t("basic")`). The mean follow-up time was `r calc_followup_time()`
+years, where `r calc_pct_full_visits()`% of participants attended all three
+visits.
 The average age of the participants was `r aide::average(ds0yr$Age)` years 
@@
 `r aide::average(ds0yr$BMI)` (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
-participants, `r fhd.npct$npct`, had a family history of diabetes.
+participants, `r famhist$npct`, had a family history of diabetes.
 Insulin sensitivity and beta-cell function measures showed a significant median
 decline between
-`r percent.change.outcomes$chg` (p`r percent.change.outcomes$p`) over the 6
+`r pct_outcomes$chg` (p`r pct_outcomes$p`) over the 6
 years in this analysis of the PROMISE cohort.
-Consistent with this decline, there were `r dm.convert`
-participants who developed diabetes while `r predm.convert` participants either
-converted to or maintained IFG or IGT status over the 6 years.
-<!-- maintain vs convert -->
+Consistent with this decline, there were `r calc_incid_dysgly("ConvertDM")`
+and `r calc_incid_dysgly("ConvertPreDM")`incident cases of diabetes and pre-diabetes (IFG and IGT),
+respectively, over the 6-years. Incident diabetes cases were excluded from further GEE
+analyses.
 
-`r cite_f('nefaDist')` shows the compositional distribution of NEFA in the study
+`r cite_f('nefa')` shows the compositional distribution of NEFA in the study
 participants. Four individual NEFA made up the vast majority 
-(`r sum(percent.nefa$pct)`%) of the total NEFA fraction. The largest contributors
-were `r percent.nefa$c[1]`, `r percent.nefa$c[2]`, `r percent.nefa$c[3]`, and 
-`r percent.nefa$c[4]`. Raw concentration values are shown in `r cite_st('nefaDist')`. 
+(`r sum(nefa$pct)`%) of the total NEFA fraction. The largest contributors
+were `r nefa$c[1]`, `r nefa$c[2]`, `r nefa$c[3]`, and 
+`r nefa$c[4]`. Raw concentration values are shown in `r cite_st('nefa')`. 
+There were no differences in individual NEFA for those who did compared to those
+who did not take medications to manage blood lipids (data not shown).
 All individual NEFA as well as the total fraction had
 correlations that ranged from weak to null (r<0.3) with participant
-characteristics (see `r cite_sf('corr')`).
+characteristics (see `r cite_sf('corr_conc')` and `r cite_sf("corr_pct")`).
 

### 2017-07-05 doc/manuscript.Rmd
updated gee model results section

@@
 in any of these three NEFA variables, there was an average predicted
-`r est.gee$est[1]`% lower IGI/IR and `r est.gee$est[2]`% lower ISSI-2 at each
+`r est_gee$est[1]`% lower IGI/IR and `r est_gee$est[2]`% lower ISSI-2 at each
 clinic visit. However, in contrast to these results using NEFA modeled as
@@
 the outcomes. Adjusting for TAG attenuated all associations with IGI/IR and
-ISSI-2 (data not shown). Sensitivity analyses to examine the association with
+ISSI-2 (data not shown), though this may be due to the fact that only a subset
+of participants at the 6 year visit had TAG measured. Sensitivity analyses to
+examine the association with
 the sum of fatty acid classes (saturated vs unsaturated) revealed similar
@@
 There were no significant interaction effects between time, sex, or ethnicity
-and the individual NEFA on any of the outcome measures, except for a single
-marginally significant (p=0.04) interaction between sex, 20:3n-6 (mol%), and
-ISI, which showed that males had a 12% higher ISI for every SD increase in
-20:3n-6 compared to females. Raw values from the GEE models are shown in `r cite_st('gee')` for
+and the individual NEFA on any of the outcome measures.
+Raw values from the GEE models are shown in `r cite_st('gee')` for
 fully adjusted models and `r cite_st('gee_unadj')` for unadjusted models.
+Consistent with the GEE modeling, results from modeling all NEFA in a
+multivariate environment using PLS showed no discernible clustering nor was it
+predictive of the outcomes (data not shown).
 

### 2017-07-05 R/calculate.R
added calc fun to calc percent returned for visit

@@ -134,3 +134,25 @@ calc_followup_time <- function(data = project_data) {
         dplyr::summarise(MeanFollowup = aide::ave_sd(YearsFromBaseline)) %>%
         .[[1]]
 }
+
+#' Calculate the number of participants who attended one, two, or three visits.
+#'
+#' @param data Project data.
+#'
+calc_n_for_visits <- function(data = project_data) {
+    data %>%
+        dplyr::select(SID, VN, f.VN) %>%
+        stats::na.omit() %>%
+        tidyr::spread(f.VN, VN) %>%
+        dplyr::mutate(Visits = paste(yr0, yr3, yr6, sep = '-')) %>%
+        .$Visits %>%
+        table()
+}
+
+
+#' Percent of participants who attended each of the 3 visits.
+#'
+calc_pct_full_visits <- function() {
+    pct_val <- calc_n_for_visits()[1] / sum(calc_n_for_visits())
+    precise_rounding(pct_val * 100)
+}

### 2017-07-11 doc/esm-figures.Rmd
updated esm figure file with dagitty images

@@
 sfig_consort <- sfig('consort', 'CONSORT diagram of sample size at each examination visit.')
-sfig_dag <- sfig("dagitty", "Directed acyclic graph taken from DAGitty (DAGitty.net).")
+sfig_dag_isi <- sfig("dag_isi", "Directed acyclic graph taken from DAGitty (DAGitty.net) for the potential causal pathway between NEFA and ISI.")
+sfig_dag_issi2 <- sfig("dag_issi2", "Directed acyclic graph taken from DAGitty (DAGitty.net) for the potential causal pathway between NEFA and ISSI-2.")
 sfig_corr_conc <- sfig('corr_conc', 'Pearson correlation heatmap of non-esterified fatty acids (*nmol/mL*) and basic PROMISE participant characteristics for the baseline visit (2004-2006).
@@
 
-![`r sfig_dag`](../img/dagitty.png)
+![`r sfig_dag_isi`](../img/dagitty-isi.png)
+
+![`r sfig_dag_issi2`](../img/dagitty-issi2.png)
 

### 2017-07-11 R/setup.R
don't need to export pipe

@@ -26,5 +26,4 @@ set_options <- function() {
 }
 
 #' @importFrom magrittr %>%
-#' @export
 `%>%` <- magrittr::`%>%`

### 2017-07-11 R/rfigshare.R
renamed rfigshare fun

@@ -4,7 +4,7 @@
 # 3. git commit the files
 # 4. git archive --format=zip --output=code-archive.zip code
 # 5. Then run the below code
-send_to_new_figshare <- function() {
+send_to_figshare_new <- function() {
     rfigshare::fs_new_article(
         "Analysis code for poster on triacylglycerol DNL on components of the Metabolic Syndrome in PROMISE",
         "Analysis code for a poster for the American Diabetes Association Scientific Sessions in 2017",

### 2017-07-11 doc/manuscript.Rmd
removed top outline and added set_options

@@
 
-<!-- outline
-
-TODO:
-- Bootstrap for the longitudinal table data... like instead of an ANOVA/GEE for
-the p-value...?
-- What about adiponectin?
-- Precise followup time for each participant?
-- [X] Interaction by sex or ethnicity
-- [X] Interaction by time
-- [X] Sensitivity analysis with TAG
-- [X] PLS-DA with dysglycemia (for anyone who got it over the years).
-- [X] Replace TAG with BaseTAG in the QIC (or use both)
-- [X] Fix time axis labels on LCMM plot
-- Describe LCMM and PLS-DA better in plots
-- [X] Use pct for PLSDA
-- [X] Explained variation in X and in Y for PLS-DA
-- Supplemental only file (separate from manuscript).. maybe?
-
--->
-
 ```{r setup, echo=FALSE}
 devtools::load_all('.')
+set_options()
 ```

### 2017-07-11 doc/manuscript.Rmd
s/pct/chg

@@
 decline between
-`r pct_outcomes$chg` (p`r pct_outcomes$p`) over the 6
+`r chg_outcomes$chg` (p`r chg_outcomes$p`) over the 6
 years in this analysis of the PROMISE cohort.

### 2017-07-11 doc/manuscript.Rmd
clarified esm fig refs

@@
 A number of associations were seen in the unadjusted GEE models 
-(see `r cite_sf('gee')`), particularly for NEFA modeled as a concentration.
-While full model adjustment (`r cite_f('gee')`) attenuated most of these
+(see `r cite_f('gee_unadj')`), particularly for NEFA modeled as a concentration.
+While full model adjustment (`r cite_f('gee_adj')`) attenuated most of these
 associations, total NEFA, 16:0, 18:1n-9, and 18:2n-6 (all as nmol/mL)
@@
 and the individual NEFA on any of the outcome measures.
-Raw values from the GEE models are shown in `r cite_st('gee')` for
-fully adjusted models and `r cite_st('gee_unadj')` for unadjusted models.
+Raw values from the GEE models are shown in `r cite_st('gee_unadj')` for
+unadjusted models and `r cite_st('gee_adj')` for fully adjusted models.
 Consistent with the GEE modeling, results from modeling all NEFA in a

### 2017-07-11 doc/esm-figures.Rmd
added png dagitty images, updated esm-figs with dagitty

--- 2017-07-03 doc/esm-figures.Rmd
@@
 sfig_consort <- sfig('consort', 'CONSORT diagram of sample size at each examination visit.')
-sfig_dag <- sfig("dagitty", "Directed acyclic graph taken from DAGitty (DAGitty.net).")
+sfig_dag_isi <- sfig("dag_isi", "Directed acyclic graph taken from DAGitty (DAGitty.net) for the potential causal pathway between NEFA and ISI.
+DAGitty recommended a minimal adjustment of waist, sex, and MET (physical activity).")
+sfig_dag_issi2 <- sfig("dag_issi2", "Directed acyclic graph taken from DAGitty (DAGitty.net) for the potential causal pathway between NEFA and ISSI-2.
+DAGitty recommended a minimal adjustment of waist, ethnicity, and MET (physical activity).")
 sfig_corr_conc <- sfig('corr_conc', 'Pearson correlation heatmap of non-esterified fatty acids (*nmol/mL*) and basic PROMISE participant characteristics for the baseline visit (2004-2006).
@@
 
-![`r sfig_dag`](../img/dagitty.png)
+![`r sfig_dag_isi`](../img/dagitty-isi.png)
 
-```{r ESM_Fig3, fig.cap=sfig_corr_conc}
+![`r sfig_dag_issi2`](../img/dagitty-issi2.png)
+
+```{r ESM_Fig4, fig.cap=sfig_corr_conc}
 plot_corr_heatmap(corr_results_conc, unit = "nmol/mL")
@@
 
-```{r ESM_Fig4, fig.cap=sfig_corr_pct}
+```{r ESM_Fig5, fig.cap=sfig_corr_pct}
 plot_corr_heatmap(corr_results_pct, unit = "mol%")

### 2017-07-11 doc/manuscript.Rmd
sample.outcome -> pct_outcome$n
removed extra empty lines
s/echo/include
inline results of significant FA from GEE

@@
 
-```{r setup, echo=FALSE}
+```{r setup, include=FALSE}
 devtools::load_all('.')
@@
     calc_gee_magnitude()
+
+sig_fa_unadj <- gee_results %>% 
+    dplyr::filter(model == "Unadjusted", p.value < 0.05) %>% 
+    dplyr::select(Yterms, Xterms, unit) %>% 
+    dplyr::mutate(Xterms = ifelse(Xterms == "Total", "total NEFA", as.character(Xterms)))
+sig_fa_adj <- gee_results %>% 
+    dplyr::filter(model == "Adjusted", p.value < 0.05) %>% 
+    dplyr::select(Yterms, Xterms, unit) %>% 
+    dplyr::mutate(Xterms = ifelse(Xterms == "Total", "total NEFA", as.character(Xterms)))
 ```
@@
 and sex groups did not change over the 6 years. Outcome variables sample size
-ranges from `r sample.outcomes` over the 3 visits.
+ranges from `r chg_outcomes$n` over the 3 visits.
 

### 2017-07-18 R/qic.R
added meds covar to qic model selection

@@ -22,7 +22,8 @@ prep_qic_data <- function(data) {
             AlcoholPerWk,
             TobaccoUse,
             FamHistDiab,
-            BaseTAG
+            BaseTAG,
+            MedsLipidsChol
         )
 }
 
@@ -60,10 +61,11 @@ qic_is <- function(data) {
     M7 <- stats::update(M3, . ~ . + FamHistDiab)
     M8 <- stats::update(M3, . ~ . + TobaccoUse)
     M9 <- stats::update(M3, . ~ . + ALT + MET)
-    M10 <- stats::update(M3, . ~ . + TotalNE:YearsFromBaseline)
-    M11 <- stats::update(M9, . ~ . + TotalNE:YearsFromBaseline)
+    M10 <- stats::update(M3, . ~ . + MedsLipidsChol)
+    M11 <- stats::update(M3, . ~ . + TotalNE:YearsFromBaseline)
+    M12 <- stats::update(M9, . ~ . + TotalNE:YearsFromBaseline)
 
-    MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M9, M10, M11,
+    MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M9, M10, M11, M12,
                      rank = MuMIn::QIC) %>%
         clean_qic()
 }
@@ -92,10 +94,11 @@ qic_bcf <- function(data) {
     M7 <- stats::update(M3, . ~ . + FamHistDiab)
     M8 <- stats::update(M3, . ~ . + TobaccoUse)
     M9 <- stats::update(M3, . ~ . + ALT + MET)
-    M10 <- stats::update(M3, . ~ . + TotalNE:YearsFromBaseline)
-    M11 <- stats::update(M9, . ~ . + TotalNE:YearsFromBaseline)
+    M10 <- stats::update(M3, . ~ . + MedsLipidsChol)
+    M11 <- stats::update(M3, . ~ . + TotalNE:YearsFromBaseline)
+    M12 <- stats::update(M9, . ~ . + TotalNE:YearsFromBaseline)
 
-    MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M9, M10, M11,
+    MuMIn::model.sel(M0, M1, M2, M3, M4, M5, M6, M7, M8, M9, M10, M11, M12,
                      rank = MuMIn::QIC) %>%
         clean_qic()
 }

### 2017-07-18 doc/esm-tables.Rmd
updated models for qic in esm-tables

@@
 - M9: M3 + ALT + physical activity
-- M10: M3 + NEFA by time interaction
-- M11: M9 + NEFA by time interaction
+- M10: M3 + medication use for lipid and cholesterol control
+- M11: M3 + NEFA by time interaction
+- M12: M9 + NEFA by time interaction
 

### 2017-07-18 doc/submission/reviewer-response.md
edits to reviewer response (incomplete, one left)

@@
 analysis) and defined by glycemic status as outcomes. While the primary analysis
-is well-described and easy to follow, **these additional analyses are described
+is well-described and easy to follow, these additional analyses are described
 mostly in the online appendix and much harder to interpret for the general
@@
 the conclusion, and I would move these to the online appendix or drop them from
-the paper.** I think that the paper needs a major revision however both the
+the paper. I think that the paper needs a major revision however both the
 dataset and the study questions are adequate and important thus conditional
@@
 
-    Given that GEE models do not take into account the high dimensionality of
+    We'd like to thank the reviewer for this suggestion. Given that GEE models
+    do not take into account the high dimensionality (i.e. intercorrelation) of
     the NEFA composition data, we ran the PLS analyses in order to confirm the
     GEE results when using a multivariate approach. Use of the LCMM analysis
-    allowed us to take include the longitudinal data in the PLS analysis.
+    allowed us to include the longitudinal data in the PLS analysis.
     However, we agree with the reviewer that these analyses, while interesting
-    in their confirmation of the GEE findings, do not contribute to the overall
-    manuscript. In light of this reviewer's comments, we decided to move {{or
-    drop?}} the analyses into the ESM, including the description of the methods
-    and results.
+    in their confirmation of the GEE findings, do not contribute substantially
+    to the overall manuscript. In light of this reviewer's comments, we decided
+    to drop these analyses from the paper to keep the overall message of the
+    manuscript clearer and more focused.
 
@@
 
-    As suggested, we've made the title results agnostic.
+    As suggested, we've made the title agnostic.
 
@@
 
-    We have made the requested changes to the methods and stated the exact
-    analyses used more explicitly. Incident diabetes cases are presented only
-    for descriptive purposes and were not included in any further analysis. We
-    thank the reviewer for the suggestion and agree that using time since
-    baseline is a better variable to adjust for; the time variable has been
-    replaced in GEE models, though the change does not substantially impact the
-    final results. {{confirm}}
+    We have made as many of the requested changes to the methods as the reviewer
+    suggested without going over the word limit. Incident diabetes cases are
+    presented for descriptive purposes though were not included in any GEE
+    analysis. We thank the reviewer for the suggestion regarding time adjustment
+    and agree that using time since baseline is a better variable to adjust for
+    in the GEE models; the time variable has been replaced in GEE models, though
+    the change does not substantially impact the final results.
 
@@
     have included the unadjusted results in the main manuscript and have 
-    described it in the results and discussion sections, in addition to making
-    the changes to the abstract.
+    described it in the results and discussion sections, in addition to briefly
+    describing them in the abstract (keeping within the word limit).
 
@@
     We have included a sensitivity analysis identifying which exact covariate
-    attenuates the results between the unadjusted and adjusted models. {{briefly
-    describe the results here}}.
+    attenuates the results between the unadjusted and adjusted models; the
+    covariate is waist circumference. While inclusion of this variable
+    attenuates most of the associations, we believe that adjustment of this
+    variable is necessary given that biologically NEFA come from the adipose
+    tissue during fasting and hypothetically more adipose tissue would indicate
+    higher NEFA. Since other aspects of adipose tissue (e.g. role in
+    inflammation and appetite regulation) could also impact insulin sensitivity
+    and beta-cell function, we believe it is an important confounder to adjust
+    for. We have included in the conclusions a discussion on how adjustment of
+    waist circumference influences the model and the interpretation.
 
@@
         We have clarified how we dealt with incident diabetes cases. Briefly, we
-        only report incident diabetes and dysglycemia cases in the text, but in
-        the modeling these were excluded. {{not sure how to address a) and b)...}}
+        only report incident diabetes and dysglycemia cases in the text to
+        describe the cohort, but in statistical analyses and modeling, these
+        cases were excluded. Regarding a) within the main PROMISE cohort,
+        prevalent diabetes cases were not excluded from the study as they
+        continued to be followed; prevalent diabetes cases did not have fatty
+        acids measured. For b) any participant is followed until they request to
+        be removed from the study; some participants missed the 3-yr visit, but
+        came in for the 6-yr visit. We have included wording within the results
+        on the sample and percentage who attended at least two visits andLastly,
+        c) because diabetes can effect the values of the OGTT-derived indices
+        due to medication the patient is taking and due to diabetes itself
+        impacting beta-cell function and insulin resistance, we excluded them
+        from all analyses; often, these participants did not even complete an
+        OGTT since it is discouraged that these patients consume that much
+        glucose. To restate, diabetes cases were not included in GEE or other
+        analyses.
 
@@
 
-        We have included the equations for the outcome measures as requested. In
-        addition, we replaced HOMA-IR with HOMA2-%S from the HOMA2 Calculator.
-        The results were not substantially different after the replacement of
-        HOMA-IR. {{confirm}}
+        Thank you for this suggestion. We have included the equations for the
+        outcome measures as requested. In addition, we replaced HOMA-IR with
+        HOMA2-%S from the HOMA2 Calculator. The results were not substantially
+        different after the replacement of HOMA-IR.
 
@@
 
-        These have been corrected.
+        Thanks for catching these issues. They have been corrected in the
+        revised manuscript.
 
@@
 
-        {{bit confused about this one. Will need to look into the wording in the
-        manuscript.}}
+        We are not entirely clear on the reviewer's comment. PLS extracts from
+        high dimensional data (multiple predictor variables, i.e. the fatty
+        acids) underlying correlations, constrained by a outcome variable (i.e.
+        beta-cell function). In this context, because mol% data is by definition
+        restricted to sums of 100% (all fatty acid values add up to 100%), the
+        underlying correlations between fatty acids is dependent on the relative
+        contributions those fatty acids have to the total and as such can
+        reflect correlations of specific fatty acids or groups of fatty acids
+        that increase or decrease in proportion together. The concentration data
+        is not contrained by this inherent correlation and thus underlying groups 
+        or clusters may not be identified using PLS.
 
@@
 
-        {{not completely sure what is meant here}} The changes over time as
-        reported in the results of the outcome variables are simple median
-        differences between the baseline and 6-year visit. The p-value {{confirm
-        this is added}} representing the difference was computed using an
-        unadjusted GEE model.
+        The changes over time as reported in the results of the outcome
+        variables are simple median differences between the baseline and 6-year
+        visit. The p-value representing the difference was computed using an
+        unadjusted (time only) GEE model.
 
@@
 
-        The reviewer makes an excellent point regarding the interaction check in
-        model building. We have incorporated this change in the analysis and
+        The reviewer makes an excellent point regarding the importance of
+        conducting an interaction check in the early stages of
+        model building. We have incorporated this new approach in the analysis and
         report on the results in the paper. In simple terms, the reviewer is
@@
         computation of the GEE estimates. This nuance provides substantially
-        more power to the model and the proceeding results.
+        more power to the model and the proceeding results. See ref {{num}} in
+        the manuscript for more detail on this statistical technique. Within
+        the manuscript we have expanded on this nuance to clarify the interpretation.
 
@@
 
-        We have included further mention of potential over-adjustment to the
-        paper. {{where would this go exactly? throughout the results and
-        discussion?}}
+        To clarify, the model building processes used two additional techniques:
+        quasi-likelihood information criteria (QIC) and directed acyclic graphs.
+        The results and output of these techniques can be found in the ESM.
+        Combining all of these processes in building the model ensures that
+        adjustment is empirically based and reduces (but not eliminates) the
+        potential for over-adjustment. However, we have added additional
+        sensitivity analyses to identify covariates that strongly influence the
+        results and discuss the results in light of these findings.
 
@@
         Given this reviewer's previous comments and suggestions, we have decided
-        to focus on the GEE analysis in the main paper and move the PLS-DA to
-        the ESM and expand on {{or remove? it seems this reviewer prefers if we
-        remove it. I could simplify it by doing what I did in the TAG as a
-        confirmation, and include it in the ESM}} it there.
+        to: focus on the GEE analysis in the main paper; remove the LCMM
+        analysis; perform the PLS analysis on the baseline (cross-sectional)
+        data using the outcomes as continuous variables; and, briefly describe
+        the findings in the Results section. To keep the ESM material
+        streamlined, we have contained the PLS results and other additional
+        analyses conducted within the code (accessible via the DOI) for the
+        interested reader. We believe this simplifies the results and primary
+        findings of this paper without sacrificing a loss in the potential use
+        these additional analyses may provide for the interested researcher.
 
@@
 
-        As requested, we have included the DAG underlying that hypothesized
+        As requested, we have included the DAG underlying the hypothesized
         associations in the ESM.
@@
 
-    {{do I need to include this?}} We thank the associate editor for the kind
-    comments. We have included and discussed the cited paper in the discussion
-    section.
+    We thank the associate editor for the kind comments. We have included and
+    discussed the cited paper in the discussion section.

### 2017-07-18 doc/manuscript.Rmd
updated captions for gee figs

@@
 X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.
-P-values were adjusted for the BH false discovery rate, with the {{largest dot}} representing a significant (p<0.05) association.')
+P-values were adjusted for the BH false discovery rate, with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, p<0.01, and p<0.001.')
 fig_gee_adj <- fig('gee_adj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
@@
 X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.
-P-values were adjusted for the BH false discovery rate, with the {{largest dot}} representing a significant (p<0.05) association.')
+P-values were adjusted for the BH false discovery rate, with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, and p<0.01.')
 ```

### 2017-07-18 doc/manuscript.Rmd
added inline code for meds use
clarified outcomes change meaning and use of gee

@@
 
+meds_use <- table(project_data$MedsLipidsChol, project_data$VN)[2, c(1, 3)]
+
 sig_fa_unadj <- gee_results %>% 
@@
 Insulin sensitivity and beta-cell function measures showed a significant median
-decline between
-`r chg_outcomes$chg` (p`r chg_outcomes$p`) over the 6
-years in this analysis of the PROMISE cohort.
+decline from baseline to 6-years between
+`r chg_outcomes$chg` (p`r chg_outcomes$p`, based on a time-adjusted GEE model)
+in this analysis.
 Consistent with this decline, there were `r calc_incid_dysgly("ConvertDM")`

### 2017-07-19 doc/manuscript.Rmd
Added phys act calculation to methods

@@
 Sociodemographic information, including age, sex, and ethnicity, were determined
-using questionnaires administered at each examination.
+using questionnaires administered at each examination. In the lifestyle 
+questionnaire, physical activity was determined using a version of the 
+Modifiable Activity Questionnaire (MAQ) [@Kriska1990]. The MAQ collects 
+information on leisure and occupational activity, including intensity, 
+frequency, and duration, over the past year. Each reported activity from the 
+MAQ was weighted by its metabolic intensity allowing for the estimation of 
+MET-hours per week.
 

### 2017-07-19 doc/submission/reviewer-response.md
updated reviewer response based on requested analyses

@@
 
-        We agree with the reviewer's suggestion and ran additional sensitivity
-        analyses stratified by sex and ethnicity. We mention the results of this
-        analysis in the results.
+        We agree with the reviewer's comment and ran stratified models by either
+        sex or ethnicity. In both cases, all significant associations from the
+        non-stratified analyses were attenuated, likely due to the reduction in
+        sample size in the stratified models. Stratifying analyses has the
+        limitation of substantially reducing sample size and is the reason why
+        interaction testing is prefered. We ran interaction tests by sex and
+        ethnicity and found no significant interactive associations. While
+        these results internally suggest no differences between sex or ethnicity,
+        our cohort was not designed to answer these research questions and the
+        results presented in the paper may still not be generalizable to other
+        populations. {{yea?}}
 
@@
 
-        As requested, we have included the DAG underlying the hypothesized
+        As requested, we have included the DAGs underlying the hypothesized
         associations in the ESM.
@@
         covariates as time-independent, to match the NEFA variables in the GEE
-        model. We found {{state findings}}.
+        model. We found that there were negligible differences in the results
+        between models with covariates as either time-dependent and
+        time-independent and models with all covariates as time-independent.
 
@@
 
-    We thank the associate editor for the kind comments. We have included and
-    discussed the cited paper in the discussion section.
+    We thank the associate editor for the kind comments. We have included the
+    cited paper in the discussion section.

### 2017-07-20 doc/submission/reviewer-response.md
updated response to med use question

@@
     on medications. However, for those who were taking medications that could
-    influence the results, sensitivity analyses revealed they did not
-    substantially influence NEFA concentrations, the outcome measures, nor the
-    GEE model results {{need to run/confirm this}}. These results have been
-    appended to the ESM {{yes?}}.
+    influence the results, additional analyses revealed that medication use did
+    not substantially influence the predictor or outcome measures, nor did 
+    medication use substantially contribute to GEE model fit compared to models 
+    without it (see QIC table).
 

### 2017-07-20 doc/manuscript.Rmd
made edits to abstract

@@
 # Abstract
-<!--243 / 250 words-->
 
@@
 
-Methods: In the PROMISE cohort, 477 participants had serum
-NEFA measured at the baseline visit and completed an OGTT at 3 time points over
-6 years. Outcome variables were calculated using the OGTT values. IS was
-assessed using HOMA2-%S and the Matsuda index, while beta-cell function was
+Methods: In the PROMISE cohort, 477 participants had serum NEFA (mol% and
+concentration) measured at the baseline visit and completed an OGTT at 3 time
+points over 6 years. Outcome variables were calculated using the OGTT values. IS
+was assessed using HOMA2-%S and the Matsuda index, while beta-cell function was 
 assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin 
@@
 used, adjusting for time, waist, sex, ethnicity, baseline age, alanine 
-aminotransferase, and family history of diabetes. NEFA were analyzed as both
+aminotransferase, and physical activity. NEFA were analyzed as both 
 concentrations (nmol/mL) and proportions (mol%) of the total fraction.
 
-Results: IS and beta-cell function declined by 14-27% while BMI did not
-significantly change over 6 years of follow-up. Total NEFA, 16:0, 18:1n-9, and
-18:2n-6 as concentrations had significant negative associations with IGI/IR and
-ISSI-2, predicting between 4-8% lower IGI/IR and ISSI-2 over the 6 years.
-However, individual NEFA modeled as mol% showed no associations with these
-outcomes. Total and individual NEFA were not associated with IS measures.
+Results: Participants (73% female, 70% with European-ancestry) had their IS and
+beta-cell function decline by 14-27% over 6 years of follow-up. While in
+unadjusted models there were several NEFA (e.g. 18:1n-7, 20:1n-9, 22:4n-6) that
+had negative associations with insulin sensitivity, nearly all of these
+associations were attenuated in fully adjusted models. In adjusted models, total
+NEFA, 16:0, 18:1n-9, and 18:2n-6 (as concentrations) had significant negative
+associations with IGI/IR and ISSI-2, while only 20:5n-3 (as mol%) had positive
+associations with HOMA2-%S.
 

### 2017-07-20 doc/manuscript.Rmd
minor removal of extra commented code

@@
 cite_st <- pryr::partial(stab, display = 'cite')
-# usage: cite_st('basicChar')
 

### 2017-07-20 doc/manuscript.Rmd
mainly reformating paragraphs (added paris ref too)

@@
 pancreatic beta-cell production of insulin [@Rachek2014a;@Cnop2008a]. 
-Observational and clinical studies have reported concordant findings, showing
-in particular that elevated total plasma NEFA associates with an increased risk
-for incident type 2 diabetes [@Paolisso1995a;@Salgin2012a].
+Observational and clinical studies have reported concordant findings, showing in
+particular that elevated total plasma NEFA associates with an increased risk for
+incident type 2 diabetes [@Paolisso1995a;@Salgin2012a].
 
@@
 high in the polyunsaturated fatty acid (PUFA) linoleic acid (18:2n-6), as the 
-exposure to characterize the metabolic impact of total NEFA. However, fatty acids comprise
-multiple molecules with diverse physiological functions, and few 
-studies have analyzed the effects of a broader spectrum of fatty acids. Notably,
+exposure to characterize the metabolic impact of total NEFA. However, fatty
+acids comprise multiple molecules with diverse physiological functions, and few 
+studies have analyzed the effects of a broader spectrum of fatty acids. For instance,
 one study suggested that eicosapentaenoic acid (20:5n-3) can protect against the
@@
 diabetes, there are important gaps in this research. The majority of previous 
-studies have used animal models or cell lines [@Maris2013a;@Deguil2011a], 
-have been short term human trials 
-[@Szendroedi2012a;@Daniele2014a;@Liang2013a;@Kehlenbrink2012a], or have
-been epidemiological studies that only looked at total NEFA and not individual
-fatty acids [@Miller2012a;@Johns2014a]. To date, there have
-been no longitudinal studies examining the role of the composition of the serum NEFA
-fraction on the pathogenesis of diabetes, which is critical given the protracted natural
-history of diabetes and the growing appreciation of the divergent effects of
-individual fatty acids. Therefore, our objective was to examine the 
-association of serum NEFA composition on changes over time in insulin
-sensitivity and beta-cell function in a longitudinal cohort. We hypothesized
-that higher palmitic acid and lower PUFA such as
-eicosapentaenoic acid in the NEFA fraction would associate with declining insulin sensitivity and
-beta-cell function over 6 years.
+studies have used animal models or cell lines [@Maris2013a;@Deguil2011a], have
+been short term human trials 
+[@Szendroedi2012a;@Daniele2014a;@Liang2013a;@Kehlenbrink2012a], or have been
+epidemiological studies that only looked at total NEFA and not individual fatty
+acids [@Miller2012a;@Johns2014a;@Charles1997a]. To date, there have been no longitudinal
+studies examining the role of the composition of the serum NEFA fraction on the
+pathogenesis of diabetes, which is critical given the protracted natural history
+of diabetes and the growing appreciation of the divergent effects of individual
+fatty acids. Therefore, our objective was to examine the association of serum
+NEFA composition on changes over time in insulin sensitivity and beta-cell
+function in a longitudinal cohort. We hypothesized that higher palmitic acid and
+lower PUFA such as eicosapentaenoic acid in the NEFA fraction would associate
+with declining insulin sensitivity and beta-cell function over 6 years.
 

### 2017-07-20 doc/manuscript.Rmd
reformat last intro paragraph
homa2 ref on validation

@@
 epidemiological studies that only looked at total NEFA and not individual fatty 
-acids [@Miller2012a;@Johns2014a;@Charles1997a]. To date, there have been no longitudinal
-studies examining the role of the composition of the serum NEFA fraction on the
-pathogenesis of diabetes, which is critical given the protracted natural history
-of diabetes and the growing appreciation of the divergent effects of individual
-fatty acids. Therefore, our objective was to examine the association of serum
-NEFA composition on changes over time in insulin sensitivity and beta-cell
-function in a longitudinal cohort. We hypothesized that higher palmitic acid and
-lower PUFA such as eicosapentaenoic acid in the NEFA fraction would associate
-with declining insulin sensitivity and beta-cell function over 6 years.
+acids [@Miller2012a;@Johns2014a;@Charles1997a]. To date, there have been no
+longitudinal studies examining the role of the composition of the serum NEFA
+fraction on the pathogenesis of diabetes, which is critical given the protracted
+natural history of diabetes and the growing appreciation of the divergent
+effects of individual fatty acids. Therefore, our objective was to examine the
+association of serum NEFA composition on changes over time in insulin
+sensitivity and beta-cell function in a longitudinal cohort. We hypothesized
+that higher palmitic acid and lower PUFA such as eicosapentaenoic acid in the
+NEFA fraction would associate with declining insulin sensitivity and beta-cell
+function over 6 years.
 
@@
 the disposition index (but using OGTT values). Each index has been validated
-against gold standard measures [@Levy1998a; @Matsuda1999; @Retnakaran2009]. {{confirm re homa2}}
+against gold standard measures [@Hermans1999a; @Matsuda1999; @Retnakaran2009]. 
 Formulas used to calculate these measures can be found in the ESM Methods.

### 2017-07-20 doc/esm-methods.Rmd
updated esm methods with more info on gee

@@
 
-Generalized estimating equations (GEE) is analogous to mixed effects models, 
-except it emphasizes population-level trends over individual-level trends that 
-mixed models use. Because it can handle autocorrelations in the outcome (as 
-would be seen in variables measured over time), GEE modeling is well suited to
-longitudinal data. Latent class mixed models (LCMM) is a form of mixed effects
-modeling that tries to categorize individuals into groups that have similar
-clusters; in the case of the present analysis are the trajectories in beta-cell
-function over time. Because mixed models use the participant as the random
-effect, LCMM can identify individuals who similarly change over time.
+Generalized estimating equations (GEE) is a technique similar to mixed effects 
+modeling, except it calculates the marginal population estimates compared to the
+subject-specific estimates in mixed effects models. GEE is well suited to and
+commonly used in longitudinal cohort studies, especially given its capacity to
+handle missed visits. GEE is able to handle longitudinal data due to the
+estimating of a working correlation matrix, which was chosen based on
+quasi-likelihood information criteria (QIC). The auto-regressive of order 1
+(AR1) matrix was chosen for the GEE models as it had the best model fit accessed
+using QIC, though other matrices (eg. exchangeable) had similar fit (data not
+shown).
+
+For the confounders, they were chosen based on previous literature, from 
+directed acyclic graph (DAG) recommendations, and from QIC. The DAG
+recommendations were obtained from using the DAGitty software,
+http://dagitty.net/. The final GEE model selected as best fitting was negligibly
+different when using insulin sensitivity or beta-cell function as an outcome and
+as such the same covariates were chosen for models with each of the outcomes.
 

### 2017-07-20 doc/manuscript.Rmd
reformat paragraphs
multiple clarifications for stats analysis in methods

@@
 glucose and insulin data. Insulin sensitivity was assessed using HOMA2-%S [@Levy1998a]
-and the Insulin Sensitivity Index (ISI)
-[@Matsuda1999]. HOMA2-%S largely reflects hepatic insulin sensitivity, while ISI
-reflects whole-body insulin sensitivity [@AbdulGhani2007].  Beta-cell function
-was assessed using the Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR)
-and the Insulin Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR
-is a measure of the first phase of insulin secretion while ISSI-2 is analogous to
-the disposition index (but using OGTT values). Each index has been validated
-against gold standard measures [@Hermans1999a; @Matsuda1999; @Retnakaran2009]. 
-Formulas used to calculate these measures can be found in the ESM Methods.
+and the Insulin Sensitivity Index (ISI) [@Matsuda1999]. HOMA2-%S largely
+reflects hepatic insulin sensitivity, while ISI reflects whole-body insulin
+sensitivity [@AbdulGhani2007].  Beta-cell function was assessed using the
+Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR) and the Insulin
+Secretion-Sensitivity Index-2 (ISSI-2) [@Retnakaran2009]. IGI/IR is a measure of
+the first phase of insulin secretion while ISSI-2 is analogous to the
+disposition index (but using OGTT values). Each index has been validated against
+gold standard measures [@Hermans1999a;@Matsuda1999;@Retnakaran2009]. Formulas
+used to calculate these measures can be found in the ESM Methods.
 
@@
 [@Zeger1986a] were used to determine the longitudinal associations between the 
-outcome variables and the predictor variables. Given the longitudinal design 
-an auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE models, 
-though other matrices (eg. exchangeable) had similar fit (data not shown). GEE
-is well suited to longitudinal cohort studies given it's capacity to handle
-missed visits. The predictor variables and continuous covariates were scaled
-(mean centered and standardized). The NEFA variables were classified as
+outcome variables and the predictor variables. Given the longitudinal design an
+auto-regressive of order 1 (AR1) correlation matrix was chosen for the GEE
+models, though other matrices (eg. exchangeable) had similar fit (data not
+shown). GEE is well suited to longitudinal cohort studies given it's capacity to
+handle missed visits. The predictor variables and continuous covariates were
+scaled (mean centered and standardized). The NEFA variables were classified as 
 *time-independent* (held constant) as they were measured only at the baseline 
-visit, while the outcome variables and covariates were set as *time-dependent*.
-No imputation was conducted on missing values.
+visit, while the outcome variables were set as *time-dependent*. No imputation
+was conducted on missing values and incident diabetes cases were excluded from 
+the analysis.
 
 Covariate selection was based on previous literature, directed acyclic graph 
-[@Greenland1999a] recommendations (see `r cite_sf("dag")`), and quasi-likelihood information criterion 
-(QIC). `r cite_st('qic')` shows the covariates compared using QIC. To reduce
-complexity between models using either insulin sensitivity or beta-cell function
-measures, only the model that had similar good fit for both was used. 
-As such, we selected the model
-that had the fewest covariates and that had similar fit between the outcome
-measures (M9). The final GEE model was adjusted for years from baseline (time),
-sex (female or male), ethnicity (European ancestry or non-European ancestry),
-baseline age, WC, ALT, and MET (physical activity). Sex, ethnicity, and baseline 
-age were time-*independent* variables. After scaling,
-log-transforming, and exponentiating, the GEE estimates are
-interpreted as an expected percent difference in the outcome variable for every
-standard deviation (SD) increase in the predictor variable given the covariates
-are held constant.
-Since TAG is a risk factor for diabetes and since NEFA contribute to TAG
-production, TAG may act as a mediator between NEFA and the outcomes. To
-determine the role of TAG in the association between NEFA and the outcomes, TAG
-was included in the GEE model as a sensitivity analysis. Lastly for the GEE
-models, we tested for an interaction with sex, ethnicity, or time by the
-predictor term for each outcome variable.
+[@Greenland1999a;@Textor2011a] recommendations (see `r cite_sf("dag_isi")` and 
+`r cite_sf("dag_issi2")`), and quasi-likelihood information criterion (QIC). See
+the ESM Methods for more detail on QIC and the directed acyclic graphs. 
+`r cite_st('qic')` shows the covariates compared using QIC. We selected the
+model (M9) that had the fewest covariates and that had similar fit between the
+outcome measures. The final GEE model included the covariates years from baseline
+(time), sex (female or male), ethnicity (European ancestry or non-European 
+ancestry), baseline age, WC, ALT, and MET (physical activity). Sex, ethnicity, 
+and baseline age were time-*independent* variables. After scaling, 
+and log-transforming the model variables, then exponentiating the GEE estimates,
+the interpretation of the GEE results is as an expected percent difference in 
+the outcome variable for every standard deviation (SD) increase in the predictor
+variable given the covariates are held constant. With the main effect 
+associations (not with a time interaction), this interpretation is applied 
+consistently at each time point and thus represents associations of baseline 
+NEFA for each future value of the outcomes. Since TAG is a risk factor for
+diabetes and since NEFA contribute to TAG production [@Barrows2006a], TAG may
+act as a mediator between NEFA and the outcomes. To determine the role of TAG in
+the association between NEFA and the outcomes, TAG was included in the GEE model
+in a sensitivity analysis. Lastly for the GEE models, we tested for an 
+interaction with sex, ethnicity, or time by the predictor term for each outcome 
+variable.
 
@@
 limited in that it cannot analyze the inherent multivariate nature of the
-composition of the NEFA fraction. Therefore, to confirm the results of the GEE
-analyses in a multivariate environment (all NEFA variables in a single model),
-partial least squares (PLS) was used with the outcomes as the independent variable.
+composition of the NEFA fraction, as each model only contains data from a single
+fatty acid. Therefore, to confirm the results of the GEE analyses in a
+multivariate environment (all NEFA variables in a single model), partial least
+squares (PLS) was used with the outcomes as the dependent variable.
 
 All analyses were performed using R `r getRversion()` [@Rbase], along with the R
-packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a].
-The R code for this manuscript is available at 
-https://dx.doi.org/10.6084/m9.figshare.3472433. Results were considered statistically significance at p<0.05,
-after adjusting for multiple testing using the Benjamini-Hochberg False
-Discovery Rate [@Benjamini1995a]. STROBE was used as a guideline for reporting [@Vandenbroucke2007b].
+packages geepack `r packageVersion('geepack')` for GEE [@Hoejsgaard2006a]
+and the package pls `r packageVersion('pls')` for the PLS analysis.
+The R code and output for all analyses of this manuscript is available at 
+https://dx.doi.org/10.6084/m9.figshare.3472433. Results were considered
+statistically significance at p<0.05, after adjusting for multiple testing using
+the Benjamini-Hochberg False Discovery Rate [@Benjamini1995a]. STROBE was used
+as a guideline for reporting [@Vandenbroucke2007b].
 

### 2017-07-20 doc/manuscript.Rmd
updated the results

@@
 
-For this study, there were 
-`r female$npct` females and `r european$npct` had European-ancestry (see 
-`r cite_t("basic")`). The mean follow-up time was `r calc_followup_time()`
-years, where `r calc_pct_full_visits()`% of participants attended all three
-visits.
-The average age of the participants was `r aide::average(ds0yr$Age)` years 
-(`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
+For this study, there were `r female$npct` females and `r european$npct` had
+European-ancestry (see `r cite_t("basic")`). The mean follow-up time was 
+`r calc_followup_time()` years, where `r calc_pct_full_visits()`% of 
+participants attended all three visits.
+The average age of the participants at baseline was `r aide::average(ds0yr$Age)`
+years (`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
 `r aide::average(ds0yr$BMI)` (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
-participants, `r famhist$npct`, had a family history of diabetes.
-Insulin sensitivity and beta-cell function measures showed a significant median
-decline from baseline to 6-years between
-`r chg_outcomes$chg` (p`r chg_outcomes$p`, based on a time-adjusted GEE model)
-in this analysis.
+participants, `r famhist$npct`, had a family history of diabetes. Insulin
+sensitivity and beta-cell function measures showed a significant median
+decline from baseline to 6-years of between `r chg_outcomes$chg` 
+(p`r chg_outcomes$p`, based on a time-adjusted GEE model) in this analysis.
 Consistent with this decline, there were `r calc_incid_dysgly("ConvertDM")`
-and `r calc_incid_dysgly("ConvertPreDM")`incident cases of diabetes and pre-diabetes (IFG and IGT),
-respectively, over the 6-years. Incident diabetes cases were excluded from further GEE
-analyses.
+and `r calc_incid_dysgly("ConvertPreDM")`incident cases of diabetes and
+pre-diabetes (IFG and IGT), respectively, over the 6-years. Incident diabetes
+cases were excluded from further GEE analyses. There were `r meds_use[1]` and
+`r meds_use[2]` participants who used lipid and cholesterol modifying medications
+at the baseline and 6-year visit, respectively. Medication use did not influence
+NEFA composition, the outcomes, or the GEE models (data not shown).
 
@@
 (`r sum(nefa$pct)`%) of the total NEFA fraction. The largest contributors
-were `r nefa$c[1]`, `r nefa$c[2]`, `r nefa$c[3]`, and 
-`r nefa$c[4]`. Raw concentration values are shown in `r cite_st('nefa')`. 
-There were no differences in individual NEFA for those who did compared to those
-who did not take medications to manage blood lipids (data not shown).
-All individual NEFA as well as the total fraction had
-correlations that ranged from weak to null (r<0.3) with participant
-characteristics (see `r cite_sf('corr_conc')` and `r cite_sf("corr_pct")`).
+were `r nefa$c[1]`, `r nefa$c[2]`, `r nefa$c[3]`, and `r nefa$c[4]`. Raw
+concentration values are shown in `r cite_st('nefa')`. All individual NEFA as
+well as the total fraction had correlations that ranged from weak to null
+(r<0.3) with participant characteristics (see `r cite_sf('corr_conc')` and 
+`r cite_sf("corr_pct")`).
 
@@
 (see `r cite_f('gee_unadj')`), particularly for NEFA modeled as a concentration.
-While full model adjustment (`r cite_f('gee_adj')`) attenuated most of these
-associations, total NEFA, 16:0, 18:1n-9, and 18:2n-6 (all as nmol/mL)
-continued to show negative associations with IGI/IR and ISSI-2. Sensitivity
-analyses revealed that adjustment for WC was responsible for attenuating the
-unadjusted results. The magnitude of association for each of these variables was
-fairly consistent for each beta-cell function measure. For every one SD increase
-in any of these three NEFA variables, there was an average predicted
-`r est_gee$est[1]`% lower IGI/IR and `r est_gee$est[2]`% lower ISSI-2 at each
-clinic visit. However, in contrast to these results using NEFA modeled as
-concentrations, none of the NEFA variables modeled as mol% were associated with
-the outcomes. Adjusting for TAG attenuated all associations with IGI/IR and
-ISSI-2 (data not shown), though this may be due to the fact that only a subset
-of participants at the 6 year visit had TAG measured. Sensitivity analyses to
-examine the association with
-the sum of fatty acid classes (saturated vs unsaturated) revealed similar
-findings as seen with the adjusted GEE model for total NEFA (data not shown).
-There were no significant interaction effects between time, sex, or ethnicity
-and the individual NEFA on any of the outcome measures.
-Raw values from the GEE models are shown in `r cite_st('gee_unadj')` for
-unadjusted models and `r cite_st('gee_adj')` for fully adjusted models.
-Consistent with the GEE modeling, results from modeling all NEFA in a
-multivariate environment using PLS showed no discernible clustering nor was it
-predictive of the outcomes (data not shown).
+There were several fatty acids (`r paste(unique(sig_fa_unadj$Xterms),collapse=", ")`)
+that had significant associations with both insulin sensitivity and beta-cell
+function outcomes. Most associations were negative, however, only 20:5n-3
+(eicosapentaenoic acid; EPA) had positive associations with insulin sensitivity
+as both a mol% and as a concentration. One fatty acid, 18:1n-7 (vaccenic acid),
+had consistent negative associations with all four outcomes.
+
+After full model adjustment (`r cite_f('gee_adj')`), most significant 
+associations were attenuated, particularly with insulin sensitivity. The
+remaining fatty acids (`r paste(unique(sig_fa_adj$Xterms),collapse=", ")`) had
+significant associations, of which only EPA had a significant association (with
+HOMA2-%S) as a mol% and was the only fatty acid with a positive association. All
+other fatty acids had negative associations with IGI/IR and ISSI-2. The
+magnitude of association for each of these variables was fairly consistent for
+each beta-cell function measure. For every one SD increase in any of the NEFA
+variables associated with beta-cell function, there was an average predicted
+`r est_gee$est[1]`% lower IGI/IR and `r est_gee$est[2]`% lower ISSI-2 at any
+given clinic visit.
+
+Sensitivity analyses revealed that adjustment for WC was responsible for 
+attenuating the unadjusted results (data not shown). Adjusting for TAG 
+attenuated all associations with IGI/IR and ISSI-2 (data not shown), though this
+may be due to the fact that only a subset of participants at the 6 year visit 
+had TAG measured. Sensitivity analyses to examine the association with the sum 
+of fatty acid classes (saturated vs unsaturated) revealed similar findings as 
+seen with the adjusted GEE model for total NEFA (data not shown). There were no 
+significant interaction effects of time, sex, or ethnicity in either unadjusted
+or adjusted models. Medication use for controlling lipid and cholesterol did not
+influence the results (data not shown). Raw values from the GEE models are shown
+in `r cite_st('gee_unadj')` for unadjusted models and `r cite_st('gee_adj')` for
+fully adjusted models. Consistent with the GEE modeling, results from modeling
+all NEFA in a multivariate environment using PLS showed no discernible
+clustering nor was it predictive of the outcomes (data not shown).
 

### 2017-07-20 doc/esm-methods.Rmd
minor clarifications to pls in esm methods
clarified dag and qic in esm methods

@@
 
-For the confounders, they were chosen based on previous literature, from 
+For the covariates, they were chosen based on previous literature, from 
 directed acyclic graph (DAG) recommendations, and from QIC. The DAG
 recommendations were obtained from using the DAGitty software,
-http://dagitty.net/. The final GEE model selected as best fitting was negligibly
-different when using insulin sensitivity or beta-cell function as an outcome and
-as such the same covariates were chosen for models with each of the outcomes.
+http://dagitty.net/. DAG is a technique that allows for identifying potential
+confounding causal pathways and to limit potential bias introduced from
+adjusting for colliding variables. A collider variable is a variable that is
+determined from two or more other variables, such that in causal graph modeling,
+the arrows from the other variables appear to 'collide' into the variable. QIC 
+is a technique that balances the goodness of fit of the model with its relative
+complexity (i.e. number of covariates) against other models being compared. This
+identifies a specific model from a number of models that is the 'best' fit and
+least complex of all the models. QIC and other information criterion methods are
+used in conjunction with other techniques such as literature and DAG based. The 
+final GEE model selected as best fitting was negligibly different when using 
+insulin sensitivity or beta-cell function as an outcome and as such the same 
+covariates were chosen for models with each of the outcomes.
 
@@
 and PLS is that PLS is a supervised method, while PCA is not. This means that
-PLS uses a response variable(s), i.e. an outcome or **Y**, to describe the
-variation in the predictors (**X**) while PCA only describes the variation 
-inherent in **X**. Because of this, PLS can be better at predicting the 
-contributions of predictor variables against an outcome variable.
+PLS uses a response variable(s), i.e. an outcome or **Y**, to 'supervise' (or
+constrain) the variation inherent in the predictors (**X**) while PCA only
+describes the variation inherent in **X**. Because of this, PLS can be better at
+predicting the contributions of predictor variables against an outcome variable.

### 2017-07-20 doc/manuscript.Rmd
add bit about refering to esm methods for covariate selection
updated the conclusions section in light of reviewer comments

@@
 ancestry), baseline age, WC, ALT, and MET (physical activity). Sex, ethnicity, 
-and baseline age were time-*independent* variables. After scaling, 
+and baseline age were time-*independent* variables. See the ESM Methods for more
+detail on the covariate inclusion. After scaling,
 and log-transforming the model variables, then exponentiating the GEE estimates,
@@
 (16:0), oleic acid (18:1n-9), and linoleic acid (18:2n-6) for the concentration
-(nmol/mL) modeling, no associations were seen for these fatty
-acids as a proportion, an observation that was further confirmed by multivariate
-cluster techniques, which found no ability of these
-individual NEFA to predict beta-cell function or dysglycemia. These observations suggest
-that the absolute size of the total NEFA fraction, rather than its specific
-composition, likely influences the pathogenesis of diabetes, at least within a
-population at-risk for diabetes.
+(nmol/mL) modeling, no associations were seen for these fatty acids as a
+proportion, an observation that was further confirmed by multivariate cluster
+techniques, which found no ability of these individual NEFA to predict beta-cell
+function. We did find that one fatty acid as a mol%, eicosapentaenoic acid
+(20:5n-3), had a positive association with insulin sensitivity. In general,
+these observations suggest that the absolute size of the total NEFA fraction,
+rather than its specific composition (aside from EPA), likely influences the 
+pathogenesis of diabetes, at least within a population at-risk for diabetes.
 
@@
 the baseline visit and as such we cannot investigate whether there were 
-concomitant changes in NEFA and the metabolic measures over time. However, we believe this
-is a strength for our specific objective, as the chance of reverse causality is 
-reduced given that fatty acid and glucose metabolism pathways are tightly 
-integrated. This is also an observational cohort, and there may be some residual
-confounding we have not considered or that could not be measured. Nonetheless,
-potential covariates were empirically analyzed prior to inclusion into the
-GEE models to best understand and minimize potential
-confounding. Finally, our cohort consists of individuals at-risk for diabetes, 
-who are primarily female and European-ancestry and as such our results may not be
-generalizable to other populations. However, given these limitations, our study
-also has several strengths, including the longitudinal design and the rigorous
-statistical techniques and methods applied, which are specifically suited to
-investigating temporal relationships and to handling the multivariate nature of
-the data. Lastly, our cohort contains highly detailed and comprehensive
-variable measurements at each collection visit, and has both concentration and
-mol% data for the fatty acids.
+concomitant changes in NEFA and the metabolic measures over time. However, we
+believe this is a strength for our specific objective, as the potential impact
+that reverse causality, due to the tight integration between fatty acid and
+glucose metabolism, may have on the results is lower. This is also an
+observational cohort, and there may be some residual confounding we have not
+considered or that could not be measured. Nonetheless, potential covariates were
+empirically analyzed prior to inclusion into the GEE models to best understand
+and minimize potential confounding. While there may be potential overadjustment
+due to controlling for WC, both QIC and directed acyclic graph outputs
+recommended including WC in the GEE model. Even still, we ran sensitivity
+analyses and identified WC as the covariate that attenuated the unadjusted
+results. This attenuation may be due to the fact that higher NEFA may be 
+reflective of greater adipose tissue mass and acting as a mediator between the 
+adipose tissue and insulin sensitivity (since adipocytes are active endocrine 
+organs).
+
+Finally, our cohort consists of individuals at-risk for diabetes, 
+who are primarily female and European-ancestry and, while sex and ethnicity did
+not influence the results presented from this cohort, our findings may not
+entirely be generalizable to other populations. However, given these
+limitations, our study also has several strengths, including the longitudinal
+design and the rigorous statistical techniques and methods applied, which are
+specifically suited to investigating temporal relationships and to handling the
+multivariate nature of the data. Lastly, our cohort contains highly detailed and
+comprehensive variable measurements at each collection visit, and has both
+concentration and mol% data for the fatty acids.
 
@@
 secretion and a higher risk for developing diabetes 
-[@Paolisso1995a;@Salgin2012a;@Pankow2004a]. In a cross-sectional analysis of the
+[@Paolisso1995a;@Salgin2012a;@Pankow2004a;@Charles1997a]. In a cross-sectional analysis of the
 RISCK cohort, total NEFA had a negative association with insulin sensitivity and
@@
 palmitic acid, oleic acid, and linoleic acid modeled as concentrations with
-lower beta-cell function. However, in modeling these fatty acids as a mol% and
-using novel clustering analysis approaches, no specific fatty acids in the NEFA
-fraction strongly predicted lower beta-cell function, insulin sensitivity, or
-dysglycemia status. Taken together, these results suggest that it is the
+lower beta-cell function. No fatty acid as a mol% had associations with
+beta-cell function. Taken together, these results suggest that it is the
 absolute size of the circulating NEFA fraction, irrespective of any specific 
@@
 [@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this longitudinal
-analysis, we saw no association of any individual or total NEFA with hepatic
-(HOMA2-%S) or whole-body insulin sensitivity (ISI). We identified
-WC as having a powerful attenuating effect on the associations between
-unadjusted and adjusted models. There are some possible explanations for our
-null findings for the insulin sensitivity measures and the influence of WC on
-the results. First, WC may be a strong causal link in previously reported
-associations between NEFA and insulin sensitivity, given the role adipocytes
-play in metabolism (e.g. leptin, adiponectin). Second, there may be differences
-in physiology between fasting and postprandial NEFA kinetics that we are not
-able to investigate but that may explain our null findings for fasting NEFA and
-insulin sensitivity. For instance, some experimental studies using clamp
-protocols found that fasting NEFA was a weak predictor of insulin sensitivity
-compared to postprandial concentrations of NEFA
-[@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA uptake into the adipose
-tissue (e.g. as seen in obesity and larger WC) following postprandial TAG
-lipolysis via lipoprotein lipase may result in NEFA spillover into the blood and
-a subsequent increase in circulating NEFA [@Almandoz2013a], which may be more
-metabolically active given postprandial activity. Third, the null findings seen
-may be due to the high risk population examined in PROMISE, as the majority of subjects had a 
-high WC and BMI. It may be that in this population, IR has become well
-established and NEFA may not contribute to IR at this somewhat more advanced
-stage in the pathogenesis of diabetes.
+analysis, we saw in general no association of any individual or total NEFA with
+hepatic (HOMA2-%S) or whole-body insulin sensitivity (ISI), except for EPA. 
+There is some evidence [@Steffen2015a;@Virtanen2014;@Sarbolouki2013] that EPA
+may be involved with lowering the risk for diabetes, potentially due to its role
+in inflammation [@Calder2009]. However, given we found the association only with
+HOMA2-%S and not with ISI, the significant association could be chance finding,
+which will require future studies to confirm. We also identified WC as having a
+powerful attenuating effect on the associations between unadjusted and adjusted
+models. There are some possible explanations for our null findings for the
+insulin sensitivity measures and the influence of WC on the results. First, WC
+may be a strong causal link in previously reported associations between NEFA and
+insulin sensitivity, given the role adipocytes play in metabolism (e.g. leptin,
+adiponectin). Second, there may be differences in physiology between fasting and
+postprandial NEFA kinetics that we are not able to investigate but that may
+explain our null findings for fasting NEFA and insulin sensitivity. For
+instance, some experimental studies using clamp protocols found that fasting
+NEFA was a weak predictor of insulin sensitivity compared to postprandial
+concentrations of NEFA [@Magkos2012a;@Kehlenbrink2012a]. Inefficiencies in NEFA
+uptake into the adipose tissue (e.g. as seen in obesity and larger WC) following
+postprandial TAG lipolysis via lipoprotein lipase may result in NEFA spillover
+into the blood and a subsequent increase in circulating NEFA [@Almandoz2013a],
+which may be more metabolically active given postprandial activity. Third, the
+null findings seen may be due to the high risk population examined in PROMISE, 
+as the majority of subjects had a high WC and BMI. It may be that in this
+population, insulin resistance has become well established and NEFA may not
+contribute to insulin resistance at this somewhat more advanced stage in the
+pathogenesis of diabetes.
 
@@
 through medication and/or lifestyle strategies to reduce the risk of diabetes 
-are well warranted. While future
-studies are needed to confirm these findings, our results reinforce the
-importance of continuing to investigate the role of circulating NEFA
-concentration on the natural history of diabetes.
+are well warranted. We also found that higher EPA was associated with higher
+hepatic insulin sensitivity. While future studies are needed to confirm these
+findings, our results reinforce the importance of continuing to investigate the
+role of circulating NEFA concentration on the natural history of diabetes.
 

### 2017-07-20 doc/submission/reviewer-response.md
added explanation of types of covars to response
minor edits for clarification

@@
 
-    We thank the reviewer for these comments. {{not sure about comment on
-    pre-diabetes}}. In the PROMISE cohort, we collected medication data from
-    all participants. While most participants took some form of medication over
-    the 6 years, during the times of data collection, few ({{get number}}) were
-    on medications. However, for those who were taking medications that could
-    influence the results, additional analyses revealed that medication use did
-    not substantially influence the predictor or outcome measures, nor did 
-    medication use substantially contribute to GEE model fit compared to models 
-    without it (see QIC table).
+    We thank the reviewer for these comments. In the PROMISE cohort, we
+    collected medication data from all participants. While most participants
+    took some form of medication over the 6 years, during the times of data
+    collection, many (n=60-103 between the three visits) were on taking lipid
+    and cholesterol controlling medications. However, for those who were taking
+    medications that could influence the results, additional analyses revealed
+    that medication use did not substantially influence the predictor or outcome
+    measures, nor did medication use substantially contribute to GEE model fit
+    compared to models without it (see QIC table). We have included this
+    information within the manuscript.
 
@@
     described it in the results and discussion sections, in addition to briefly
-    describing them in the abstract (keeping within the word limit).
+    describing them in the abstract (keeping within the word limit). An
+    explanation of what each covariate is in regard to NEFA and the outcomes
+    (e.g. mediator or confounder) has been added in the ESM Methods.
 
@@
 
-        These have been added as requested.
+        These have been added as requested. {{Tony, not sure about how I feel
+        adding signif to Table 1. My analysis isn't on changes in these
+        measures, so there is not need to test if there is a difference. It's
+        purely descriptive and I'd prefer not to generate p-values for the sake
+        of generating p-values. Thoughts? I could always just do a baseline table...}}
 
@@
         computation of the GEE estimates. This nuance provides substantially
-        more power to the model and the proceeding results. See ref {{num}} in
-        the manuscript for more detail on this statistical technique. Within
-        the manuscript we have expanded on this nuance to clarify the interpretation.
+        more power to the model and the proceeding results. See ref 27 in the
+        manuscript for more detail on this statistical technique. Within the
+        manuscript we have expanded on this nuance to clarify the
+        interpretation.
 
@@
         these results internally suggest no differences between sex or ethnicity,
-        our cohort was not designed to answer these research questions and the
-        results presented in the paper may still not be generalizable to other
-        populations. {{yea?}}
+        there may still be features unique to sex and ethnicity in this cohort
+        that limit generalizability.
 

### 2017-07-20 R/plots.R
no dots in boxplot fig

@@ -12,7 +12,7 @@ plot_nefa_distribution <- function() {
         dplyr::mutate(Measure = forcats::fct_inorder(Measure))
 
     nefa_dist %>%
-        seer::view_boxplots(dots = TRUE, "Measure", "Value") +
+        seer::view_boxplots(dots = FALSE, "Measure", "Value") +
         ggplot2::labs(y = 'Concentration (nmol/mL)',
                       x = 'Non-esterified fatty acid') +
         graph_theme(ticks = FALSE)

### 2017-07-20 doc/esm-methods.Rmd
added info on what each covariate is in esm methods
cat for formulas for proper printing

@@
 ```{r}
-PROMISE.methods::insert_formula("ISI")
+cat(PROMISE.methods::insert_formula("ISI"))
 ```
@@
 ```{r}
-PROMISE.methods::insert_formula("HOMA")
+cat(PROMISE.methods::insert_formula("HOMA"))
 ```
@@
 ```{r}
-PROMISE.methods::insert_formula("IGIIR")
+cat(PROMISE.methods::insert_formula("IGIIR"))
 ```
@@
 ```{r}
-PROMISE.methods::insert_formula("ISSI2")
+cat(PROMISE.methods::insert_formula("ISSI2"))
 ```
@@
 insulin sensitivity or beta-cell function as an outcome and as such the same 
-covariates were chosen for models with each of the outcomes.
+covariates were chosen for models with each of the outcomes. The covariates
+medication use, WC, ALT, and MET would be considered classical confounders while
+TAG would be considered a mediating variable.
 

### 2017-07-20 R/renaming.R
homair to homa2 in table 1

@@ -6,7 +6,7 @@ renaming_table_rows <- function(x) {
         gsub('HOMA2_S', 'HOMA2-%S', .) %>%
         gsub('IGIIR', 'IGI/IR', .) %>%
         gsub('ISSI2', 'ISSI-2', .) %>%
-        gsub('HOMA', 'HOMA-IR', .) %>%
+        gsub('^HOMA$', 'HOMA-IR', .) %>%
         gsub('TAG', 'TAG (mmol/L)', .) %>%
         gsub('Chol', 'Chol (mmol/L)', .) %>%
         gsub('ALT', 'ALT (U/L)', .) %>%

### 2017-07-20 R/tables.R

@@ -13,7 +13,7 @@ table_basic <- function(data, caption = NULL) {
                       Sex = ifelse(VN == 0, as.character(Sex), NA),
                       Sex = as.factor(Sex)) %>%
         carpenter::outline_table('f.VN') %>%
-        carpenter::add_rows(c('HOMA', 'ISI', 'IGIIR', 'ISSI2'),
+        carpenter::add_rows(c('HOMA2_S', 'ISI', 'IGIIR', 'ISSI2'),
                             carpenter::stat_medianIQR, digits = 1) %>%
         carpenter::add_rows(c('BMI', 'Waist', 'Age', 'MET', 'ALT', 'TAG', 'Chol', 'HDL',
                               'BaseTotalNE'),

### 2017-07-20 doc/submission/reviewer-response.md
fixed so c) doesn't start a list

@@
         came in for the 6-yr visit. We have included wording within the results
-        on the sample and percentage who attended at least two visits andLastly,
-        c) because diabetes can effect the values of the OGTT-derived indices
-        due to medication the patient is taking and due to diabetes itself
-        impacting beta-cell function and insulin resistance, we excluded them
-        from all analyses; often, these participants did not even complete an
-        OGTT since it is discouraged that these patients consume that much
+        on the sample and percentage who attended at least two visits. Lastly,
+        for c) because diabetes can effect the values of the OGTT-derived 
+        indices due to medication the patient is taking and due to diabetes 
+        itself impacting beta-cell function and insulin resistance, we excluded 
+        them from all analyses; often, these participants did not even complete 
+        an OGTT since it is discouraged that these patients consume that much 
         glucose. To restate, diabetes cases were not included in GEE or other 

### 2017-07-20 doc/submission/reviewer-response.md
updated my comment on p-values for table 1

@@
         These have been added as requested. {{Tony, not sure about how I feel
-        adding signif to Table 1. My analysis isn't on changes in these
-        measures, so there is not need to test if there is a difference. It's
-        purely descriptive and I'd prefer not to generate p-values for the sake
-        of generating p-values. Thoughts? I could always just do a baseline table...}}
+        adding p-values to Table 1. I'm showing it purely for descriptive
+        purposes and it might detract from discussion since, as you may recall,
+        some variables (e.g. HDL) have a significant difference between visits,
+        but don't really change over the visits... thoughts? I'm not too keen on
+        generating p-values for the sake of generating p-values, esp. when it is
+        not really relevant to the main analysis.}}
 

### 2017-07-25 doc/esm-figures.Rmd
generated gee results with no waist, added to esm

@@
 
-sfig_consort <- sfig('consort', 'CONSORT diagram of sample size at each examination visit.')
+sfig_consort <- sfig('consort', 'CONSORT diagram of sample size at each examination visit for the main PROMISE cohort in addition of the sample size determination for the present analysis.')
 sfig_dag_isi <- sfig("dag_isi", "Directed acyclic graph taken from DAGitty (DAGitty.net) for the potential causal pathway between NEFA and ISI.
@@
 Darkness of the colour indicates the magnitude of the correlation, with orange indicating positive and blue indicating negative correlations.')
+sfig_gee_adj_nowaist <- sfig('gee_adj_nowaist', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
+Generalized estimating equation models were fully adjusted, *excluding waist circumference*.
+X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.
+P-values were adjusted for the BH false discovery rate, with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, and p<0.01.')
 ```
@@
 ```
+
+```{r ESM_Fig6, fig.cap=sfig_gee_adj_nowaist}
+gee_results %>% 
+    dplyr::filter(model == "Adjusted no waist") %>% 
+    plot_gee_main()
+```

### 2017-07-25 R/generate_results.R

@@ -32,6 +32,21 @@ generate_results_gee <- function() {
         ) %>%
         tidy_gee_results()
 
+    covar_nowaist <- covariates[which(!grepl("Waist", covariates))]
+    gee_results_adj_nowaist <- dplyr::bind_rows(
+        analyze_gee(gee_data, y = outcomes[1:2], x = ne_pct, covar = covar_nowaist,
+                    unit = 'mol%', mod = "Adjusted no waist"),
+        analyze_gee(gee_data, y = outcomes[3:4], x = ne_pct, covar = covar_nowaist,
+                    unit = 'mol%', mod = "Adjusted no waist"),
+        analyze_gee(gee_data, y = outcomes[1:2], x = ne_conc, covar = covar_nowaist,
+                    unit = 'nmol/mL', mod = "Adjusted no waist"),
+        analyze_gee(gee_data, y = outcomes[3:4], x = ne_conc, covar = covar_nowaist,
+                    unit = 'nmol/mL', mod = "Adjusted no waist"),
+        analyze_gee(gee_data, y = outcomes, x = ne_total, covar = covar_nowaist,
+                    unit = 'Total', mod = "Adjusted no waist")
+    ) %>%
+        tidy_gee_results()
+
     gee_results_unadj <- dplyr::bind_rows(
         analyze_gee(gee_data, y = outcomes[1:2], x = ne_pct, covar = 'YearsFromBaseline',
                     unit = 'mol%', mod = "Unadjusted"),
@@ -47,7 +62,7 @@ generate_results_gee <- function() {
         tidy_gee_results()
 
     # Save output of results into dataset
-    gee_results <- dplyr::bind_rows(gee_results_adj, gee_results_unadj)
+    gee_results <- dplyr::bind_rows(gee_results_adj, gee_results_unadj, gee_results_adj_nowaist)
     devtools::use_data(gee_results, overwrite = TRUE)
 }
 

### 2017-07-25 doc/submission/reviewer-response.md
updated response as per Tony's edits

--- 2017-07-20 doc/submission/reviewer-response.md
@@
     collected medication data from all participants. While most participants 
-    took some form of medication over the 6 years, during the times of data
-    collection, many (n=60-103 between the three visits) were on taking lipid
-    and cholesterol controlling medications. However, for those who were taking
-    medications that could influence the results, additional analyses revealed
-    that medication use did not substantially influence the predictor or outcome
-    measures, nor did medication use substantially contribute to GEE model fit
-    compared to models without it (see QIC table). We have included this
-    information within the manuscript.
+    took some form of medication over the 6 years, many (n=60-103 between the 
+    three visits) were on taking lipid and cholesterol controlling medications.
+    However, for those who were taking medications that could influence the
+    results, additional analyses revealed that there were no differences in the
+    predictor or outcome measures with medication use, nor did adjustment for
+    medication use substantially contribute to GEE model fit compared to models
+    without this variable (see QIC table). We have included this information in
+    the revised manuscript.
 
@@
     in the GEE models; the time variable has been replaced in GEE models, though
-    the change does not substantially impact the final results.
+    the change does not substantially impact the final results. This change has
+    been included in the statistical analysis section.
 
@@
 
-    The specifics of the covariates are expanded upon in the methods section. We
+    The specifics of the covariates have been expanded upon in the methods
+    section and the ESM Methods of the revised manuscript. We
     have included the unadjusted results in the main manuscript and have 
@@
     describing them in the abstract (keeping within the word limit). An
-    explanation of what each covariate is in regard to NEFA and the outcomes
+    explanation of our interpretation of each covariate in regard to NEFA and the outcomes
     (e.g. mediator or confounder) has been added in the ESM Methods.
@@
     for. We have included in the conclusions a discussion on how adjustment of
-    waist circumference influences the model and the interpretation.
+    waist circumference influences the model and the interpretation, as well as
+    show the results of models without waist circumference in the ESM.
 
@@
 
-        We have clarified how we dealt with incident diabetes cases. Briefly, we
-        only report incident diabetes and dysglycemia cases in the text to
-        describe the cohort, but in statistical analyses and modeling, these
-        cases were excluded. Regarding a) within the main PROMISE cohort,
-        prevalent diabetes cases were not excluded from the study as they
-        continued to be followed; prevalent diabetes cases did not have fatty
-        acids measured. For b) any participant is followed until they request to
-        be removed from the study; some participants missed the 3-yr visit, but
-        came in for the 6-yr visit. We have included wording within the results
-        on the sample and percentage who attended at least two visits andLastly,
-        c) because diabetes can effect the values of the OGTT-derived indices
-        due to medication the patient is taking and due to diabetes itself
-        impacting beta-cell function and insulin resistance, we excluded them
-        from all analyses; often, these participants did not even complete an
-        OGTT since it is discouraged that these patients consume that much
-        glucose. To restate, diabetes cases were not included in GEE or other
-        analyses.
+        In the revised manuscript, we have clarified how we dealt with incident
+        diabetes cases. Briefly, we only report incident diabetes and
+        dysglycemia cases in the text to describe the cohort, but in statistical
+        analyses and modeling, these cases were excluded. Regarding point a)
+        above prevalent diabetes cases at baseline were not included in this
+        analysis as not prevalent diabetes cases had fatty acids measured. For
+        point b) above, not all participants attended each visit; some
+        participants missed the 3-yr visit, but came in for the 6-yr visit. We
+        have included wording and numbers within the Results section of the
+        revised manuscript regarding the sample and percentage who attended at
+        least two visits. Lastly, regarding point c) above because diabetes can
+        effect the values of the OGTT-derived indices due to medication the
+        patient is taking and due to diabetes itself impacting beta-cell
+        function and insulin resistance, we excluded them from all analyses. To
+        restate, diabetes cases were not included in GEE or other analyses.
 
@@
 
-        We are not entirely clear on the reviewer's comment. PLS extracts from
-        high dimensional data (multiple predictor variables, i.e. the fatty
-        acids) underlying correlations, constrained by a outcome variable (i.e.
-        beta-cell function). In this context, because mol% data is by definition
-        restricted to sums of 100% (all fatty acid values add up to 100%), the
-        underlying correlations between fatty acids is dependent on the relative
-        contributions those fatty acids have to the total and as such can
-        reflect correlations of specific fatty acids or groups of fatty acids
-        that increase or decrease in proportion together. The concentration data
-        is not contrained by this inherent correlation and thus underlying groups 
-        or clusters may not be identified using PLS.
+        For the PLS analysis, we used the mol% data to input into the PLS model.
+        PLS extracts underlying correlations from high dimensional data
+        (multiple predictor variables, i.e. the fatty acids), constrained by a
+        outcome variable (i.e. beta-cell function). In this context, because
+        mol% data is by definition restricted to sums of 100% (all fatty acid
+        values add up to 100%), the underlying correlations between fatty acids
+        is dependent on the relative contributions those fatty acids have to the
+        total and as such can reflect correlations of specific fatty acids or 
+        groups of fatty acids that increase or decrease in proportion together. 
+        The concentration data is not contrained by this inherent correlation 
+        and thus underlying groups or clusters may not be identified using PLS.
 
@@
 
-        These have been added as requested. {{Tony, not sure about how I feel
-        adding signif to Table 1. My analysis isn't on changes in these
-        measures, so there is not need to test if there is a difference. It's
-        purely descriptive and I'd prefer not to generate p-values for the sake
-        of generating p-values. Thoughts? I could always just do a baseline table...}}
+        The number of participants who attended all three visits has been added
+        as requested. Regarding including 'ns' in Table 1, we intentionally did
+        not include p-values for this table for several reasons. Our objective
+        was not to examine significant differences in the basic characteristics
+        of the cohort and this table is included to show a purely descriptive
+        summary of the cohort over the 6 years. Our analysis currently generates
+        a large number of tests and subsequent p-values and, even though we
+        adjusted these p-values for the false discovery rate, we did not want to
+        generate more p-values (and the resulting increase in Type 1 error) than
+        necessary if they are not primary to our objective.
 
@@
 
-        The unadjusted results and figure has been added to the main paper and
+        The unadjusted results and figure have been added to the main paper and
         described in the results section. We've also added a sensitivity
-        analysis confirming which covariate(s) attenuates the associations.
+        analysis confirming which covariate(s) attenuates the associations and
+        present the GEE model results of that sensitivity analysis in the ESM.
 
@@
         model building. We have incorporated this new approach in the analysis and
-        report on the results in the paper. In simple terms, the reviewer is
+        report on the results in the revised paper. In fundamental terms, the reviewer is
         correct that without an interaction term the model is essentially
         "repeated cross-sectional analyses", however, GEE handles the analysis
-        in a slightly more nuanced approach. GEE calculates the estimates of the
-        association of NEFA on the outcomes at any given time point, taking into
+        with a slightly more nuanced approach. GEE calculates the estimates of the
+        association of NEFA with the outcomes at any given time point, taking into
         consideration the inherent multivariate structure of the repeated
@@
         manuscript we have expanded on this nuance to clarify the
-        interpretation.
+        interpretation in the statistical analysis section of the Methods in
+        the revised manuscript.
 
@@
         sensitivity analyses to identify covariates that strongly influence the
-        results and discuss the results in light of these findings.
+        results and discuss the results in light of these findings. We have
+        presented the results of excluding the covariate (waist circumferen) in
+        the ESM.
 
@@
 
-        We agree with the reviewer's comment and ran stratified models by either
-        sex or ethnicity. In both cases, all significant associations from the
-        non-stratified analyses were attenuated, likely due to the reduction in
+        We agree with the reviewer's comment and is the reason we originally 
+        tested for interaction tests by sex and ethnicity. In both cases, we
+        found no statistically significant interactions. While these results
+        internally suggest no differences between sex or ethnicity, there may
+        still be features unique to sex and ethnicity in this cohort that limit
+        generalizability. Regardless, in light of the reviewers request, we ran
+        stratified models by either sex or ethnicity. In these findings, all 
+        significant associations from the non-stratified analyses were 
+        attenuated in the stratified analyses, likely due to the reduction in
         sample size in the stratified models. Stratifying analyses has the
         limitation of substantially reducing sample size and is the reason why
-        interaction testing is prefered. We ran interaction tests by sex and
-        ethnicity and found no significant interactive associations. While
-        these results internally suggest no differences between sex or ethnicity,
-        there may still be features unique to sex and ethnicity in this cohort
-        that limit generalizability.
+        interaction testing is prefered.
 
@@
     We thank the associate editor for the kind comments. We have included the
-    cited paper in the discussion section.
+    cited paper in the revised Discussion section.

### 2017-08-08 doc/manuscript.Rmd
updated manuscript word, table, and figure count
added ref to gee esm fig for no waist
minor edits before submission

@@
 
-Word count - Abstract: 242 / 250
+Word count - Main text: 3869 / 4000 (with figure captions)
 
-Word count - Main text: 3726 / 4000
+<!-- Table and figure count: 4 / 4 -->
 
-<!-- Table and figure count: 3 / 4 -->
-
-<!-- Reference limit 39 / 50 -->
+<!-- Reference limit 49 / 50 -->
 
@@
 
-Aim: Our aim was to determine the longitudinal associations of individual
-non-esterified fatty acids (NEFA) on changes in insulin sensitivity (IS) and
-beta-cell function over 6 years in a cohort of individuals who are at-risk for
-diabetes.
+Aim: Our aim was to determine the longitudinal associations of individual non-esterified fatty acids (NEFA) on changes in insulin sensitivity (IS) and beta-cell function over 6 years in a cohort of individuals who are at-risk for diabetes.
 
-Methods: In the PROMISE cohort, 477 participants had serum NEFA (mol% and
-concentration) measured at the baseline visit and completed an OGTT at 3 time
-points over 6 years. Outcome variables were calculated using the OGTT values. IS
-was assessed using HOMA2-%S and the Matsuda index, while beta-cell function was 
-assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin 
-Secretion-Sensitivity Index-2 (ISSI-2). Generalized estimating equations were 
-used, adjusting for time, waist, sex, ethnicity, baseline age, alanine 
-aminotransferase, and physical activity. NEFA were analyzed as both 
-concentrations (nmol/mL) and proportions (mol%) of the total fraction.
+Methods: In the PROMISE cohort, 477 participants had serum NEFA (mol% and concentration) measured at the baseline visit and completed an OGTT at 3 time points over 6 years.
+Outcome variables were calculated using the OGTT values.
+IS was assessed using HOMA2-%S and the Matsuda index, while beta-cell function was  assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin  Secretion-Sensitivity Index-2 (ISSI-2).
+Generalized estimating equations were  used, adjusting for time, waist, sex, ethnicity, baseline age, alanine  aminotransferase, and physical activity.
+NEFA were analyzed as both  concentrations (nmol/mL) and proportions (mol%) of the total fraction.
 
-Results: Participants (73% female, 70% with European-ancestry) had their IS and
-beta-cell function decline by 14-27% over 6 years of follow-up. While in
-unadjusted models there were several NEFA (e.g. 18:1n-7, 20:1n-9, 22:4n-6) that
-had negative associations with insulin sensitivity, nearly all of these
-associations were attenuated in fully adjusted models. In adjusted models, total
-NEFA, 16:0, 18:1n-9, and 18:2n-6 (as concentrations) had significant negative
-associations with IGI/IR and ISSI-2, while only 20:5n-3 (as mol%) had positive
-associations with HOMA2-%S.
+Results: Participants (73% female, 70% with European-ancestry) had their IS and beta-cell function decline by 14-27% over 6 years of follow-up.
+While in unadjusted models there were several NEFA (e.g.
+18:1n-7, 20:1n-9, 22:4n-6) that had negative associations with insulin sensitivity, nearly all of these associations were attenuated in fully adjusted models.
+In adjusted models, total NEFA, 16:0, 18:1n-9, and 18:2n-6 (as concentrations) had significant negative associations with IGI/IR and ISSI-2, while only 20:5n-3 (as mol%) had positive associations with HOMA2-%S.
 
-Conclusions: Total NEFA concentration was a strong predictor of lower beta-cell 
-function over 6 years. Our results suggest that the association with beta-cell 
-function is due to the absolute size of the serum NEFA fraction, rather than the
-specific fatty acid composition.
+Conclusions: Total NEFA concentration was a strong predictor of lower beta-cell  function over 6 years.
+Our results suggest that the association with beta-cell  function is due to the absolute size of the serum NEFA fraction, rather than the specific fatty acid composition.
 
@@
 exposure to high concentrations of NEFA can induce insulin resistance in 
-insulin-sensitive tissues such as muscle and liver, and in addition can impair 
+insulin-sensitive tissues such as muscle and liver, and can impair 
 pancreatic beta-cell production of insulin [@Rachek2014a;@Cnop2008a]. 
@@
 Sensitivity analyses revealed that adjustment for WC was responsible for 
-attenuating the unadjusted results (data not shown). Adjusting for TAG 
+attenuating the unadjusted results (see `r cite_sf("gee_adj_nowaist")`). Adjusting for TAG 
 attenuated all associations with IGI/IR and ISSI-2 (data not shown), though this
@@
 in inflammation [@Calder2009]. However, given we found the association only with
-HOMA2-%S and not with ISI, the significant association could be chance finding,
+HOMA2-%S and not with ISI, the significant association could be a chance finding,
 which will require future studies to confirm. We also identified WC as having a
@@
 
-```{r figure2_gee_adj, fig.cap=fig_gee_adj}
+```{r figure3_gee_adj, fig.cap=fig_gee_adj}
 gee_results %>% 

### 2017-08-08 R/calculate.R
Updated code, added esm and vignettes

@@ -1,0 +1,158 @@
+
+#' Calculate the conversion of participants to DM or pre DM (incidence).
+#'
+#' @param variable which dysgly variable to use
+#'
+calc_incid_dysgly <- function(variable = c('ConvertDM', 'ConvertPreDM')) {
+    variable <- match.arg(variable)
+
+    conversion <-
+        table(dysgly_data[variable])[2] %>%
+        {
+            paste0(., ' (', round((. / 477) * 100, 0), '%)')
+        }
+
+    return(conversion)
+}
+
+#' Calculate the percentage of contribution for NEFA that are >10 percent.
+#'
+#' @param data project data
+#'
+calc_pct_nefa <- function(data = project_data) {
+    over_10pct_nefa <- data %>%
+        dplyr::filter(VN == 0) %>%
+        dplyr::select(dplyr::matches('pct_ne')) %>%
+        tidyr::gather(fat, value) %>%
+        dplyr::group_by(fat) %>%
+        dplyr::summarise(pct = round(mean(value, na.rm = TRUE), 1)) %>%
+        dplyr::mutate(fat = renaming_fa(fat),
+               c = paste0(fat, ' (', pct, '%)')) %>%
+        dplyr::arrange(dplyr::desc(pct)) %>%
+        dplyr::filter(pct >= 10)
+
+    return(over_10pct_nefa)
+}
+
+#' Calculate the percent change in the outcomes (plus sample size and p-value).
+#'
+#' @param data project data
+#'
+calc_outcomes_chg_n <- function(data = project_data) {
+    prep_data <- data %>%
+        dplyr::select(f.VN, HOMA2_S, ISI, IGIIR, ISSI2) %>%
+        tidyr::gather(Measure, Value,-f.VN) %>%
+        stats::na.omit() %>%
+        dplyr::group_by(Measure, f.VN) %>%
+        dplyr::summarise(med = stats::median(Value),
+                         n = n()) %>%
+        dplyr::ungroup()
+
+    sample_size <- prep_data$n %>%
+        {paste0(min(.), '-', max(.))}
+
+    change_over_time <- prep_data %>%
+        dplyr::select(-n) %>%
+        tidyr::spread(f.VN, med) %>%
+        dplyr::mutate(pctChg = ((yr6 - yr0) / yr0) * 100) %>%
+        dplyr::select(pctChg) %>%
+        abs() %>%
+        round(0) %>%
+        {paste0(min(.), '% to ', max(.), '%')}
+
+    pval <- mason::design(data, 'gee') %>%
[…]

### 2017-08-08 R/corr.R

@@ -1,0 +1,44 @@
+#' Compute correlation coefficients (pearson) results dataset.
+#'
+#' @return Saves a dataset with the correlation coefficient results.
+#' @param x Variables on the X-axis
+#' @param y Variables on the Y-axis
+analyze_corr <- function(x = c(outcomes, 'BMI', 'Waist', 'Age', 'lALT', 'lTAG', 'Chol', 'HDL', 'LDL'),
+                         y) {
+
+    baseline_data <- project_data %>%
+        dplyr::filter(VN == 0)
+
+    corr_results <- baseline_data %>%
+        mason::design('cor') %>%
+        mason::add_settings(method = 'pearson', use = 'complete.obs') %>%
+        mason::add_variables('yvars', y) %>%
+        mason::add_variables('xvars', x) %>%
+        mason::construct() %>%
+        mason::scrub() %>%
+        dplyr::mutate_at("Vars2", renaming_fa) %>%
+        dplyr::mutate_at("Vars1", function(x)
+            gsub('l(ALT|TAG|IGIIR|HOMA2_S|invHOMA|ISI|ISSI2)', '\\1', x) %>%
+                renaming_outcomes()) %>%
+        order_by_fattyacid(fa_col = 'Vars2') %>%
+        dplyr::mutate(Vars2 = forcats::fct_inorder(Vars2),
+                      Vars1 = forcats::fct_inorder(Vars1),
+                      Correlations = round(Correlations, 2))
+
+    return(corr_results)
+}
+
+#' Plot the correlations into a heatmap.
+#'
+#' @param results Correlation results
+plot_corr_heatmap <- function(results, unit) {
+    results %>%
+        seer::view_heatmap(
+            y = 'Vars2',
+            x = 'Vars1',
+            ylab = paste0('Non-esterified fatty acids (', unit, ")"),
+            number.colours = 5,
+            values.text = FALSE
+        ) +
+        graph_theme(ticks = FALSE, legend.pos = 'right')
+}

### 2017-08-08 R/data_prep.R

@@ -1,0 +1,54 @@
+
+#' Order fatty acid variables that are in a column (long form) by degree of desaturation.
+#'
+#' @param data Dataset with fatty acid variables in long form.
+#' @param fa_col The column that contains the fatty acid variables.
+#'
+order_by_fattyacid <- function(data, fa_col) {
+    data %>%
+        dplyr::rename_("Measure" = fa_col) %>%
+        dplyr::mutate(
+            order1 = substr(Measure, nchar(Measure), nchar(Measure)),
+            order1 = ifelse(order1 == 0, 10, order1),
+            order1 = ifelse(order1 == 'l', 20, order1),
+            order1 = as.integer(order1)
+        ) %>%
+        dplyr::arrange(dplyr::desc(order1)) %>%
+        dplyr::select(-order1) %>%
+        dplyr::rename_(.dots = stats::setNames(names(.), gsub("Measure", fa_col, names(.))))
+}
+
+#' Calculate incident dysglycemia over the 6 years.
+#'
+#' @param data project data
+#'
+incident_dysglycemia_data <- function(data) {
+    dysgly_data <-
+        dplyr::left_join(
+            data %>%
+                dplyr::filter(VN == 0),
+            data %>%
+                dplyr::filter(!is.na(TotalNE)) %>%
+                dplyr::mutate_at(c("IFG", "IGT"), dplyr::funs(ifelse(is.na(.), 0, .))) %>%
+                dplyr::mutate(PreDM = as.numeric(rowSums(.[c('IFG', 'IGT')], na.rm = TRUE))) %>%
+                dplyr::mutate(FactorDysgly = ifelse(
+                    PreDM == 1, 'PreDM',
+                    ifelse(DM == 1, 'DM',
+                           'NGT')
+                )) %>%
+                dplyr::select(SID, VN, FactorDysgly) %>%
+                tidyr::spread(VN, FactorDysgly) %>%
+                dplyr::mutate(
+                    DetailedConvert = as.factor(paste0(`0`, '-', `1`, '-', `2`)),
+                    ConvertDysgly = as.numeric(!grepl('NGT$|NGT-NA$|NGT-NGT-NA$|NGT-NA-NA$',
+                                                      DetailedConvert)),
+                    ConvertDM = as.numeric(grepl('-DM$|-DM-DM$|-DM-NA$|-DM-NGT$',
+                                                 DetailedConvert)),
[…]

### 2017-08-08 R/fetch_data.R

@@ -7,40 +7,39 @@
 #' @return Saves the wrangled data into the data/ folder.
 #' @export
 #'
-#' @examples
-#' \dontrun{
-#' fetch_data()
-#' }
 fetch_data <- function() {
     # Load the master dataset,
-    ds.prep <- PROMISE::PROMISE_data %>%
+    ds.prep <- PROMISE::PROMISE %>%
         dplyr::filter(VN %in% c(1, 3, 6)) %>%
-        ## Kick out Canoers
+        # Kick out Canoers
         dplyr::filter(is.na(Canoe)) %>%
         dplyr::tbl_df()
 
     print(paste0('Original dataset rows are ', dim(ds.prep)[1], ' and columns are ', dim(ds.prep)[2]))
 
-    ##' Munge and wrangle the data into the final version.
-    ds <-
-        dplyr::full_join(ds.prep,
-                  ds.prep %>%
-                      dplyr::filter(VN == 1) %>%
-                      dplyr::select(SID, BaseAge = Age, BaseTAG = TAG)) %>%
+    # Munge and wrangle the data into the final version.
+    ds <- ds.prep %>%
         dplyr::select(
             SID, VN, BMI, Waist, HOMA, ISI, IGIIR, ISSI2, TAG, LDL, HDL, Chol,
             ALT, CRP, FamHistDiab, dplyr::matches('meds'), Age, Sex, Ethnicity,
-            IFG, IGT, DM, MET, BaseAge, AlcoholPerWk, TobaccoUse, SelfEdu, Occupation,
-            TotalNE, dplyr::matches('^ne\\d+'), Glucose0, Glucose120, BaseTAG
+            IFG, IGT, DM, MET, Age, AlcoholPerWk, TobaccoUse, SelfEdu, Occupation,
+            TotalNE, dplyr::matches('^ne\\d+'), Glucose0, Glucose120, TAG,
+            MonthsFromBaseline, dplyr::matches("HOMA")
         ) %>%
         dplyr::mutate(
-            BaseTotalNE = TotalNE,
+            YearsFromBaseline = MonthsFromBaseline / 12,
             FamHistDiab =
                 plyr::mapvalues(FamHistDiab, c(0, 1:12),
                                 c('No', rep('Yes', 12))) %>%
                 as.factor(),
+            BaseTotalNE = TotalNE,
+            BaseTAG = ifelse(VN == 1, TAG, NA),
[…]

### 2017-08-08 R/figshare.R

@@ -1,0 +1,33 @@
+library(rfigshare)
+
+# Steps to take:
+# 1. make new orphan branch
+# 2. keep only R files and purl(manuscript.Rmd)
+# 3. git commit the files
+# 4. git archive --format=zip --output=git-nefa-code.zip manuscript-code #(branch name)
+# 5. then run the below code
+send_to_figshare <- function() {
+    fs_new_article(
+        "Analysis code for study on non-esterified fatty acids and the pathogenesis of diabetes",
+        "R code used for the analysis in the NEFA manuscript.",
+        type = "fileset",
+        # there was a problem with the authors...
+        # authors = c("Luke Johnston", "Stewart Harris", "Ravi Retnakaran",
+        #             "Adria Giacca", "Zhen Liu", "Richard P. Bazinet",
+        #             "Anthony J. Hanley"),
+        tags = c(
+            "insulin sensitivity",
+            "beta-cell function",
+            "fatty acids",
+            "non-esterified fatty acids",
+            "free fatty acids",
+            "cohort",
+            "longitudinal"
+        ),
+        categories = c("Diseases", "Pathogenesis", "Epidemiology"),
+        #links = "link to github",
+        links = "https://github.com/lwjohnst86/nefa-code",
+        files = "code-archive.zip",
+        visibility = "draft"
+    )
+}

### 2017-08-08 R/gee.R

@@ -1,0 +1,147 @@
+#' Get the dataset organized and wrangled into a format for the GEE analysis.
+#'
+prep_gee_data <- function(data) {
+    data <- data %>%
+        dplyr::filter(DM != 1)
+
+    data_without_fa <- dplyr::select(data, -dplyr::matches('pct_ne|^ne'), -TotalNE)
+
+    data_with_fa_scaled <- data %>%
+        dplyr::filter(VN == 0) %>%
+        dplyr::select(SID, TotalNE, dplyr::matches('pct_ne|^ne')) %>%
+        dplyr::mutate_at(dplyr::vars(-SID), dplyr::funs(as.numeric(scale(.))))
+
+    full_data_fa_scaled <-
+        dplyr::full_join(data_without_fa,
+                         data_with_fa_scaled,
+                         by = 'SID')
+
+    data_fa_covar_scaled <- full_data_fa_scaled %>%
+        dplyr::group_by(VN) %>%
+        dplyr::mutate_at(dplyr::vars(Waist, ALT, BaseAge, MET, TAG),
+                         dplyr::funs(as.numeric(scale(.))))
+
+    gee_prepped_data <- data_fa_covar_scaled %>%
+        dplyr::ungroup() %>%
+        dplyr::arrange(SID, VN)
+
+    return(gee_prepped_data)
+}
+
+
+#' Run GEE on data and generate the model results.
+#'
+#' @param data project data
+#' @param y outcomes
+#' @param x predictors
+#' @param covar covariates
+#' @param unit unit for the fatty acids (percent or conc)
+#' @param intvar interaction variable
+#' @param mod adjusted or unadjusted
+#'
+analyze_gee <- function(data, y, x, covar, unit, intvar = NULL, mod = NA) {
+
+    # What to filter from the dataset
+    int <- !is.null(intvar)
+    if (int) {
+        extract_term <- ':'
+    } else {
+        extract_term <- 'Xterm$'
+    }
+
+    gee_results <- data %>%
+        mason::design('gee') %>%
+        mason::add_settings(family = stats::gaussian('identity'), corstr = 'ar1', cluster.id = 'SID') %>%
+        mason::add_variables('yvars', y) %>%
+        mason::add_variables('xvars', x) %>%
+        mason::add_variables('covariates', covar) %>% {
+            if (int) {
+                mason::add_variables(., 'interaction', intvar)
+            } else {
+                .
+            }
+        } %>%
+        mason::construct() %>%
+        mason::scrub() %>%
[…]

### 2017-08-08 R/generate_results.R

@@ -1,0 +1,107 @@
+
+#' Generate and save all the outputs of the analyses as datasets.
+#'
+#' @return Saves the various outputs of the analyses as datasets.
+#' @export
+generate_results <- function() {
+    generate_results_corr()
+    generate_results_gee()
+    generate_results_lcmm()
+    generate_data_dysgly()
+    generate_results_pls()
+
+    invisible()
+}
+
+#' Generate and save the results from the GEE modeling as a dataset.
+#'
+generate_results_gee <- function() {
+    gee_data <- prep_gee_data(project_data)
+
+    gee_results_adj <- dplyr::bind_rows(
+        analyze_gee(gee_data, y = outcomes[1:2], x = ne_pct, covar = covariates,
+                    unit = 'mol%', mod = "Adjusted"),
+        analyze_gee(gee_data, y = outcomes[3:4], x = ne_pct, covar = covariates,
+                    unit = 'mol%', mod = "Adjusted"),
+        analyze_gee(gee_data, y = outcomes[1:2], x = ne_conc, covar = covariates,
+                    unit = 'nmol/mL', mod = "Adjusted"),
+        analyze_gee(gee_data, y = outcomes[3:4], x = ne_conc, covar = covariates,
+                    unit = 'nmol/mL', mod = "Adjusted"),
+        analyze_gee(gee_data, y = outcomes, x = ne_total, covar = covariates,
+                    unit = 'Total', mod = "Adjusted")
+        ) %>%
+        tidy_gee_results()
+
+    covar_nowaist <- covariates[which(!grepl("Waist", covariates))]
+    gee_results_adj_nowaist <- dplyr::bind_rows(
+        analyze_gee(gee_data, y = outcomes[1:2], x = ne_pct, covar = covar_nowaist,
+                    unit = 'mol%', mod = "Adjusted no waist"),
+        analyze_gee(gee_data, y = outcomes[3:4], x = ne_pct, covar = covar_nowaist,
+                    unit = 'mol%', mod = "Adjusted no waist"),
+        analyze_gee(gee_data, y = outcomes[1:2], x = ne_conc, covar = covar_nowaist,
+                    unit = 'nmol/mL', mod = "Adjusted no waist"),
+        analyze_gee(gee_data, y = outcomes[3:4], x = ne_conc, covar = covar_nowaist,
+                    unit = 'nmol/mL', mod = "Adjusted no waist"),
[…]

### 2017-08-08 R/lcmm.R

@@ -1,0 +1,72 @@
+
+#' Compute LCMM groups.
+#'
+#' @param data project data
+#' @param lc_var latent class variable
+#'
+analyze_lcmm <- function(data, lc_var = 'lISSI2') {
+    data_prep <- data %>%
+        dplyr::select_('SID', lc_var, 'VN') %>%
+        stats::na.omit() %>%
+        dplyr::arrange(SID, VN)
+
+    lcmm_formula <- as.formula(paste0(lc_var, " ~ VN"))
+    lcmm_results <- lcmm::lcmm(
+        lcmm_formula,
+        mixture = ~ VN, random = ~ VN, subject = 'SID',
+        ng = 3, idiag = TRUE, link = '3-equi-splines',
+        data = data_prep
+    )
+
+    return(lcmm_results)
+}
+
+#' Combine together the project data and the lcmm results.
+#'
+#' @param data project data
+#' @param results lcmm results
+#'
+combine_lcmm_results_with_data <- function(data, results) {
+    data <- data %>%
+        dplyr::full_join(results$pprob %>%
+                             as.data.frame() %>%
+                             dplyr::select(SID, class)) %>%
+        dplyr::filter(!is.na(class)) %>%
+        dplyr::mutate(class = factor(
+            class,
+            levels = c(2, 1, 3),
+            labels = c('High', 'Mid', 'Low')
+        )) %>%
+        dplyr::rename(ClassLCMM = class)
+
+    return(data)
+}
+
+#' Plot the results of the LCMM modeling.
+#'
+#' @param data project data
+#' @param results lcmm results
+#' @param lc_var the latent class variable
+#'
+plot_lcmm_results <- function(data, results, lc_var = "lISSI2") {
+    data %>%
+        combine_lcmm_results_with_data(results) %>%
+        ggplot2::ggplot(ggplot2::aes_string("f.VN", lc_var, group = "SID", colour = "ClassLCMM")) +
+        ggplot2::geom_line(ggplot2::aes(group = SID, colour = ClassLCMM), size = 0.1, alpha = 0.9) +
+        ggplot2::geom_smooth(ggplot2::aes(group = ClassLCMM), method = "loess", size = 2)  +
+        ggplot2::labs(x = "Time", y = "log(ISSI-2)", colour = "Latent Class") +
+        ggplot2::scale_x_discrete(labels = c('Baseline', 'Year 3', 'Year 6'), expand = c(0.05, 0.05)) +
[…]

### 2017-08-08 R/plots.R

@@ -1,0 +1,19 @@
+
+#' Plot the individual fatty acids in NEFA to see the distribution.
+#'
+plot_nefa_distribution <- function() {
+    nefa_dist <- project_data %>%
+        dplyr::filter(VN == 0) %>%
+        dplyr::select(dplyr::matches('^ne\\d')) %>%
+        tidyr::gather(Measure, Value) %>%
+        dplyr::mutate(Measure = renaming_fa(Measure) %>%
+                   gsub('ne_', '', .)) %>%
+        order_by_fattyacid(fa_col = "Measure") %>%
+        dplyr::mutate(Measure = forcats::fct_inorder(Measure))
+
+    nefa_dist %>%
+        seer::view_boxplots(dots = FALSE, "Measure", "Value") +
+        ggplot2::labs(y = 'Concentration (nmol/mL)',
+                      x = 'Non-esterified fatty acid') +
+        graph_theme(ticks = FALSE)
+}

### 2017-08-08 R/pls.R

@@ -1,0 +1,68 @@
+# Functions for the PLS analysis
+#
+# Grab or combine data ----------------------------------------------------
+
+#' Prepare the project data for PLS analysis.
+#'
+#' @param data Project data.
+#' @param y Y or outcome variable.
+#' @param x X or exposure variables of interest.
+#'
+prep_pls_data <- function(data, y, x) {
+    data %>%
+        dplyr::filter(VN == 0) %>%
+        dplyr::select_(.dots = c(y, x)) %>%
+        stats::na.omit()
+}
+
+# Analyze -----------------------------------------------------------------
+
+#' Compute PLS analysis.
+#'
+#' @param data Project data.
+#' @param y Outcome variable.
+#' @param x Exposure variables.
+#' @param ncomp Number of components.
+#' @param cv Whether to use CV.
+#'
+analyze_pls <- function(data, y, x = ne_pct, ncomp = NULL, cv = TRUE) {
+    data %>%
+        prep_pls_data(y = y, x = x) %>%
+        mason::design('pls') %>%
+        mason::add_settings(ncomp = ncomp, validation = 'CV', cv.data = cv, cv.seed = 5436) %>%
+        mason::add_variables('yvars', y) %>%
+        mason::add_variables('xvars', x) %>%
+        mason::construct() %>%
+        mason::scrub()
+}
+
+# Plots -------------------------------------------------------------------
+
+#' Plot the X loadings of the PLS results.
+#'
+#' @param data PLS output results.
+#'
+plot_pls <- function(data) {
+    seer::view_pls_xloadings(data, renaming.x = renaming_fa, dot.colour = "black") +
+        graph_theme(minor.grid.lines = FALSE)
+}
+
+# Calculations ------------------------------------------------------------
+
+#' Calculate the correlation between the predicted and actual outcome.
+#'
+#' @param model The PLS output results.
+#' @param test The test data to predict from.
+#' @param ncomps The component to predict on.
+#'
+calc_pred_corr <- function(model, test, ncomps = 1) {
+    predicted <- stats::predict(model, ncomp = ncomps, newdata = test)
+    measured <- as.matrix(stats::model.response(
+        stats::model.frame(formula(model), data = test))
+        )
+    corr <- broom::tidy(stats::cor.test(predicted, measured))[c(1, 3)]
+    r <- precise_rounding(corr[1], 2)
+    p <- format_p(corr[2])
+    list(r = r, p = p, r_p = paste0('r=', r, ', p', p))
+
+}

### 2017-08-08 R/qic.R

@@ -1,0 +1,136 @@
+
+#' Prepare the data for using in the QIC model comparisons.
+#'
+#' @param data Project data
+#'
+prep_qic_data <- function(data) {
+    data %>%
+        prep_gee_data() %>%
+        dplyr::select(
+            SID,
+            YearsFromBaseline,
+            lISSI2,
+            lISI,
+            TotalNE,
+            Sex,
+            BiEthnicity,
+            MET,
+            ALT,
+            BaseAge,
+            Waist,
+            BMI,
+            AlcoholPerWk,
+            TobaccoUse,
+            FamHistDiab,
+            BaseTAG,
+            MedsLipidsChol
+        )
+}
+
+#' Clean data output from the QIC model selection functions.
+#'
+#' @param data QIC output data.
+#'
+clean_qic <- function(data) {
+    data %>%
+        tibble::rownames_to_column() %>%
+        dplyr::select(Model = rowname, QIC = IC, Delta = delta) %>%
+        dplyr::mutate(id = seq_len(nrow(.)))
+}
+
+qic_is <- function(data) {
+    qic_data <<- data %>%
+        prep_qic_data() %>%
+        dplyr::select(-lISSI2) %>%
+        stats::na.omit()
+
+    M0 <- geepack::geeglm(
+        lISI ~ TotalNE + YearsFromBaseline,
+        data = qic_data,
+        id = SID,
+        family = stats::gaussian,
+        corstr = 'ar1'
+    )
+
+    M1 <- stats::update(M0, . ~ . + TotalNE:YearsFromBaseline)
+    M2 <- stats::update(M0, . ~ . + BaseAge + BiEthnicity + Sex)
+    M3 <- stats::update(M2, . ~ . + Waist)
+    M4 <- stats::update(M3, . ~ . + ALT)
+    M5 <- stats::update(M3, . ~ . + MET)
+    M6 <- stats::update(M3, . ~ . + AlcoholPerWk)
+    M7 <- stats::update(M3, . ~ . + FamHistDiab)
+    M8 <- stats::update(M3, . ~ . + TobaccoUse)
+    M9 <- stats::update(M3, . ~ . + ALT + MET)
+    M10 <- stats::update(M3, . ~ . + MedsLipidsChol)
+    M11 <- stats::update(M3, . ~ . + TotalNE:YearsFromBaseline)
[…]

### 2017-08-08 R/renaming.R

@@ -1,0 +1,51 @@
+
+# Renaming ----------------------------------------------------------------
+
+renaming_table_rows <- function(x) {
+    x %>%
+        gsub('HOMA2_S', 'HOMA2-%S', .) %>%
+        gsub('IGIIR', 'IGI/IR', .) %>%
+        gsub('ISSI2', 'ISSI-2', .) %>%
+        gsub('^HOMA$', 'HOMA-IR', .) %>%
+        gsub('TAG', 'TAG (mmol/L)', .) %>%
+        gsub('Chol', 'Chol (mmol/L)', .) %>%
+        gsub('ALT', 'ALT (U/L)', .) %>%
+        gsub('LDL', 'LDL (mmol/L)', .) %>%
+        gsub('HDL', 'HDL (mmol/L)', .) %>%
+        gsub('BaseTotalNE', 'NEFA (nmol/mL)', .) %>%
+        gsub('Age', 'Age (yrs)', .) %>%
+        gsub('BMI', 'BMI (kg/m^2^)', .) %>%
+        gsub('Waist', 'WC (cm)', .)
+}
+
+renaming_outcomes <- function(x) {
+    x %>%
+        gsub('lHOMA2_S', 'log(HOMA2-%S)', .) %>%
+        gsub('linvHOMA', 'log(1/HOMA-IR)', .) %>%
+        gsub('lHOMA2_S', 'log(HOMA2-%S)', .) %>%
+        gsub('lISI', 'log(ISI)', .) %>%
+        gsub('lIGIIR', 'log(IGI/IR)', .) %>%
+        gsub('lISSI2', 'log(ISSI-2)', .) %>%
+        gsub('^invHOMA$', '1/HOMA-IR', .) %>%
+        gsub('^ISSI2$', 'ISSI-2', .) %>%
+        gsub('^IGIIR$', 'IGI/IR', .)
+}
+
+renaming_fa <- function(x) {
+    x %>%
+        gsub('.*(\\d\\d)(\\d)', '\\1:\\2', .) %>%
+        gsub('n(\\d)$', 'n-\\1', .) %>%
+        gsub('D(\\d\\d)$', 'D-\\1', .) %>%
+        gsub('^pct_', '', .) %>%
+        gsub('TotalNE', 'Total', .)
+}
+
+renaming_fraction <- function(x) {
+    gsub('ne', 'Non-esterified', x)
+}
+
+renaming_list <- function(x) {
+    x %>%
+        renaming_fa() %>%
+        renaming_outcomes()
+}

### 2017-08-08 R/rfigshare.R

@@ -1,0 +1,26 @@
+# Steps to take in the Git repo:
+# 1. Make new orphan branch (git checkout --orphan code)
+# 2. Keep only R files and purl('vignette/maindocument.Rmd')
+# 3. git commit the files
+# 4. git archive --format=zip --output=code-archive.zip code
+# 5. Then run the below code
+send_to_figshare_new <- function() {
+    rfigshare::fs_new_article(
+        "Analysis code for poster on triacylglycerol DNL on components of the Metabolic Syndrome in PROMISE",
+        "Analysis code for a poster for the American Diabetes Association Scientific Sessions in 2017",
+        type = "fileset",
+        # there was a problem with the authors...
+        # authors = c("First Last"),
+        tags = c(
+            "metabolic syndrome",
+            "fatty acids",
+            "triacylglycerols",
+            "cohort",
+            "longitudinal"
+        ),
+        categories = c("Diseases", "Pathogenesis", "Epidemiology"),
+        links = "https://github.com/lwjohnst86/ada.2017",
+        files = "code-archive.zip",
+        visibility = "draft"
+    )
+}

### 2017-08-08 R/setup.R

@@ -8,13 +8,6 @@
 
 #' Set options for all documents and scripts.
 #'
-#' @export
-#' @importFrom magrittr %>%
-#'
-#' @examples
-#'
-#' set_options()
-#'
 set_options <- function() {
     # Set the options here for individual packages
 
@@ -27,9 +20,10 @@ set_options <- function() {
 
     # For the document (knitr)
     knitr::opts_chunk$set(
-        warning = FALSE, message = FALSE, collapse = TRUE,
-        #fig.showtext = TRUE,
-        echo = FALSE, dev = c('cairo_ps')
+        warning = FALSE, message = FALSE, collapse = TRUE, echo = FALSE,
+        dev = c("cairo_ps", "pdf", "png"), fig.showtext = TRUE
     )
 }
 
+#' @importFrom magrittr %>%
+`%>%` <- magrittr::`%>%`

### 2017-08-08 R/tables.R

@@ -1,0 +1,53 @@
+
+# Tables ------------------------------------------------------------------
+
+#' Create a basic descriptive characteristics table.
+#'
+#' @param data project data
+#' @param caption table caption
+#'
+table_basic <- function(data, caption = NULL) {
+    data %>%
+        dplyr::mutate(Ethnicity = ifelse(VN == 0, as.character(Ethnicity), NA),
+                      Ethnicity = as.factor(Ethnicity),
+                      Sex = ifelse(VN == 0, as.character(Sex), NA),
+                      Sex = as.factor(Sex)) %>%
+        carpenter::outline_table('f.VN') %>%
+        carpenter::add_rows(c('HOMA2_S', 'ISI', 'IGIIR', 'ISSI2'),
+                            carpenter::stat_medianIQR, digits = 1) %>%
+        carpenter::add_rows(c('BMI', 'Waist', 'Age', 'MET', 'ALT', 'TAG', 'Chol', 'HDL',
+                              'BaseTotalNE'),
+                            carpenter::stat_meanSD,
+                            digits = 1) %>%
+        carpenter::add_rows(c('Ethnicity', 'Sex'), carpenter::stat_nPct, digits = 0) %>%
+        carpenter::renaming('rows', renaming_table_rows) %>%
+        carpenter::renaming('header', function(x) c('Measure', 'Baseline', '3-yr', '6-yr')) %>%
+        carpenter::build_table(caption = caption)
+}
+
+#' Table of raw values for the concentration and proportion for NEFA.
+#'
+#' @param data project data
+#' @param caption table caption
+#'
+table_distribution <- function(data, caption = NULL) {
+    grep_nefa <- function(pattern, x = c(ne_conc, ne_total))
+        rev(grep(pattern, x, value = TRUE))
+
+    nefa <- c(grep_nefa('3$'), grep_nefa('6$'), grep_nefa('7$'),
+              grep_nefa('9$'), grep_nefa('0$'), grep_nefa('TotalNE$'))
+
+    data %>%
+        dplyr::filter(VN == 0) %>%
+        tidyr::gather(FA, Value, dplyr::matches('ne\\d')) %>%
+        dplyr::mutate(unit = as.factor(ifelse(grepl('pct_', FA), 'mol%', 'nmol/mL')),
+                      FA = gsub('pct_', '', FA)) %>%
+        tidyr::spread(FA, Value) %>%
[…]

### 2017-08-08 R/utils.R

@@ -1,0 +1,49 @@
+
+# Misc --------------------------------------------------------------------
+
+#' Round a number to precisely the digit indicated (e.g. "1.0", not "1").
+#'
+#' @param x numeric vector
+#' @param digits digits to round to
+#'
+precise_rounding <- function(x, digits = 1) {
+    trimws(format(round(x, digits = digits), nsmall = digits))
+}
+
+graph_theme <- function(base.plot, ticks = TRUE, minor.grid.lines = FALSE, legend.pos = 'bottom') {
+    graph.theme <-
+        ggplot2::"%+replace%"(
+            ggthemes::theme_tufte(base_size = 10, base_family = 'sans'),
+            ggplot2::theme(
+                axis.line = ggplot2::element_line('black', size = 0.75),
+                axis.line.x = ggplot2::element_line('black', size = 0.75),
+                axis.line.y = ggplot2::element_line('black', size = 0.75),
+                legend.key.width = grid::unit(0.7, "line"),
+                legend.key.height = grid::unit(0.7, "line"),
+                strip.background = ggplot2::element_blank(),
+                plot.margin = grid::unit(c(0.5, 0, 0, 0), "cm"),
+                legend.position = legend.pos
+            )
+        )
+
+    if (!ticks) {
+        graph.theme <- ggplot2::"%+replace%"(graph.theme,
+                                            ggplot2::theme(axis.ticks.y = ggplot2::element_blank()))
+    }
+
+    if (minor.grid.lines) {
+        graph.theme <- ggplot2::"%+replace%"(
+            graph.theme,
+            ggplot2::theme(
+                panel.grid = ggplot2::element_line(),
+                panel.grid.minor = ggplot2::element_blank()
+            )
+        )
+    }
+
+    return(graph.theme)
+}
+
+format_p <- function(x) {
+   format.pval(x, digits = 2, eps = 0.001)
+}

### 2017-08-08 R/variables.R

@@ -1,0 +1,5 @@
+outcomes <- c('lHOMA2_S', 'lISI', 'lIGIIR', 'lISSI2')
+ne_pct <- grep('^pct_ne', vars, value = TRUE)
+ne_conc <- grep('^ne\\d\\d', vars, value = TRUE)
+ne_total <- "TotalNE"
+covariates <- c("YearsFromBaseline", 'Waist', 'Sex', 'BiEthnicity', 'BaseAge', 'ALT', 'MET')

### 2017-08-08 R/zzz.R

@@ -1,0 +1,3 @@
+.onLoad <- function(libname, pkgname) {
+    devtools::load_data()
+}

### 2017-08-08 doc/manuscript.R
updated manuscript.R from rmd

@@ -1,14 +1,8 @@
-## ----knit_setup, echo=FALSE----------------------------------------------
-#knitr::opts_knit$set(root.dir = '../')
-
-## ----setup, message=FALSE, echo=FALSE------------------------------------
-#source('.Rprofile')
-devtools::load_all()
+## ----setup, include=FALSE------------------------------------------------
+devtools::load_all('.')
 set_options()
-ds <- project_data
-ds0yr <- dplyr::filter(ds, VN == 0)
 
-## ----caption_setup-------------------------------------------------------
+## ----captions------------------------------------------------------------
 ## Include captions below using `captioner` package
 library(captioner)
 fig <- captioner(prefix = 'Fig.')
@@ -19,190 +13,48 @@ tab <- captioner(prefix = 'Table')
 cite_t <- pryr::partial(tab, display = 'cite')
 stab <- captioner(prefix = 'ESM Table')
 cite_st <- pryr::partial(stab, display = 'cite')
-# usage: cite_st('basicChar')
 
-tab_basic <- tab('basicChar', 'Basic characteristics of the PROMISE participants at each of the 3 clinic visits.')
-fig_dist <- fig('nefaDist', 'Concentrations (nmol/mL) of each NEFA in PROMISE participants at the baseline visit (2004-2006).')
-fig_gee <- fig('gee', 'Longitudinal associations of individual NEFA (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort. Fully adjusted generalized estimating equation models. X-axis values represent a percent difference in the outcome per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, with the largest dot representing a significant (p<0.05) association.')
+tab_basic <- tab('basic', 'Basic characteristics of the PROMISE participants at each of the three clinic visits.')
+fig_dist <- fig('nefa', 'Concentrations (nmol/mL) of each NEFA in PROMISE participants at the baseline visit (2004-2006).')
[…]

### 2017-09-25 doc/submission/reviewer-response-2.md
added reviewer comments

---
output: fost::default_manuscript
---

# Response to reviewers

## Referee: 1

- *"This paper presents an investigation of the relationship between NEFAs measured
at baseline and subsequent development of markers of insulin sensitivity. The
analysis is very thorough and the conclusions, in particular the interpretation
of results with biological considerations is excellent. I am not an expert on
this subject area and although I am a statistician, GEE is not an approach that
I am familiar with, so my comments are from the perspective of a
technically-minded but non-expert reader. Therefore some of my comments may not
be wholly relevant and are given more as possible considerations; or areas that
may require clarification for other non-experts."*

- *"A key question for me is the choice of using the GEE approach, which is of
course an established and appropriate method for many research questions, but I
think could be better justified here. The main advantages I am aware of are:
Richer correlation structures, easily handles missing data, robust standard
errors."*

- *"Here, the number of follow up visits is not large and the amount of missing data
is relatively small. My first thought when approaching the analysis would
therefore be to consider a standard linear regression analysis with
robust/clustered standard errors to incorporate repeated measures, and potential
multiple imputation of missing values and a comparison with complete case
analysis. So it would be nice to briefly spell out the things that GEE offers
here over other approaches to make this clear (in the ESM if need be so as not
to add to the word count). Currently the main statement is the ease of handling
missing data – which is handy but perhaps not essential – and some comments on
the working correlation matrix, which are not terribly informative. So some
statements about the improvement in estimates under a specific correlation
structure that is not easily handled otherwise, for instance, might be helpful.
Of note is that in the ESM it is stated that "other … matrices had similar fit"
– it may be worth noting that the model fit may be similar, but estimates may be
more efficient under a particular correlation structure."*

- *"Another thing that might be commented on briefly is the different nature of
estimates from GEE models, which are population averaged and have a slightly
different interpretation. This difference might be seen as a disadvantage to GEE
modelling – or do the authors feel this is more appropriate for this particular
question? This is something I do not have a thorough understanding of, but I do
note that the authors comment in the methods "…the interpretation of the GEE
results is as an expected percent difference in the outcome variable for every
standard deviation (SD) increase in the predictor variable given the covariates
are held constant" does not appear to be consistent with what I have read in
terms of keeping other factors constant, e.g.: "However, because the GEE
approach does not contain explicit terms for the between-cluster variation, the
resulting parameter estimates for the contrast of interest do not have the usual
"keeping other factors constant" interpretation." (Hanley et al, Am J Epidemiol
2003;157:364–375)."*

- *"Another area that is unclear is the exact model specification. The unadjusted
results I can assume are the outcome variables regressed via GEE against each
NEFA measure. Then adjusted results include age, gender, WC and so on. Now, for
the proportion of NEFA results, the approach seems to be that adjusted models
were fitted with each NEFA as a predictor in turn while adjusting for the
anthropomorphic and other covariates, as a PLS approach was used to examine all
the NEFA components together as proportions. Was this approach the same for the
analysis of NEFA as a concentration, or in this analysis were all NEFA
components included in a single multivariable model? I think that this should be
spelled out: if each NEFA component was considered in turn, and the variables
adjusted for, for the concentration/proportion analysis. At present the
assumption might be that multivariable models include all of the NEFA components
in one model."*

- *"If so, I do not fully understand why there is not a model that includes the
joint effect of all NEFA components together. Could this not be done for the
concentration analysis? Also, it could perhaps be done for the proportion too,
if one of the components is set as the baseline – similar to how categorical
variables are handled in regression analyses, but allowing percentages for each
indicator rather than the usual 0/1 dichotomy, vs. a "baseline" component. Such
an analysis could also adjust for total NEFA, with the interpretation that
proportion of each NEFA component is assessed while accounting for other NEFA
components, and also the total NEFA. This is also something I struggled to
understand a little in general – why should *proportions* of NEFA components be
important? There may be a relatively high proportion of a particular NEFA but if
the patient has a generally low overall NEFA, is there any reason to think a
relatively high proportion of something will increase insulin sensitivity? These
ideas may be wholly implausible of course and I may have missed the point – as I
said my background knowledge of the biology is not ideal."*

- *"I do also wonder somewhat whether there may be over-adjustment with some of the
covariates included, and agree with another reviewer who pointed out the
importance of the unadjusted results. A potential issue is confounders that may
be on the causal pathway, which is a judgement that must be based on expert
knowledge. The authors note for instance that adjusting for WC improved the fit
of the model, and was indicated for inclusion by the DAG. Neither in my mind is
reason alone for its inclusion, and there is no guarantee on this basis that
adjustment for this is necessarily appropriate – this must be based on
subject-specific knowledge of whether adjustment is appropriate to determine the
independent effect of the variable of interest."*

- *"Which brings me to the DAGs. "Directed acyclic graphs" just means a collection
of boxes representing different bits of information with arrows between them
that do not form loops; DAGs are used in Bayesian analysis for instance to
represent how data and parameters in a model are linked. A DAG therefore is a
representation for how information in linked, not a process to in itself to
determine appropriate model building or make a "recommendation". I think that if
the approach employed is to be used as justification of the model building
process, it needs to be better described: presumably there is some algorithm
being used that determines interrelationships of variables in the data, of which
I would like to see at least a brief outline – a link to a website is not really
sufficient here. Further, the graphs provided give no indication of what the
different arrows and colours are."*

- *"Last of all, the analysis excluded diabetes cases, which is justified. However,
could incident cases be included with follow up time included to the last visit
prior to developing diabetes? This may not be so straightforward in practice of
course. It’s not really clear if any of any pre-incident follow up time is
included at present, or if all data on incident cases is discarded "incident
diabetes cases were excluded from further GEE analysis" – does the "further" bit
mean post-incidence here?"*

- *"MINOR COMMENTS:"*
    - *"Abstract and elsewhere: OGTT not spelled out, even if this is obvious to
    most readers."*
    - *"List of abbreviations: if this is to be an exhaustive list many are not
    given, OGTT, HOMA, NEFA"*

    - *"QIC - presumably this is quite like the AIC and other information criteria -
    maybe mention this, and how the calculation differs? Is the QIC specifically for
    GEE models? Also in the ESM: "This identifies a specific model from a number of
    models that is the 'best' fit and least complex of all the models" - it can’t be
    the best fit AND least complex, as mentioned in the previous sentence it’s a
    trade-off - so maybe omit this sentence."*

    - *"ESM - collider variables. As mentioned in main comments this is all a
    bit vague - "arrows appearing to collide" and so on doesn’t really convey
    what is going on here. This is a technical description so please be a little
    more precise!"*

## Referee: 2

- *"In this manuscript, the authors investigated the link between total and several
fractions of serum NEFA and the clinical indexes of insulin secretion and
sensitivity by using GEE model in a diabetes risk group over 6-years
longitudinal follow-up."* ... *"Overall, this study is interesting and novel to
show the long-term exposures to some specific NEFA fractions are negatively
associated with mainly beta-cell function in human, which convincingly confirms
and advances the previous basic and clinical studies in this field."*

1. *"Serum insulin and/or NEFA levels could be affected by several factors: I
would suggest excluding the participants who have endocrine, hepatic or renal
disorders. Also, I suggest mentioning the detail of medications, which could
affect insulin/glucose/NEFA measurement; page 24, line 7, "Medication use did
not influence NEFA composition, the outcomes, or the GEE models (data not
shown)" is not clear enough."*

2. *"Glucose clamp test is the standard to evaluate insulin sensitivity instead
of the indexes used in this study. I know it is very hard to perform it in a
large number of participants, but it may be one of the reasons why the effect of
NEFAs on insulin sensitivity was limited. I would suggest discussing the
limitation of these indexes."*

3. *"Most of the indexes in this study are based on glucose and insulin levels.
I think the reader would know the data of baseline and follow-up glucose and
insulin levels as well as indexes (in Table 1)."*

4. *"I understand several NEFA fractions are statistically related to indexes of
beta-cell function; however, does each individual show the same trend? For
example, the graph which shows NEFA concentrations and the individual decline
ratio of beta-cell function over 6-years (value of slope, if compatible) may
convince the reader more, at least for the representative NEFA fractions."*

5. *"Referring previous longitudinal studies showing the link between NEFA level
and beta cell deterioration, even which were investigated in the diabetic
patients, could confirm the results of this study."*

## Referee: 3

- *"The authors addressed all major concerns related to the manuscript. I feel
that the paper improved by the changes and became more focused by omitting the
PLS-DA analysis. By adding a detailed explanation to the discussion on the
potential importance of potential confounders and mediators, the interpretation
of the findings became much clearer."*

- *"Some minor comments regarding the wording of the conclusions and missing
information on the DAGs remain but I only have the following comments:"*

- *"Minor comments"*
    1. *"Introduction: "Therefore, our objective was to examine the association of
    serum NEFA composition on changes over time in insulin sensitivity and beta-cell
    function in a longitudinal cohort. We hypothesized that higher palmitic acid and
    lower PUFA such as eicosapentaenoic acid in the NEFA fraction would associate
    with declining insulin sensitivity and beta-cell function over 6 years." I
    would prefer to use the word differences instead of changes, as the main
    focus of the paper is not on the time interaction but on the main effects."*

    2. *"Conclusions: "In a Canadian population of adults who were at-risk for
    diabetes, we found that higher total NEFA concentrations independently predicted
    lower beta-cell function after 6 years." I would prefer to use over 6 years
    instead of after 6 years, as the difference was present from the measurement
    point to the end of follow-up."*

    3. *"Please add the number of observations for each clinic visit in Table 1."*

    4. *"Please add the number of incident and prevalent diabetes cases to the
    respective exclusions in ESM Figure 1. For example: n=736 had baseline data
    collected, n=XX diabetes cases were excluded; n=619 had 3-year data, n=XX
    incident diabetes cases were excluded..."*

    5. *"Please add the colour codes for DAGs (i.e.: exposure, outcome, ancestor
    of exposure, ancestor of outcome, ancestor of both, causal path, biasing
    path)."*

## Referee: 4

- *"This is an interesting analysis on NEFA ,insulin sensitivity and beta cell
function."* ... *"My major concern is that with the current approach the
complexity of metabolic regulation may be lost to some extent. This makes it
hard to interpret these findings with respect to biological significance:"*

    - *"With respect to peripheral insulin sensitivity (referred to in the
    discussion) the effects of saturated fatty acids may strongly depend on
    metabolic phenotype, ie more pronounced in insulin resistant subjects, was
    initial metabolic phenotype taken into account in the present analysis?"*

    - *"it is a pity that the analysis of fatty acids composition in
    phospholipids and serum cholesteryl esters have been published in a separate
    paper, altogether this would add very much to the total interpretation."*

    - *"some discussion on what determines plasma fatty acid profile may be
    useful, beside spillover also (selective) release by endogenous lipolysis.
    To what extent does it reflect dietary composition, was habitual fat intake
    taken into account?"*

- *"Altogether, this paper would gain strength by addressing the limitations in
interpretation in more detail."*

## Associate Editor

- *"The reviewers provide numerous minor and major comments, for example regarding
motivation of the statistical approach, which require revision or careful
response."*

### 2017-09-29 doc/manuscript.Rmd
changes to differences

@@
 
-Aim: Our aim was to determine the longitudinal associations of individual non-esterified fatty acids (NEFA) on changes in insulin sensitivity (IS) and beta-cell function over 6 years in a cohort of individuals who are at-risk for diabetes.
+Aim: Our aim was to determine the longitudinal associations of individual non-esterified fatty acids (NEFA) on differences in insulin sensitivity (IS) and beta-cell function over 6 years in a cohort of individuals who are at-risk for diabetes.
 
@@
 effects of individual fatty acids. Therefore, our objective was to examine the
-association of serum NEFA composition on changes over time in insulin
+association of serum NEFA composition on differences over time in insulin
 sensitivity and beta-cell function in a longitudinal cohort. We hypothesized

### 2017-09-29 doc/manuscript.Rmd
after to over

@@
 higher total NEFA concentrations independently predicted lower beta-cell
-function after 6 years. While we found negative associations with palmitic acid
+function over 6 years. While we found negative associations with palmitic acid
 (16:0), oleic acid (18:1n-9), and linoleic acid (18:2n-6) for the concentration

### 2017-09-29 doc/manuscript.Rmd
Remove total label from facet_grid (gee)

@@
     dplyr::filter(model == "Unadjusted") %>% 
+    dplyr::mutate(unit = ifelse(unit == "Total", " ", unit)) %>% 
     plot_gee_main()
@@
     dplyr::filter(model == "Adjusted") %>% 
+    dplyr::mutate(unit = ifelse(unit == "Total", " ", unit)) %>% 
     plot_gee_main()

### 2017-09-29 doc/esm.Rmd
Merged esm into one file.

---
output: fost::diabetologia_manuscript
---

```{r setup, include=FALSE}
devtools::load_all()
set_options()
knitr::opts_chunk$set(results = "asis")
```

# ESM Methods

## Equations for outcomes

The equations for calculating the outcome variables are listed below. HOMA2-%S was 
calculated using the Excel found at https://www.dtu.ox.ac.uk/homacalculator/.

```{r}
cat(PROMISE.methods::insert_formula("ISI"))
```

```{r}
cat(PROMISE.methods::insert_formula("HOMA"))
```

```{r}
cat(PROMISE.methods::insert_formula("IGIIR"))
```

```{r}
cat(PROMISE.methods::insert_formula("ISSI2"))
```

Area under the curve (AUC) for both glucose (G) and insulin (I) was computed
using the trapezoidal rule. ISI~OGTT~ (ISI) uses all the glucose and insulin
values from the oral glucose tolerance (the overall mean).

## Further description of statistical methods

Generalized estimating equations (GEE) is a technique similar to mixed effects 
modeling, except it calculates the marginal population estimates compared to the
subject-specific estimates in mixed effects models. GEE is well suited to and
commonly used in longitudinal cohort studies, especially given its capacity to
handle missed visits. GEE is able to handle longitudinal data due to the
estimating of a working correlation matrix, which was chosen based on
quasi-likelihood information criteria (QIC). The auto-regressive of order 1
(AR1) matrix was chosen for the GEE models as it had the best model fit accessed
using QIC, though other matrices (eg. exchangeable) had similar fit (data not
shown).

For the covariates, they were chosen based on previous literature, from 
directed acyclic graph (DAG) recommendations, and from QIC. The DAG
recommendations were obtained from using the DAGitty software,
http://dagitty.net/. DAG is a technique that allows for identifying potential
confounding causal pathways and to limit potential bias introduced from
adjusting for colliding variables. A collider variable is a variable that is
determined from two or more other variables, such that in causal graph modeling,
the arrows from the other variables appear to 'collide' into the variable. QIC 
is a technique that balances the goodness of fit of the model with its relative
complexity (i.e. number of covariates) against other models being compared. This
identifies a specific model from a number of models that is the 'best' fit and
least complex of all the models. QIC and other information criterion methods are
used in conjunction with other techniques such as literature and DAG based. The 
final GEE model selected as best fitting was negligibly different when using 
insulin sensitivity or beta-cell function as an outcome and as such the same 
covariates were chosen for models with each of the outcomes. The covariates
medication use, WC, ALT, and MET would be considered classical confounders while
TAG would be considered a mediating variable.

Partial least squares (PLS) regression is a technique similar to principal
component analysis (PCA) that is designed to extract meaningful information from
multivariate, high dimensionality data (e.g. as in metabolomic and other -omic type 
analyses). These types of methods try to extract as much of the variance in the 
data into a smaller number of components or factors. The difference between PCA 
and PLS is that PLS is a supervised method, while PCA is not. This means that
PLS uses a response variable(s), i.e. an outcome or **Y**, to 'supervise' (or
constrain) the variation inherent in the predictors (**X**) while PCA only
describes the variation inherent in **X**. Because of this, PLS can be better at
predicting the contributions of predictor variables against an outcome variable.

```{r setup, include=FALSE}
devtools::load_all()
set_options()
```

```{r captions}
# Include captions below using `captioner` package via library(captioner)
stab <- captioner::captioner(prefix = 'ESM Table')
cite_st <- pryr::partial(stab, display = 'cite')

stab_qic <- stab('qic', 'Comparing generalized estimating equation models adjusting for different covariates using Quasi-Likelihood Information Criterion.')
stab_dist <- stab('nefa', 'Raw concentration values of each non-esterified fatty acid in the PROMISE cohort at the baseline visit (2004-2006). Values are presented as mean (SD).')
stab_gee_unadj <- stab('gee_unadj', 'Unadjusted generalized estimating equation models of the longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.  GEE models are only adjusted for time. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values were adjusted for the BH false discovery rate, with significant (p<0.05) associations indicated by asterisk.')
stab_gee_adj <- stab('gee_adj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort. GEE models were adjusted for time, sex, ethnicity, baseline age, WC, ALT, and family history of diabetes. Outcome variables were log-transformed, predictor variables were scaled, and x-axis values were exponentiated to represent percent difference per SD increase in the fatty acid.  P-values were adjusted for the BH false discovery rate, with significant (p<0.05) associations indicated by asterisk (*).')
```

# ESM Tables

```{r ESM_Table1, results="asis"}
analyze_qic() %>% 
    table_qic(stab_qic)
```

Given the number of possible combinations of outcome and predictor variables,
only ISI and ISSI-2 with total non-esterified fatty acids (nmol/mL) were used to
compare various GEE models and to select a final model.  Baseline age was used
as including both the original age and the time variable would result in
collinearity.  Column names are: QIC is the quasi-likelihood information criteria
(smaller values, eg. larger negative values, indicate a better fit compared to
other models), Delta is the QIC minus the lowest QIC (models that have a delta <10 
between them are considered equivalent).  Models were:

- M0: log(ISSI-2) or log(ISI) = total non-esterified fatty acids (nmol/mL) + years from baseline
- M1: M0 + NEFA by time interaction
- M2: M0 + sex + ethnicity + baseline age
- M3: M2 + waist
- M4: M3 + ALT
- M5: M3 + physical activity
- M6: M3 + alcohol intake
- M7: M3 + family history of diabetes
- M8: M3 + smoking status
- M9: M3 + ALT + physical activity
- M10: M3 + medication use for lipid and cholesterol control
- M11: M3 + NEFA by time interaction
- M12: M9 + NEFA by time interaction

```{r ESM_Table2}
table_distribution(project_data, stab_dist)
```

```{r ESM_Table3_4}
gee_results %>% 
    dplyr::filter(model == "Unadjusted") %>% 
    table_gee_main(stab_gee_unadj)
gee_results %>% 
    dplyr::filter(model == "Adjusted") %>% 
    table_gee_main(stab_gee_adj)
```

```{r setup, include=FALSE}
devtools::load_all()
set_options()
knitr::opts_chunk$set(dev = "png")

# Include captions below using `captioner` package via library(captioner)
sfig <- captioner::captioner(prefix = 'ESM Fig.')
cite_sf <- pryr::partial(sfig, display = 'cite')

sfig_consort <- sfig('consort', 'CONSORT diagram of sample size at each examination visit for the main PROMISE cohort in addition of the sample size determination for the present analysis.')
sfig_dag_isi <- sfig("dag_isi", "Directed acyclic graph taken from DAGitty (DAGitty.net) for the potential causal pathway between NEFA and ISI. DAGitty recommended a minimal adjustment of waist, sex, and MET (physical activity).")
sfig_dag_issi2 <- sfig("dag_issi2", "Directed acyclic graph taken from DAGitty (DAGitty.net) for the potential causal pathway between NEFA and ISSI-2. DAGitty recommended a minimal adjustment of waist, ethnicity, and MET (physical activity).")
sfig_corr_conc <- sfig('corr_conc', 'Pearson correlation heatmap of non-esterified fatty acids (*nmol/mL*) and basic PROMISE participant characteristics for the baseline visit (2004-2006). Darkness of the colour indicates the magnitude of the correlation, with orange indicating positive and blue indicating negative correlations.')
sfig_corr_pct <- sfig('corr_pct', 'Pearson correlation heatmap of non-esterified fatty acids (*mol%*) and basic PROMISE participant characteristics for the baseline visit (2004-2006). Darkness of the colour indicates the magnitude of the correlation, with orange indicating positive and blue indicating negative correlations.')
sfig_gee_adj_nowaist <- sfig('gee_adj_nowaist', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort. Generalized estimating equation models were fully adjusted, *excluding waist circumference*. X-axis values represent a percent difference in the outcome per SD increase in the fatty acid. P-values were adjusted for the BH false discovery rate, with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, and p<0.01.')
```

# ESM Figures

![`r sfig_consort`](../img/flowDiagramSample.png)

![`r sfig_dag_isi`](../img/dagitty-isi.png)

![`r sfig_dag_issi2`](../img/dagitty-issi2.png)

```{r ESM_Fig4, fig.cap=sfig_corr_conc}
plot_corr_heatmap(corr_results_conc, unit = "nmol/mL")
```

```{r ESM_Fig5, fig.cap=sfig_corr_pct}
plot_corr_heatmap(corr_results_pct, unit = "mol%")
```

```{r ESM_Fig6, fig.cap=sfig_gee_adj_nowaist}
gee_results %>% 
    dplyr::filter(model == "Adjusted no waist") %>% 
    dplyr::mutate(unit = ifelse(unit == "Total", " ", unit)) %>% 
    plot_gee_main()
```

### 2017-10-02 doc/manuscript.Rmd
data availability statement
beta-cell -> beta cell

@@
 Title: Association of non-esterified fatty acid composition with insulin
-sensitivity and beta-cell dysfunction in the Prospective Metabolism and Islet Cell
+sensitivity and beta cell dysfunction in the Prospective Metabolism and Islet Cell
 Evaluation (PROMISE) cohort
@@
 
-Aim: Our aim was to determine the longitudinal associations of individual non-esterified fatty acids (NEFA) on differences in insulin sensitivity (IS) and beta-cell function over 6 years in a cohort of individuals who are at-risk for diabetes.
+Aim: Our aim was to determine the longitudinal associations of individual non-esterified fatty acids (NEFA) on differences in insulin sensitivity (IS) and beta cell function over 6 years in a cohort of individuals who are at-risk for diabetes.
 
@@
 Outcome variables were calculated using the OGTT values.
-IS was assessed using HOMA2-%S and the Matsuda index, while beta-cell function was  assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin  Secretion-Sensitivity Index-2 (ISSI-2).
+IS was assessed using HOMA2-%S and the Matsuda index, while beta cell function was  assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin  Secretion-Sensitivity Index-2 (ISSI-2).
 Generalized estimating equations were  used, adjusting for time, waist, sex, ethnicity, baseline age, alanine  aminotransferase, and physical activity.
@@
 
-Results: Participants (73% female, 70% with European-ancestry) had their IS and beta-cell function decline by 14-27% over 6 years of follow-up.
+Results: Participants (73% female, 70% with European-ancestry) had their IS and beta cell function decline by 14-27% over 6 years of follow-up.
 While in unadjusted models there were several NEFA (e.g.
@@
 
-Conclusions: Total NEFA concentration was a strong predictor of lower beta-cell  function over 6 years.
-Our results suggest that the association with beta-cell  function is due to the absolute size of the serum NEFA fraction, rather than the specific fatty acid composition.
+Conclusions: Total NEFA concentration was a strong predictor of lower beta cell  function over 6 years.
+Our results suggest that the association with beta cell  function is due to the absolute size of the serum NEFA fraction, rather than the specific fatty acid composition.
 
@@
 
-insulin sensitivity; beta-cell function; longitudinal cohort; non-esterified
+insulin sensitivity; beta cell function; longitudinal cohort; non-esterified
 fatty acids; diabetes pathogenesis; fatty acid composition
@@
 fig_dist <- fig('nefa', 'Concentrations (nmol/mL) of each NEFA in PROMISE participants at the baseline visit (2004-2006).')
-fig_gee_unadj <- fig('gee_unadj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over 6 years in the PROMISE cohort.
+fig_gee_unadj <- fig('gee_unadj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta cell function over 6 years in the PROMISE cohort.
 Models are only adjusted for time.
@@
 P-values were adjusted for the BH false discovery rate, with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, p<0.01, and p<0.001.')
-fig_gee_adj <- fig('gee_adj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta-cell function over the 6 years in the PROMISE cohort.
+fig_gee_adj <- fig('gee_adj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta cell function over the 6 years in the PROMISE cohort.
 Generalized estimating equation models were fully adjusted.
@@
 insulin-sensitive tissues such as muscle and liver, and can impair 
-pancreatic beta-cell production of insulin [@Rachek2014a;@Cnop2008a]. 
+pancreatic beta cell production of insulin [@Rachek2014a;@Cnop2008a]. 
 Observational and clinical studies have reported concordant findings, showing in
@@
 one study suggested that eicosapentaenoic acid (20:5n-3) can protect against the
-lipotoxic effect of palmitic acid in the beta-cells [@Kato2008].
+lipotoxic effect of palmitic acid in the beta cells [@Kato2008].
 
@@
 association of serum NEFA composition on differences over time in insulin
-sensitivity and beta-cell function in a longitudinal cohort. We hypothesized
+sensitivity and beta cell function in a longitudinal cohort. We hypothesized
 that higher palmitic acid and lower PUFA such as eicosapentaenoic acid in the
-NEFA fraction would associate with declining insulin sensitivity and beta-cell
+NEFA fraction would associate with declining insulin sensitivity and beta cell
 function over 6 years.
@@
 
-Insulin sensitivity and beta-cell function indices were computed using the OGTT 
+Insulin sensitivity and beta cell function indices were computed using the OGTT 
 glucose and insulin data. Insulin sensitivity was assessed using HOMA2-%S [@Levy1998a]
@@
 reflects hepatic insulin sensitivity, while ISI reflects whole-body insulin
-sensitivity [@AbdulGhani2007].  Beta-cell function was assessed using the
+sensitivity [@AbdulGhani2007].  beta cell function was assessed using the
 Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR) and the Insulin
@@
 as a guideline for reporting [@Vandenbroucke2007b].
+The cohort datasets analysed during the current study are available from the
+corresponding author on reasonable request.
 
@@
 participants, `r famhist$npct`, had a family history of diabetes. Insulin
-sensitivity and beta-cell function measures showed a significant median
+sensitivity and beta cell function measures showed a significant median
 decline from baseline to 6-years of between `r chg_outcomes$chg` 
@@
 There were several fatty acids (`r paste(unique(sig_fa_unadj$Xterms),collapse=", ")`)
-that had significant associations with both insulin sensitivity and beta-cell
+that had significant associations with both insulin sensitivity and beta cell
 function outcomes. Most associations were negative, however, only 20:5n-3
@@
 magnitude of association for each of these variables was fairly consistent for
-each beta-cell function measure. For every one SD increase in any of the NEFA
-variables associated with beta-cell function, there was an average predicted
+each beta cell function measure. For every one SD increase in any of the NEFA
+variables associated with beta cell function, there was an average predicted
 `r est_gee$est[1]`% lower IGI/IR and `r est_gee$est[2]`% lower ISSI-2 at any
@@
 In a Canadian population of adults who were at-risk for diabetes, we found that 
-higher total NEFA concentrations independently predicted lower beta-cell
+higher total NEFA concentrations independently predicted lower beta cell
 function over 6 years. While we found negative associations with palmitic acid
@@
 proportion, an observation that was further confirmed by multivariate cluster
-techniques, which found no ability of these individual NEFA to predict beta-cell
+techniques, which found no ability of these individual NEFA to predict beta cell
 function. We did find that one fatty acid as a mol%, eicosapentaenoic acid
@@
 RISCK cohort, total NEFA had a negative association with insulin sensitivity and
-a particularly strong negative association with beta-cell function
+a particularly strong negative association with beta cell function
 [@Johns2014a]. Experimentally, several potential mechanisms have been elucidated
-for the role of NEFA on beta-cell function, particularly for palmitic acid.
-Prolonged exposure to elevated NEFA can induce apoptosis in the beta-cells,
+for the role of NEFA on beta cell function, particularly for palmitic acid.
+Prolonged exposure to elevated NEFA can induce apoptosis in the beta cells,
 possibly through endoplasmic reticulum stress, formation of ceramides, and
@@
 present analysis is the first study to our knowledge to examine the longitudinal
-association of a broad spectrum of individual NEFA on beta-cell function in a
+association of a broad spectrum of individual NEFA on beta cell function in a
 large cohort. We found that there was a strong signal of higher total NEFA,
 palmitic acid, oleic acid, and linoleic acid modeled as concentrations with
-lower beta-cell function. No fatty acid as a mol% had associations with
-beta-cell function. Taken together, these results suggest that it is the
+lower beta cell function. No fatty acid as a mol% had associations with
+beta cell function. Taken together, these results suggest that it is the
 absolute size of the circulating NEFA fraction, irrespective of any specific 
 composition of fatty acids, that is responsible for the hypothesized lipotoxic 
-effects of NEFA on the beta-cells.
+effects of NEFA on the beta cells.
 
@@
 factor for diabetes. In the sensitivity analysis adjusting for TAG, all
-associations with beta-cell function were attenuated, suggesting that NEFA may
-in fact be mediating its association with beta-cell dysfunction through higher
+associations with beta cell function were attenuated, suggesting that NEFA may
+in fact be mediating its association with beta cell dysfunction through higher
 TAG.
@@
 In conclusion, we found that total NEFA was a strong predictor for lower 
-beta-cell function over 6 years, irrespective of the specific composition of the
+beta cell function over 6 years, irrespective of the specific composition of the
 NEFA fraction, suggesting that efforts at lowering circulating total NEFA 

### 2017-10-02 doc/manuscript.Rmd
naming format change (acknow), duality of interest, and guarantor

@@
 
-The authors thank Jan Neuman, Paula Van Nostrand, Stella Kink, and Annette
-Barnie of the Leadership Sinai Centre for Diabetes, Mount Sinai Hospital,
-Toronto, Canada and Sheila Porter and Mauricio Marin of the Centre for Studies
-in Family Medicine, University of Western Ontario, London, Canada for their 
-expert technical assistance and dedication in their work for PROMISE.
+The authors thank J Neuman, PV Nostrand, S Kink, and A Barnie (all of the
+Leadership Sinai Centre for Diabetes, Mount Sinai Hospital, Canada) and
+S Porter and M Marin (both of the Centre for Studies in Family Medicine,
+University of Western Ontario, Canada) for their expert technical
+assistance and dedication in their work for PROMISE.
 
@@
 This study was supported by grants from the Canadian Diabetes Association (CDA),
-the Canadian Institutes for Health Research, and the University of
-Toronto Banting and Best Diabetes Centre; LWJ is supported by a CDA
-Doctoral Student Research Award; RR is supported by a Heart and Stroke
-Foundation of Ontario Mid-Career Investigator Award; SBH holds the CDA Chair in
-National Diabetes Management and the Ian McWhinney Chair of Family Medicine
-Studies at the University of Western Ontario; RBP holds a Tier II Canada
-Research Chair in Brain Lipid Metabolism; AJH holds a Tier II Canada Research
-Chair in Diabetes Epidemiology.
+the Canadian Institutes for Health Research, and the University of Toronto
+Banting and Best Diabetes Centre; LWJ was supported by a CDA Doctoral Student
+Research Award; RR was supported by a Heart and Stroke Foundation of Ontario
+Mid-Career Investigator Award; SBH holds the CDA Chair in National Diabetes
+Management and the Ian McWhinney Chair of Family Medicine Studies at the
+University of Western Ontario; RPB holds a Tier II Canada Research Chair in
+Brain Lipid Metabolism; AJH holds a Tier II Canada Research Chair in Diabetes
+Epidemiology.
 
@@
 
-The authors report no potential conflicts of interest relevant to this study.
+The authors declare that there is no duality of interest associated with this
+manuscript.
 
@@
 feedback on all versions of the paper; LWJ and AJH had primary responsibility
-for final content. All authors read and approved the final manuscript.
+for final content and are the guarantors of this work. All authors read and
+approved the final manuscript.
 

### 2017-10-02 doc/esm.Rmd
more detail about ar1/corr matrix in esm

@@
 using QIC, though other matrices (eg. exchangeable) had similar fit (data not
-shown).
+shown). One main advantage of using methods such as GEE is the inclusion of the
+working correlation matrix that takes into account the inherent covariance
+between values measured at different time points when computing model estimates
+and standard errors. AR1 is the most appropriate correlation matrix for
+longitudinal data since the correlation matrix assumes lower intercorrelation of
+outcomes within an individual as time passes. For example, BMI measured at the
+baseline visit would be more highly correlated to BMI measured at the 3rd year
+visit than compared to the 6th year visit.
 

### 2017-10-02 R/tables.R
glucose and insulin values in table 1

@@ -16,7 +16,7 @@ table_basic <- function(data, caption = NULL) {
         carpenter::add_rows(c('HOMA2_S', 'ISI', 'IGIIR', 'ISSI2'),
                             carpenter::stat_medianIQR, digits = 1) %>%
         carpenter::add_rows(c('BMI', 'Waist', 'Age', 'MET', 'ALT', 'TAG', 'Chol', 'HDL',
-                              'BaseTotalNE'),
+                              'BaseTotalNE', "Glucose0", "Glucose120", "Insulin0", "Insulin120"),
                             carpenter::stat_meanSD,
                             digits = 1) %>%
         carpenter::add_rows(c('Ethnicity', 'Sex'), carpenter::stat_nPct, digits = 0) %>%

### 2017-10-02 doc/submission/reviewer-response-2.md
responses to the reviewers

@@
 
+    We thank the reviewer for the comments and hope that the below responses
+    adequately address issues or suggestions raised.
+
 - *"A key question for me is the choice of using the GEE approach, which is of
@@
 Richer correlation structures, easily handles missing data, robust standard
-errors."*
-
-- *"Here, the number of follow up visits is not large and the amount of missing data
-is relatively small. My first thought when approaching the analysis would
-therefore be to consider a standard linear regression analysis with
+errors. Here, the number of follow up visits is not large and the amount of
+missing data is relatively small. My first thought when approaching the analysis
+would therefore be to consider a standard linear regression analysis with
 robust/clustered standard errors to incorporate repeated measures, and potential
@@
 to add to the word count). Currently the main statement is the ease of handling
-missing data – which is handy but perhaps not essential – and some comments on
+missing data - which is handy but perhaps not essential - and some comments on
 the working correlation matrix, which are not terribly informative. So some
@@
 structure that is not easily handled otherwise, for instance, might be helpful.
-Of note is that in the ESM it is stated that "other … matrices had similar fit"
-– it may be worth noting that the model fit may be similar, but estimates may be
+Of note is that in the ESM it is stated that "other ... matrices had similar fit"
+- it may be worth noting that the model fit may be similar, but estimates may be
 more efficient under a particular correlation structure."*
+
+    The reviewer's suggestion to more fully describe the justification for using
+    GEE over other methods is warranted and we will expand on this in the
+    manuscript in the Methods and the ESM Methods sections. We chose the GEE
+    technique for a variety of reasons. As the reviewer states, GEE is able to
+    handle visit dates that differ between participants (e.g. some participants
+    come exactly 3 years later for a followup while others come some months
+    before or after 3 years) and for dealing with the inherent correlational
+    structure underlying longitudinal data. Because of these features, among
+    others, GEE can compute robust standard errors of the estimates that more
+    accurate reflect the underlying data. The ability to specify a working
+    correlation matrix makes this a powerful technique for longitudinal data
+    compared to standard linear regression, as it removes the underlying
+    assumption that variables are independent (which is not the case with
+    longitudinal data). While choosing the specific correlation matrix is
+    important, GEE still produces robust estimates and standard errors even with
+    a misspecified correlation matrix. In the case of longitudinal data, an
+    auto-regressive correlation matrix is the most appropriate statistically and
+    biologically since measurements on an individual over time will change (e.g.
+    people gain or lose weight over time).
 
@@
 different interpretation. This difference might be seen as a disadvantage to GEE
-modelling – or do the authors feel this is more appropriate for this particular
+modelling - or do the authors feel this is more appropriate for this particular
 question? This is something I do not have a thorough understanding of, but I do
-note that the authors comment in the methods "…the interpretation of the GEE
+note that the authors comment in the methods "...the interpretation of the GEE
 results is as an expected percent difference in the outcome variable for every
@@
 2003;157:364–375)."*
+
+    We were only concerned with population level averages, and not
+    individual-level estimates. GEE is a marginal model, which estimates the
+    population average, and is one of the other reasons we chose GEE. However,
+    even though GEE is a marginal model, for continuous outcomes (e.g. insulin
+    sensitivity) the estimates derived from GEE's counter part, mixed effects
+    models, are more or less equivalent in both the value and in the
+    interpretation. For nonlinear outcomes (e.g. binary values such as diabetes
+    incidence), the interpretation and estimates are substantially different
+    from mixed effects models. We did not model nonlinear outcomes and as such
+    this difference is {{irrelevant?}}.
+
+    Regarding the cited paper, the authors state the paper is an
+    *"...orientation focused on correlated data arising from the relatedness of
+    **several individuals** in the **same** cluster, rather than several
+    "longitudinal" observations in the same individual"* (emphasis added). This
+    is an important distinction and their statement about the "keeping other
+    factors constant" and the lack of terms for the "between-cluster variation"
+    is true in this case. However, this present analysis has the repeated
+    measures at the individual level; the cluster is the individual and the
+    within-cluster variation is the individual over time. In this context,
+    interpreting the estimates derived from GEE is comparable to the classical
+    linear regression equations:
+
+    y = B~0~ + B~1~ X~1~ + ... + B~n~ X~n~
+
+    Such that the interpretation of e.g. total NEFA with ISSI-2 is that for
+    every one SD increase in total NEFA there is an average 5% lower ISSI-2 for
+    individuals at the same time point and of the same age, waist, ethnicity,
+    etc.
 
@@
 in one model."*
+
+    The reviewer brings up excellent points. The reason NEFA were included as a
+    proportion in the PLS analysis rather than as a concentration was due to the
+    inherent interdependence between NEFA values as a proportion. They all must
+    equal to 100. It is this dependency structure in the data that allows PLS to
+    more accurately identify clusters within the data. Concentration data does
+    not have this inherent dependency (all NEFA values do *not* equal to 100).
+    If one NEFA goes up in concentration, other NEFA values do not necessarily
+    go down, as would be the case with the proportion. We have clarified this in
+    the methods. {{not all of this comment is clear to me}}.
 
@@
 
+    {{Don't fully understand this comment, but I'll try to answer}}. We hope the
+    response to the point above partially addresses this comment. To reiterate,
+    the reason NEFA proportion was used in the PLS analysis was due to the
+    inherent interdependency of the values as a proportion. When they are
+    proportions, PLS would be more likely to pick up potential clusters since
+    changes in one value *must* lead to changes in another value as they all
+    equal 100.
+
+    The reason we did not include all NEFA variables (as proportion or as
+    concentration) into a single GEE model is because GEE can not handle this
+    level of complexity, or at least, the interpretation for us and the readers
+    would be nearly totally meaningless or incomprehensible. How the reviewer
+    described setting the variables in the regression equation may be useful for
+    some research questions and with fewer variables. However, we had 22
+    different fatty acids (44 if including mol% and nmol/ml) and 4 different
+    outcomes. The high dimensionality of this analysis makes interpretation
+    using standard linear regression approaches extremely challenging, not to
+    mention introducing potential bias due to correlation among the NEFA
+    (especially as a proportion). These reasons are why we chose to run multiple
+    GEE models for each NEFA-outcome pairing in addition to running a PLS
+    analysis with all NEFA in a single model (since PLS is designed for this
+    particular analysis).
+
+    The reviewer is correct that proportions *alone* are not enough to draw a
+    meaningful conclusion, as per how the reviewer described it. This, we
+    believe, is a major strength of our study since previous literature using other
+    fatty acid fractions (e.g. phospholipid) generally only investigate
+    proportions but do not report on or measure the concentration. Concentration
+    data is more technically challenging to quantify and so is understandable
+    why large cohort studies do not measure this data. Using both concentration
+    and proportion, we can identify either it is *both* the total size of any
+    individual NEFA and a specific proportion of a single or multiple individual
+    NEFA that associates with risk for diabetes. Or whether it is only the total
+    concentration, as we found in this study.
+
 - *"I do also wonder somewhat whether there may be over-adjustment with some of the
@@
 reason alone for its inclusion, and there is no guarantee on this basis that
-adjustment for this is necessarily appropriate – this must be based on
+adjustment for this is necessarily appropriate - this must be based on
 subject-specific knowledge of whether adjustment is appropriate to determine the
 independent effect of the variable of interest."*
+
+    {{I'm not entirely clear how to address this. Since WC hypothetically could
+    lead to higher NEFA as well as lower IS (in particular), and since we are
+    testing whether higher NEFA lowers IS, how can we not adjust for it? It's a
+    classic confounder. I'll take a stab at answering it though.}} There may be
+    some potential over-adjustment in our models. However, we have taken
+    extensive steps to understand the biological and causal pathways underlying
+    our associations and to empirically determine which covariates to include in
+    the model.
+
+    In the case of WC, we feel there is substantial grounds for adjusting for it
+    when studying NEFA on metabolic capacity: 1) biologically NEFA comes from
+    adipose tissue; 2) intra-abdominal fat tissue is more metabolically active;
+    3) adipose tissue is an active endocrine organ that contributes to insulin
+    sensitivity and metabolism in general; and 4) WC is a direct measure of
+    abdominal (and indirectly intra-abdominal) fat mass. It is for these reasons
+    that adjusting for WC is, we believe, strongly warranted.
+
+    Having said that, because of comments from previous reviewers about this
+    very issue, we included an analyses to address this issue. We ran models
+    adjusted for all covariates except for WC and reported on the results in the
+    ESM material. We had also included a discussion about this issue in
+    conclusions, which we hope will inform the interested reader about this
+    particular issue and allow them to draw their own interpretations of the
+    results based on this raised issue.
 
@@
 
+    We agree with the reviewer that the wording around DAG use in model building
+    was not entirely clear and we have worked to make it clearer in the ESM and
+    methods. DAGs *are* in fact a process to identify potential covariates to
+    include in a model when used algorithmically. This mathematically construct
+    underlying DAGs and model building is the reason the DAGitty software was
+    developed, so that this mathematical process could be automated. Much
+    greater detail on how DAGs are used to build models can be seen in the two
+    references listed below. The DAG figures in the manuscript will be described
+    in more detail, including what each colour and arrow indicates.
+
+    - Greenland S, Pearl J, Robins JM (1999) Causal diagrams for epidemiologic
+    research. Epidemiology 10:37–48.
+    - Textor J, Hardt J, Knüppel S (2011) DAGitty: A graphical tool for
+    analyzing causal diagrams. Epidemiology 22:745.
+
 - *"Last of all, the analysis excluded diabetes cases, which is justified. However,
@@
 
+    Yes, all follow-up data is included *until* the visit with the incident
+    diabetes case (e.g. a participant is diagnosed with diabetes at the 6 year
+    visit, but the baseline and 3rd year visit data are still included in the
+    analysis). This reason, among others, is why GEE and similar techniques are
+    so powerful. The wording has been fixed to make this point clearer.
+
 - *"MINOR COMMENTS:"*
@@
     most readers."*
+
+        This suggestion has been added to the abstract.
+
     - *"List of abbreviations: if this is to be an exhaustive list many are not
     given, OGTT, HOMA, NEFA"*
+
+        These have been added as suggested.
 
@@
 
+        The reviewer is correct that QIC is similar to AIC. We have expanded on
+        the explanation about QIC vs other information criterion. The reviewer
+        is correct that QIC is specific to GEE-type models. Because GEE uses
+        quasi-likelihood rather than maximum likelihood estimation (as other
+        techniques use such as mixed effects models) information criteria such
+        as AIC can not be used since they rely on maximum likelihood. We agree
+        that the wording surrounding 'fit' and 'complex' could be clearer.
+        However, we disagree that it should be omitted since techniques such as
+        QIC *do* select models that have both a good fit and are less complex.
+        While techniques such as AIC are powerful model selection methods, they
+        have an inherent limitation in that models with more parameters will
+        have a better fit. So models with more parameters (more 'complex') be
+        more likely to be selected by default when using AIC. QIC and other
+        similar information criteria techniques on the other hand try to balance
+        this limitation by explicitly including terms that penalize a more
+        'complex' model (more parameters) while still emphasizing a good fit. We
+        have clarified this in the manuscript with the wording: "... that has
+        both a good model fit while keeping model complexity low (less variables
+        included)."
+
     - *"ESM - collider variables. As mentioned in main comments this is all a
@@
     more precise!"*
+
+        We have added this to the ESM Methods section.
 
@@
 
+    We thank the reviewer for this comments and the suggested edits below. We
+    have incorporated most of the suggestions and have provided responses that
+    hopefully clarify other questions and comments.
+
 1. *"Serum insulin and/or NEFA levels could be affected by several factors: I
@@
 
+    {{confirm that participants didn't have other disease}}. We have expanded on
+    the details about medication use (i.e. general types taken).
+
 2. *"Glucose clamp test is the standard to evaluate insulin sensitivity instead
@@
 
+    We agree, this suggestion is pertinent and has been included in the
+    limitations section of the conclusion.
+
 3. *"Most of the indexes in this study are based on glucose and insulin levels.
@@
 insulin levels as well as indexes (in Table 1)."*
+
+    Both glucose and insulin have been added to Table 1.
 
@@
 
+    {{This comment is not entirely clear to me}} This is an interesting
+    question. GEE does not model on the individual level, only on the population
+    as a whole. As such, we cannot answer research questions specific to
+    individuals. There are statistical techniques that can extract individual
+    level associations, however, our objective was (at least initially) to
+    examine population level associations. We therefore chose statistical
+    techniques suitable to the study design and our research questions, which
+    was in this case GEE modelling.
+
 5. *"Referring previous longitudinal studies showing the link between NEFA level
@@
 patients, could confirm the results of this study."*
+
+    We have added some relevant references and associated discussions to the
+    conclusions.
 
@@
 of the findings became much clearer."*
+
+    We thank the reviewer for the comments and have made the suggested changes
+    as requested below.
 
@@
 
+        We agree with this wording. It has been edited.
+
     2. *"Conclusions: "In a Canadian population of adults who were at-risk for
@@
 
+        We agree with this wording. It has been edited.
+
     3. *"Please add the number of observations for each clinic visit in Table 1."*
+
+        As requested, this has been added lto Table 1.
 
@@
 
+        This has been added to ESM Figure 1.
+
     5. *"Please add the colour codes for DAGs (i.e.: exposure, outcome, ancestor
@@
     path)."*
+
+        This has been added to the DAG ESM Figures, as requested.
 
@@
 
+    The initial metabolic phenotype and subsequent influence of saturated fatty
+    acids is taken into account in the time-by-fatty acid interaction. {{expand}}
+
     - *"it is a pity that the analysis of fatty acids composition in
@@
     paper, altogether this would add very much to the total interpretation."*
+
+    While we agree with the reviewer that inclusion of the phospholipid and
+    cholesteryl ester farctions would have added to the interpretation, the
+    volume and complexity of the analyses would not have been publishable with
+    current journal word and figure count limits. This was the reason we decided
+    to publish the fractions in separate manuscripts, to maintain a clear and
+    focus narrative within the constraints of the journals limits {{rephrase}}.
+    {{mention my thesis?}}
 
@@
 response."*
+
+    We hope that the revisions made and the responses to the reviewers
+    adequately address all of their concerns or comments.

### 2017-10-02 doc/manuscript.Rmd
superscript affil
remove fax number for tony

@@
 <!-- First, middle initial, last -->
-Author: Luke W. Johnston (1); Stewart B. Harris (2); Ravi Retnakaran
-(3,4); Adria Giacca (5); Zhen Liu (1);
-Richard P. Bazinet (1); and Anthony J. Hanley (1,3,6)
+Author: Luke W. Johnston^1^; Stewart B. Harris^2^; Ravi Retnakaran
+^3,4^; Adria Giacca^5^; Zhen Liu^1^;
+Richard P. Bazinet^1^; and Anthony J. Hanley^1,3,6^
 
@@
 - Phone number: 416.978.3616
-- Fax number:
 - Email: anthony.hanley@utoronto.ca

### 2017-10-03 doc/manuscript.Rmd
IS -> insulin sensitivity, analyze -> se

@@
 
-Aim: Our aim was to determine the longitudinal associations of individual non-esterified fatty acids (NEFA) on differences in insulin sensitivity (IS) and beta cell function over 6 years in a cohort of individuals who are at-risk for diabetes.
+Aim: Our aim was to determine the longitudinal associations of individual non-esterified fatty acids (NEFA) on differences in insulin sensitivity and beta cell function over 6 years in a cohort of individuals who are at-risk for diabetes.
 
@@
 Outcome variables were calculated using the OGTT values.
-IS was assessed using HOMA2-%S and the Matsuda index, while beta cell function was  assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin  Secretion-Sensitivity Index-2 (ISSI-2).
+Insulin sensitivity was assessed using HOMA2-%S and the Matsuda index, while beta cell function was  assessed using the Insulinogenic index over HOMA-IR (IGI/IR) and the Insulin  Secretion-Sensitivity Index-2 (ISSI-2).
 Generalized estimating equations were  used, adjusting for time, waist, sex, ethnicity, baseline age, alanine  aminotransferase, and physical activity.
-NEFA were analyzed as both  concentrations (nmol/mL) and proportions (mol%) of the total fraction.
+NEFA were analysed as both  concentrations (nmol/mL) and proportions (mol%) of the total fraction.
 
-Results: Participants (73% female, 70% with European-ancestry) had their IS and beta cell function decline by 14-27% over 6 years of follow-up.
+Results: Participants (73% female, 70% with European-ancestry) had their insulin sensitivity and beta cell function decline by 14-27% over 6 years of follow-up.
 While in unadjusted models there were several NEFA (e.g.

### 2017-10-03 doc/manuscript.Rmd
analyze -> analyse

@@
 acids comprise multiple molecules with diverse physiological functions, and few 
-studies have analyzed the effects of a broader spectrum of fatty acids. For instance,
+studies have analysed the effects of a broader spectrum of fatty acids. For instance,
 one study suggested that eicosapentaenoic acid (20:5n-3) can protect against the
@@
 points. Specific insulin was measured using the Elecsys 1010 (Roche Diagnostics,
-Basel, Switzerland) immunoassay analyzer and electrochemiluminescence
+Basel, Switzerland) immunoassay analyser and electrochemiluminescence
 immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
@@
 While GEE accounts for the longitudinal design of the data, this approach is
-limited in that it cannot analyze the inherent multivariate nature of the
+limited in that it cannot analyse the inherent multivariate nature of the
 composition of the NEFA fraction, as each model only contains data from a single
@@
 considered or that could not be measured. Nonetheless, potential covariates were
-empirically analyzed prior to inclusion into the GEE models to best understand
+empirically analysed prior to inclusion into the GEE models to best understand
 and minimize potential confounding. While there may be potential overadjustment
@@
 Few studies have examined the composition of NEFA on metabolic functioning. One 
-recent, well-analyzed study used a variety of advanced fatty acid measurement 
+recent, well-analysed study used a variety of advanced fatty acid measurement 
 and statistical techniques to explore the multivariate relationship between the 
@@
 
-The authors had the following responsibility: LWJ conducted research, analyzed
+The authors had the following responsibility: LWJ conducted research, analysed
 data, and wrote the paper; RR, ZL, and SBH designed research, conducted

### 2017-10-03 doc/manuscript.Rmd
UK spelling

@@
 Much of the previous experimental work on the role of NEFA in type 2 diabetes 
-utilized individual fatty acids such as palmitic acid (16:0) or oleic acid 
+utilised individual fatty acids such as palmitic acid (16:0) or oleic acid 
 (18:1n-9), or alternatively used specific oils such as soybean oil, which is 
 high in the polyunsaturated fatty acid (PUFA) linoleic acid (18:2n-6), as the 
-exposure to characterize the metabolic impact of total NEFA. However, fatty
+exposure to characterise the metabolic impact of total NEFA. However, fatty
 acids comprise multiple molecules with diverse physiological functions, and few 
@@
 each visit is shown in `r cite_sf('consort')`. At each examination,
-participants undergo metabolic characterization, anthropometric measurements,
+participants undergo metabolic characterisation, anthropometric measurements,
 and questionnaires on lifestyle and sociodemographics. Research ethics approval
@@
 and all participants provided written informed consent.
-Data collection methods were standardized across the 2 centres and research
+Data collection methods were standardised across the 2 centres and research
 nurses were centrally trained.
@@
 Impaired fasting glucose (IFG), impaired glucose tolerance (IGT), and diabetes
-were categorized using the 2006 WHO criteria [@WHO2006].
+were categorised using the 2006 WHO criteria [@WHO2006].
 
@@
 and TAG) was isolated using thin layer chromatography; each fraction was
-visualized under UV light after lightly spraying with 8-anilino-1-naphthalene
+visualised under UV light after lightly spraying with 8-anilino-1-naphthalene
 sulfonic acid (0.1% wt/vol) and then converted to fatty acid methyl esters with
@@
 handle missed visits. The predictor variables and continuous covariates were
-scaled (mean centered and standardized). The NEFA variables were classified as 
+scaled (mean centered and standardised). The NEFA variables were classified as 
 *time-independent* (held constant) as they were measured only at the baseline 
@@
 empirically analysed prior to inclusion into the GEE models to best understand
-and minimize potential confounding. While there may be potential overadjustment
+and minimise potential confounding. While there may be potential overadjustment
 due to controlling for WC, both QIC and directed acyclic graph outputs
@@
 not influence the results presented from this cohort, our findings may not
-entirely be generalizable to other populations. However, given these
+entirely be generalisable to other populations. However, given these
 limitations, our study also has several strengths, including the longitudinal
@@
 NEFA and the subsequent increase in insulin resistance via impairment of insulin 
-signaling cascades, as reviewed in previously published articles 
+signalling cascades, as reviewed in previously published articles 
 [@Martins2012a;@Ebbert2013a;@Capurso2012a]. However, in this longitudinal

### 2017-10-03 doc/esm.Rmd
UK spelling

@@
 Generalized estimating equations (GEE) is a technique similar to mixed effects 
-modeling, except it calculates the marginal population estimates compared to the
+modelling, except it calculates the marginal population estimates compared to the
 subject-specific estimates in mixed effects models. GEE is well suited to and
@@
 adjusting for colliding variables. A collider variable is a variable that is
-determined from two or more other variables, such that in causal graph modeling,
+determined from two or more other variables, such that in causal graph modelling,
 the arrows from the other variables appear to 'collide' into the variable. QIC 

### 2017-10-03 R/gee.R
changes to gee figure as per journal standard

@@ -95,19 +95,24 @@ tidy_gee_results <- function(results) {
 #' @param results Results from the GEE modeling.
 #'
 plot_gee_main <- function(results) {
+    legend_title <- ~ atop(
+        paste("FDR-adjusted"),
+        paste(italic(p), "-value          ")
+    )
     results %>%
         seer::view_main_effect(
             graph.options = "dot.grey",
             groups = 'unit~Yterms',
-            legend.title = 'FDR-adjusted\np-value',
-            xlab = 'Percent difference with 95% CI in the outcomes\nfor each SD increase in fatty acid',
-            ylab = 'Non-esterified fatty acids'
+            legend.title = legend_title,
+            xlab = 'Percent difference with 95% CI in the outcomes\nfor each SD increase in NEFA',
+            ylab = 'NEFA'
             ) +
         graph_theme(ticks = FALSE, legend.pos = "right") +
-        ggplot2::facet_grid(unit~Yterms, switch = 'y',
+        ggplot2::facet_grid(unit~Yterms, switch = 'both',
                             scales = 'free_y',
-                            space = 'free_y') +
-        ggplot2::theme(panel.spacing = grid::unit(0.6, "lines"))
+                            space = 'free_y', as.table = FALSE) +
+        ggplot2::theme(panel.spacing = grid::unit(0.6, "lines"),
+                       strip.placement = "outside")
 }
 
 #' Create a table from the GEE results.

### 2017-10-03 R/renaming.R
L -> l

@@ -7,14 +7,19 @@ renaming_table_rows <- function(x) {
         gsub('IGIIR', 'IGI/IR', .) %>%
         gsub('ISSI2', 'ISSI-2', .) %>%
         gsub('^HOMA$', 'HOMA-IR', .) %>%
-        gsub('TAG', 'TAG (mmol/L)', .) %>%
-        gsub('Chol', 'Chol (mmol/L)', .) %>%
-        gsub('ALT', 'ALT (U/L)', .) %>%
-        gsub('LDL', 'LDL (mmol/L)', .) %>%
-        gsub('HDL', 'HDL (mmol/L)', .) %>%
-        gsub('BaseTotalNE', 'NEFA (nmol/mL)', .) %>%
+        gsub('TAG', 'TAG (mmol/l)', .) %>%
+        gsub('Chol', 'Chol (mmol/l)', .) %>%
+        gsub('ALT', 'ALT (U/l)', .) %>%
+        gsub('LDL', 'LDL (mmol/l)', .) %>%
+        gsub('HDL', 'HDL (mmol/l)', .) %>%
+        gsub('BaseTotalNE', 'NEFA (nmol/ml)', .) %>%
         gsub('Age', 'Age (yrs)', .) %>%
+        gsub('Glucose0', 'Fasting glucose (mmol/l)', .) %>%
+        gsub('Glucose120', '2-hour glucose (mmol/l)', .) %>%
+        gsub('Insulin0', 'Fasting insulin (pmol/l)', .) %>%
+        gsub('Insulin120', '2-hour insulin (pmol/l)', .) %>%
         gsub('BMI', 'BMI (kg/m^2^)', .) %>%
+        gsub('FG', 'BMI (kg/m^2^)', .) %>%
         gsub('Waist', 'WC (cm)', .)
 }
 

### 2017-10-14 doc/esm.Rmd
Added more explanation of GEE to manuscript + esm

@@
 commonly used in longitudinal cohort studies, especially given its capacity to
-handle missed visits. GEE is able to handle longitudinal data due to the
+handle missed visits and for the correlation of variables across time. 
+GEE is able to handle the correlation of longitudinal data due to the
 estimating of a working correlation matrix, which was chosen based on
-quasi-likelihood information criteria (QIC). The auto-regressive of order 1
+quasi-likelihood information criteria (QIC; more detail on QIC below). The auto-regressive of order 1
 (AR1) matrix was chosen for the GEE models as it had the best model fit accessed
 using QIC, though other matrices (eg. exchangeable) had similar fit (data not
-shown). One main advantage of using methods such as GEE is the inclusion of the
+shown). While the choice of the correlation matrix is important, it should be noted
+that even with misspecification GEE still computes robust estimates and standard errors.
+One main advantage of using methods such as GEE is the inclusion of the
 working correlation matrix that takes into account the inherent covariance
@@
 longitudinal data since the correlation matrix assumes lower intercorrelation of
-outcomes within an individual as time passes. For example, BMI measured at the
+values within an individual as time passes. For example, BMI measured at the
 baseline visit would be more highly correlated to BMI measured at the 3rd year

### 2017-10-14 doc/manuscript.Rmd
Added more explanation of GEE to manuscript + esm

@@
 shown). GEE is well suited to longitudinal cohort studies given it's capacity to
-handle missed visits. The predictor variables and continuous covariates were
+handle any missed visits and for the correlation of variables between time points. 
+See the ESM Methods for more detail on GEE.
+The predictor variables and continuous covariates were
 scaled (mean centered and standardised). The NEFA variables were classified as 
@@
 [@Greenland1999a;@Textor2011a] recommendations (see `r cite_sf("dag_isi")` and 
-`r cite_sf("dag_issi2")`), and quasi-likelihood information criterion (QIC). See
+`r cite_sf("dag_issi2")`), and quasi-likelihood information criterion (QIC) [@Pan2001a]. See
 the ESM Methods for more detail on QIC and the directed acyclic graphs. 

### 2017-10-14 doc/submission/reviewer-response-2.md
(incomplete) edits to response

@@
 
-    We thank the reviewer for the comments and hope that the below responses
-    adequately address issues or suggestions raised.
+    We thank the reviewer for these supportive comments and hope that our responses below
+    adequately address the issues and suggestions raised.
 
@@
 structure that is not easily handled otherwise, for instance, might be helpful.
-Of note is that in the ESM it is stated that "other ... matrices had similar fit"
-- it may be worth noting that the model fit may be similar, but estimates may be
+Of note is that in the ESM it is stated that "other ... matrices had similar fit" -
+it may be worth noting that the model fit may be similar, but estimates may be
 more efficient under a particular correlation structure."*
 
-    The reviewer's suggestion to more fully describe the justification for using
-    GEE over other methods is warranted and we will expand on this in the
-    manuscript in the Methods and the ESM Methods sections. We chose the GEE
+    We agree entirely with the reviewer's suggestion to more fully describe the justification for using
+    GEE over other methods. In the revised manuscript, we have expanded our rationale
+    for using GEE in both the Methods and the ESM Methods sections. We chose the GEE
     technique for a variety of reasons. As the reviewer states, GEE is able to
@@
     others, GEE can compute robust standard errors of the estimates that more
-    accurate reflect the underlying data. The ability to specify a working
+    accurately reflect the underlying data. The ability to specify a working
     correlation matrix makes this a powerful technique for longitudinal data
     compared to standard linear regression, as it removes the underlying
-    assumption that variables are independent (which is not the case with
+    assumption that observations are independent (which is not the case with
     longitudinal data). While choosing the specific correlation matrix is
@@
     a misspecified correlation matrix. In the case of longitudinal data, an
-    auto-regressive correlation matrix is the most appropriate statistically and
-    biologically since measurements on an individual over time will change (e.g.
-    people gain or lose weight over time).
+    auto-regressive correlation matrix is the most appropriate, statistically and
+    biologically, since measurements on an individual over time will change (e.g.
+    people gain or lose weight over time) but still be highly correlated to their 
+    previous measurement (e.g. people don't gain or lose too much weight from their
+    original weight).
 
@@
 
-    We were only concerned with population level averages, and not
+    Thanks for this comment. We were only concerned with population level averages, and not
     individual-level estimates. GEE is a marginal model, which estimates the
@@
     even though GEE is a marginal model, for continuous outcomes (e.g. insulin
-    sensitivity) the estimates derived from GEE's counter part, mixed effects
+    sensitivity) the estimates derived from GEE's counter-part, namely mixed effects
     models, are more or less equivalent in both the value and in the
-    interpretation. For nonlinear outcomes (e.g. binary values such as diabetes
-    incidence), the interpretation and estimates are substantially different
-    from mixed effects models. We did not model nonlinear outcomes and as such
-    this difference is {{irrelevant?}}.
+    interpretation. For binary outcomes (e.g. diabetes cases), the
+    interpretation and estimates are substantially different from mixed effects
+    models.
 
@@
     interpreting the estimates derived from GEE is comparable to the classical
-    linear regression equations:
+    regression equation:
 
@@
 assumption might be that multivariable models include all of the NEFA components
-in one model."*
+in one model." ... "If so, I do not fully understand why there is not a model
+that includes the joint effect of all NEFA components together. Could this not
+be done for the concentration analysis? Also, it could perhaps be done for the
+proportion too, if one of the components is set as the baseline - similar to how
+categorical variables are handled in regression analyses, but allowing
+percentages for each indicator rather than the usual 0/1 dichotomy, vs. a
+"baseline" component. Such an analysis could also adjust for total NEFA, with
+the interpretation that proportion of each NEFA component is assessed while
+accounting for other NEFA components, and also the total NEFA. This is also
+something I struggled to understand a little in general - why should proportions
+of NEFA components be important? There may be a relatively high proportion of a
+particular NEFA but if the patient has a generally low overall NEFA, is there
+any reason to think a relatively high proportion of something will increase
+insulin sensitivity? These ideas may be wholly implausible of course and I may
+have missed the point - as I said my background knowledge of the biology is not
+ideal."*
 
@@
     equal to 100. It is this dependency structure in the data that allows PLS to
-    more accurately identify clusters within the data. Concentration data does
+    more accurately identify clusters within the data. Concentration data do
     not have this inherent dependency (all NEFA values do *not* equal to 100).
-    If one NEFA goes up in concentration, other NEFA values do not necessarily
-    go down, as would be the case with the proportion. We have clarified this in
-    the methods. {{not all of this comment is clear to me}}.
-
-- *"If so, I do not fully understand why there is not a model that includes the
-joint effect of all NEFA components together. Could this not be done for the
-concentration analysis? Also, it could perhaps be done for the proportion too,
-if one of the components is set as the baseline – similar to how categorical
-variables are handled in regression analyses, but allowing percentages for each
-indicator rather than the usual 0/1 dichotomy, vs. a "baseline" component. Such
-an analysis could also adjust for total NEFA, with the interpretation that
-proportion of each NEFA component is assessed while accounting for other NEFA
-components, and also the total NEFA. This is also something I struggled to
-understand a little in general – why should *proportions* of NEFA components be
-important? There may be a relatively high proportion of a particular NEFA but if
-the patient has a generally low overall NEFA, is there any reason to think a
-relatively high proportion of something will increase insulin sensitivity? These
-ideas may be wholly implausible of course and I may have missed the point – as I
-said my background knowledge of the biology is not ideal."*
-
-    {{Don't fully understand this comment, but I'll try to answer}}. We hope the
-    response to the point above partially addresses this comment. To reiterate,
-    the reason NEFA proportion was used in the PLS analysis was due to the
-    inherent interdependency of the values as a proportion. When they are
-    proportions, PLS would be more likely to pick up potential clusters since
-    changes in one value *must* lead to changes in another value as they all
-    equal 100.
+    If one NEFA increases in concentration, other NEFA values do not necessarily
+    decrease, as would be the case with the proportion. We have clarified this in
+    the methods.
 
@@
     concentration) into a single GEE model is because GEE can not handle this
-    level of complexity, or at least, the interpretation for us and the readers
-    would be nearly totally meaningless or incomprehensible. How the reviewer
+    level of complexity particular due to the collinearity between NEFA leading
+    to biased estimates and model instability. How the reviewer
     described setting the variables in the regression equation may be useful for
@@
     proportions but do not report on or measure the concentration. Concentration
-    data is more technically challenging to quantify and so is understandable
-    why large cohort studies do not measure this data. Using both concentration
-    and proportion, we can identify either it is *both* the total size of any
-    individual NEFA and a specific proportion of a single or multiple individual
-    NEFA that associates with risk for diabetes. Or whether it is only the total
-    concentration, as we found in this study.
+    data ara more technically challenging to quantify in the lab and so it is understandable
+    why large cohort studies do not measure these data. Using both concentration
+    and proportion, we can identify what specific role the absolute size and
+    specific proportion of individual NEFA have on the risk for diabetes.
 
@@
 
-    {{I'm not entirely clear how to address this. Since WC hypothetically could
-    lead to higher NEFA as well as lower IS (in particular), and since we are
-    testing whether higher NEFA lowers IS, how can we not adjust for it? It's a
-    classic confounder. I'll take a stab at answering it though.}} There may be
-    some potential over-adjustment in our models. However, we have taken
+    Given that the underlying physiology of NEFA's role in diabetes pathogenesis
+    has not been fully elucidated, it is possible that there may be some
+    over-adjustment in our models. However, we have taken
     extensive steps to understand the biological and causal pathways underlying
@@
 
-    In the case of WC, we feel there is substantial grounds for adjusting for it
-    when studying NEFA on metabolic capacity: 1) biologically NEFA comes from
-    adipose tissue; 2) intra-abdominal fat tissue is more metabolically active;
-    3) adipose tissue is an active endocrine organ that contributes to insulin
+    In the case of WC, we feel there is substantial evidence to support
+    adjusting for this variable
+    when studying NEFA's role in metabolic capacity: 1) biologically NEFA comes from
+    adipose tissue; 2) intra-abdominal fat tissue is more metabolically active; 3) 
+    adipose tissue is an active endocrine organ that contributes to insulin
     sensitivity and metabolism in general; and 4) WC is a direct measure of
@@
     Having said that, because of comments from previous reviewers about this
-    very issue, we included an analyses to address this issue. We ran models
-    adjusted for all covariates except for WC and reported on the results in the
+    very issue, we included an analyses to address this concern. We ran models
+    adjusted for all covariates *except for WC* and reported on the results in the
     ESM material. We had also included a discussion about this issue in
-    conclusions, which we hope will inform the interested reader about this
-    particular issue and allow them to draw their own interpretations of the
-    results based on this raised issue.
+    the Conclusions of the original manuscript, which we hope will inform the
+    interested reader about this particular issue and allow them to draw their
+    own interpretations of the results.
 
@@
     was not entirely clear and we have worked to make it clearer in the ESM and
-    methods. DAGs *are* in fact a process to identify potential covariates to
-    include in a model when used algorithmically. This mathematically construct
+    methods of the revised manuscript. DAGs *are* in fact a process to identify potential covariates to
+    include in a model when used algorithmically. The mathematical construct
     underlying DAGs and model building is the reason the DAGitty software was
@@
     greater detail on how DAGs are used to build models can be seen in the two
-    references listed below. The DAG figures in the manuscript will be described
+    references listed below. The DAG figures in the revised manuscript will be described
     in more detail, including what each colour and arrow indicates.
@@
 
-    Yes, all follow-up data is included *until* the visit with the incident
+    Yes, all follow-up data are included *until* the visit with the incident
     diabetes case (e.g. a participant is diagnosed with diabetes at the 6 year
@@
     analysis). This reason, among others, is why GEE and similar techniques are
-    so powerful. The wording has been fixed to make this point clearer.
+    so powerful. The wording has been fixed in the revised manuscript to make
+    this point clearer.
 
@@
         quasi-likelihood rather than maximum likelihood estimation (as other
-        techniques use such as mixed effects models) information criteria such
+        techniques use such as mixed effects models), information criteria such
         as AIC can not be used since they rely on maximum likelihood. We agree
@@
         have an inherent limitation in that models with more parameters will
-        have a better fit. So models with more parameters (more 'complex') be
+        have a better fit. So models with more parameters (more 'complex') will be
         more likely to be selected by default when using AIC. QIC and other
         similar information criteria techniques on the other hand try to balance
-        this limitation by explicitly including terms that penalize a more
+        this limitation by more strongly penalizing a more
         'complex' model (more parameters) while still emphasizing a good fit. We
@@
         both a good model fit while keeping model complexity low (less variables
-        included)."
+        included)." See the reference below for more information on QIC:
+
+        Pan W. (2001) Akaike's Information Criterion in Generalized Estimating
+        Equations. Biometrics 57: 120-125
 
@@
 
-    We thank the reviewer for this comments and the suggested edits below. We
-    have incorporated most of the suggestions and have provided responses that
-    hopefully clarify other questions and comments.
+    We thank the reviewer for these supportive comments and the suggested edits below. We
+    have incorporated the vast majority of the suggestions and have provided responses that
+    hopefully clarify these concerns.
 
@@
 
-    {{confirm that participants didn't have other disease}}. We have expanded on
-    the details about medication use (i.e. general types taken).
+    At the baseline visit, individuals were excluded from participating in the
+    study if they CKD, liver disease, or other chronic conditions.
+    We have expanded on the details about medication use (i.e. general types
+    taken).
 
@@
     We agree, this suggestion is pertinent and has been included in the
-    limitations section of the conclusion.
+    limitations section of the conclusion in the revised manuscript.
 
@@
 
-    {{This comment is not entirely clear to me}} This is an interesting
+    This is an interesting
     question. GEE does not model on the individual level, only on the population
@@
 
-    We thank the reviewer for the comments and have made the suggested changes
+    We thank the reviewer for these comments and have made the suggested changes
     as requested below.
@@
 
-        We agree with this wording. It has been edited.
+        We agree with this wording. It has been edited in the revised manuscript.
 
@@
         The initial metabolic phenotype and subsequent influence of saturated fatty
-    acids is taken into account in the time-by-fatty acid interaction. {{expand}}
+        acids is taken into account in the time-by-fatty acid interaction. In
+        the present analysis, there were no significant time-by-NEFA
+        interactions.
 
@@
         While we agree with the reviewer that inclusion of the phospholipid and
-    cholesteryl ester farctions would have added to the interpretation, the
-    volume and complexity of the analyses would not have been publishable with
-    current journal word and figure count limits. This was the reason we decided
-    to publish the fractions in separate manuscripts, to maintain a clear and
-    focus narrative within the constraints of the journals limits {{rephrase}}.
-    {{mention my thesis?}}
+        cholesteryl ester (as well as the triacylglycerol) fractions would have
+        added to the interpretation, the volume and complexity of the analyses
+        would not have been publishable given current journal word and figure
+        count limits. This was the reason we decided to publish the fractions in
+        separate manuscripts, to maintain a clear and focused narrative within
+        the constraints of the journals limits.
 
@@
 
+        {{add from the notes}}
+
 - *"Altogether, this paper would gain strength by addressing the limitations in
 interpretation in more detail."*
+
+    Based on comments from multiple reviewers, the limitation section has been expanded.
 

### 2017-10-14 doc/manuscript.Rmd
keywords in title/abstract, keywords in alpha order

@@
 
-Aim: Our aim was to determine the longitudinal associations of individual non-esterified fatty acids (NEFA) on differences in insulin sensitivity and beta cell function over 6 years in a cohort of individuals who are at-risk for diabetes.
+Aim: Our aim was to determine the longitudinal associations of individual non-esterified fatty acids (NEFA) on the pathogenesis of diabetes, specifically on differences in the insulin sensitivity and beta cell function over 6 years in a cohort of individuals who are at-risk for diabetes.
 
-Methods: In the PROMISE cohort, 477 participants had serum NEFA (mol% and concentration) measured at the baseline visit and completed an OGTT at 3 time points over 6 years.
+Methods: In the PROMISE longitudinal cohort, 477 participants had serum NEFA (mol% and concentration) measured at the baseline visit and completed an OGTT at 3 time points over 6 years.
 Outcome variables were calculated using the OGTT values.
@@
 
-insulin sensitivity; beta cell function; longitudinal cohort; non-esterified
-fatty acids; diabetes pathogenesis; fatty acid composition
+beta cell function; diabetes pathogenesis; fatty acid composition insulin sensitivity; longitudinal cohort; non-esterified fatty acids;
 

### 2017-10-15 doc/manuscript.Rmd
added all abbrev + rearrange order

@@
 
-beta cell function; diabetes pathogenesis; fatty acid composition insulin sensitivity; longitudinal cohort; non-esterified fatty acids;
+beta cell function; diabetes pathogenesis; fatty acid composition; insulin sensitivity; longitudinal cohort; non-esterified fatty acids;
 
@@
 
+- AR1: Autoregressive of order 1
+- ALT: Alanine aminotransferase
+- EPA: Eicosapentaenoic acid
+- FDR: False Discovery Rate
 - GEE: Generalized estimating equations
+- HOMA: Homeostatic model assessment
+- HOMA-IR: Homeostatic model assessment of insulin resistance
+- HOMA2-%S: Homeostatic model assessment 2 of insulin sensitivity
+- IFG: Impaired fasting glucose
 - IGI/IR: Insulinogenic index over HOMA-IR
+- IGT: Impaired glucose tolerance
 - ISI: Insulin Sensitivity Index (Matsuda index)
 - ISSI-2: Insulin Secretion-Sensitivity Index-2
-- ALT: Alanine aminotransferase
+- MAQ: Modified Activity Questionnaire
+- MET: Metabolic Equivalent of Task
+- NEFA: Non-esterified fatty acids
+- OGTT: Oral glucose tolerance test
+- PLS: Partial least squares regression
+- PROMISE: Prospective Metabolism and Islet Cell Evaluation Cohort
+- PUFA: Polyunsaturated fatty acids
+- QIC: Quasi-Likelihood Information Criterion
 - TAG: Triacylglyceride
-- PROMISE: Prospective Metabolism and Islet Cell Evaluation Cohort
-- QIC: Quasi-Likelihood Information Criterion
-- MAQ: Modified Activity Questionnaire
 - WC: Waist circumference
@@
 X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.
-P-values were adjusted for the BH false discovery rate, with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, p<0.01, and p<0.001.')
+P-values were adjusted for the BH false discovery rate (FDR), with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, p<0.01, and p<0.001.')
 fig_gee_adj <- fig('gee_adj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta cell function over the 6 years in the PROMISE cohort.
@@
 X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.
-P-values were adjusted for the BH false discovery rate, with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, and p<0.01.')
+P-values were adjusted for the BH false discovery rate (FDR), with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, and p<0.01.')
 ```
@@
 statistically significance at p<0.05, after adjusting for multiple testing using
-the Benjamini-Hochberg False Discovery Rate [@Benjamini1995a]. STROBE was used
+the Benjamini-Hochberg False Discovery Rate (FDR) [@Benjamini1995a]. STROBE was used
 as a guideline for reporting [@Vandenbroucke2007b].

### 2017-10-15 doc/manuscript.Rmd
added details about manufactorers

@@
 procedures. Cholesterol, HDL, and triacylglycerides (TAG) were measured using Roche
-Modular's enzymatic colorimetric tests (Mississauga, ON).  Both insulin
+Modular's enzymatic colorimetric tests (Roche Canada, Mississauga, Ontario, Canada).  Both insulin
 and glucose were derived from the OGTT at fasting, 30 minute, and 2 hour time
@@
 immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
-and the Des 31,32 circulating split form (Linco Res. Inc), and has a coefficient of variation (CV) of 9.3%.
+and the Des 31,32 circulating split form (Linco Res. Inc, St. Charles, Missouri, USA), and has a coefficient of variation (CV) of 9.3%.
 Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
-Diagnostics) with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The
+Diagnostics, Mississauga, Ontario, Canada) with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The
 inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were

### 2017-10-15 doc/manuscript.Rmd
mL -> l

@@
 Generalized estimating equations were used, adjusting for time, waist, sex, ethnicity, baseline age, alanine aminotransferase, and physical activity.
-NEFA were analysed as both concentrations (nmol/mL) and proportions (mol%) of the total fraction.
+NEFA were analysed as both concentrations (nmol/ml) and proportions (mol%) of the total fraction.
 
@@
 tab_basic <- tab('basic', 'Basic characteristics of the PROMISE participants at each of the three clinic visits.')
-fig_dist <- fig('nefa', 'Concentrations (nmol/mL) of each NEFA in PROMISE participants at the baseline visit (2004-2006).')
-fig_gee_unadj <- fig('gee_unadj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta cell function over 6 years in the PROMISE cohort.
+fig_dist <- fig('nefa', 'Concentrations (nmol/ml) of each NEFA in PROMISE participants at the baseline visit (2004-2006).')
+fig_gee_unadj <- fig('gee_unadj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/ml) with insulin sensitivity and beta cell function over 6 years in the PROMISE cohort.
 Models are only adjusted for time.
@@
 P-values were adjusted for the BH false discovery rate (FDR), with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, p<0.01, and p<0.001.')
-fig_gee_adj <- fig('gee_adj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/mL) with insulin sensitivity and beta cell function over the 6 years in the PROMISE cohort.
+fig_gee_adj <- fig('gee_adj', 'Longitudinal associations of individual non-esterified fatty acids (mol% and nmol/ml) with insulin sensitivity and beta cell function over the 6 years in the PROMISE cohort.
 Generalized estimating equation models were fully adjusted.
@@
 22 individual NEFA as mole percent (mol%) of the total fraction and as
-concentration (nmol/mL). Pearson correlation coefficients were computed to
+concentration (nmol/ml). Pearson correlation coefficients were computed to
 assess the relationships of individual NEFA with other continuous variables.
@@
 years (`r aide::stddev(ds0yr$Age)` SD) and the average BMI was
-`r aide::average(ds0yr$BMI)` (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
+`r aide::average(ds0yr$BMI)` kg/m^2^ (`r aide::stddev(ds0yr$BMI)` SD). Most of the 
 participants, `r famhist$npct`, had a family history of diabetes. Insulin
@@
 (16:0), oleic acid (18:1n-9), and linoleic acid (18:2n-6) for the concentration
-(nmol/mL) modeling, no associations were seen for these fatty acids as a
+(nmol/ml) modeling, no associations were seen for these fatty acids as a
 proportion, an observation that was further confirmed by multivariate cluster

### 2017-10-15 doc/manuscript.Rmd
revised contrib section

@@
 
-The authors had the following responsibility: LWJ conducted research, analysed
-data, and wrote the paper; RR, ZL, and SBH designed research, conducted
-research, and provided essential materials (infrastructure and clinical
-resources); RR, SBH, RPB, and AG provided intellectual feedback on the
-paper; RPB conducted research, provided essential reagents and materials; AJH
-designed research, assisted with interpretation, and provided intellectual
-feedback on all versions of the paper; LWJ and AJH had primary responsibility
-for final content and are the guarantors of this work. All authors read and
-approved the final manuscript.
+The authors had the following responsibility: LWJ analysed and interpreted the
+data and drafted the article; RR and SBH contributed to the conception and
+design and revised the article on intellectual content; ZL acquired the data and
+revised the article on intellectual content; AG assisted with interpretation of
+the data and revised the article on intellectual content; AJH and RPB
+substantially contributed to the study conception and design, assisted with
+interpretation of the data, and revised the article on intellectual content. All
+authors read and approved the final version. LWJ and AJH have primary
+responsibility for final content and are the guarantors of this work.
 

### 2017-10-15 doc/manuscript.Rmd
expanded on pls multivariable vs gee

@@
 was conducted on missing values and incident diabetes cases were excluded from 
-the analysis.
+the analysis. A GEE model is applied for each combination of outcome and NEFA (as
+either mol% or nmol/ml).
 
@@
 the outcome variable for every standard deviation (SD) increase in the predictor
-variable given the covariates are held constant. With the main effect 
+variable, for the same covariate values (e.g. same age or sex). With the main effect 
 associations (not with a time interaction), this interpretation is applied 
@@
 fatty acid. Therefore, to confirm the results of the GEE analyses in a
-multivariate environment (all NEFA variables in a single model), partial least
+multivariate environment (with all NEFA variables in a single model), partial least
 squares (PLS) was used with the outcomes as the dependent variable.

### 2017-10-15 doc/manuscript.Rmd
added another ref to DAG

@@
 Covariate selection was based on previous literature, directed acyclic graph 
-[@Greenland1999a;@Textor2011a] recommendations (see `r cite_sf("dag_isi")` and 
+[@Greenland1999a;@Textor2011a;@Shrier2008a] recommendations (see `r cite_sf("dag_isi")` and 
 `r cite_sf("dag_issi2")`), and quasi-likelihood information criterion (QIC) [@Pan2001a]. See

### 2017-10-15 doc/esm.Rmd
expanded on pls multivariable vs gee
added more detail about DAG

@@
 
-For the covariates, they were chosen based on previous literature, from 
-directed acyclic graph (DAG) recommendations, and from QIC. The DAG
-recommendations were obtained from using the DAGitty software,
-http://dagitty.net/. DAG is a technique that allows for identifying potential
-confounding causal pathways and to limit potential bias introduced from
-adjusting for colliding variables. A collider variable is a variable that is
-determined from two or more other variables, such that in causal graph modelling,
-the arrows from the other variables appear to 'collide' into the variable. QIC 
-is a technique that balances the goodness of fit of the model with its relative
-complexity (i.e. number of covariates) against other models being compared. This
-identifies a specific model from a number of models that is the 'best' fit and
-least complex of all the models. QIC and other information criterion methods are
-used in conjunction with other techniques such as literature and DAG based. The 
-final GEE model selected as best fitting was negligibly different when using 
-insulin sensitivity or beta-cell function as an outcome and as such the same 
-covariates were chosen for models with each of the outcomes. The covariates
-medication use, WC, ALT, and MET would be considered classical confounders while
-TAG would be considered a mediating variable.
+For the covariates, they were chosen based on previous literature, from directed
+acyclic graph (DAG) recommendations, and from QIC. The DAG recommendations were
+obtained from using the DAGitty software, http://dagitty.net/. DAG is a
+technique that allows for identifying potential confounding causal pathways and
+to limit potential bias. DAGs are composed on nodes (variables) connected by
+edges (arrows; causal pathway between two variables). Using DAGs to identify
+variables to adjust for done algorithmically, as implemented in the DAGitty
+software. DAGitty calculates possible sets of variables to include in a model to
+adjust for potential confounding in the causal network. Use of DAGs in this
+manner can identify potential sources of bias, such as from inclusion of
+colliding variables. A collider variable is a variable that has two (or more)
+'parent' nodes/variables (connected causally by DAG edges pointing into the
+'child' node, which is the collider). The collider variable and parent variables
+are directed linked to the exposure and the outcome. Including all variables
+involved with the collider could introduce bias as it does not generate an
+unconditional association between all involved variables in the model.
+
+QIC is a technique that balances the goodness of fit of the model with its
+relative complexity (i.e. number of covariates) against other models being
+compared. This identifies a specific model from a number of models that is the
+'best' fit and least complex of all the models. QIC and other information
+criterion methods are used in conjunction with other techniques such as
+literature and DAG based. The final GEE model selected as best fitting was
+negligibly different when using insulin sensitivity or beta-cell function as an
+outcome and as such the same covariates were chosen for models with each of the
+outcomes. The covariates medication use, WC, ALT, and MET would be considered
+classical confounders while TAG would be considered a mediating variable.
 
@@
 predicting the contributions of predictor variables against an outcome variable.
+In this analysis, PLS was computed on the NEFA as a proportion due to the
+inherent interdependence of the values (all must equal 100%). This 
+interdependence allows PLS to identify patterns in the composition, since
+the concentration data is all highly positively correlated with each other.
 

### 2017-10-15 doc/esm.Rmd
described DAG colors

@@
 sfig_dag_isi <- sfig("dag_isi", "Directed acyclic graph taken from DAGitty (DAGitty.net) for the potential causal pathway between NEFA and ISI.
+Node and arrow description: Green with a triangle is the exposure; blue with a line/I is the outcome; empty blue is an ancestor of the outcome; empty green is an ancestor of the exposure; and empty red is an ancestor of both the outcome and the exposure; green arrow is on causal pathway; red arrow is a biasing pathway; black arrow is not along the pathway from the exposure to the outcome.
 DAGitty recommended a minimal adjustment of waist, sex, and MET (physical activity).")
 sfig_dag_issi2 <- sfig("dag_issi2", "Directed acyclic graph taken from DAGitty (DAGitty.net) for the potential causal pathway between NEFA and ISSI-2.
+Node and arrow description: Green with a triangle is the exposure; blue with a line/I is the outcome; empty blue is an ancestor of the outcome; empty green is an ancestor of the exposure; and empty red is an ancestor of both the outcome and the exposure; green arrow is on causal pathway; red arrow is a biasing pathway; black arrow is not along the pathway from the exposure to the outcome.
 DAGitty recommended a minimal adjustment of waist, ethnicity, and MET (physical activity).")

### 2017-10-15 doc/esm.Rmd
more detail on QIC

@@
 relative complexity (i.e. number of covariates) against other models being
-compared. This identifies a specific model from a number of models that is the
-'best' fit and least complex of all the models. QIC and other information
+compared. QIC, more so then the more common AIC, penalizes the inclusion of more
+variables in the model. This technique is used to compare multiple models to
+identify the model with the highest fit and with the lowest complexity (variables),
+*of the models compared*. QIC is similar to AIC, except AIC is based on maximum
+likelihood of models such as mixed effects models. Since GEE does not use
+maximum likelihood, AIC can not be used. QIC and other information
 criterion methods are used in conjunction with other techniques such as

### 2017-10-16 R/renaming.R
sample size to table

@@ -3,6 +3,7 @@
 
 renaming_table_rows <- function(x) {
     x %>%
+        gsub('sample_size', 'Sample size', .) %>%
         gsub('HOMA2_S', 'HOMA2-%S', .) %>%
         gsub('IGIIR', 'IGI/IR', .) %>%
         gsub('ISSI2', 'ISSI-2', .) %>%

### 2017-10-16 R/tables.R

@@ -8,11 +8,15 @@
 #'
 table_basic <- function(data, caption = NULL) {
     data %>%
+        dplyr::group_by(f.VN) %>%
+        dplyr::mutate(sample_size = sum(!is.na(TotalNE))) %>%
+        dplyr::ungroup() %>%
         dplyr::mutate(Ethnicity = ifelse(VN == 0, as.character(Ethnicity), NA),
                       Ethnicity = as.factor(Ethnicity),
                       Sex = ifelse(VN == 0, as.character(Sex), NA),
                       Sex = as.factor(Sex)) %>%
         carpenter::outline_table('f.VN') %>%
+        carpenter::add_rows("sample_size", unique) %>%
         carpenter::add_rows(c('HOMA2_S', 'ISI', 'IGIIR', 'ISSI2'),
                             carpenter::stat_medianIQR, digits = 1) %>%
         carpenter::add_rows(c('BMI', 'Waist', 'Age', 'MET', 'ALT', 'TAG', 'Chol', 'HDL',

### 2017-10-16 doc/manuscript.Rmd
clarified that diabetes cases were not used only at/after visit
clarified med us

@@
 was conducted on missing values and incident diabetes cases were excluded from 
-the analysis. A GEE model is applied for each combination of outcome and NEFA (as
-either mol% or nmol/ml).
+the analysis (follow-up data were included until date of diagnosis). A GEE model
+is applied for each combination of outcome and NEFA (as either mol% or nmol/ml).
 
@@
 pre-diabetes (IFG and IGT), respectively, over the 6-years. Incident diabetes
-cases were excluded from further GEE analyses. There were `r meds_use[1]` and
-`r meds_use[2]` participants who used lipid and cholesterol modifying medications
-at the baseline and 6-year visit, respectively. Medication use did not influence
-NEFA composition, the outcomes, or the GEE models (data not shown).
+cases were excluded from GEE analyses (follow-up data were included until date
+of diagnosis). Certain medications (e.g. lipid medications such as statins or
+fibrates) could influence the association between NEFA and the outcomes.
+There were `r meds_use[1]` and `r meds_use[2]` participants who used lipid and
+cholesterol modifying medications at the baseline and 6-year visit,
+respectively. Use of these medications did not influence NEFA composition, the
+outcomes, or the GEE models (data not shown).
 

### 2017-10-16 doc/esm.Rmd
rearranged setup code chunks in esm

@@
 
-```{r setup, include=FALSE}
-devtools::load_all()
-set_options()
-```
+# ESM Tables
 
-```{r captions}
+```{r tab-captions}
 # Include captions below using `captioner` package via library(captioner)
+knitr::opts_chunk$set(results = "markup")
 stab <- captioner::captioner(prefix = 'ESM Table')
@@
 ```
-
-# ESM Tables
 
@@
 
-```{r setup, include=FALSE}
-devtools::load_all()
-set_options()
+# ESM Figures
+
+```{r fig-captions, include=FALSE}
 knitr::opts_chunk$set(dev = "png")
@@
 
-# ESM Figures
-
 ![`r sfig_consort`](../img/flowDiagramSample.png)

### 2017-10-16 doc/manuscript.Rmd
added numbers to abstract

@@
 While in unadjusted models there were several NEFA (e.g.
-18:1n-7, 20:1n-9, 22:4n-6) that had negative associations with insulin sensitivity, nearly all of these associations were attenuated in fully adjusted models.
-In adjusted models, total NEFA, 16:0, 18:1n-9, and 18:2n-6 (as concentrations) had significant negative associations with IGI/IR and ISSI-2, while only 20:5n-3 (as mol%) had positive associations with HOMA2-%S.
+18:1n-7, 20:1n-9, 22:4n-6) that associated with lower insulin sensitivity, nearly all of these associations were attenuated in fully adjusted models.
+In adjusted models, total NEFA, 16:0, 18:1n-9, and 18:2n-6 (as concentrations) associated with 3.7 to 8.0% lower IGI/IR and ISSI-2, while only 20:5n-3 (as mol%) associated with 7.7% lower HOMA2-%S.
 

### 2017-10-16 doc/manuscript.Rmd
limitation of outcomes as estimates

@@
 
+We used indices of insulin sensitivity and beta cell function estimated from
+glucose and insulin values and as such could that lead to measurement bias.
+While these indices have all been validated against gold standard measures,
+it is possible our measures are not sensitivity enough to identify potential
+associations with NEFA.
 Finally, our cohort consists of individuals at-risk for diabetes,

### 2017-10-16 doc/submission/reviewer-response-2.md
added last response

@@
 
-        {{add from the notes}}
+        Fasting NEFA derives from adipose tissue, which contains fat stores from
+        both diet and de novo lipogenesis. Besides potential dysfunction in the
+        adipose tissue for the release of NEFA, other factors that could determine
+        the NEFA profile include genetics and long term dietary fat intake
+        (adipose tissue fat stores have a half-life of ~1.5 years). These other
+        factors were not considered and discussed in the present paper due to
+        the difficulty in measuring long term diet accurately enough on the
+        specific type of fat consumed, the lack of genetic information contained
+        in the cohort, and the fact that disentangling the specific contribution
+        of de novo lipogenesis from dietary fat on adipose tissue stores is well
+        outside the scope of this paper.
 

### 2017-10-17 R/setup.R
removed png from options

@@ -21,7 +21,7 @@ set_options <- function() {
     # For the document (knitr)
     knitr::opts_chunk$set(
         warning = FALSE, message = FALSE, collapse = TRUE, echo = FALSE,
-        dev = c("cairo_ps", "pdf", "png"), fig.showtext = TRUE
+        dev = c("cairo_ps", "pdf"), fig.showtext = TRUE
     )
 }
 

### 2017-10-17 R/utils.R
slight margins to figures

@@ -21,7 +21,7 @@ graph_theme <- function(base.plot, ticks = TRUE, minor.grid.lines = FALSE, legen
                 legend.key.width = grid::unit(0.7, "line"),
                 legend.key.height = grid::unit(0.7, "line"),
                 strip.background = ggplot2::element_blank(),
-                plot.margin = grid::unit(c(0.5, 0, 0, 0), "cm"),
+                plot.margin = grid::unit(c(0.5, 0.1, 0.1, 0.1), "cm"),
                 legend.position = legend.pos
             )
         )

### 2017-10-17 doc/submission/reviewer-response-2.md
updated response re nums for excluded dm

@@
 
-        This has been added to ESM Figure 1.
+        The exact number of excluded cases of diabetes is dependent on this analysis
+        and doesn't reflect the main PROMISE cohort. Since the CONSORT diagram 
+        represents the main cohort, we included the number of excluded diabetes
+        cases in the text of the Results section since the text reflects the 
+        present analysis.
 

### 2017-10-17 doc/manuscript.Rmd
started text on excluded cases at each visit
added other refs on beta cell decline

@@
 pre-diabetes (IFG and IGT), respectively, over the 6-years. Incident diabetes
-cases were excluded from GEE analyses (follow-up data were included until date
-of diagnosis). Certain medications (e.g. lipid medications such as statins or
+cases were excluded from GEE analyses (n={{num}} at 3rd year, n=
+{{num}} at 6th year). Follow-up data were included until date of
+diagnosis.
+Certain medications (e.g. lipid medications such as statins or
 fibrates) could influence the association between NEFA and the outcomes.
@@
 a particularly strong negative association with beta cell function
-[@Johns2014a]. Experimentally, several potential mechanisms have been elucidated
+[@Johns2014a]. Another cross-sectional study found a null association of
+total NEFA on beta cell function [@Rebelos2015a], however they used C-peptide to
+estimate beta cell function.
+Experimentally, several potential mechanisms have been elucidated
 for the role of NEFA on beta cell function, particularly for palmitic acid.
@@
 generation of nitric oxide, as well as impairment of proinsulin production and
-mitochondrial function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a]. The
+mitochondrial function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a]. 
+One longitudinal studies also found an inverse association between total NEFA
+and beta cell function over 10 years in 77 Japanese patients with type 2
+diabetes [@Morita2012a], though they used C-peptide to estimate beta cell
+function. The
 present analysis is the first study to our knowledge to examine the longitudinal

### 2017-10-17 doc/manuscript.Rmd
minor typos/edits before sending to Tony

@@
 
+- ALT: Alanine aminotransferase
 - AR1: Autoregressive of order 1
-- ALT: Alanine aminotransferase
 - EPA: Eicosapentaenoic acid
@@
 Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
-Diagnostics, Mississauga, Ontario, Canada) with a detection range of 0.11 (2 mg/dL) to 41.6 mmol/L.  The
+Diagnostics, Mississauga, Ontario, Canada) with a detection range of 0.11 (2 mg/dl) to 41.6 mmol/l.  The
 inter-assay %CV is <1.1% and intra-assay %CV is < 1.9%.  All assays were
@@
 reflects hepatic insulin sensitivity, while ISI reflects whole-body insulin
-sensitivity [@AbdulGhani2007].  beta cell function was assessed using the
+sensitivity [@AbdulGhani2007].  Beta cell function was assessed using the
 Insulinogenic Index [@Wareham1995] over HOMA-IR (IGI/IR) and the Insulin
@@
 Consistent with this decline, there were `r calc_incid_dysgly("ConvertDM")`
-and `r calc_incid_dysgly("ConvertPreDM")`incident cases of diabetes and
+and `r calc_incid_dysgly("ConvertPreDM")` incident cases of diabetes and
 pre-diabetes (IFG and IGT), respectively, over the 6-years. Incident diabetes

### 2017-10-20 doc/submission/reviewer-response-2.md
slight edits to response as per Tony's suggestions

--- 2017-10-16 doc/submission/reviewer-response-2.md
@@
     handle visit dates that differ between participants (e.g. some participants
-    come exactly 3 years later for a followup while others come some months
+    come exactly 3 years later for a follow up while others come some months
     before or after 3 years) and for dealing with the inherent correlational
@@
     The reason we did not include all NEFA variables (as proportion or as
-    concentration) into a single GEE model is because GEE can not handle this
+    concentration) into a single GEE model is because GEE cannot handle this
     level of complexity particular due to the collinearity between NEFA leading
@@
     proportions but do not report on or measure the concentration. Concentration
-    data ara more technically challenging to quantify in the lab and so it is understandable
+    data are more technically challenging to quantify in the lab and so it is understandable
     why large cohort studies do not measure these data. Using both concentration
@@
         techniques use such as mixed effects models), information criteria such
-        as AIC can not be used since they rely on maximum likelihood. We agree
+        as AIC cannot be used since they rely on maximum likelihood. We agree
         that the wording surrounding 'fit' and 'complex' could be clearer.
@@
 
-        This has been added to ESM Figure 1.
+        The exact number of excluded cases of diabetes is dependent on this analysis
+        and doesn't reflect the main PROMISE cohort. Since the CONSORT diagram 
+        represents the main cohort, we included the number of excluded diabetes
+        cases in the text of the Results section as the text reflects the 
+        present analysis. (See first paragraph of Results section.)
 
@@
         separate manuscripts, to maintain a clear and focused narrative within
-        the constraints of the journals limits.
+        the constraints of the journals limits. If interested, the reference is:
+
+        Johnston LW, Harris SB, Retnakaran R, et al (2016) Longitudinal
+        associations of phospholipid and cholesteryl ester fatty acids with
+        disorders underlying diabetes. J Clin Endocrinol Metab 101:2536–2544.
 

### 2017-10-20 doc/esm.Rmd
removed gen png for esm
minor typo fix in esm caption

@@
 ```{r fig-captions, include=FALSE}
-knitr::opts_chunk$set(dev = "png")
 
@@
 X-axis values represent a percent difference in the outcome per SD increase in the fatty acid.
-P-values were adjusted for the BH false discovery rate, with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, and p<0.01.')
+P-values were adjusted for the BH false discovery rate, with each increase in dot size and blackness representing a p-value significance of p>0.05, p<0.05, p<0.01, and p<0.001.')
 ```

### 2017-10-20 R/calculate.R
calc number of dm at each visit

@@ -156,3 +156,20 @@ calc_pct_full_visits <- function() {
     pct_val <- calc_n_for_visits()[1] / sum(calc_n_for_visits())
     precise_rounding(pct_val * 100)
 }
+
+#' Calculate the number of DM cases at each clinic visit.
+#'
+calc_dm_each_visit <- function() {
+    project_data %>%
+        incid_dysgly() %>%
+        dplyr::select(SID, DetailedConvert, `0`, `1`, `2`) %>%
+        tidyr::gather(VN, Dysgly, -SID, -DetailedConvert) %>%
+        dplyr::mutate(DM = as.numeric(Dysgly == "DM")) %>%
+        dplyr::arrange(SID, VN) %>%
+        dplyr::filter(DM == 1) %>%
+        dplyr::group_by(SID) %>%
+        dplyr::mutate(DM_cumsum = cumsum(DM)) %>%
+        dplyr::ungroup() %>%
+        dplyr::filter(DM_cumsum == 1) %>%
+        dplyr::count(VN)
+}

### 2017-10-20 R/gee.R
remove legend

@@ -107,7 +107,7 @@ plot_gee_main <- function(results) {
             xlab = 'Percent difference with 95% CI in the outcomes\nfor each SD increase in NEFA',
             ylab = 'NEFA'
             ) +
-        graph_theme(ticks = FALSE, legend.pos = "right") +
+        graph_theme(ticks = FALSE, legend.pos = "none") +
         ggplot2::facet_grid(unit~Yterms, switch = 'both',
                             scales = 'free_y',
                             space = 'free_y', as.table = FALSE) +

### 2017-10-20 doc/manuscript.Rmd
edits to manuscript as per Tony

@@
 Title: Association of non-esterified fatty acid composition with insulin
-sensitivity and beta cell dysfunction in the Prospective Metabolism and Islet Cell
+sensitivity and beta cell function in the Prospective Metabolism and Islet Cell
 Evaluation (PROMISE) cohort
@@
 
-Word count - Main text: 3869 / 4000 (with figure captions)
-
-<!-- Table and figure count: 4 / 4 -->
-
-<!-- Reference limit 49 / 50 -->
-
 # Abstract
 
-Aim: Our aim was to determine the longitudinal associations of individual non-esterified fatty acids (NEFA) on the pathogenesis of diabetes, specifically on differences in the insulin sensitivity and beta cell function over 6 years in a cohort of individuals who are at-risk for diabetes.
+Aim: Our aim was to determine the longitudinal associations of individual non-esterified fatty acids (NEFA) on the pathogenesis of diabetes, specifically on differences in insulin sensitivity and beta cell function over 6 years in a cohort of individuals who are at-risk for diabetes.
 
@@
 immunoassay. This assay shows 0.05% cross-reactivity to intact human pro-insulin
-and the Des 31,32 circulating split form (Linco Res. Inc, St. Charles, Missouri, USA), and has a coefficient of variation (CV) of 9.3%.
+and the Des 31,32 circulating split form, and has a coefficient of variation (CV) of 9.3%.
 Glucose was determined using an enzymatic hexokinase (Roche Modular, Roche
@@
 models, though other matrices (eg. exchangeable) had similar fit (data not
-shown). GEE is well suited to longitudinal cohort studies given it's capacity to
+shown). GEE is well suited to longitudinal cohort studies given its capacity to
 handle any missed visits and for the correlation of variables between time points. 
@@
 as a guideline for reporting [@Vandenbroucke2007b].
-The cohort datasets analysed during the current study are available from the
-corresponding author on reasonable request.
 
@@
     dplyr::mutate(Xterms = ifelse(Xterms == "Total", "total NEFA", as.character(Xterms)))
+
+dm_each_visit <- calc_dm_each_visit()
 ```
@@
 pre-diabetes (IFG and IGT), respectively, over the 6-years. Incident diabetes
-cases were excluded from GEE analyses (n={{num}} at 3rd year, n=
-{{num}} at 6th year). Follow-up data were included until date of
-diagnosis.
+cases were excluded from GEE analyses (1, 28, and 13 cases were excluded at
+0, 3, and 6 year visits, respectively).
+Follow-up data were included until date of diagnosis.
 Certain medications (e.g. lipid medications such as statins or
 fibrates) could influence the association between NEFA and the outcomes.
-There were `r meds_use[1]` and `r meds_use[2]` participants who used lipid and
-cholesterol modifying medications at the baseline and 6-year visit,
+There were `r meds_use[1]` and `r meds_use[2]` participants who used lipid 
+modifying medications at the baseline and 6-year visit,
 respectively. Use of these medications did not influence NEFA composition, the
-outcomes, or the GEE models (data not shown).
+outcomes, or their associations as analysed from the GEE models (data not shown).
 
@@
 We used indices of insulin sensitivity and beta cell function estimated from
-glucose and insulin values and as such could that lead to measurement bias.
-While these indices have all been validated against gold standard measures,
-it is possible our measures are not sensitivity enough to identify potential
-associations with NEFA.
+glucose and insulin values during OGTTs. While these indices have all been
+validated against gold standard measures, it is possible our measures may
+contain some degree of measurement misclassification and are thus not sensitive
+enough to identify potential associations with NEFA.
 Finally, our cohort consists of individuals at-risk for diabetes,
@@
 [@Johns2014a]. Another cross-sectional study found a null association of
-total NEFA on beta cell function [@Rebelos2015a], however they used C-peptide to
-estimate beta cell function.
+total NEFA on beta cell function [@Rebelos2015a], however C-peptide was used to
+estimate beta cell function in this study.
 Experimentally, several potential mechanisms have been elucidated
@@
 mitochondrial function [@Cnop2008a; @Maris2013a; @Giacca2011a; @Xiao2009a]. 
-One longitudinal studies also found an inverse association between total NEFA
+One longitudinal study also found an inverse association between total NEFA
 and beta cell function over 10 years in 77 Japanese patients with type 2
-diabetes [@Morita2012a], though they used C-peptide to estimate beta cell
+diabetes [@Morita2012a]; in this study C-peptide was used to estimate beta cell
 function. The
@@
 
-This study was supported by grants from the Canadian Diabetes Association (CDA),
-the Canadian Institutes for Health Research, and the University of Toronto
+This study was supported by grants from the Canadian Diabetes Association (CDA;
+Grant number: OG-3-14-4574-AH), the Canadian Institutes for Health Research
+(Grant number: MOP-130458), and the University of Toronto
 Banting and Best Diabetes Centre; LWJ was supported by a CDA Doctoral Student
@@
 Epidemiology.
+
+## Data availability
+
+The cohort datasets analysed during the current study are available from the
+corresponding author on request.
 
